### 2020-11-30 docs/manuscript/manuscript.tex
added manuscript shell

\documentclass[11pt,twoside]{article}
\usepackage{etex, soul}
\newcommand{\num}{6{} }

\raggedbottom

%geometry (sets margin) and other useful packages
\usepackage{geometry}
\geometry{top=1.2in, left=1in,right=1in,bottom=1in,headsep=6pt}
 \usepackage{graphicx,booktabs,calc}

%=== GRAPHICS PATH ===========
\graphicspath{{./images/}}
% Marginpar width
%Marginpar width
\newcommand{\pts}[1]{\marginpar{ \small\hspace{0pt} \textit{[#1]} } }
\setlength{\marginparwidth}{.5in}
%\reversemarginpar
%\setlength{\marginparsep}{.02in}

%% Fonts
% \usepackage{fourier}
% \usepackage[T1]{pbsi}

\usepackage{lmodern}
\usepackage[T1]{fontenc}
%\usepackage{minted}

\usepackage{rotating}
%% Cite Title
\usepackage[style=numeric,backend=biber,natbib,sorting=none,maxcitenames=2,maxbibnames=99,doi=false,isbn=false,url=false,eprint=false]{biblatex}
\addbibresource{../ai-trees-references.bib}

%%% Counters
\usepackage{chngcntr,mathtools}
\counterwithout{figure}{section}
\counterwithout{table}{section}

\numberwithin{equation}{section}

%% Captions
\usepackage{caption}
\captionsetup{
  labelsep=quad,
  justification=raggedright,
  labelfont=sc
}

%AMS-TeX packages
\usepackage{amssymb,amsmath,amsthm}
\usepackage{bm}
\usepackage[mathscr]{eucal}
\usepackage{colortbl}
\usepackage{color}

\DeclareRobustCommand{\hlcyan}[1]{{\sethlcolor{cyan}\hl{#1}}}

\usepackage{epstopdf,subfigure,hyperref,enumerate,polynom,polynomial}
\usepackage{multirow,minitoc,fancybox,array,multicol}

\definecolor{slblue}{rgb}{0,.3,.62}
\hypersetup{
    colorlinks,%
    citecolor=blue,%
    filecolor=blue,%
    linkcolor=blue,
    urlcolor=slblue
}

%%%TIKZ
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{pgfgantt}
\pgfplotsset{compat=newest}

\usetikzlibrary{arrows,shapes,positioning}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{shadows,automata}
\usetikzlibrary{patterns,fit}
%\usetikzlibrary{circuits.ee.IEC}
\usetikzlibrary{decorations.text}
% For Sagnac Picture
\usetikzlibrary{%
    decorations.pathreplacing,%
    decorations.pathmorphing%
}

%
%Redefining sections as problems
%
\makeatletter
\newenvironment{question}{\@startsection
	{section}
	{1}
	{-.2em}
	{-3.5ex plus -1ex minus -.2ex}
    	{1.3ex plus .2ex}
    	{\pagebreak[3]%forces pagebreak when space is small; use \eject for better results
	\large\bf\noindent{Question }
	}
	}
	%{\vspace{1ex}\begin{center} \rule{0.3\linewidth}{.3pt}\end{center}}
	%\begin{center}\large\bf \ldots\ldots\ldots\end{center}}
\makeatother

%
%Fancy-header package to modify header/page numbering
%
%\renewcommand{\chaptermark}[1]{ \markboth{#1}{} }
\renewcommand{\sectionmark}[1]{ \markright{#1}{} }

\usepackage{fancyhdr}
\pagestyle{fancy}
%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
%\addtolength{\headwidth}{\marginparwidth}
%\fancyheadoffset{30pt}
%\fancyfootoffset{30pt}
\fancyhead[LO,RE]{\small  \it \nouppercase{\leftmark}}
\fancyhead[RO,LE]{\small Page \thepage}
\fancyfoot[RO,LE]{\small }% PR \num S-2015}
\fancyfoot[LO,RE]{\small }%\scshape MODL}
\cfoot{}
\renewcommand{\headrulewidth}{0.1pt}
\renewcommand{\footrulewidth}{0pt}
%\setlength\voffset{-0.25in}
%\setlength\textheight{648pt}

\usepackage{paralist}

%%% FORMAT PYTHON CODE
\usepackage{listings}
% Default fixed font does not support bold face
\DeclareFixedFont{\ttb}{T1}{txtt}{bx}{n}{8} % for bold
\DeclareFixedFont{\ttm}{T1}{txtt}{m}{n}{8}  % for normal

% Custom colors
\usepackage{color}
\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}

%\usepackage{listings}

% % Python style for highlighting
% \newcommand\pythonstyle{\lstset{
% language=Python,
% basicstyle=\footnotesize\ttm,
% otherkeywords={self},             % Add keywords here
% keywordstyle=\footnotesize\ttb\color{deepblue},
% emph={MyClass,__init__},          % Custom highlighting
% emphstyle=\footnotesize\ttb\color{deepred},    % Custom highlighting style
% stringstyle=\color{deepgreen},
% frame=tb,                         % Any extra options here
% showstringspaces=false            %
% }}

% % Python environment
% \lstnewenvironment{python}[1][]
% {
% \pythonstyle
% \lstset{#1}
% }
% {}

% % Python for external files
% \newcommand\pythonexternal[2][]{{
% \pythonstyle
% \lstinputlisting[#1]{#2}}}

% % Python for inline
% \newcommand\pythoninline[1]{{\pythonstyle\lstinline!#1!}}

\newcommand{\osn}{\oldstylenums}
\newcommand{\dg}{^{\circ}}
\newcommand{\lt}{\left}
\newcommand{\rt}{\right}
\newcommand{\pt}{\phantom}
\newcommand{\tf}{\therefore}
\newcommand{\?}{\stackrel{?}{=}}
\newcommand{\fr}{\frac}
\newcommand{\dfr}{\dfrac}
%\newcommand{\ul}{\underline}
\newcommand{\tn}{\tabularnewline}
\newcommand{\nl}{\newline}
\newcommand\relph[1]{\mathrel{\phantom{#1}}}
\newcommand{\cm}{\checkmark}
\newcommand{\ol}{\overline}
\newcommand{\rd}{\color{red}}
\newcommand{\bl}{\color{blue}}
\newcommand{\pl}{\color{purple}}
\newcommand{\og}{\color{orange!90!black}}
\newcommand{\gr}{\color{green!40!black}}
\newcommand{\nin}{\noindent}
\newcommand{\la}{\lambda}
\renewcommand{\th}{\theta}
\newcommand{\al}{\alpha}
\newcommand{\G}{\Gamma}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,thick,inner sep=1pt] (char) {\small #1};}}

\newcommand{\bc}{\begin{compactenum}[\quad--]}
\newcommand{\ec}{\end{compactenum}}

\newcommand{\p}{\partial}
\newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\dpd}[2]{\dfrac{\partial{#1}}{\partial{#2}}}
\newcommand{\pdd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
%\vspace{-16ex}
\title{Predicting tree risks to power lines using a convlutional neural network}
%\author{}
\date{}
\maketitle

\section{Introduction}
\hl{@Nasko: write first draft of introduction/motivation for the work}

\section{Background}
\hl{@Nasko: write first draft of literature review. I will edit.}

\section{Data}

\section{Methods}

\section{Results}

\section{Discussion}

\section{Conclusion}

\appendix

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2021-01-04 docs/presentations/20201104-Meeting/20201104-Meeting.tex
added presentations

\documentclass[usenames,dvipsnames,smaller,handout
]{beamer}
% \usepackage[T1]{fontenc} 
% \usepackage{lmodern} 
%\usepackage{etex}
%\newcommand{\num}{6{} }

% \usetheme[
%   outer/progressbar=foot,
%   outer/numbering=fraction,
%   block=fill,
%   inner/subsectionpage=progressbar
% ]{metropolis}
\usetheme{Madrid}
\useoutertheme[subsection=false]{miniframes} % Alternatively: miniframes, infolines, split
\useinnertheme{circles}
%\useoutertheme{Frankfurt}
\usecolortheme{beaver}
%\useoutertheme{crane}
%\useoutertheme{metropolis}
\usepackage[backend=biber,style=authoryear,maxcitenames=2,maxbibnames=99,safeinputenc,url=false,
eprint=false]{biblatex}
\addbibresource{bib/references.bib}
\AtEveryCitekey{\iffootnote{{\tiny}\tiny}{\tiny}}

\usepackage{pgfpages}
%\setbeameroption{hide notes} % Only slides
%\setbeameroption{show only notes} % Only notes
\setbeameroption{hide notes} % Only notes
%\setbeameroption{show notes on second screen=right} % Both
\usepackage{appendixnumberbeamer}
% \usepackage[sfdefault]{Fira Sans}

% \setsansfont[BoldFont={Fira Sans}]{Fira Sans Light}
% \setmonofont{Fira Mono}

%\usepackage{fira}
%\setsansfont{Fira}
%\setmonofont{Fira Mono}
% To give a presentation with the Skim reader (http://skim-app.sourceforge.net) on OSX so
% that you see the notes on your laptop and the slides on the projector, do the following:
% 
% 1. Generate just the presentation (hide notes) and save to slides.pdf
% 2. Generate onlt the notes (show only nodes) and save to notes.pdf
% 3. With Skim open both slides.pdf and notes.pdf
% 4. Click on slides.pdf to bring it to front.
% 5. In Skim, under "View -> Presentation Option -> Synhcronized Noted Document"
%    select notes.pdf.
% 6. Now as you move around in slides.pdf the notes.pdf file will follow you.
% 7. Arrange windows so that notes.pdf is in full screen mode on your laptop
%    and slides.pdf is in presentation mode on the projector.

% Give a slight yellow tint to the notes page
\setbeamertemplate{note page}{\pagecolor{yellow!5}\insertnote}\usepackage{palatino}

%\usetheme{metropolis}
%\usecolortheme{beaver}
\usepackage{tipa}
\definecolor{darkcandyapplered}{HTML}{A40000}
\definecolor{lightcandyapplered}{HTML}{e74c3c}

%\setbeamercolor{title}{fg=darkcandyapplered}

\definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667} % UBC Blue (primary)
\definecolor{UBCgrey}{rgb}{0.3686, 0.5255, 0.6235} % UBC Grey (secondary)

% \setbeamercolor{palette primary}{bg=darkcandyapplered,fg=white}
% \setbeamercolor{palette secondary}{bg=darkcandyapplered,fg=white}
% \setbeamercolor{palette tertiary}{bg=darkcandyapplered,fg=white}
% \setbeamercolor{palette quaternary}{bg=darkcandyapplered,fg=white}
% \setbeamercolor{structure}{fg=darkcandyapplered} % itemize, enumerate, etc
% \setbeamercolor{section in toc}{fg=darkcandyapplered} % TOC sections
% \setbeamercolor{frametitle}{fg=darkcandyapplered,bg=white} % TOC sections
% \setbeamercolor{title in head/foot}{bg=white,fg=white} % TOC sections
% \setbeamercolor{button}{fg=darkcandyapplered} % TOC sections

% % Override palette coloring with secondary
% \setbeamercolor{subsection in head/foot}{bg=lightcandyapplered,fg=white}

%\usecolortheme{crane}
% \makeatletter
% \setbeamertemplate{headline}{%
%   \begin{beamercolorbox}[colsep=1.5pt]{upper separation line head}
%   \end{beamercolorbox}
%   \begin{beamercolorbox}{section in head/foot}
%     \vskip1pt\insertsectionnavigationhorizontal{\paperwidth}{}{}\vskip1pt
%   \end{beamercolorbox}%
%   \ifbeamer@theme@subsection%
%     \begin{beamercolorbox}[colsep=1.5pt]{middle separation line head}
%     \end{beamercolorbox}
%     \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,%
%       leftskip=.3cm,rightskip=.3cm plus1fil]{subsection in head/foot}
%       \usebeamerfont{subsection in head/foot}\insertsubsectionhead
%     \end{beamercolorbox}%
%   \fi%
%   \begin{beamercolorbox}[colsep=1.5pt]{lower separation line head}
%   \end{beamercolorbox}
% }
% \makeatother

\setbeamertemplate{frametitle}{%
    \nointerlineskip%
    \begin{beamercolorbox}[wd=\paperwidth,ht=2.0ex,dp=0.6ex]{frametitle}
        \hspace*{1ex}\insertframetitle%
    \end{beamercolorbox}%
}

%\setbeamercolor{frametitle}{bg=darkcandyapplered!80!black!90!white}
%\setbeamertemplate{frametitle}{\bf\insertframetitle}

%\setbeamercolor{footnote mark}{fg=darkcandyapplered}
%\setbeamercolor{footnote}{fg=darkcandyapplered!70}
%\Raggedbottom
%\setbeamerfont{page number in head/foot}{size=\tiny}
%\usepackage[tracking]{microtype}

%\usepackage[sc,osf]{mathpazo}   % With old-style figures and real smallcaps.
%\linespread{1.025}              % Palatino leads a little more leading

% Euler for math and numbers
%\usepackage[euler-digits,small]{eulervm}
%\AtBeginDocument{\renewcommand{\hbar}{\hslash}}
\usepackage{graphicx,multirow,paralist,booktabs}

\mode<presentation> { \setbeamercovered{transparent} }

\setbeamertemplate{navigation symbols}{}
\makeatletter
\def\beamerorig@set@color{%
  \pdfliteral{\current@color}%
  \aftergroup\reset@color
}
\def\beamerorig@reset@color{\pdfliteral{\current@color}}
\makeatother

%=== GRAPHICS PATH ===========
\graphicspath{{./images/}}
% Marginpar width
%Marginpar width
%\setlength{\marginparsep}{.02in}

%% Captions
% \usepackage{caption}
% \captionsetup{
%   labelsep=quad,
%   justification=raggedright,
%   labelfont=sc
% }

\setbeamerfont{caption}{size=\footnotesize}
\setbeamercolor{caption name}{fg=darkcandyapplered}

%AMS-TeX packages

\usepackage{amssymb,amsmath,amsthm,mathtools} 
\usepackage{bm}
\usepackage{color}

%https://tex.stackexchange.com/a/31370/2269
\usepackage{mathtools,cancel}

\renewcommand{\CancelColor}{\color{red}} %change cancel color to red

\makeatletter
\let\my@cancelto\cancelto %copy over the original cancelto command
\newcommand<>{\cancelto}[2]{\alt#3{\my@cancelto{#1}{#2}}{\mathrlap{#2}\phantom{\my@cancelto{#1}{#2}}}}
% redefine the cancelto command, using \phantom to assure that the
% result doesn't wiggle up and down with and without the arrow
\makeatother

\usepackage{comment}
%\usepackage{hyperref,enumerate}
\usepackage{minitoc,array,enumerate}

\definecolor{slblue}{rgb}{0,.3,.62}
% \hypersetup{
%     colorlinks,%
%     citecolor=blue,%
%     filecolor=blue,%
%     linkcolor=blue,
%     urlcolor=slblue
% }

\usepackage{epstopdf}
\epstopdfDeclareGraphicsRule{.gif}{png}{.png}{convert gif:#1 png:\OutputFile}
\AppendGraphicsExtensions{.gif}

\usepackage{listings}

%%% TIKZ
\usepackage{forest}
\usepackage{tikz}
\usepackage{tikz-3dplot}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
% \usepackage{pgfgantt}
\usepackage{neuralnetwork}
\pgfplotsset{compat=newest}

\usetikzlibrary{fit,arrows,shapes,positioning,chains,shapes.geometric}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{shadows,automata}
\usetikzlibrary{patterns}
\usetikzlibrary{trees,mindmap,backgrounds}
%\usetikzlibrary{circuits.ee.IEC}
\usetikzlibrary{decorations.text}
% For Sagnac Picture
\usetikzlibrary{%
    decorations.pathreplacing,%
    decorations.pathmorphing%
}
\tikzset{no shadows/.style={general shadow/.style=}}
%
%\usepackage{paralist}

\tikzset{
  font=\Large\sffamily\bfseries,
  red arrow/.style={
    midway,red,sloped,fill, minimum height=3cm, single arrow, single arrow head extend=.5cm, single arrow head indent=.25cm,xscale=0.3,yscale=0.15,
    allow upside down
  },
  black arrow/.style 2 args={-stealth, shorten >=#1, shorten <=#2},
  black arrow/.default={1mm}{1mm},
  tree box/.style={draw, rounded corners, inner sep=1em},
  node box/.style={white, draw=black, text=black, rectangle, rounded corners},
}

%%% FORMAT PYTHON CODE
%\usepackage{listings}
% Default fixed font does not support bold face
\DeclareFixedFont{\ttb}{T1}{txtt}{bx}{n}{8} % for bold
\DeclareFixedFont{\ttm}{T1}{txtt}{m}{n}{8}  % for normal

% Custom colors
\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}

\usepackage{animate}

% Python style for highlighting
% \newcommand\pythonstyle{\lstset{
% language=Python,
% basicstyle=\footnotesize\ttm,
% otherkeywords={self},             % Add keywords here
% keywordstyle=\footnotesize\ttb\color{deepblue},
% emph={MyClass,__init__},          % Custom highlighting
% emphstyle=\footnotesize\ttb\color{deepred},    % Custom highlighting style
% stringstyle=\color{deepgreen},
% frame=tb,                         % Any extra options here
    % showstringspaces=false            % 
% }}

% % Python environment
% \lstnewenvironment{python}[1][]
% {
% \pythonstyle
% \lstset{#1}
% }
% {}

% % Python for external files
% \newcommand\pythonexternal[2][]{{
% \pythonstyle
% \lstinputlisting[#1]{#2}}}

% Python for inline
% 
% \newcommand\pythoninline[1]{{\pythonstyle\lstinline!#1!}}

%\usepackage{algorithm2e}

\newcommand{\eps}{\epsilon}
\newcommand{\bX}{\mb X}
\newcommand{\by}{\mb y}
\newcommand{\bbe}{\bm\beta}
\newcommand{\beps}{\bm\epsilon}
\newcommand{\bY}{\mb Y}

\newcommand{\osn}{\oldstylenums}
\newcommand{\dg}{^{\circ}}
\newcommand{\lt}{\left}
\newcommand{\rt}{\right}
\newcommand{\pt}{\phantom}
\newcommand{\tf}{\therefore}
\newcommand{\?}{\stackrel{?}{=}}
\newcommand{\fr}{\frac}
\newcommand{\dfr}{\dfrac}
\newcommand{\ul}{\underline}
\newcommand{\tn}{\tabularnewline}
\newcommand{\nl}{\newline}
\newcommand\relph[1]{\mathrel{\phantom{#1}}}
\newcommand{\cm}{\checkmark}
\newcommand{\ol}{\overline}
\newcommand{\rd}{\color{red}}
\newcommand{\bl}{\color{blue}}
\newcommand{\pl}{\color{purple}}
\newcommand{\og}{\color{orange!90!black}}
\newcommand{\gr}{\color{green!40!black}}
\newcommand{\lbl}{\color{CornflowerBlue}}
\newcommand{\dca}{\color{darkcandyapplered}}
\newcommand{\nin}{\noindent}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,thick,inner sep=1pt] (char) {\small #1};}}

\newcommand{\bc}{\begin{compactenum}[\quad--]}
\newcommand{\ec}{\end{compactenum}}

\newcommand{\p}{\partial}
\newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\dpd}[2]{\dfrac{\partial{#1}}{\partial{#2}}}
\newcommand{\pdd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}
\newcommand{\pde}[3]{\frac{\partial^2{#1}}{\partial{#2}\partial{#3}}}
\newcommand{\nmfr}[3]{\Phi\left(\frac{{#1} - {#2}}{#3}\right)}
\newcommand{\Err}{\text{Err}}
\newcommand{\err}{\text{err}}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

%%%% GREEK LETTER SHORTCUTS %%%%%
\newcommand{\la}{\lambda}
\renewcommand{\th}{\theta}
\newcommand{\al}{\alpha}
\newcommand{\G}{\Gamma}
\newcommand{\si}{\sigma}
\newcommand{\Si}{\Sigma}

\pgfmathdeclarefunction{poiss}{1}{%
  \pgfmathparse{(#1^x)*exp(-#1)/(x!)}%
  }

\pgfmathdeclarefunction{gauss}{2}{%
  \pgfmathparse{1/(#2*sqrt(2*pi))*exp(-((x-#1)^2)/(2*#2^2))}%
}

% \usepackage{pst-plot}

% \usepackage{pstricks-add}
% \usepackage{auto-pst-pdf}

% \psset{unit = 3}

% \def\target(#1,#2){%
%  {\psset{fillstyle = solid}
%   \rput(#1,#2){%
%     \pscircle[fillcolor = white](0.7,0.7){0.7}
%     \pscircle[fillcolor = blue!60](0.7,0.7){0.5}
%     \pscircle[fillcolor = white](0.7,0.7){0.3}
%     \pscircle[fillcolor = red!80](0.7,0.7){0.1}}}}
% \def\dots[#1](#2,#3){%
%     \psRandom[
%       dotsize = 2pt,
%       randomPoints = 25
%     ](!#2 #1 0.04 sub sub #3 #1 0.04 sub sub)%
%      (!#2 #1 0.04 sub add #3 #1 0.04 sub add)%
%      {\pscircle[linestyle = none](#2,#3){#1}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title[AI Trees: Intro]{ {\normalsize Artificial Intelligence for Tree Failure Identification and Risk Quantification}
  \\ Introductory Meeting}
\date[\today]{\footnotesize \today}
\author{{\bf NARS Lab}}
\institute[UMass Amherst]{
%\titlegraphic{\hfill
  \begin{tikzpicture}[baseline=(current bounding box.center)]
    \node[anchor=base] at (-7,0) (its) {\includegraphics[scale=.3]{UMassEngineering_vert}} ;
  \end{tikzpicture}
  % \hfill\includegraphics[height=1.5cm]{logo}
}

%https://tex.stackexchange.com/questions/55806/mindmap-tikzpicture-in-beamer-reveal-step-by-step
  \tikzset{
    invisible/.style={opacity=0},
    visible on/.style={alt={#1{}{invisible}}},
    alt/.code args={<#1>#2#3}{%
      \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}} % \pgfkeysalso doesn't change the path
    },
  }

% https://tex.stackexchange.com/questions/446468/labels-with-arrows-for-an-equation
% https://tex.stackexchange.com/a/402466/121799
\newcommand{\tikzmark}[3][]{
\ifmmode
\tikz[remember picture,baseline=(#2.base)] \node [inner sep=0pt,#1](#2) {$#3$};
\else
\tikz[remember picture,baseline=(#2.base)] \node [inner sep=0pt,#1](#2) {#3};
\fi
}

\lstset{language=matlab,
                basicstyle=\scriptsize\ttfamily,
                keywordstyle=\color{blue}\ttfamily,
                stringstyle=\color{blue}\ttfamily,
                commentstyle=\color{gray}\ttfamily,
                morecomment=[l][\color{gray}]{\#}
              }

\begin{document}

\maketitle

% \begin{frame}
%   \frametitle{Second half of the semester}
%   \pause
%   Modules to be covered in the next 4 weeks: \\ \pause
%   \begin{enumerate}[<+->]\setcounter{enumi}{6}
%   \item Tree-based methods
%   \item Support Vector Machines (SVM)
%   \item Neural Networks
%   \item Unsupervised Learning
%     \begin{itemize}[<+->]
%     \item Principal components analysis (PCA)
%     \item Clustering analysis
%     \end{itemize}
%   \end{enumerate}

%   \pause

%   \textbf{Project}\pause
%   \begin{itemize}
%   \item Check Moodle for instructions; proposals due April 10
%   \end{itemize}
%   \pause

%   \medskip

%   \textbf{Final Exam}\pause
%   \begin{itemize}
%   \item Cumulative 24-hr open-book take-home exam.
%   \end{itemize}
% \end{frame}

\begin{frame}
  \frametitle{Outline for this lecture}\small
  \tableofcontents
\end{frame}

%\section{Introduction}

% \begin{frame}
%   \frametitle{Variable importance in random forests}

%   %   \begin{figure}[h!]
%   %     \centering
%   %     \visible<6->{\includegraphics[width=.75\textwidth, trim={4.5cm 8cm 4cm 4.4cm}, clip]{ESL15-5}}
%   %     \caption{}
%   % \end{figure}

% \end{frame}

% \begin{frame}
%   \frametitle{Introduction}

% \end{frame}

% \begin{frame}
%   \frametitle{Neural networks}
%   \pause

%   networks
% \end{frame}

\section{Introduction}
\begin{frame}
  \frametitle{Biological neuron}
  \pause

  \visible<2->{\begin{figure}[h!]
    \centering
    \includegraphics[width=.8\textwidth]{neuron}
    \caption{Biological neuron (Source: \url{https://cs231n.github.io/neural-networks-1/})}
  \end{figure}}

  \begin{itemize}[<+->]
  \item $\sim$86 billion neurons are found in the human nervous system
  \item These neurons are connected by 10$^{14}$ to $10^{15}$ synapses
  \item Each neuron receives input signals from its dendrites and outputs signals along a single axon
  \item The axon in turn connects to other neurons via synapses
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Computational neuron model}
  \pause
  \begin{figure}[h!]
    \centering

    \begin{tikzpicture}[>=stealth,scale=1]
      \visible<13->{\begin{scope}
        \clip[radius=2cm] (0,0) circle;
        \draw[thick,blue] (1.4,-2) -- (1.4,2) node [right,pos=.5] {$f$};
      \end{scope}}
      \visible<12->{\node (O) at (0,0) {$\sum_i w_i x_i + b$ };}
      \visible<11->{\node[above=5mm of O,blue] {\small\normalfont cell body};}
      \visible<10->{\node[blue, thick, draw, circle,minimum width=4cm] at (O) (N) {} ;}
      \visible<14->{\node[circle,fill=purple,minimum width=1mm,xshift=1.5mm] (A) at (2,0) {};}
      \visible<15->{\node[below right of=A,text width=8ex, yshift=-.5cm,purple,] {\small\normalfont activation function};}
      \visible<16->{\draw[->, green!50!black,thick] (A) -- +(3,0) node[below,pos=.5] {\small\normalfont output axon};}
      \visible<17->{\path[->, green!50!black,thick] (A) -- +(3,0) node[above,pos=.5] {$f\lt(\sum_iw_ix_i + b\rt)$};}
      \visible<8->{\draw[->, thick] (-5,0) -- (-2,0) node[above,pos=.5] {$w_1x_1$};}
      \visible<9->{\draw[<-, thick] (N.225) -- (225:5cm) node [above,sloped,pos=.5] {$w_2x_2$};}
      \visible<6->{\draw[<-, thick,orange] (N.135) -- (135:5cm)   node[below,sloped,pos=.5] {\small \normalfont dendrite};}
      \visible<7->{\path[] (N.135) -- (135:5cm)  node[above,sloped,pos=.5] {$w_0x_0$};}
      \visible<4->{\path[] (N.135) -- (135:5cm)  node[opacity=1,circle,fill=red, minimum width=1mm] (W) {};
      \node[above] at (W) (S) {$w_0$};}
      \visible<5->{\node[below right,xshift=2mm,yshift=-2mm] at (S) {\rd \small\normalfont synapse};}
      \visible<2->{\draw[<-, thick,green!50!black ] (135:5cm) (W) -- +(-3,0) node[below,pos=.5] {\small\normalfont axon from a neuron};}
      \visible<3->{\draw[<-, thick,green!50!black ] (135:5cm) (W) -- +(-3,0) node[above,pos=.5] {$x_0$};}
      % \end{scope}
    \end{tikzpicture}

    %\caption{Computational neuron components}
  \end{figure}

\end{frame}

\begin{frame}
  \frametitle{Neural networks - basic ML terminology}
  \pause

  \begin{itemize}[<+->]
  \item $x_i$: \pause  signals traveling along axons (inputs)
  \item $w_i$: \pause  measure of synaptic strength, which is learned;\\ \pause
    \begin{itemize}[<+->]\item 
$w_i > 0 \rightarrow$ excitory influence
\item    $w_i <0 \rightarrow$ inhibitory influence
  \end{itemize}
  \item Dendrites carry signals $w_ix_i$ to the cell body, where they are summed.
  \item If the final sum $w_ix_i + b > t$ where $t$ is a threshold\footnote{intercept $b$ is referred to as the ``bias'' in ML literature},
   the neuron sends a spike along its axon (i.e.\ fires)
 \item Computationally, the firing rate of a neuron is represented by an \textbf{\pl activation function $f$}
 \item The output of a neuron is also called the \textit{\pl activation}
 \item   The output neurons have no activation function. Instead, they perform a final transformation of outputs from the penultimate layer
 \item  [Artificial] Neural networks (ANNs) are typically  modeled as connected layers of neuron in an acyclic graph (no loops).
  \item $N$-layer neural network (number of hidden layers $+$ output layer)
  \end{itemize}
\end{frame}

% \begin{frame}
%   \frametitle{Nomenclature}
%   \pause

%   [Artificial] Neural networks (ANNs) are modeled as connected layers of neuron in an acyclic graph (no loops).

%   \begin{itemize}[<+->]
%   \item ANNs are organized into layers of neurons (or ``units'')
%   \item Fully-connected layers are common
%   \item The basic ANN architecture with multiple hidden layers is called the \textbf{multilayer perceptron (MLP}
%   \item An ANN with only one hidden layer is called the \textbf{single layer perceptron}
%   \item $N$-layer neural network (number of hidden layers $+$ output layer)
%   \item The output neurons have no activation function. Instead, they perform a final transformation of outputs from the penultimate layer
%   \end{itemize}
% \end{frame}
% \begin{frame}
%   \frametitle{Examples of activation functions}
%   \pause
%   \begin{figure}[t!]
%     \begin{tikzpicture}[scale=1.7]
%       \begin{axis}[width=5cm,height=5.5cm,
%         ylabel=,
%         xlabel=$z$,ymin=-1.25,ymax=1.25,xmin=-5,xmax=5,
%         style={font=\tiny\normalfont},
%         legend style ={at={(1,.7)},anchor=west,align=left}
%         ]
%         \addplot[thick,blue,smooth] {1/(1+exp(-x))}; \addlegendentry{Logistic sigmoid: $\sigma(z)$}
%         \addplot[thick,red,smooth] {tanh(x)}; \addlegendentry{Hyperbolic tangent: $\tanh(z)$}
%         \addplot[thick,green,smooth,domain=-5:-0.001] {max(0,x)}; \addlegendentry{ReLU: $\max(0,z)$}
%         \addplot[thick,green,smooth,domain=0.001:5] {max(0,x)}; 
%         \end{axis}
%       \end{tikzpicture}

%     	% \caption[Sigmoidal activation functions.]{Common used activation functions include the logistic sigmoid $\sigma(z)$ and the hyperbolic tangent $tanh(z)$. More recently used activation functions are the softsign and the rectified hyperbolic tangent.}
%     	% \label{fig:sigmoid-tanh}
% \end{figure}
% \end{frame}

\begin{frame}[fragile]
  \frametitle{Three-layer neural network (with bias neurons)}
  \begin{figure}[h!]
    \centering
    \begin{neuralnetwork}[height=4]
    \newcommand{\x}[2]{$x_#2$}
    \newcommand{\y}[2]{$\hat{y}_#2$}
    \newcommand{\hfirst}[2]{\small $h^{(1)}_#2$}
    \newcommand{\hsecond}[2]{\small $h^{(2)}_#2$}
    \inputlayer[count=3, bias=true, title=I, text=\x]
    \hiddenlayer[count=4, bias=true, title=H1, text=\hfirst] \linklayers
    \hiddenlayer[count=3, bias=true, title=H2, text=\hsecond] \linklayers
    \outputlayer[count=2, title=O, text=\y] \linklayers
  \end{neuralnetwork}
\end{figure}

\pause
  \begin{itemize}[<+->]
  \item \textbf{Layers}: 3; \pause  \textbf{Hidden layers:} 2
  \item \textbf{Neurons}: 9
  \item \textbf{Learnable parameters:} $(4\times 4) + (5\times 3) + (4\times 2) = \pause 39$ weights
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Matrix operations in neural networks}
  \pause
%  The activation in the next layer can be constructed as a matrix operation as follows.\pause

  Given the activation vector ($m+1$ neurons) in the zeroth (input) layer:
  \begin{equation}
    a^{(0)} =   \begin{bmatrix}
    a_0^{0}\\[2mm]
    a_1^{0}\\[2mm]
    \vdots \\[2mm]
    a_m^{0}\\
    \end{bmatrix}
  \end{equation}
  \pause
  Then the activations in the next layer ($j+1$ neurons) are given by:\pause
  \begin{equation}
    a^{(1)}=
    \sigma\left(
    \boldsymbol{W}a^{0}+b
  \right)
  = \pause
  \sigma \left(
    \begin{bmatrix}
    w_{0,0} & w_{0,1} & \cdots & w_{0,k}\\
    w_{1,0} & w_{1,1} & \cdots & w_{1,k}\\
    \vdots & \vdots & \ddots & \vdots \\
    w_{j,0} & w_{j,1} & \cdots & w_{j,k}\\
    \end{bmatrix}
    \, 
    \begin{bmatrix}
    a_0^{0}\\
    a_1^{0}\\
    \vdots \\
    a_n^{0}\\
    \end{bmatrix}
    +
    \begin{bmatrix}
    b_0\\
    b_1\\
    \vdots \\
    b_n\\
    \end{bmatrix}
    \right)
  \end{equation}
  \pause
  Example: If Layer 1 had only two neurons, then the weight matrix $\bm W$ would have only 2 rows.
\end{frame}

\section{Training overview}
\begin{frame}
  \frametitle{Another look at the neuron}
  \pause
  Given the notation formalization, we revisit the neuron.\pause

  \begin{center}
  \begin{tikzpicture}[>=latex,scale=1.2]
\path
(0,0)     node[circle,draw,scale=2,inner sep=2pt] (S) {$\Sigma$}
% +(90:2.5) node[circle,draw,inner sep=2.5pt,opacity=0] (b) {}
%           node[above=1mm] {}
+(-3.5,1.5)  node[circle,fill=green!50]  (x1) {$x_0$}
+(-3.5,0)    node[circle,fill=green!50]  (x2) {$x_1$}
+(-3.5,-1.5) node[circle,fill=green!50]  (x3) {$x_2$}
(2,0)    node[draw] (g) {$\sigma$} node[above=3mm]{\small\normalfont Activation}
+(15:3)  node[circle,fill=red!50]  (y1) {$\hat{y}_0$}
+(-15:3) node[circle,fill=red!50]  (y2) {$\hat{y}_1$};
\draw[->] (S)--(g);
%\draw[->] (b)--(S);
\draw[->] (g)--(y1) node[pos=.6,above]{$w_{0,1}^{(2)}$};
\draw[->] (g)--(y2) node[pos=.6,below]{$w_{1,1}^{(2)}$};
\draw[->] (x1)--(S) node[pos=.4,above]{$w_{0,0}^{(1)}$};
\draw[->] (x2)--(S) node[pos=.4,above]{$w_{0,1}^{(1)}$};
\draw[->] (x3)--(S) node[pos=.4,above]{$w_{0,2}^{(1)}$};
\draw[blue] (1,0) circle(2);
\end{tikzpicture}
\end{center}

\end{frame}

\begin{frame}
  \frametitle{Neural network notation}
  \pause

  The {\bf \bl activation (output)} of a neuron is denoted:\pause
  \begin{equation}\bl
    \sigma(w_1a_1 + w_2a_2 + \cdots + w_n a_n + b) = \sigma(\sum w_ia_i + b) = \text{new neuron}
  \end{equation}
  \pause
  Further, we denote each activation as $a_{neuron}^{(layer)}$, e.g.\pause
  \begin{itemize}[<+->]
  \item $a_3^{(1)}$: fourth neuron in first layer (layers are counted from first hidden layer)
  \end{itemize}
  \pause

  \bigskip

  \begin{minipage}[t]{.47\linewidth}
  \textbf{\rd Weights} are denoted as $\rd w_{to,from}$, e.g.\pause
  \begin{itemize}[<+->]
  \item $w_{1,2}^2$: from neuron 3 to neuron 2 in second layer
  \item The superscript is not often used, as it is clear from the context which layer we are dealing with
  \end{itemize}
\end{minipage}  \pause
\begin{minipage}[t]{.47\linewidth}
  \begin{figure}[h!]
    \centering
    \includegraphics<9->[width=.8\textwidth]{nn-1-3}
  \end{figure}
\end{minipage}
\end{frame}

\begin{frame}
  \frametitle{Training a neural network}\pause

  % Let us describe a single hidden layer neural network using the following equations:\pause

  % \begin{align}
  %   Z_m &= \sigma(b_m + w_m^TX), \quad m = 1, \ldots, M \\
  %   T_k &= b_k + w_k^TZ, \quad k =1\ldots, K \\
  %   y_k(X) &= g_k(T), \quad k= 1,\ldots, K
  % \end{align}

  % \begin{itemize}[<+->]
  % \item $p$ is number of input units/neurons
  % \item $M$ is number of hidden units/neurons
  % \item $K$ is number of output units/neurons
  % \item Note: $w_m^TX = \sum_i w_{mi}x_i$
  % \item Total number of learnable parameters: $M(p+1) + K(M+1)$ weights
  % \end{itemize}

  \begin{itemize}[<+->]
  \item A neural network is trained or fitted by \textit{learning} the optimal values of the weights.
  \item This learning is done via optimization (e.g.\ gradient descent)
  \item Gradient descent: \pause
    \begin{equation}
      w^{\text{new}} = w^{\text{old}} - \eta \pd{C}{w^{\text{old}}}
    \end{equation}
    \pause
    where:
    \begin{itemize}[<+->]
    \item $\eta$ is the learning rate
    \item $C$ is the cost function (e.g.\ residual sum of squares):\pause
      \begin{equation}
        C = \sum_{i=1}^n (y_i - \hat y_i)^2
      \end{equation}
    \item $w$ the weight
  \end{itemize}

  \item In neural networks, the gradients are computed via \textbf{\bl backpropagation}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Neural network loss function}
  \pause
  Given $k$ output neurons and $i$ observations (where $\theta$ represents the weights and $f_k$ the output),
  we can compute the loss functions as follows. \pause

  \medskip

  \textbf{For regression:}
  \begin{equation}
    R(\theta) = \sum_{k=1}^K\sum_{i=1}^n(y_{ik} - f_k(x_i))^2
  \end{equation}

  \pause

  \medskip

  \textbf{For classification}, we use the cross-entropy (deviance):\pause
  \begin{equation}
    R(\theta) = - \sum_{i=1}^n\sum_{k=1}^Ky_{ik} \log f_k(x_i))
  \end{equation}

  \pause

  In modern ML terminology, this loss is usually referred to as a cost function $C$. \pause

  Thus, we can write, where $k=1$:

  \begin{equation}
    C = \sum_{i=1}^n (y_i - \hat y_i)^2
  \end{equation}
\end{frame}

\begin{frame}
  \frametitle{Summary of backpropagation}\pause

  \begin{enumerate}[<+->]
  \item Fix initial weights $w^{(l),0}$, $b^{(l),0}$ and perform a forward sweep/pass through the network computing the activations $a$ (outputs) of each layer $l$ as:\pause
    \begin{equation}
      a^{(l)} = \sigma(\bm W^{(l)} a^{(l-1)} + b^{(l)})
    \end{equation}
  \item At the output layer, we compute the cost function $C$ (what we want to minimize)
  \item Then, we {\it backpropagate} the errors through each layer in order to compute the gradients $\pd{C}{w^{(l)}}$, $\pd{C}{b^{(l)}}$ \pause
    and  weight updates  $w^{(l),r+1}$ and $b^{(l),r+1}$
\item Repeat the forward and backward passes until cost is sufficiently minimized
  \end{enumerate}

\end{frame}

\begin{frame}
  \frametitle{Example: backpropagation for 3-layer network}
  \pause
%  Consider the ANN below. What are the cost function gradients? \pause
\visible<2->{  \begin{figure}[h!]
    \centering
    \includegraphics[width=.6\textwidth]{4layer-nn}
  \end{figure}
}
\pause
\vspace{-2ex}
\begin{align}
 \pd{C}{w^{(3)}} &=  {\rd \pd{C}{a^{(3)}}\pd{a^{(3)}}{z^{(3)}}}\pause \pd{z^{(3)}}{w^{(3)}}\\\pause
\pd{C}{w^{(2)}} &=  {\rd \pd{C}{a^{(3)}}\pd{a^{(3)}}{z^{(3)}}} \pause {\bl \pd{z^{(3)}}{a^{(2)}}\pd{a^{(2)}}{z^{(2)}}} \pause                   
                    \pd{z^{(2)}}{w^{(2)}}\\\pause  
\pd{C}{w^{(1)}} &=  {\rd \pd{C}{a^{(3)}}\pd{a^{(3)}}{z^{(3)}}} \pause {\bl \pd{z^{(3)}}{a^{(2)}}\pd{a^{(2)}}{z^{(2)}}} \pause                   
                    \pd{z^{(2)}}{a^{(1)}}                          
                    \pd{a^{(1)}}{z^{(1)}}\pd{z^{(1)}}{w^{(1)}}
\end{align}
\end{frame}

 \begin{frame}
  \frametitle{Batch learning}
  The gradient descent method was described without referencing the observations. \pause

  \medskip

  In reality, the forward and backward passes are performed for each observation in the training set, with parameter updates obtained by
  \textbf{\bl averaging} the gradients: \pause

  \begin{align}
    w^{(l), r+1}  &= \pause  w^{(l),r} - \pause \eta{\bl  \fr1n \sum_{i=1}^n}\pd{C_i}{w^{(l)}}  \\\pause
    b^{(l), r+1}  &=  \pause b^{(l),r} - \pause \eta {\bl \fr1n \sum_{i=1}^n} \pd{C_i}{b^{(l)}}  
  \end{align}

  \pause

  This approach is called \textbf{\bl batch learning}. \pause

  \begin{itemize}[<+->]
  \item One sweep through all the training observations is called an \textit{epoch}
  \item In batch learning, only one update results from a training epoch
  \item Thus, several epochs are required for convergence
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Stochastic gradient descent}
  \pause

  In standard gradient descent (batch learning), the updates are performed only after the gradient is computed for \textit{all} training observations.

  \pause

  \medskip

  The stochastic gradient descent approach approximates the gradient using a {\rd randomly selected}  observation: \pause

    \begin{align}
    w^{(l), r+1}  &= \pause  w^{(l),r} - \pause \eta \pd{C_{\rd i}}{w^{(l)}}  \\\pause
    b^{(l), r+1}  &=  \pause b^{(l),r} - \pause \eta  \pd{C_{\rd i}}{b^{(l)}}  
  \end{align}

  \pause

  \begin{itemize}[<+->]
  \item Each iteration over observation $i$ results in a weight/bias update
  \item Thus, one sweep through the entire training set (an epoch) produces $n$ updates
  \end{itemize}
  \pause
  This procedure is also known as \textbf{\rd online learning}
\end{frame}

\begin{frame}
  \frametitle{Mini-batch learning}
  \pause
  To introduce stability, we can compute weight updates over a \textit{\pl subset} of training observations

  \begin{enumerate}[<+->]
  \item Randomly partition the training set into $M$ mini-batches, each with $m$ observations
  \item Perform a forward and backward pass through each mini-batch to update the weights:\pause
      \begin{align}
    w^{(l), r+1}  &= \pause  w^{(l),r} - \pause \eta{\pl  \fr1m \sum_{i=1}^m} \pd{C_{\pl i}}{w^{(l)}}  \\\pause
    b^{(l), r+1}  &=  \pause b^{(l),r} - \pause \eta{\pl  \fr1m \sum_{i=1}^m}  \pd{C_{\pl i}}{b^{(l)}}  
  \end{align}
  \pause
  Thus, for each iteration, average the gradient over a \textit{\pl mini-batch} of $m$ observations
\item Repeat step 2 until convergence
  \end{enumerate}  
\end{frame}

\begin{frame}
  \frametitle{Performance and robustness considerations }
  \pause
  \begin{itemize}[<+->]
  \item Online learning is more efficient for very large datasets
  \item In practice, mini-batch learning is employed, as batch sizes (usually, $m=16$ or $m=32$) can be chosen to take advantage of parallel computing architectures
  \item An \textit{\bf \rd adpative} learning rate $\eta$ can guarantee convergence, e.g. $\rd \eta_r = \fr1r$
  \item \textit{Overfitting} can be mitigated via \textbf{\bl weight decay} (regularization of weights):\pause
    \begin{equation}\bl 
      C \leftarrow \pause C + \la J = \pause C + \la\lt(\sum w^2 + \sum b^2\rt)
    \end{equation}
    \pause where $\la$ is tuning parameter estimated via cross-validation\\
    \textit{Early stopping} and \textit{dropout} are also used.
  \item Input \textbf{standardization} (mean zero, SD 1) is recommended for consistent weight initialization and regularization
  \item Weights/biases are initialized to \textit{small} values near 0 for better performance
  \item Cost function $C$ is nonlinear and nonconvex; other optimization approaches (e.g.\ conjugate gradient) can provide faster convergence compared to
    stochastic gradient descent
  \end{itemize}
\end{frame}

%\section{Further architectures}
\begin{frame}
  \frametitle{Other types of neural networks}
  \pause

  The standard ANN architecture (MLP) we have studied is also called the feed-forward network.

  \medskip

  Other architectures have been shown to give better performance for various applications: \pause

  \medskip
  \begin{itemize}[<+->]
  \item Recurrent neural networks (RNNs): time-series forecasting

  \item Convolutional neural networks (CNNs): image classification

  \item Long short-term memory networks (LSTMs): time-series, pattern identification, etc.
  \end{itemize}
\end{frame}

\section{Convolutional NNs}
\begin{frame}
  \frametitle{The convolutional neural network (CNN)}
  \pause

  \begin{itemize}
  \item Motivated by the image recognition process of the brain's visual cortex.

  \item Groundbreaking study on cats revealed the importance of \textit{local receptive fields} for activating neurons
    in the visual cortex. (Hubel \& Wiesel, 1958; 1959)

  \item Earliest neural network for image recognitron introduced: \textit{neocognitron} (Fukushima, 1980)

  \item Milestone: introduction of \textit{LeNet-5} architecture for handwritten digit recognition (Yann LeCun et al., 1998)
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Building blocks of a CNN}

  \begin{itemize}
  \item \textbf{Input layer:} the image to be classified 
  \item \textbf{Convolutional layer:} represents the action of a filter transmitting signals (features) from various
    portions (receptive fields) of the preceding layer. The size of the receptive field is specified by the
    \textit{convolutional kernel}. Each layer can have multiple feature maps representing different filters.
  \item \textbf{Pooling layer:} subsamples signals from preceding layer to reduce dimensionality and extract dominant
    features (subsample space determined by kernel size)
  \item \textbf{Dense layer:} neuron outputs are flattened and fully connected %(usually with decreasing number of neurons)
  \item \textbf{Output layer:} neurons equal to number of classes; with softmax activation
  \end{itemize}

  \begin{center}
    \includegraphics[width=.6\textwidth]{cnn-ex1}
  \end{center}

  {\tiny Source: \url{https://neurdiness.wordpress.com/2018/05/17/deep-convolutional-neural-networks-as-models-of-the-visual-system-qa/}}

\end{frame}

\begin{frame}
  \frametitle{Training hyperparameters in a CNN}
  Several decisions must be made in selecting hyperparameters for training a CNN.

  \begin{itemize}
  \item Number of convolutional layers and feature maps in each layer
  \item Convolutional kernel size
  \item Stride length (spacing of filters)
  \item Choice of pooling function (max, average, etc)
  \item Number of dense layers
  \item Activation function in each layer (ReLU, tanh, etc)
  \end{itemize}

  Various high-performing architectures have been developed in recent years that can be adapted for other problems.

  Training the network involves finding the weights for the various layers (using mini-batch gradient descent or other variants)
\end{frame}

\begin{frame}
  \frametitle{Another CNN example}
  \begin{center}
    \includegraphics[width=.9\textwidth]{cnn-ex2}
  \end{center}
  {\tiny Source: \url{https://towardsdatascience.com/mnist-handwritten-digits-classification-using-a-convolutional-neural-network-cnn-af5fafbc35e9}}
\end{frame}

\section{AI for Tree Risk}
\begin{frame}
  \frametitle{Automating tree risk prediction using AI}

  \begin{itemize}
  \item \textbf{Objective:}  Automate tree risk classification using a convolutional neural network?
  \item \textbf{Question 1:} Can we elicit the visual features from a trained network that indicate various levels of risk?
  \item \textbf{Question 2:} Can we map estimated CNN weights to structural relationships governing tree stability?
  \item \textbf{Question 3:} How well can a trained CNN be transferred to another set of images and perform similarly well for risk prediction?
  \item \textbf{Question 4:} How can this process be adapted for Google Maps images (or drone photography)?
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{CNN Model 1}
  \begin{itemize}
  \item Cropped and resized images to $108\times 108\times 1$ (grayscale)
  \item Binary classification; 103 probable/possible and 259 improbable
  \item Used a simple architecture (5 convolutional layers, 2 hidden dense layers, 1 output layer)
%     \begin{quote}\tiny
%  \begin{verbatim}
% model = models.Sequential([
%     layers.Conv2D(64, 7, activation="relu", padding="same", input_shape = input_image_shape),
%     layers.MaxPooling2D(2),
%     layers.Conv2D(128, 3, activation="relu", padding="same"),
%     layers.Conv2D(128, 3, activation="relu", padding="same"),
%     layers.MaxPooling2D(2),
%     layers.Conv2D(256, 3, activation="relu", padding="same"),
%     layers.Conv2D(256, 3, activation="relu", padding="same"),
%     layers.MaxPooling2D(2),
%     layers.Flatten(),
%     layers.Dense(128, activation="relu"),
%     layers.Dropout(0.5),
%     layers.Dense(64, activation="relu"),
%     layers.Dropout(0.5),
%     layers.Dense(2, activation="softmax")
% ])
% \end{verbatim}
%        \end{quote}
  \item Validation accuracy: 74\% (10 epochs)
  \item Increasing input image size to  $216\times 216$ worsens outcome with accuracy of 68\%
  \item If the stride in the first convolutional layer is increased to 2 (with the higher resolution), this does not improve the result.
  \item However, with a stride of 2 and $108\times 108$ image resolution, we obtain a validation accuracy of 77\%.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{CNN Model 2}

  \begin{itemize}
  \item Images are cropped but not converted to grayscale.
  \item Three classes: 40 probable, 63 possible , 259 improbable
  \item CNN architecture unchanged
  \item Validation accuracy: 74\% (10 epochs)        
  \end{itemize}
\end{frame}
\appendix
\section{Backpropagation}

\begin{frame}
  \frametitle{Equation summary: outer layer}
  \pause
  At the outer layer $L$ (without indexing by neuron):\pause
  \begin{align}
    z^{(L)} &= w^{(L)} \times a^{(L-1)} + b^{(L)} \\\pause
    a^{(L)} &= \sigma(z^{(L)}) \\\pause
    C &= (a^{(L)}  - y)^2
  \end{align}
  \pause

  The gradient of the cost function with respect to $w^{(L)}$ is:\pause
  \begin{equation}
    \frac{\partial C}{\partial w^{(L)}}
    =\pause
    \frac{\partial C}{\partial a^{(L)}}
    \frac{\partial a^{(L)}}{\partial z^{(L)}}
    \frac{\partial z^{(L)}}{\partial w^{(L)}}
    =\pause
    2 \left(a^{(L)} - y \right) \sigma' \left(z^{(L)}\right) a^{(L-1)}
  \end{equation}
  \pause
  Thus, we see that this gradient depends on the activation from the previous layer $a^{(L-1)}$. \pause Also wrt to the bias:
  \begin{equation}
    \pd{C}{b^{(L)}} = \pause \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{b^{(L)}} =\pause 2\lt(a^{(L)} - y\rt)\sigma'(z^{(L)})(1)
  \end{equation}
\end{frame}

\begin{frame}
  \frametitle{Updating weights}
  \pause
  We can then update the weights for the last layer for the next iteration $r+1$:\pause
  \begin{align}
    w^{(L), r+1}  &=  w^{(L),r} - \eta \pd{C}{w^{(L)}}  \\\pause
    b^{(L), r+1}  &=  b^{(L),r} - \eta \pd{C}{b^{(L)}}  
  \end{align}
  \pause
  To update the weights for layer $L-1$, we need to find the gradients $\pd{C}{w^{(L-1)}}$ and $\pd{C}{b^{(L-1)}}$.

  \pause

  \medskip

  Using the chain rule again, we write:\pause
  \begin{align}
    \pd{C}{w^{(L-1)}} &= {\rd \pd{C}{a^{(L-1)}}}\pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{w^{(L-1)}} \\\pause
    \pd{C}{b^{(L-1)}} &= {\rd \pd{C}{a^{(L-1)}}}\pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{b^{(L-1)}} 
  \end{align}
\end{frame}

\begin{frame}
  \frametitle{Backward pass}
  \pause
  But we recall that $C$ is not \textit{explicitly} dependent on $a^{(L-1)}$ as $C = (a^{(L)} - y)^2$.
  \pause

  However, it is \textit{implicitly} dependent, since\pause
  \begin{equation}
    C \propto a^{(L)},
  \end{equation}
  \pause
  \begin{equation}
    a^{(L)} \propto z^{(L)} 
  \end{equation}
  \pause
  and\pause
  \begin{equation}
    z^{(L)} \propto a^{(L-1)}
  \end{equation}
  \pause

  So, we use the chain rule to expand $\pd{C}{a^{(L-1)}}$ as follows:\pause
  \begin{equation}\rd
    \pd{C}{a^{(L-1)}} = \pause \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}}
  \end{equation}
\end{frame}

\begin{frame}
  \frametitle{Backward pass (cont.)}
  \pause
  We can then expand the cost function gradient wrt to weights for layer $L-1$ as:\pause
  \begin{align}
    \pd{C}{w^{(L-1)}} &= {\rd \pd{C}{a^{(L-1)}}}\pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{w^{(L-1)}} \pause
                        = {\rd \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}}} \pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{w^{(L-1)}} \\[2mm]\pause
        \pd{C}{b^{(L-1)}} &= {\rd \pd{C}{a^{(L-1)}}}\pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{b^{(L-1)}} 
                            = {\rd \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}}} \pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{b^{(L-1)}} 
  \end{align}

  \pause

  Once these gradients are computed, we update the weights for the $r+1$th iteration using: \pause
   \begin{align}
    w^{(L-1), r+1}  &= \pause  w^{(L-1),r} - \pause \eta \pd{C}{w^{(L-1)}}  \\\pause
    b^{(L-1), r+1}  &=  \pause b^{(L-1),r} - \pause \eta \pd{C}{b^{(L-1)}}  
  \end{align}
\end{frame}

\begin{frame}
  \frametitle{Summary: forward pass}
  \begin{enumerate}[<+->]
  \item Initialize weights and biases: $w^{(l),0}$, $b^{(l),0}$
  \item Perform forward pass to compute activations:
    \begin{align}
      z^{(l),0} &= w^{(l),0} \times a^{(l-1),0} + b^{(l),0} \\\pause
    a^{(l)} &= \sigma(z^{(l),0}) 
    \end{align}
    \pause
    At output layer:
    \begin{align}      
      z^{(L),0} &= w^{(L),0} \times a^{(L-1),0} + b^{(L),0} \\\pause
      a^{(L),0} &= \sigma(z^{(L),0}) \\
      C &= (a^{(L),0}  - y)^2
    \end{align}
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{Summary: backward pass---outer layer }
  \begin{enumerate}[<+->]\setcounter{enumi}{2}
  \item Backward pass, outer layer $(L)$:\pause
    \begin{enumerate}[<+->]
    \item Compute gradients: \pause
      \begin{align}
      \pd{C}{w^{(L)}} &= \pause \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{w^{(L)}}\\\pause
      \pd{C}{b^{(L)}} &= \pause \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{b^{(L)}}\pause
      \end{align}
    \item Update weights: \pause
      \begin{align}
        w^{(L), 1}  &=  w^{(L),0} - \eta \pd{C^0}{w^{(L)}}  \\\pause
      b^{(L), 1}  &=  b^{(L),0} - \eta \pd{C^0}{b^{(L)}}  
      \end{align}
  \end{enumerate}
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{Summary: backward pass---last hidden layer}
  \begin{enumerate}[<+->]\setcounter{enumi}{2}
  \item Backward pass, layer $(L-1)$:\pause
    \begin{enumerate}[<+->]\setcounter{enumii}{2}
    \item Compute gradients: \pause
    \begin{align}
      \pd{C}{w^{(L-1)}} &=  \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}}  \pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{w^{(L-1)}}\\\pause
      \pd{C}{b^{(L-1)}} &=   \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}} \pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{b^{(L-1)}}
    \end{align}
    \pause
  \item Update weights: \pause
    \begin{align}
      w^{(L-1), 1}  &= \pause  w^{(L-1),0} - \pause \eta \pd{C}{w^{(L-1)}}  \\\pause
      b^{(L-1), 1}  &=  \pause b^{(L-1),0} - \pause \eta \pd{C}{b^{(L-1)}}  
    \end{align}
  \end{enumerate}
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{Summary: backward pass---second-to-last hidden layer}
  \begin{enumerate}[<+->]\setcounter{enumi}{2}
  \item Backward pass, layer $(L-2)$:\pause
    \begin{enumerate}[<+->] \setcounter{enumii}{4}
    \item Compute gradients: \pause
    \begin{align}
      \pd{C}{w^{(L-2)}} &=  \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}} \pause
                        \pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{a^{(L-2)}} \pause
                        \pd{a^{(L-2)}}{z^{(L-2)}}\pd{z^{(L-2)}}{w^{(L-2)}}\\\pause
      \pd{C}{b^{(L-2)}} &=   \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}} \pause
                        \pd{a^{(L-1)}}{z^{(L-1)}}\pd{z^{(L-1)}}{a^{(L-2)}} \pause
                        \pd{a^{(L-2)}}{z^{(L-2)}}\pd{z^{(L-2)}}{b^{(L-2)}}\\\pause
    \end{align}
    \pause
  \item Update weights: \pause
    \begin{align}
      w^{(L-2), 1}  &= \pause  w^{(L-2),0} - \pause \eta \pd{C}{w^{(L-2)}}  \\\pause
      b^{(L-2), 1}  &=  \pause b^{(L-2),0} - \pause \eta \pd{C}{b^{(L-2)}}  
    \end{align}
  \end{enumerate}
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{Summary: backward pass---first hidden layer}
  \begin{enumerate}[<+->]\setcounter{enumi}{2}
  \item Backward pass, layer $(1)$:\pause
        \begin{enumerate}[<+->] \setcounter{enumii}{4}
    \item Compute gradients: \pause
    \begin{align}
      \pd{C}{w^{(1)}} &=  \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}}\pd{a^{(L-1)}}{z^{(L-1)}} \pause
                          \pause\cdots  
                          \pd{z^{(2)}}{a^{(1)}}                          
                        \pd{a^{(1)}}{z^{(1)}}\pd{z^{(1)}}{w^{(1)}}\\\pause
      \pd{C}{b^{(1)}} &=  \pd{C}{a^{(L)}}\pd{a^{(L)}}{z^{(L)}}\pd{z^{(L)}}{a^{(L-1)}}\pd{a^{(L-1)}}{z^{(L-1)}} \pause
                          \pause\cdots  
                          \pd{z^{(2)}}{a^{(1)}}                          
                        \pd{a^{(1)}}{z^{(1)}}\pd{z^{(1)}}{b^{(1)}}\\\pause
    \end{align}
    \pause
  \item Update weights: \pause
    \begin{align}
      w^{(1), 1}  &= \pause  w^{(1),0} - \pause \eta \pd{C}{w^{(1)}}  \\\pause
      b^{(1), 1}  &=  \pause b^{(1),0} - \pause \eta \pd{C}{b^{(1)}}  
    \end{align}
  \end{enumerate}
  \end{enumerate}
\end{frame}
% \section{Summary}
% \begin{frame}[fragile]
%   \frametitle{Summary}
%   \pause
%   Today, we considered an introduction to neural networks. We will finalize this on Friday with real examples.
%   \begin{itemize}[<+->]
%   \item For now, you may play around with \href{http://playground.tensorflow.org}{\bl \bf this}
%     \item Reading: ESL 11.3 -- 11.6 (complete before next lecture); you will notice differences in notation used.
%   \end{itemize}
% \end{frame}

% \begin{frame}
%   \frametitle{Reading}\pause
%   Today's lecture:\pause

% \begin{itemize}[<+->]
% \item Support vector machines: ISLR 9.3, 9.4; ESL 12.3.1
% \item 
%   % \item Boosting: ESL 10 (Very dense chapter. You may want to focus on 10.1--10.3, 10.7, 10.10, 10.13, 10.14.)
% \end{itemize}
% \pause
% \bigskip

% \end{frame}

%http://www.di.fc.ul.pt/~jpn/r/rbf/rbf.html
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2021-01-13 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
 %\vspace{-16ex}
-\title{Predicting tree risks to power lines using a convlutional neural network}
+\title{Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network}
 %\author{}
@@
 \section{Introduction}
-\hl{@Nasko: write first draft of introduction/motivation for the work}
+Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions).
+In general, tree-related outages are due to contact between tree parts and lines.
+Sometimes, the contact is due to growth of branches into the lines; sometimes it is due to branch failure; sometimes it is due to tree failure (either uprooting or trunk failure).
+When outages occur, the economic costs can be substantial.
+From 2005 – 2015, Graziano et al.
+(2020) computed the annual economic cost of outages in the state of Connecticut; the mean was approximately \$8.3 bn.
+Outages due to contact from branch growth or branch failure can be mitigated by line clearance tree trimming.
+Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in the billions of dollars (Guggenmoos 2003).
+Despite the comparatively high annual cost of tree trimming, studies have repeatedly shown that trimming branches away from the wires reduces the number of outages (sometimes classified as “preventable” outages).
+For example, in Massachusetts, Simpson and Van Bossuyt (1996) reported that tree failure caused 40\% of preventable tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\%.
+In a more recent (and sophisticated) analysis in Connecticut, Parent et al.
+(2019) showed that enhanced tree trimming reduced outages during storms.
+And using a statistical model to predict outages based on data from two Gulf Coast states in the USA, Nateghi et al.
+(2014) demonstrated that prediction models were less uncertain when they included the effect of tree trimming; models including wind speed without considering the effect of trimming were not as accurate.
+
+Trimming can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging the lines in proximity are removed, reducing the likelihood of contact. 
+% Trimming also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of whole-tree failure, whether by uprooting or stem breakage. 
+% Although the annual cost of trimming is comparatively large, the economic cost of outages can far exceed it.
+A recent benefit cost analysis demonstrated a highly favorable benefit cost ratio for trimming relative to the economic costs of outages (Graziano et al.
+2020).
+% Trimming,
+However, it cannot eliminate tree-related outages.
+Failure of trees away from the right-of-way can still impact the lines and cause outages (Guggenmoos 2003).
+The proportion of tree failures away from the wires that cause outages varies and has not been rigorously quantified.
+Guggenmoos (2011) estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and Wismer (2018) reported approximately 25\% of interruptions in Illinois, USA were caused by trees that uprooted or broke in the stem.
+Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed (Smiley et al.
+2017; Goodfellow 2020).
+Assessing risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences.
+Likelihood of failure depends on the anticipated loads on the tree and its load-bearing capacity.
+The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware—“infrastructure”—in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees.
+Severity of consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs and disruption associated with outages—this, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure.
+
+Assessing risk of individual trees can be costly because of the time required to complete the assessment.
+In some situations, a less time-consuming assessment may be justified to reduce costs—a ``Level 1'' assessment (Smiley et al.
+2017).
+Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA (Rooney et al.
+2005) and Florida, USA (Koeser et al.
+2016).
+The utility of Level 1 assessments in Rhode Island and Florida suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
+
+AI-based image analysis is relatively widely used, even in engineering applications such as earthquake risk assessment (Jiao and Alavi 2019; Salehi and Burgueno 2018) and structural health monitoring (Spencer et al.
+2019).
+A relevant application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated (Frick et al.
+2019).
+Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
+However, the flexibility and power of CNNs, appears promising.
+In this paper, we demons
+In the remainder of this paper, we first provide a background on CNN
+which is critical for sustainability in critical infrastructural systems.
+Our goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
 
 \section{Background}
-\hl{@Nasko: write first draft of literature review. I will edit.}
-
 \section{Data}
@@
 
-\section{Results}
-
-\section{Discussion}
+\section{Results and Discussion}
 

### 2021-01-13 docs/manuscript/manuscript.tex
updated bib file and fixed references in manuscript intro

@@
 \section{Introduction}
-Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions).
-In general, tree-related outages are due to contact between tree parts and lines.
-Sometimes, the contact is due to growth of branches into the lines; sometimes it is due to branch failure; sometimes it is due to tree failure (either uprooting or trunk failure).
-When outages occur, the economic costs can be substantial.
-From 2005 – 2015, Graziano et al.
-(2020) computed the annual economic cost of outages in the state of Connecticut; the mean was approximately \$8.3 bn.
-Outages due to contact from branch growth or branch failure can be mitigated by line clearance tree trimming.
-Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in the billions of dollars (Guggenmoos 2003).
-Despite the comparatively high annual cost of tree trimming, studies have repeatedly shown that trimming branches away from the wires reduces the number of outages (sometimes classified as “preventable” outages).
-For example, in Massachusetts, Simpson and Van Bossuyt (1996) reported that tree failure caused 40\% of preventable tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\%.
-In a more recent (and sophisticated) analysis in Connecticut, Parent et al.
-(2019) showed that enhanced tree trimming reduced outages during storms.
-And using a statistical model to predict outages based on data from two Gulf Coast states in the USA, Nateghi et al.
-(2014) demonstrated that prediction models were less uncertain when they included the effect of tree trimming; models including wind speed without considering the effect of trimming were not as accurate.
+Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been
+done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions). In general,
+tree-related outages are due to contact between tree parts and lines. Sometimes, the contact is due to growth of
+branches into the lines; sometimes it is due to branch failure; sometimes it is due to tree failure (either uprooting or
+trunk failure). When outages occur, the economic costs can be substantial. From 2005 to 2015, \cite{graziano2020wider}
+computed the annual economic cost of outages in the state of Connecticut; the mean was approximately \$8.3 bn.  Outages
+due to contact from branch growth or branch failure can be mitigated by line clearance tree trimming. Estimates vary,
+but the annual amount that utilities spend on tree trimming operations is typically in the billions of dollars
+\cite{guggenmoos2003effects}. Despite the comparatively high annual cost of tree trimming, studies have repeatedly shown
+that trimming branches away from the wires reduces the number of outages (sometimes classified as “preventable”
+outages). For example, in Massachusetts, \citet{simpson1996treecaused} reported that tree failure caused 40\% of
+preventable tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to
+30\%. In a more recent (and sophisticated) analysis in Connecticut, \citet{parent2019analysis} showed that enhanced tree
+trimming reduced outages during storms. And using a statistical model to predict outages based on data from two Gulf
+Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less uncertain when they
+included the effect of tree trimming; models including wind speed without considering the effect of trimming were not as
+accurate.
 
-Trimming can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging the lines in proximity are removed, reducing the likelihood of contact. 
-% Trimming also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of whole-tree failure, whether by uprooting or stem breakage. 
-% Although the annual cost of trimming is comparatively large, the economic cost of outages can far exceed it.
-A recent benefit cost analysis demonstrated a highly favorable benefit cost ratio for trimming relative to the economic costs of outages (Graziano et al.
-2020).
-% Trimming,
-However, it cannot eliminate tree-related outages.
-Failure of trees away from the right-of-way can still impact the lines and cause outages (Guggenmoos 2003).
-The proportion of tree failures away from the wires that cause outages varies and has not been rigorously quantified.
-Guggenmoos (2011) estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and Wismer (2018) reported approximately 25\% of interruptions in Illinois, USA were caused by trees that uprooted or broke in the stem.
-Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed (Smiley et al.
-2017; Goodfellow 2020).
-Assessing risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences.
-Likelihood of failure depends on the anticipated loads on the tree and its load-bearing capacity.
-The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware—“infrastructure”—in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees.
-Severity of consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs and disruption associated with outages—this, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure.
+Trimming can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging
+the lines in proximity are removed, reducing the likelihood of contact.
+% Trimming also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of
+% whole-tree failure, whether by uprooting or stem breakage.  Although the annual cost of trimming is comparatively
+% large, the economic cost of outages can far exceed it. A recent benefit cost analysis demonstrated a highly favorable
+% benefit cost ratio for trimming relative to the economic costs of outages (Graziano et al. 2020).  Trimming,
+However, it cannot eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the
+lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that cause
+outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of
+tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citet{wismer2018targeted}
+reported approximately 25\% of interruptions in Illinois, USA were caused by trees that uprooted or broke in the stem.
+Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been
+developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Assessing risk includes assessing the likelihood of
+failure, the likelihood of impact, and the severity of consequences. Likelihood of failure depends on the anticipated
+loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines,
+poles, and other hardware—“infrastructure”—in the case of utility tree risk assessment), the target’s occupancy rate
+(which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of
+consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs
+and disruption associated with outages—this, in turn, is partially related to the size of the tree or tree part that
+fails, and how much momentum it has when it impacts the infrastructure.
 
-Assessing risk of individual trees can be costly because of the time required to complete the assessment.
-In some situations, a less time-consuming assessment may be justified to reduce costs—a ``Level 1'' assessment (Smiley et al.
-2017).
-Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA (Rooney et al.
-2005) and Florida, USA (Koeser et al.
-2016).
-The utility of Level 1 assessments in Rhode Island and Florida suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
+Assessing risk of individual trees can be costly because of the time required to complete the assessment. In some
+situations, a less time-consuming assessment may be justified to reduce costs—a ``Level 1'' assessment (Smiley et
+al. 2017). Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk
+assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA
+\cite{koeser2016frequency} . The utility of Level 1 assessments in Rhode Island and Florida suggests that artificial
+intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
 
-AI-based image analysis is relatively widely used, even in engineering applications such as earthquake risk assessment (Jiao and Alavi 2019; Salehi and Burgueno 2018) and structural health monitoring (Spencer et al.
-2019).
-A relevant application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated (Frick et al.
-2019).
-Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
-However, the flexibility and power of CNNs, appears promising.
-In this paper, we demons
-In the remainder of this paper, we first provide a background on CNN
-which is critical for sustainability in critical infrastructural systems.
-Our goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
+AI-based image analysis is relatively widely used, even in engineering applications such as earthquake risk assessment
+\cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}. A relevant
+application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated
+\cite{fricker2019convolutional}. Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is
+complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and
+associated risk and local meteorological conditions. However, the flexibility and power of CNNs, appears promising. In
+this paper, we demons In the remainder of this paper, we first provide a background on CNN which is critical for
+sustainability in critical infrastructural systems. Our goal is to further demonstrate an innovative automated approach
+to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually
+improved through subsequent training on new datasets.
 
@@
 
+\printbibliography
+
 \appendix

### 2021-01-13 bin/python/label-image-files.py
created four classification scenarios in label-image-files.py

@@ -26,54 +26,98 @@ def getListOfFiles(dirName):
                 
     return allFiles
 
-def saveImageFilesBinary(image_file_list):
-    """Serially labels all images by class (probable or improbable) and saves them to data/tidy/labeled_images"""
-    probable_counter = 1
-    improbable_counter = 1
-    tidy_image_dir = '../../data/tidy/labeled_images_2_classes/'
-    if not os.path.exists(tidy_image_dir):
-        os.makedirs(tidy_image_dir)
-    for filename in image_file_list:
-        if '.JPG' in filename or '.jpg' in filename:        
-            if any(re.findall(r'improbable', filename, re.IGNORECASE)):
-                save_name = tidy_image_dir + 'improbable' + '-' + str(improbable_counter) + '.jpg'
-                improbable_counter += 1
-            elif any(re.findall(r'probable|possible', filename, re.IGNORECASE)):
-                save_name = tidy_image_dir + 'probable' + '-' + str(probable_counter) + '.jpg'
-                probable_counter += 1   
-            imageio.imwrite(save_name, np.array(Image.open(filename)))
-    print('Number of probable images saved:', probable_counter-1)
-    print('Number of improbable images saved:', improbable_counter-1) 
+def saveImageFiles(image_file_list, classification_scenario):
+    """Serially labels all images by class:
+    Classification Scenario A: {probable, possible, improbable}
+    Classification Scenario B: {probable, improbable}
+    Classification Scenario C: {{probable, possible}, improbable}
+    Classification Scenario D: {probable, {possible, improbable}}
+    and saves them to data/tidy/labeled_images"""
+    
+    if classification_scenario=="A":
+        improbable_counter = 1
+        possible_counter = 1
+        probable_counter = 1
+        tidy_image_dir = '../../data/tidy/labeled_images_scenario_A/'
+        if not os.path.exists(tidy_image_dir):
+            os.makedirs(tidy_image_dir)
+        for filename in image_file_list:
+            if '.JPG' in filename or '.jpg' in filename:        
+                if any(re.findall(r'improbable', filename, re.IGNORECASE)):
+                    save_name = tidy_image_dir + 'improbable' + '-' + str(improbable_counter) + '.jpg'
+                    improbable_counter += 1
+                elif any(re.findall(r'probable', filename, re.IGNORECASE)):
+                    save_name = tidy_image_dir + 'probable' + '-' + str(probable_counter) + '.jpg'
+                    probable_counter += 1 
[…]

### 2021-01-13 bin/python/label-image-files.py
updated label-image-files.py and prototype-cnn.ipynb

@@ -26,97 +26,34 @@ def getListOfFiles(dirName):
                 
     return allFiles
 
-def saveImageFiles(image_file_list, classification_scenario):
-    """Serially labels all images by class:
-    Classification Scenario A: {probable, possible, improbable}
-    Classification Scenario B: {probable, improbable}
-    Classification Scenario C: {{probable, possible}, improbable}
-    Classification Scenario D: {probable, {possible, improbable}}
-    and saves them to data/tidy/labeled_images"""
+def saveImageFiles(image_file_list):
+    """Serially labels all images by class:  and saves them to data/tidy/labeled_images"""
     
-    if classification_scenario=="A":
-        improbable_counter = 1
-        possible_counter = 1
-        probable_counter = 1
-        tidy_image_dir = '../../data/tidy/labeled_images_scenario_A/'
-        if not os.path.exists(tidy_image_dir):
-            os.makedirs(tidy_image_dir)
-        for filename in image_file_list:
-            if '.JPG' in filename or '.jpg' in filename:        
-                if any(re.findall(r'improbable', filename, re.IGNORECASE)):
-                    save_name = tidy_image_dir + 'improbable' + '-' + str(improbable_counter) + '.jpg'
-                    improbable_counter += 1
-                elif any(re.findall(r'probable', filename, re.IGNORECASE)):
-                    save_name = tidy_image_dir + 'probable' + '-' + str(probable_counter) + '.jpg'
-                    probable_counter += 1 
-                elif any(re.findall(r'possible', filename, re.IGNORECASE)):
-                    save_name = tidy_image_dir + 'possible' + '-' + str(possible_counter) + '.jpg'
-                    possible_counter += 1 
-                imageio.imwrite(save_name, np.array(Image.open(filename)))
-        print('Number of improbable images saved:', improbable_counter - 1)    
-        print('Number of possible images saved:', possible_counter - 1)
-        print('Number of probable images saved:', probable_counter - 1)
-        
-    elif classification_scenario=="B":
-        improbable_counter = 1
-        probable_counter = 1        
-        tidy_image_dir = '../../data/tidy/labeled_images_scenario_B/'
-        if not os.path.exists(tidy_image_dir):
-            os.makedirs(tidy_image_dir)
-            for filename in image_file_list:
-                if '.JPG' in filename or '.jpg' in filename:        
-                    if any(re.findall(r'improbable', filename, re.IGNORECASE)):
[…]

### 2021-01-13 docs/manuscript/manuscript.tex
additions to manuscript outline

@@
 \section{Background}
-\section{Data}
 
-\section{Methods}
+\section{Data and Methods}
+\subsection{Image data description}
+Based on the  ISA’s Tree Risk Assessment Qualification (TRAQ) protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
+\begin{itemize}
+\item \textbf{Probable}: failure expected under normal weather conditions within a given timeframe
+\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
+\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
+\end{itemize}
+
+\subsection{Data pre-processing}
+
+\subsection{Convolutional neural network}
 

### 2021-01-18 bin/python/remove-output.py
Overwrite existing notebook with cleared cell output

@@ -1,5 +1,5 @@
 """
-Usage: python remove-output.py notebook.ipynb [ > notebook_output_removed.ipynb ]
+Usage: python remove-output.py notebook.ipynb [ > notebook.ipynb #output removed ]
 Modified from remove-output by Minrk
 
 """
@@ -22,7 +22,7 @@ if __name__ == '__main__':
         nb = read(f, 'json')
     remove_outputs(nb)
     base, ext = os.path.splitext(fname)
-    new_ipynb = "%s_output_removed%s" % (base, ext)
+    new_ipynb = "%s%s" % (base, ext)
     with io.open(new_ipynb, 'w', encoding='utf8') as f:
         write(nb, f, 'json')
     print("wrote %s" % new_ipynb)   

### 2021-01-18 docs/manuscript/manuscript.tex
first effort at Background section

@@
 done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions). In general,
-tree-related outages are due to contact between tree parts and lines. Sometimes, the contact is due to growth of
+tree-related outages are due to contact between tree parts and power lines. Sometimes, the contact is due to growth of
 branches into the lines; sometimes it is due to branch failure; sometimes it is due to tree failure (either uprooting or
-trunk failure). When outages occur, the economic costs can be substantial. From 2005 to 2015, \cite{graziano2020wider}
-computed the annual economic cost of outages in the state of Connecticut; the mean was approximately \$8.3 bn.  Outages
-due to contact from branch growth or branch failure can be mitigated by line clearance tree trimming. Estimates vary,
-but the annual amount that utilities spend on tree trimming operations is typically in the billions of dollars
-\cite{guggenmoos2003effects}. Despite the comparatively high annual cost of tree trimming, studies have repeatedly shown
-that trimming branches away from the wires reduces the number of outages (sometimes classified as “preventable”
-outages). For example, in Massachusetts, \citet{simpson1996treecaused} reported that tree failure caused 40\% of
-preventable tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to
-30\%. In a more recent (and sophisticated) analysis in Connecticut, \citet{parent2019analysis} showed that enhanced tree
-trimming reduced outages during storms. And using a statistical model to predict outages based on data from two Gulf
-Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less uncertain when they
-included the effect of tree trimming; models including wind speed without considering the effect of trimming were not as
-accurate.
+trunk failure). When outages occur, the economic costs can be substantial. For example, from 2005 to 2015,
+\cite{graziano2020wider} computed the annual economic cost of outages in the US state of Connecticut; the mean was
+approximately \$8.3 bn.  Outages due to contact from branch growth or branch failure can be mitigated by line clearance
+tree trimming. Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in
+the billions of dollars \cite{guggenmoos2003effects}. Despite the comparatively high annual cost of tree trimming,
+studies have repeatedly shown that trimming branches away from the wires reduces the number of outages (sometimes
+classified as “preventable” outages). In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable
+tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\%
+\cite{simpson1996treecaused}. A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree
+trimming reduced outages during storms \cite{parent2019analysis}. And using a statistical model to predict outages based
+on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less
+uncertain when they included the effect of tree trimming; models including wind speed without considering the effect of
+trimming were not as accurate.
 
@@
 outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of
-tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citet{wismer2018targeted}
-reported approximately 25\% of interruptions in Illinois, USA were caused by trees that uprooted or broke in the stem.
+tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted}
+reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
 Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been
-developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Assessing risk includes assessing the likelihood of
-failure, the likelihood of impact, and the severity of consequences. Likelihood of failure depends on the anticipated
+developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Risk includes assessing the likelihood of
+failure, the likelihood of impact, and the severity of consequences. The likelihood of failure depends on the anticipated
 loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines,
-poles, and other hardware—“infrastructure”—in the case of utility tree risk assessment), the target’s occupancy rate
+poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate
 (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of
 consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs
-and disruption associated with outages—this, in turn, is partially related to the size of the tree or tree part that
+and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that
 fails, and how much momentum it has when it impacts the infrastructure.
 
-Assessing risk of individual trees can be costly because of the time required to complete the assessment. In some
-situations, a less time-consuming assessment may be justified to reduce costs—a ``Level 1'' assessment (Smiley et
-al. 2017). Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk
-assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA
-\cite{koeser2016frequency} . The utility of Level 1 assessments in Rhode Island and Florida suggests that artificial
-intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
+Individual tree risk assessment can be costly because of the time it requires. In some situations, a less time-consuming
+assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment (Smiley et al. 2017). Studies have shown
+that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving
+vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency} . The utility of
+Level 1 assessments in these areas suggests that artificial intelligence (AI) tools may be an effective way to reduce
+the cost of tree risk assessment.
 
-AI-based image analysis is relatively widely used, even in engineering applications such as earthquake risk assessment
+AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment
 \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}. A relevant
 application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated
-\cite{fricker2019convolutional}. Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is
-complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and
+\cite{fricker2019convolutional}. Yet, AI has not been applied to the problem of tree-utility line risk assessment---one
+that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and
 associated risk and local meteorological conditions. However, the flexibility and power of CNNs, appears promising. In
@@
 \section{Background}
+%Neural networks developed over the past several decades from the introduction of the single-layer perceptron.
+The groundbreaking study of \citet{hubel1959receptive} showed that visual perception in cats was a result of the
+activation or inhibition of groups of cells in the visual cortex known as ``receptive fields.''  Further, they attempted
+to map the cortical architecture in cats and monkeys \cite{hubel1962receptive,hubel1965receptive,hubel1968receptive}.
+Subsequent attempts were then made to model neural networks that could be trained to automatically recognize visual
+patterns with modest performance \cite{rosenblatt1962principles,kabrisky1966proposed,giebel1971feature,fukushima1975cognitron}.
+However, the breakthrough came with ``neocognitron'' \cite{fukushima1980neocognitron}, which
+was a self-learning neural network for pattern recognition that was robust to changes in position and shape distortion,
+a problem that plagued earlier efforts, including ``cognitron'' \cite{fukushima1975cognitron} proposed a few years
+earlier.
+
+A few notable efforts demonstrated the neural networks for handwritten digit recognition
+\cite{fukushima1988neocognitron,denker1988neural}, but these required significant preprocessing and feature
+extraction. \cite{lecun1989handwritten} soon afterward introduced a multilayer neural network that mapped a feature in
+each neuron (representing a ``local receptive field'') via convolution. This network could also be trained by
+backpropagation like other existing neural networks and featured pooling operations for better distortion and
+translation invariance. Further developments from this milestone yielded the LeNet-5 convolutional neural network which
+attained accuracy levels that rendered it commercially viable.
+
+The big data revolution coupled with technological advancements that have made it possible to capture and store high
+resolution images have raised challenges that continue to be surmounted with successively high-performing
+architectures. Over the past decade, some of these efforts resulted in significant breakthroughs in performance. AlexNet
+\cite{krizhevsky2012imageneta}, with 5 convolutional layers and 3 dense layers---one of the largest CNNs of its time,
+won the ILSVRC-2012\footnote{ImageNet Large Scale Visual Recognition Challenge; held annually from 2010 through 2017.}
+competition with a top-5 error rate of 15.3\% and served as a landmark in the Deep Learning subdomain.
+\citet{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization
+techniques that were foundational for model inference and interpretability.  In the same year, GoogLeNet, a 22-layer
+network, was proposed \cite{szegedy2014going}, featuring the novel ``Inception module,'' which allowed for efficiency
+and accuracy in a very deep network. Subsequent improvements have been proposed to the original inception framework
+\cite{szegedy2015rethinking,szegedy2016inceptionv4}.  VGGNet \cite{simonyan2015very} also pushed the boundaries of depth
+with up 19 layers, achieving state-of-the-art performance at ILSVRC-2014. Finally, ResNet \cite{he2015deep} addressed
+the accuracy degradation problem that arises with increasing depth in a network by succesively fitting smaller sets of
+layers to the residual and employing skip connections. With these innovations, an unprecedented level of depth was
+achieved. Implementations with with 34, 50, 101 and 152 layers were demonstrated. ResNet-152 won first place in
+ILSVRC-2015.
+
+Along with these developments in their architectures, CNNs have demonstrated viability for applications to image
+classification, object and text detection, object and document tracking, labeling, speech, among several other related
+fields \cite{gu2018recent}. \hl{Paragraph on applications to be completed---with particular attention paid to tree-related efforts.}
 
@@
 \subsection{Image data description}
+The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels. 
 Based on the  ISA’s Tree Risk Assessment Qualification (TRAQ) protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
@@
 
-\subsection{Data pre-processing}
+\subsection{Pre-processing}
 

### 2021-01-19 docs/manuscript/manuscript.tex
added sketch of results comparison table

@@
 
+\begin{table}[h!]
+  \centering
+  \begin{tabular}{l l l l l l l }\toprule
+    \bf Model & \multicolumn{3}{c}{\bf Training metrics} &\multicolumn{3}{c}{\bf Validation metrics}  \\\midrule
+    & Error & Precision & Recall     & Error & Precision & Recall \\
+    SafeTree & & & & & & \\
+    GoogleNet (InceptionV3) & & & & & & \\
+    ResNet50 & & & & & & \\
+        VGGNet & & & & & & \\
+    AlexNet & & & & & & \\\bottomrule
+  \end{tabular}
+  \caption{Comparing our model SafeTree with state-of-the-art CNN architectures trained on our data}
+  \label{tab:comp}
+\end{table}
+
 \section{Conclusion}

### 2021-01-21 docs/manuscript/manuscript.tex

@@
 tree-related outages are due to contact between tree parts and power lines. Sometimes, the contact is due to growth of
-branches into the lines; sometimes it is due to branch failure; sometimes it is due to tree failure (either uprooting or
+branches into the lines; sometimes it is due to branch failure; sometimes it is due to whole-tree failure (either uprooting or
 trunk failure). When outages occur, the economic costs can be substantial. For example, from 2005 to 2015,
@@
 approximately \$8.3 bn.  Outages due to contact from branch growth or branch failure can be mitigated by line clearance
-tree trimming. Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in
-the billions of dollars \cite{guggenmoos2003effects}. Despite the comparatively high annual cost of tree trimming,
-studies have repeatedly shown that trimming branches away from the wires reduces the number of outages (sometimes
+pruning. Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in
+the billions of dollars \cite{guggenmoos2003effects}. Despite the comparatively high annual cost of pruning,
+studies have repeatedly shown that pruning branches away from the wires reduces the number of outages (sometimes
 classified as “preventable” outages). In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable
@@
 \cite{simpson1996treecaused}. A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree
-trimming reduced outages during storms \cite{parent2019analysis}. And using a statistical model to predict outages based
+pruning reduced outages during storms \cite{parent2019analysis}. And using a statistical model to predict outages based
 on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less
-uncertain when they included the effect of tree trimming; models including wind speed without considering the effect of
+uncertain when they included the effect of utility pruning; models including wind speed without considering the effect of
 trimming were not as accurate.
 
-Trimming can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging
-the lines in proximity are removed, reducing the likelihood of contact.
-% Trimming also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of
-% whole-tree failure, whether by uprooting or stem breakage.  Although the annual cost of trimming is comparatively
+%Utility pruning can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging
+%the lines in proximity are removed or reduced, reducing the likelihood of contact.
+Utility pruning aims to reduce the frequency of tree-caused outages through the reduction or removal of branches in proximity to overhead electrical lines.
+% Pruning also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of
+% whole-tree failure, whether by uprooting or stem breakage.  Although the annual cost of utility vegetation management is 
 % large, the economic cost of outages can far exceed it. A recent benefit cost analysis demonstrated a highly favorable
-% benefit cost ratio for trimming relative to the economic costs of outages (Graziano et al. 2020).  Trimming,
-However, it cannot eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the
+% benefit cost ratio for pruning relative to the economic costs of outages (Graziano et al. 2020).  
+However, it cannot completely eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the
 lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that cause
@@
 The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels. 
-Based on the  ISA’s Tree Risk Assessment Qualification (TRAQ) protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
+Based on the  ISA’s Tree Risk Assessment protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
 \begin{itemize}

### 2021-01-21 docs/manuscript/manuscript.tex
Updates from Overleaf
minor additions to manuscript
re-formatted source text

@@
 \section{Introduction}
-Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been
-done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions). In general,
-tree-related outages are due to contact between tree parts and power lines. Sometimes, the contact is due to growth of
-branches into the lines; sometimes it is due to branch failure; sometimes it is due to whole-tree failure (either uprooting or
-trunk failure). When outages occur, the economic costs can be substantial. For example, from 2005 to 2015,
-\cite{graziano2020wider} computed the annual economic cost of outages in the US state of Connecticut; the mean was
-approximately \$8.3 bn.  Outages due to contact from branch growth or branch failure can be mitigated by line clearance
-pruning. Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in
-the billions of dollars \cite{guggenmoos2003effects}. Despite the comparatively high annual cost of pruning,
-studies have repeatedly shown that pruning branches away from the wires reduces the number of outages (sometimes
-classified as “preventable” outages). In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable
-tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\%
-\cite{simpson1996treecaused}. A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree
-pruning reduced outages during storms \cite{parent2019analysis}. And using a statistical model to predict outages based
-on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less
-uncertain when they included the effect of utility pruning; models including wind speed without considering the effect of
-trimming were not as accurate.
+Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions).
+In general, tree-related outages are due to contact between tree parts and power lines.
+Sometimes, the contact is due to growth of branches into the lines; sometimes it is due to branch failure; sometimes it is due to whole-tree failure (either uprooting or trunk failure).
+When outages occur, the economic costs can be substantial.
+For example, from 2005 to 2015, \cite{graziano2020wider} computed the annual economic cost of outages in the US state of Connecticut; the mean was approximately \$8.3 bn.
+Outages due to contact from branch growth or branch failure can be mitigated by line clearance pruning.
+Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in the billions of dollars \cite{guggenmoos2003effects}.
+Despite the comparatively high annual cost of pruning, studies have repeatedly shown that pruning branches away from the wires reduces the number of outages (sometimes classified as “preventable” outages).
+In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\% \cite{simpson1996treecaused}.
+A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree pruning reduced outages during storms \cite{parent2019analysis}.
+And using a statistical model to predict outages based on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less uncertain when they included the effect of utility pruning; models including wind speed without considering the effect of trimming were not as accurate.
 
@@
 % benefit cost ratio for pruning relative to the economic costs of outages (Graziano et al. 2020).  
-However, it cannot completely eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the
-lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that cause
-outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of
-tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted}
-reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
-Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been
-developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Risk includes assessing the likelihood of
-failure, the likelihood of impact, and the severity of consequences. The likelihood of failure depends on the anticipated
-loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines,
-poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate
-(which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of
-consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs
-and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that
-fails, and how much momentum it has when it impacts the infrastructure.
+However, it cannot completely eliminate tree-related outages.
+Failure of trees away from the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}.
+The proportion of tree failures away from the wires that cause outages varies and has not been rigorously quantified.
+\citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
+Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}.
+Risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences.
+The likelihood of failure depends on the anticipated loads on the tree and its load-bearing capacity.
+The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees.
+Severity of consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure.
 
-Individual tree risk assessment can be costly because of the time it requires. In some situations, a less time-consuming
-assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment (Smiley et al. 2017). Studies have shown
-that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving
-vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency} . The utility of
-Level 1 assessments in these areas suggests that artificial intelligence (AI) tools may be an effective way to reduce
-the cost of tree risk assessment.
+Individual tree risk assessment can be costly because of the time it requires.
+In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment (Smiley et al.
+2017).
+Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency} .
+The utility of Level 1 assessments in these areas suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
 
-AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment
-\cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}. A relevant
-application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated
-\cite{fricker2019convolutional}. Yet, AI has not been applied to the problem of tree-utility line risk assessment---one
-that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and
-associated risk and local meteorological conditions. However, the flexibility and power of CNNs, appears promising. In
-this paper, we demons In the remainder of this paper, we first provide a background on CNN which is critical for
-sustainability in critical infrastructural systems. Our goal is to further demonstrate an innovative automated approach
-to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually
-improved through subsequent training on new datasets.
+AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}.
+A relevant application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated \cite{fricker2019convolutional}.
+Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
+However, the flexibility and power of CNNs, appears promising.
+In this paper, we demons In the remainder of this paper, we first provide a background on CNN which is critical for sustainability in critical infrastructural systems.
+Our goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
 
@@
 to map the cortical architecture in cats and monkeys \cite{hubel1962receptive,hubel1965receptive,hubel1968receptive}.
-Subsequent attempts were then made to model neural networks that could be trained to automatically recognize visual
-patterns with modest performance \cite{rosenblatt1962principles,kabrisky1966proposed,giebel1971feature,fukushima1975cognitron}.
+Subsequent attempts were then made to model neural networks that could be trained to automatically recognize visual patterns with modest performance \cite{rosenblatt1962principles,kabrisky1966proposed,giebel1971feature,fukushima1975cognitron}.
 However, the breakthrough came with ``neocognitron'' \cite{fukushima1980neocognitron}, which
-was a self-learning neural network for pattern recognition that was robust to changes in position and shape distortion,
-a problem that plagued earlier efforts, including ``cognitron'' \cite{fukushima1975cognitron} proposed a few years
-earlier.
+was a self-learning neural network for pattern recognition that was robust to changes in position and shape distortion, a problem that plagued earlier efforts, including ``cognitron'' \cite{fukushima1975cognitron} proposed a few years earlier.
 
@@
 \cite{fukushima1988neocognitron,denker1988neural}, but these required significant preprocessing and feature
-extraction. \cite{lecun1989handwritten} soon afterward introduced a multilayer neural network that mapped a feature in
-each neuron (representing a ``local receptive field'') via convolution. This network could also be trained by
-backpropagation like other existing neural networks and featured pooling operations for better distortion and
-translation invariance. Further developments from this milestone yielded the LeNet-5 convolutional neural network which
-attained accuracy levels that rendered it commercially viable.
+extraction.
+\cite{lecun1989handwritten} soon afterward introduced a multilayer neural network that mapped a feature in each neuron (representing a ``local receptive field'') via convolution.
+This network could also be trained by backpropagation like other existing neural networks and featured pooling operations for better distortion and
+translation invariance. Further developments from this milestone yielded the LeNet-5 convolutional neural network which attained accuracy levels that rendered it commercially viable.
 
-The big data revolution coupled with technological advancements that have made it possible to capture and store high
-resolution images have raised challenges that continue to be surmounted with successively high-performing
-architectures. Over the past decade, some of these efforts resulted in significant breakthroughs in performance. AlexNet
-\cite{krizhevsky2012imageneta}, with 5 convolutional layers and 3 dense layers---one of the largest CNNs of its time,
-won the ILSVRC-2012\footnote{ImageNet Large Scale Visual Recognition Challenge; held annually from 2010 through 2017.}
-competition with a top-5 error rate of 15.3\% and served as a landmark in the Deep Learning subdomain.
-\citet{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization
-techniques that were foundational for model inference and interpretability.  In the same year, GoogLeNet, a 22-layer
-network, was proposed \cite{szegedy2014going}, featuring the novel ``Inception module,'' which allowed for efficiency
-and accuracy in a very deep network. Subsequent improvements have been proposed to the original inception framework
-\cite{szegedy2015rethinking,szegedy2016inceptionv4}.  VGGNet \cite{simonyan2015very} also pushed the boundaries of depth
-with up 19 layers, achieving state-of-the-art performance at ILSVRC-2014. Finally, ResNet \cite{he2015deep} addressed
-the accuracy degradation problem that arises with increasing depth in a network by succesively fitting smaller sets of
-layers to the residual and employing skip connections. With these innovations, an unprecedented level of depth was
-achieved. Implementations with with 34, 50, 101 and 152 layers were demonstrated. ResNet-152 won first place in
-ILSVRC-2015.
+The big data revolution coupled with technological advancements that have made it possible to capture and store high resolution images have raised challenges that continue to be surmounted with successively high-performing
+architectures.
+Over the past decade, some of these efforts resulted in significant breakthroughs in performance.
+AlexNet \cite{krizhevsky2012imageneta}, with 5 convolutional layers and 3 dense layers---one of the largest CNNs of its time, won the ILSVRC-2012\footnote{ImageNet Large Scale Visual Recognition Challenge; held annually from 2010 through 2017.} competition with a top-5 error rate of 15.3\% and served as a landmark in the Deep Learning subdomain.
+\citet{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization techniques that were foundational for model inference and interpretability.
+In the same year, GoogLeNet, a 22-layer network, was proposed \cite{szegedy2014going}, featuring the novel ``Inception module,'' which allowed for efficiency and accuracy in a very deep network.
+Subsequent improvements have been proposed to the original inception framework \cite{szegedy2015rethinking,szegedy2016inceptionv4}.
+VGGNet \cite{simonyan2015very} also pushed the boundaries of depth with up 19 layers, achieving state-of-the-art performance at ILSVRC-2014.
+Finally, ResNet \cite{he2015deep} addressed the accuracy degradation problem that arises with increasing depth in a network by succesively fitting smaller sets of layers to the residual and employing skip connections.
+With these innovations, an unprecedented level of depth was achieved.
+Implementations with with 34, 50, 101 and 152 layers were demonstrated.
+ResNet-152 won first place in ILSVRC-2015.
 
@@
 \begin{itemize}
-\item \textbf{Probable}: failure expected under normal weather conditions within a given timeframe
+\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
 \item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
@@
 \subsection{Convolutional neural network}
-
-\section{Results and Discussion}
 
@@
 
+\section{Results}
+
+\subsection{Sensitivity to training resolution}
+
 \section{Conclusion}

### 2021-01-23 bin/python/preprocess-images.py
created and test preprocess-images script/notebook

@@ -1,0 +1,135 @@
+"""
+The purpose of this script is to pre-process the tree images using the following approaches:
+    - one-hot vector encoding
+    - resizing
+    - grayscaling 
+    - normalization (scaling pixels from 0 to 1)
+    - random cropping (based on specified expansion factor to multiply images)
+    - horizontal flipping
+"""
+
+from PIL import Image # used for loading images
+import numpy as np
+import os # used for navigating to image path
+import imageio # used for writing images
+import random
+import matplotlib.pyplot as plt
+#from timeit import default_timer as timer
+
+## GLOBAL VARIABLES
+IMG_SIZE = 108
+IMG_SIZE_LIST = [64, 128, 224, 384]
+NUM_CHANNELS = 1
+CLASSIFICATION_SCENARIO = "Pr_Im"
+CLASSIFICATION_SCENARIO_LIST = ["Pr_Im", "Pr_Im", "PrPo_Im", "Pr_PoIm"]  
+EXPANSION_FACTOR = 5 # of augmented images
+LABELED_IMAGES_DIR = '../../data/tidy/labeled_images'
+PROCESSED_IMAGES_DIR = '../../data/tidy/preprocessed_images'
+SEED = 10 # seed for repeatability
+
+
+def getImageOneHotVector(image_file_name, classification_scenario = "Pr_Im"):
+    """Returns one-hot vector encoding for each image based on specified classification scenario:
+    Classification Scenario Pr_Po_Im (3 classes): {probable, possible, improbable}
+    Classification Scenario Pr_Im (2 classes): {probable, improbable}
+    Classification Scenario PrPo_Im (2 classes): {{probable, possible}, improbable}
+    Classification Scenario Pr_PoIm (2 classes): {probable, {possible, improbable}}
+    """
+    word_label = image_file_name.split('-')[0]
+    if classification_scenario == "Pr_Po_Im":
+        if word_label == 'probable' : 
+            return np.array([1, 0, 0])
+        elif word_label == 'possible' : 
+            return np.array([0, 1, 0])    
+        elif word_label == 'improbable':
+            return np.array([0, 0, 1])
+        else :
+            return np.array([0, 0, 0]) # if label is not present for current image
+    elif classification_scenario == "Pr_Im":
+        if word_label == 'probable' : 
+            return np.array([1, 0])
+        elif word_label == 'improbable' : 
+            return np.array([0, 1])
+        else :
+            return np.array([0, 0]) # if label is not present for current image
+    elif classification_scenario == "PrPo_Im":
[…]

### 2021-02-01 docs/manuscript/manuscript.tex
quick notes added in Results section

@@
 \section{Introduction}
-Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions).
-In general, tree-related outages are due to contact between tree parts and power lines.
-Sometimes, the contact is due to growth of branches into the lines; sometimes it is due to branch failure; sometimes it is due to whole-tree failure (either uprooting or trunk failure).
-When outages occur, the economic costs can be substantial.
-For example, from 2005 to 2015, \cite{graziano2020wider} computed the annual economic cost of outages in the US state of Connecticut; the mean was approximately \$8.3 bn.
-Outages due to contact from branch growth or branch failure can be mitigated by line clearance pruning.
-Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in the billions of dollars \cite{guggenmoos2003effects}.
-Despite the comparatively high annual cost of pruning, studies have repeatedly shown that pruning branches away from the wires reduces the number of outages (sometimes classified as “preventable” outages).
-In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\% \cite{simpson1996treecaused}.
-A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree pruning reduced outages during storms \cite{parent2019analysis}.
-And using a statistical model to predict outages based on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less uncertain when they included the effect of utility pruning; models including wind speed without considering the effect of trimming were not as accurate.
+Power outages due to contact between tree parts and power lines annually cause tens of billions of dollars in economic and other disruptions throughout the United States, despite extensive efforts by utilities to mitigate and prevent those contacts.
+Presently, the identification of potential contact between trees and power lines is labor intensive and time-consuming.
+This paper describes an artificial intelligence and machine learning approach that automatically classifies trees, using only a single photograph and with a high degree of accuracy, into categories corresponding to probable failure, possible failure and improbable failure--categories corresponding to those used by utility arborists to assess risk of contact between trees and power lines.
+This preliminary study demonstrates the possible efficacy of AI approaches to tree risk assessment and, following further development of the approach has the potential to reduce power outages and utility costs by allowing utilities to more effectively target their pruning and mitigation efforts.
 
-%Utility pruning can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging
-%the lines in proximity are removed or reduced, reducing the likelihood of contact.
-Utility pruning aims to reduce the frequency of tree-caused outages through the reduction or removal of branches in proximity to overhead electrical lines.
-% Pruning also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of
-% whole-tree failure, whether by uprooting or stem breakage.  Although the annual cost of utility vegetation management is 
-% large, the economic cost of outages can far exceed it. A recent benefit cost analysis demonstrated a highly favorable
-% benefit cost ratio for pruning relative to the economic costs of outages (Graziano et al. 2020).  
-However, it cannot completely eliminate tree-related outages.
+Contact between tree parts and power lines can take several forms: tree branches can grow into lines; branches can fail and fall onto lines; whole-tree failure can occur due to uprooting or trunk failure.
+A study in the state of Connecticut provides some context for the amount of economic, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.
+That extremely high cost occurs despite extensive efforts on the part of utilities to mitigate such disruptions through active and aggressive pruning programs that, on their own cost billions of dollars annually \cite{guggenmoos2003effects}.
+
+Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called 'preventable' contact incidents between trees and power lines.
+For example, in Massachusetts, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
+The efficacy of pruning has also been shown in a study of two US Gulf Coast states that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citet{nateghi2014power}.
+
+Even effective pruning cannot, however, completely eliminate tree-related outages.
 Failure of trees away from the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}.
@@
 \citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
+
 Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}.
@@
 Individual tree risk assessment can be costly because of the time it requires.
-In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment (Smiley et al.
+In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment (XXXSmiley et al.
 2017).
@@
 
+The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible and improbably.
+The data used for training, testing and illustration of the method consists of XXX tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.
+
+The remainder of the paper provides background AI and risk assessment, utility tree risk assessment and CNNs.
+The goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
+
+\section{Background}
+
+%\subsection{AI and risk assessment}
 AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}.
+Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by 10.1177/8755293020919419 XXX), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
+Neural networks have also been used to interrogate remote sensing data of the landscape to assess landslide risk ( 10.1007/s10346-020-01557-6 XXX)
+
 A relevant application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated \cite{fricker2019convolutional}.
 Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
-However, the flexibility and power of CNNs, appears promising.
-In this paper, we demons In the remainder of this paper, we first provide a background on CNN which is critical for sustainability in critical infrastructural systems.
-Our goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
 
-\section{Background}
 %Neural networks developed over the past several decades from the introduction of the single-layer perceptron.
@@
 The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels.
-Based on the  ISA’s Tree Risk Assessment protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
+Images were captured over a single field season in Massachusetts, USA, between May and September 2020 to limit any potential influence of changes in tree appearance due to seasonal leaf senescence on image processing.
+ESRI ArcMaps was used to randomly distribute sampling sites across the state.
+Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Beast Management Practices (Smiley et al.
+2017) and ISA’s Utility Tree Risk Assessment Best Management Practices (Goodfellow 2020).
+This method is commonly used to assess trees in the United States.
+This level assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
+Level 2 or Level 3 (Smiley et al.
+2017), given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) rights may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW (Goodfellow 2020); (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects (Rooney et al.
+2005, Koeser et al.
+2016) leading to more severe likelihood of failure ratings.
+The four categories of tree failure likelihood are defined as follows: (Smiley et al.
+2017 (I have not figured out how to embed citations yet, sorry -Ryan):
 \begin{itemize}
+\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
+\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
 \item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
-\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
-\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
+\item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load.
+This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect people from harm
 \end{itemize}
-
+	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of the \textit{imminent} category trees.
 \subsection{Pre-processing}
@@
 \subsection{Convolutional neural network}
+
+We can apply cut-out (occluding portions of the image) for improved performance \cite{devries2017improved}.
+Also, it has been shown that training with lower resolution improves performance on higher resolution test images \cite{touvronfixing}.
 

### 2021-02-01 docs/manuscript/manuscript.tex
updated citations

@@
 Individual tree risk assessment can be costly because of the time it requires.
-In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment (XXXSmiley et al.
-2017).
+In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{e.thomassmiley2017best}.
 Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency} .
@@
 The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible and improbably.
-The data used for training, testing and illustration of the method consists of XXX tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.
+The data used for training, testing and illustration of the method consists of 505 tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.
 
@@
 AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}.
-Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by 10.1177/8755293020919419 XXX), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
-Neural networks have also been used to interrogate remote sensing data of the landscape to assess landslide risk ( 10.1007/s10346-020-01557-6 XXX)
+Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citet{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
+Neural networks have also been used to interrogate remote sensing data of the landscape to assess landslide risk \cite{su2020deep}.
 
@@
 ESRI ArcMaps was used to randomly distribute sampling sites across the state.
-Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Beast Management Practices (Smiley et al.
-2017) and ISA’s Utility Tree Risk Assessment Best Management Practices (Goodfellow 2020).
+Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Beast Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.
 This method is commonly used to assess trees in the United States.
 This level assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
-Level 2 or Level 3 (Smiley et al.
-2017), given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) rights may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW (Goodfellow 2020); (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects (Rooney et al.
-2005, Koeser et al.
-2016) leading to more severe likelihood of failure ratings.
-The four categories of tree failure likelihood are defined as follows: (Smiley et al.
-2017 (I have not figured out how to embed citations yet, sorry -Ryan):
+Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) rights may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to more severe likelihood of failure ratings.
+The four categories of tree failure likelihood are defined as follows \cite{e.thomassmiley2017best}:
 \begin{itemize}

### 2021-02-01 docs/manuscript/manuscript.tex
further updates to methods and addition of classification scenarios table

@@
 	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of the \textit{imminent} category trees.
-\subsection{Pre-processing}
+
+\subsection{Pre-processing and data augmentation}
+In the original set of training images, the class distribution is given in \autoref{tab:classdist}.
+
+\begin{table}[h!]\small
+    \centering
+    \begin{tabular}{l l}\toprule
+    \bf Class Label     & \bf Number of images  \\ \midrule
+    Improbable & 322\\
+    Possible & 80 \\
+    Probable & 56 \\\midrule
+    Total & 505 \\\bottomrule
+    \end{tabular}
+    \caption{Class distribution of images in training set}
+    \label{tab:classdist}
+\end{table}
+
+To achieve robustness in training, and given the relatively small number of training images, we randomly cropped each image on either axis to 3024 $\times$ 3024 pixels, generating five instances for each one.
+Thus, we increased the size of our training set from 505 to 2525 images.
+Further, we performed horizontal flipping with a 50\% probability on each of the generated images.
+For efficiency, converted the images to grayscale and scaled the pixel values from 0 to 1.
+Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
 
@@
 
-\begin{table}[h!]
+\begin{table}[h!]\small
   \centering
@@
 
+\subsection{Hyperparameter optimization}
+We used the Hyperband approach \cite{li2018hyperband} to perform a grid search to find the optimal values of the following hyperparameters:
+\begin{compactitem}
+    \item kernel size in first convolutional layer
+    \item number of units in first densely connected layer
+    \item dropout rate applied to outputs of first dense layer
+    \item activation function for first dense layer
+    \item number of units in second densely connected layer
+    \item dropout rate applied to outputs of second dense layer
+    \item activation function for second dense layer    
+    \item learning rate (of Adam optimizer)
+\end{compactitem}
+
 \section{Results}
+
+\subsection{Classification experiments}
+
+We define four classification scenarios in \autoref{tab:scenarios}.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l l}\toprule
+    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
+    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
+    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
+    \texttt{PrPo\_Im}        & \{Probable|Possible, Improbable\}  & 2 \\
+    \texttt{Pr\_PoIm}        & \{Probable, Possible|Impossible\}  & 2 \\\bottomrule
+    \end{tabular}
+    \caption{Classification scenarios}
+    \label{tab:scenarios}
+\end{table}
 

### 2021-02-01 docs/manuscript/manuscript.tex
minor changes to results

@@
 
-\usepackage{fancyhdr}
-\pagestyle{fancy}
+% \usepackage{fancyhdr}
+% \pagestyle{fancy}
 %\addtolength{\headwidth}{\marginparsep} %these change header-rule width
@@
 %\fancyfootoffset{30pt}
-\fancyhead[LO,RE]{\small  \it \nouppercase{\leftmark}}
-\fancyhead[RO,LE]{\small Page \thepage}
-\fancyfoot[RO,LE]{\small }% PR \num S-2015}
-\fancyfoot[LO,RE]{\small }%\scshape MODL}
-\cfoot{}
-\renewcommand{\headrulewidth}{0.1pt}
-\renewcommand{\footrulewidth}{0pt}
+% \fancyhead[LO,RE]{\small  \it \nouppercase{\leftmark}}
+% \fancyhead[RO,LE]{\small Page \thepage}
+% \fancyfoot[RO,LE]{\small }% PR \num S-2015}
+% \fancyfoot[LO,RE]{\small }%\scshape MODL}
+% \cfoot{}
+% \renewcommand{\headrulewidth}{0.1pt}
+% \renewcommand{\footrulewidth}{0pt}
 %\setlength\voffset{-0.25in}
@@
 
-Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called 'preventable' contact incidents between trees and power lines.
+Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called ``preventable'' contact incidents between trees and power lines.
 For example, in Massachusetts, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
@@
 In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{e.thomassmiley2017best}.
-Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency} .
+Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}.
 The utility of Level 1 assessments in these areas suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
@@
 Subsequent attempts were then made to model neural networks that could be trained to automatically recognize visual patterns with modest performance \cite{rosenblatt1962principles,kabrisky1966proposed,giebel1971feature,fukushima1975cognitron}.
-However, the breakthrough came with ``neocognitron'' \cite{fukushima1980neocognitron}, which
+However, the breakthrough came with the ``neocognitron'' \cite{fukushima1980neocognitron}, which
 was a self-learning neural network for pattern recognition that was robust to changes in position and shape distortion, a problem that plagued earlier efforts, including ``cognitron'' \cite{fukushima1975cognitron} proposed a few years earlier.
@@
 
-\begin{table}[h!]\small
-  \centering
-  \begin{tabular}{l l l l l l l }\toprule
-    \bf Model & \multicolumn{3}{c}{\bf Training metrics} &\multicolumn{3}{c}{\bf Validation metrics}  \\\midrule
-    & Error & Precision & Recall     & Error & Precision & Recall \\
-    SafeTree & & & & & & \\
-    GoogleNet (InceptionV3) & & & & & & \\
-    ResNet50 & & & & & & \\
-        VGGNet & & & & & & \\
-    AlexNet & & & & & & \\\bottomrule
-  \end{tabular}
-  \caption{Comparing our model SafeTree with state-of-the-art CNN architectures trained on our data}
-  \label{tab:comp}
-\end{table}
-
 \subsection{Hyperparameter optimization}
@@
 
+\subsection{Model visualization and inference}
+
+\subsection{Comparison with state-of-the-art architectures}
+We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \autoref{tab:comp}.
+
+\begin{table}[h!]\small
+  \centering
+  \begin{tabular}{l l l l l l l }\toprule
+    \bf Model & \multicolumn{3}{c}{\bf Training metrics} &\multicolumn{3}{c}{\bf Validation metrics}  \\\midrule
+    & Error & Precision & Recall     & Error & Precision & Recall \\
+    SafeTree & & & & & & \\
+    GoogleNet (InceptionV3) & & & & & & \\
+    ResNet50 & & & & & & \\
+        VGGNet & & & & & & \\
+    AlexNet & & & & & & \\\bottomrule
+  \end{tabular}
+  \caption{Comparing our model SafeTree with state-of-the-art CNN architectures trained on our data}
+  \label{tab:comp}
+\end{table}
 \section{Conclusion}

### 2021-02-02 docs/manuscript/manuscript.tex

@@
 Contact between tree parts and power lines can take several forms: tree branches can grow into lines; branches can fail and fall onto lines; whole-tree failure can occur due to uprooting or trunk failure.
-A study in the state of Connecticut provides some context for the amount of economic, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.
-That extremely high cost occurs despite extensive efforts on the part of utilities to mitigate such disruptions through active and aggressive pruning programs that, on their own cost billions of dollars annually \cite{guggenmoos2003effects}.
+A study in the state of Connecticut, USA provides some context for the amount of economic disruption, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.
+That extremely high cost occurred despite extensive efforts on the part of utilities to mitigate conflicts between trees and power lines through active and aggressive pruning programs that, on their own cost billions of dollars annually \cite{guggenmoos2003effects}.
 
 Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called ``preventable'' contact incidents between trees and power lines.
-For example, in Massachusetts, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
-The efficacy of pruning has also been shown in a study of two US Gulf Coast states that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citet{nateghi2014power}.
+For example, in Massachusetts, USA, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
+The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citet{nateghi2014power}.
 
-Even effective pruning cannot, however, completely eliminate tree-related outages.
+Even effective pruning cannot, however, completely eliminate tree-caused outages.
 Failure of trees away from the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}.
-The proportion of tree failures away from the wires that cause outages varies and has not been rigorously quantified.
-\citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
+The proportion of tree failures away from the wires that causes outages varies and has not been rigorously quantified.
+\citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
 
@@
 The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees.
-Severity of consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure.
+Severity of consequences depends on the damage done to the infrastructure---which, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure---and, more importantly in some cases, the economic costs and disruption associated with outages.
 
@@
 Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}.
-The utility of Level 1 assessments in these areas suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
+The utility of Level 1 assessments in these states suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
 
-The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible and improbably.
+The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible, and improbable.
 The data used for training, testing and illustration of the method consists of 505 tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.
 
-The remainder of the paper provides background AI and risk assessment, utility tree risk assessment and CNNs.
+The remainder of the paper provides a brief history and background of AI and its use in earthquake risk assessment and tree identification (section 2); describes the methods used to train and validate a novel CNN to categorize likelihood of tree failure (section 3); and presents and discusses the output of the novel CNN (sections 4 and 5).
 The goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
@@
 
-A relevant application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated \cite{fricker2019convolutional}.
+A relevant application for tree species identification using a convolutional neural network (CNN) was recently demonstrated \cite{fricker2019convolutional}.
 Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
@@
 VGGNet \cite{simonyan2015very} also pushed the boundaries of depth with up 19 layers, achieving state-of-the-art performance at ILSVRC-2014.
-Finally, ResNet \cite{he2015deep} addressed the accuracy degradation problem that arises with increasing depth in a network by succesively fitting smaller sets of layers to the residual and employing skip connections.
+Finally, ResNet \cite{he2015deep} addressed the accuracy degradation problem that arises with increasing depth in a network by successively fitting smaller sets of layers to the residual and employing skip connections.
 With these innovations, an unprecedented level of depth was achieved.
@@
 \subsection{Image data description}
-The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels.
+The training dataset consisted of 505 images, each having an original size of $4032\times3024$ pixels.
 Images were captured over a single field season in Massachusetts, USA, between May and September 2020 to limit any potential influence of changes in tree appearance due to seasonal leaf senescence on image processing.
 ESRI ArcMaps was used to randomly distribute sampling sites across the state.
-Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Beast Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.
+Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.
 This method is commonly used to assess trees in the United States.
-This level assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
-Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) rights may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to more severe likelihood of failure ratings.
-The four categories of tree failure likelihood are defined as follows \cite{e.thomassmiley2017best}:
+A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
+Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.
+The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{e.thomassmiley2017best}:
 \begin{itemize}
@@
 \end{itemize}
-	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of the \textit{imminent} category trees.
+	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category.
 
@@
 Further, we performed horizontal flipping with a 50\% probability on each of the generated images.
-For efficiency, converted the images to grayscale and scaled the pixel values from 0 to 1.
+For efficiency, we converted the images to grayscale and scaled the pixel values from 0 to 1.
 Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.

### 2021-02-04 docs/manuscript/manuscript.tex

@@
-\documentclass[11pt,twoside]{article}
-\usepackage{etex, soul}
-\newcommand{\num}{6{} }
+\documentclass[Journal, letterpaper]{ascelike-new}
+%\usepackage{etex, soul}
+%\newcommand{\num}{6{} }
 
-\raggedbottom
+%\raggedbottom
 
 %geometry (sets margin) and other useful packages
-\usepackage{geometry}
-\geometry{top=1.2in, left=1in,right=1in,bottom=1in,headsep=6pt}
+% \usepackage{geometry}
+% \geometry{top=1.2in, left=1in,right=1in,bottom=1in,headsep=6pt}
  \usepackage{graphicx,booktabs,calc}
@@
 %Marginpar width
-\newcommand{\pts}[1]{\marginpar{ \small\hspace{0pt} \textit{[#1]} } }
-\setlength{\marginparwidth}{.5in}
+% \newcommand{\pts}[1]{\marginpar{ \small\hspace{0pt} \textit{[#1]} } }
+% \setlength{\marginparwidth}{.5in}
 %\reversemarginpar
@@
 % \usepackage[T1]{pbsi}
+\usepackage[utf8]{inputenc}
 
@@
 %\usepackage{minted}
-
+\usepackage[figurename=Fig.,labelfont=bf,labelsep=period]{caption}
+\usepackage{subcaption}
+\usepackage{amsmath}
 \usepackage{rotating}
 %% Cite Title
-\usepackage[style=numeric,backend=biber,natbib,sorting=none,maxcitenames=2,maxbibnames=99,doi=false,isbn=false,url=false,eprint=false]{biblatex}
-\addbibresource{../ai-trees-references.bib}
+%\usepackage[style=numeric,backend=biber,natbib,sorting=none,maxcitenames=2,maxbibnames=99,doi=false,isbn=false,url=false,eprint=false]{biblatex}
+%\addbibresource{../ai-trees-references.bib}
 
 %%% Counters
-\usepackage{chngcntr,mathtools}
-\counterwithout{figure}{section}
-\counterwithout{table}{section}
+% \usepackage{chngcntr,mathtools}
+% \counterwithout{figure}{section}
+% \counterwithout{table}{section}
 
-\numberwithin{equation}{section}
+% \numberwithin{equation}{section}
 
-%% Captions
-\usepackage{caption}
-\captionsetup{
-  labelsep=quad,
-  justification=raggedright,
-  labelfont=sc
-}
+% %% Captions
+% \usepackage{caption}
+% \captionsetup{
+%   labelsep=quad,
+%   justification=raggedright,
+%   labelfont=sc
+% }
 
 %AMS-TeX packages
-\usepackage{amssymb,amsmath,amsthm}
-\usepackage{bm}
-\usepackage[mathscr]{eucal}
-\usepackage{colortbl}
-\usepackage{color}
+%\usepackage{amssymb,amsmath,amsthm}
+%\usepackage{bm}
+%\usepackage[mathscr]{eucal}
+%\usepackage{colortbl}
+%\usepackage{color}
 
-\DeclareRobustCommand{\hlcyan}[1]{{\sethlcolor{cyan}\hl{#1}}}
+%\DeclareRobustCommand{\hlcyan}[1]{{\sethlcolor{cyan}\hl{#1}}}
 
-\usepackage{epstopdf,subfigure,hyperref,enumerate,polynom,polynomial}
+\usepackage{epstopdf,enumerate}
 \usepackage{multirow,minitoc,fancybox,array,multicol}
 
-\definecolor{slblue}{rgb}{0,.3,.62}
-\hypersetup{
-    colorlinks,%
-    citecolor=blue,%
-    filecolor=blue,%
-    linkcolor=blue,
-    urlcolor=slblue
-}
+\usepackage{newtxtext,newtxmath}
+\usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
+%
+%\definecolor{slblue}{rgb}{0,.3,.62}
+% \hypersetup{
+%     colorlinks,%
+%     citecolor=blue,%
+%     filecolor=blue,%
+%     linkcolor=blue,
+%     urlcolor=slblue
+% }
 
@@
 }
-
-%
-%Redefining sections as problems
-%
-\makeatletter
-\newenvironment{question}{\@startsection
-	{section}
-	{1}
-	{-.2em}
-	{-3.5ex plus -1ex minus -.2ex}
-    	{1.3ex plus .2ex}
-    	{\pagebreak[3]%forces pagebreak when space is small; use \eject for better results
-	\large\bf\noindent{Question }
-	}
-	}
-	%{\vspace{1ex}\begin{center} \rule{0.3\linewidth}{.3pt}\end{center}}
-	%\begin{center}\large\bf \ldots\ldots\ldots\end{center}}
-\makeatother
-
-%
-%Fancy-header package to modify header/page numbering
-%
-%\renewcommand{\chaptermark}[1]{ \markboth{#1}{} }
-\renewcommand{\sectionmark}[1]{ \markright{#1}{} }
-
-% \usepackage{fancyhdr}
-% \pagestyle{fancy}
-%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
-%\addtolength{\headwidth}{\marginparwidth}
-%\fancyheadoffset{30pt}
-%\fancyfootoffset{30pt}
-% \fancyhead[LO,RE]{\small  \it \nouppercase{\leftmark}}
-% \fancyhead[RO,LE]{\small Page \thepage}
-% \fancyfoot[RO,LE]{\small }% PR \num S-2015}
-% \fancyfoot[LO,RE]{\small }%\scshape MODL}
-% \cfoot{}
-% \renewcommand{\headrulewidth}{0.1pt}
-% \renewcommand{\footrulewidth}{0pt}
-%\setlength\voffset{-0.25in}
-%\setlength\textheight{648pt}
-
-\usepackage{paralist}
-
-%%% FORMAT PYTHON CODE
-\usepackage{listings}
-% Default fixed font does not support bold face
-\DeclareFixedFont{\ttb}{T1}{txtt}{bx}{n}{8} % for bold
-\DeclareFixedFont{\ttm}{T1}{txtt}{m}{n}{8}  % for normal
-
-% Custom colors
-\usepackage{color}
-\definecolor{deepblue}{rgb}{0,0,0.5}
-\definecolor{deepred}{rgb}{0.6,0,0}
-\definecolor{deepgreen}{rgb}{0,0.5,0}
-
-%\usepackage{listings}
-
-% % Python style for highlighting
-% \newcommand\pythonstyle{\lstset{
-% language=Python,
-% basicstyle=\footnotesize\ttm,
-% otherkeywords={self},             % Add keywords here
-% keywordstyle=\footnotesize\ttb\color{deepblue},
-% emph={MyClass,__init__},          % Custom highlighting
-% emphstyle=\footnotesize\ttb\color{deepred},    % Custom highlighting style
-% stringstyle=\color{deepgreen},
-% frame=tb,                         % Any extra options here
-% showstringspaces=false            %
-% }}
-
-% % Python environment
-% \lstnewenvironment{python}[1][]
-% {
-% \pythonstyle
-% \lstset{#1}
-% }
-% {}
-
-% % Python for external files
-% \newcommand\pythonexternal[2][]{{
-% \pythonstyle
-% \lstinputlisting[#1]{#2}}}
-
-% % Python for inline
-% \newcommand\pythoninline[1]{{\pythonstyle\lstinline!#1!}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\NameTag{AuthorOneLastName, \today}
 
 \begin{document}
-%\vspace{-16ex}
+% ALL CAPS TITLE REQUIRED
 \title{Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network}
-%\author{}
-\date{}
+\author[1]{Author One}
+\author[2]{Author Two}
+
+\affil[1]{First affiliation address, with corresponding author email. Email: author.one@email.com}
+\affil[2]{Second affiliation address}
+
 \maketitle
 
+\begin{abstract}
+
+\end{abstract}
 \section{Introduction}
@@
 For example, in Massachusetts, USA, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
-The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citet{nateghi2014power}.
+The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citeN{nateghi2014power}.
 
@@
 The proportion of tree failures away from the wires that causes outages varies and has not been rigorously quantified.
-\citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
+\citeN{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citeN{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
 
@@
 AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}.
-Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citet{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
+Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citeN{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
 Neural networks have also been used to interrogate remote sensing data of the landscape to assess landslide risk \cite{su2020deep}.
@@
 %Neural networks developed over the past several decades from the introduction of the single-layer perceptron.
-The groundbreaking study of \citet{hubel1959receptive} showed that visual perception in cats was a result of the
+The groundbreaking study of \citeN{hubel1959receptive} showed that visual perception in cats was a result of the
 activation or inhibition of groups of cells in the visual cortex known as ``receptive fields.''  Further, they attempted
@@
 AlexNet \cite{krizhevsky2012imageneta}, with 5 convolutional layers and 3 dense layers---one of the largest CNNs of its time, won the ILSVRC-2012\footnote{ImageNet Large Scale Visual Recognition Challenge; held annually from 2010 through 2017.} competition with a top-5 error rate of 15.3\% and served as a landmark in the Deep Learning subdomain.
-\citet{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization techniques that were foundational for model inference and interpretability.
+\citeN{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization techniques that were foundational for model inference and interpretability.
 In the same year, GoogLeNet, a 22-layer network, was proposed \cite{szegedy2014going}, featuring the novel ``Inception module,'' which allowed for efficiency and accuracy in a very deep network.
@@
 
-\printbibliography
+\bibliography{../ai-trees-references}
 

### 2021-02-04 docs/manuscript/manuscript.tex
updated manuscript to fix compile errors

@@
-\documentclass[Journal, letterpaper]{ascelike-new}
-%\usepackage{etex, soul}
+\documentclass[Journal,letterpaper]{ascelike-new}
+%\usepackage{etex}
 %\newcommand{\num}{6{} }
@@
 % \geometry{top=1.2in, left=1in,right=1in,bottom=1in,headsep=6pt}
- \usepackage{graphicx,booktabs,calc}
 
-%=== GRAPHICS PATH ===========
-\graphicspath{{./images/}}
 % Marginpar width
@@
 % \usepackage[T1]{pbsi}
-\usepackage[utf8]{inputenc}
-
+%\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage{lmodern}
-\usepackage[T1]{fontenc}
 %\usepackage{minted}
@@
 \usepackage{amsmath}
-\usepackage{rotating}
+%\usepackage{rotating}
+
+\usepackage{graphicx,booktabs}
+
+%=== GRAPHICS PATH ===========
+\graphicspath{{./images/}}
 %% Cite Title
@@
 
-\usepackage{epstopdf,enumerate}
-\usepackage{multirow,minitoc,fancybox,array,multicol}
+\usepackage{enumerate}
 
@@
 \usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
+
+\usepackage{multirow,array,multicol}
+
 %
@@
 %%%TIKZ
-\usepackage{tikz}
-\usepackage{pgfplots}
-\usepackage{pgfplotstable}
-\usepackage{pgfgantt}
-\pgfplotsset{compat=newest}
+% \usepackage{tikz}
+% \usepackage{pgfplots}
+% \usepackage{pgfplotstable}
+% \usepackage{pgfgantt}
+% \pgfplotsset{compat=newest}
 
-\usetikzlibrary{arrows,shapes,positioning}
-\usetikzlibrary{decorations.markings}
-\usetikzlibrary{shadows,automata}
-\usetikzlibrary{patterns,fit}
-%\usetikzlibrary{circuits.ee.IEC}
-\usetikzlibrary{decorations.text}
-% For Sagnac Picture
-\usetikzlibrary{%
-    decorations.pathreplacing,%
-    decorations.pathmorphing%
-}
+% \usetikzlibrary{arrows,shapes,positioning}
+% \usetikzlibrary{decorations.markings}
+% \usetikzlibrary{shadows,automata}
+% \usetikzlibrary{patterns,fit}
+% %\usetikzlibrary{circuits.ee.IEC}
+% \usetikzlibrary{decorations.text}
+% % For Sagnac Picture
+% \usetikzlibrary{%
+%     decorations.pathreplacing,%
+%     decorations.pathmorphing%
+% }
 
-\newcommand{\osn}{\oldstylenums}
-\newcommand{\dg}{^{\circ}}
-\newcommand{\lt}{\left}
-\newcommand{\rt}{\right}
-\newcommand{\pt}{\phantom}
-\newcommand{\tf}{\therefore}
-\newcommand{\?}{\stackrel{?}{=}}
-\newcommand{\fr}{\frac}
-\newcommand{\dfr}{\dfrac}
-%\newcommand{\ul}{\underline}
-\newcommand{\tn}{\tabularnewline}
-\newcommand{\nl}{\newline}
-\newcommand\relph[1]{\mathrel{\phantom{#1}}}
-\newcommand{\cm}{\checkmark}
-\newcommand{\ol}{\overline}
-\newcommand{\rd}{\color{red}}
-\newcommand{\bl}{\color{blue}}
-\newcommand{\pl}{\color{purple}}
-\newcommand{\og}{\color{orange!90!black}}
-\newcommand{\gr}{\color{green!40!black}}
-\newcommand{\nin}{\noindent}
-\newcommand{\la}{\lambda}
-\renewcommand{\th}{\theta}
-\newcommand{\al}{\alpha}
-\newcommand{\G}{\Gamma}
-\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-            \node[shape=circle,draw,thick,inner sep=1pt] (char) {\small #1};}}
+% \newcommand{\osn}{\oldstylenums}
+% \newcommand{\dg}{^{\circ}}
+% \newcommand{\lt}{\left}
+% \newcommand{\rt}{\right}
+% \newcommand{\pt}{\phantom}
+% \newcommand{\tf}{\therefore}
+% \newcommand{\?}{\stackrel{?}{=}}
+% \newcommand{\fr}{\frac}
+% \newcommand{\dfr}{\dfrac}
+% %\newcommand{\ul}{\underline}
+% \newcommand{\tn}{\tabularnewline}
+% \newcommand{\nl}{\newline}
+% \newcommand\relph[1]{\mathrel{\phantom{#1}}}
+% \newcommand{\cm}{\checkmark}
+% \newcommand{\ol}{\overline}
+% \newcommand{\rd}{\color{red}}
+% \newcommand{\bl}{\color{blue}}
+% \newcommand{\pl}{\color{purple}}
+% \newcommand{\og}{\color{orange!90!black}}
+% \newcommand{\gr}{\color{green!40!black}}
+% \newcommand{\nin}{\noindent}
+% \newcommand{\la}{\lambda}
+% \renewcommand{\th}{\theta}
+% \newcommand{\al}{\alpha}
+% \newcommand{\G}{\Gamma}
+% \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
+%             \node[shape=circle,draw,thick,inner sep=1pt] (char) {\small #1};}}
 
-\newcommand{\bc}{\begin{compactenum}[\quad--]}
-\newcommand{\ec}{\end{compactenum}}
-
-\newcommand{\p}{\partial}
-\newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
-\newcommand{\dpd}[2]{\dfrac{\partial{#1}}{\partial{#2}}}
-\newcommand{\pdd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}
+% \newcommand{\p}{\partial}
+% \newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
+% \newcommand{\dpd}[2]{\dfrac{\partial{#1}}{\partial{#2}}}
+% \newcommand{\pdd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}
 
@@
 \begin{abstract}
+    Abstract here.
+\end{abstract}
 
-\end{abstract}
 \section{Introduction}
@@
 Presently, the identification of potential contact between trees and power lines is labor intensive and time-consuming.
-This paper describes an artificial intelligence and machine learning approach that automatically classifies trees, using only a single photograph and with a high degree of accuracy, into categories corresponding to probable failure, possible failure and improbable failure--categories corresponding to those used by utility arborists to assess risk of contact between trees and power lines.
+This paper describes an artificial intelligence and machine learning approach that automatically classifies trees, using only a single photograph and with a high degree of accuracy, into categories corresponding to probable failure, possible failure and improbable failure-categories corresponding to those used by utility arborists to assess risk of contact between trees and power lines.
 This preliminary study demonstrates the possible efficacy of AI approaches to tree risk assessment and, following further development of the approach has the potential to reduce power outages and utility costs by allowing utilities to more effectively target their pruning and mitigation efforts.
@@
 classification, object and text detection, object and document tracking, labeling, speech, among several other related
-fields \cite{gu2018recent}. \hl{Paragraph on applications to be completed---with particular attention paid to tree-related efforts.}
+fields \cite{gu2018recent}. %\hl{Paragraph on applications to be completed---with particular attention paid to tree-related efforts.}
 
@@
 \subsection{Pre-processing and data augmentation}
-In the original set of training images, the class distribution is given in \autoref{tab:classdist}.
+In the original set of training images, the class distribution is given in \ref{tab:classdist}.
 
@@
 We used the Hyperband approach \cite{li2018hyperband} to perform a grid search to find the optimal values of the following hyperparameters:
-\begin{compactitem}
+\begin{itemize}
     \item kernel size in first convolutional layer
@@
     \item learning rate (of Adam optimizer)
-\end{compactitem}
+\end{itemize}
 
@@
 
-We define four classification scenarios in \autoref{tab:scenarios}.
+We define four classification scenarios in \ref{tab:scenarios}.
 
@@
 \subsection{Comparison with state-of-the-art architectures}
-We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \autoref{tab:comp}.
+We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \ref{tab:comp}.
 
@@
 
-\bibliography{../ai-trees-references}
+\section{Data Availability Statement}
+
+When submitting a new and revised manuscript, authors are asked to include a data availability statement containing one or more of the following statements, with specific items listed as appropriate.
+Please include one or more of the statements below, deleting those which do not apply.
+This section should appear directly before the Acknowledgments section.
+
+\begin{itemize}
+\item Some or all data, models, or code generated or used during the study are available in a repository online in accordance with funder data retention policies (provide full citations that include URLS or DOIs)
+\item Some or all data, models, or code used during the study were provided by a third party (list items).
+Direct requests for these materials may be made to the provider as indicated in the Acknowledgements.
+\item Some or all data, models, or code that support the findings of this study are available from the corresponding author upon reasonable request (list items).
+\item Some or all data, models, or code generated or used during the study are proprietary or confidential in nature and may only be provided with restrictions (e.g.~anonymized data) (List items and restrictions).
+\item All data, models, and code generated or used during the study appear in the submitted article.
+\item No data, models, or code were generated or used during the study (e.g., opinion or data-less paper).
+\end{itemize}
+
+Please also see the guidelines at: \url{https://ascelibrary.org/page/dataavailability}.
+
+\section{Acknowledgments}
 
@@
 
+%\bibliographystyle{ascelike-new}
+\bibliography{../ai-trees-references}
+
 \end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2021-02-04 docs/manuscript/manuscript.tex
removed conflicting aux files
Updates from Overleaf
fixed compile errors

@@
-\documentclass[Journal,letterpaper]{ascelike-new}
+\documentclass[Journal,letterpaper, SingleSpace]{ascelike-new}
 %\usepackage{etex}
@@
 % \geometry{top=1.2in, left=1in,right=1in,bottom=1in,headsep=6pt}
+\usepackage{graphicx,booktabs, calc}
 
+%=== GRAPHICS PATH ===========
+\graphicspath{{./images/}}
 % Marginpar width
@@
 %\usepackage{rotating}
+\usepackage{newtxtext,newtxmath}
+\usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
 
-\usepackage{graphicx,booktabs}
-
-%=== GRAPHICS PATH ===========
-\graphicspath{{./images/}}
 %% Cite Title
@@
 \usepackage{enumerate}
-
-\usepackage{newtxtext,newtxmath}
-\usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\NameTag{AuthorOneLastName, \today}
+\NameTag{Author1LastName, \today}
 
@@
 \title{Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network}
-\author[1]{Author One}
-\author[2]{Author Two}
+\author[1]{Author 1}
+\author[2]{Author 2}
+\author[2]{Author 3}
+\author[2]{Author 4}
+\author[2]{Author 5}
 
-\affil[1]{First affiliation address, with corresponding author email. Email: author.one@email.com}
+\affil[1]{First affiliation address, with corresponding author email. Email: author.one@umass.edu}
 \affil[2]{Second affiliation address}
@@
 For example, in Massachusetts, USA, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
-The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citeN{nateghi2014power}.
+The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \cite{nateghi2014power}.
 
@@
 ESRI ArcMaps was used to randomly distribute sampling sites across the state.
-Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.
+Likelihood of failure assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.
 This method is commonly used to assess trees in the United States.
@@
 
-%\bibliographystyle{ascelike-new}
-\bibliography{../ai-trees-references}
+\bibliography{ai-trees-references}
 
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2021-02-04 docs/manuscript/manuscript.tex
updated citations.

@@
 
-Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}.
+Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{smiley2017best,goodfellow2020best}.
 Risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences.
@@
 We can apply cut-out (occluding portions of the image) for improved performance \cite{devries2017improved}.
-Also, it has been shown that training with lower resolution improves performance on higher resolution test images \cite{touvronfixing}.
+Also, it has been shown that training with lower resolution improves performance on higher resolution test images \cite{touvron2019fixing}.
 
@@
 \section{Data Availability Statement}
-
-When submitting a new and revised manuscript, authors are asked to include a data availability statement containing one or more of the following statements, with specific items listed as appropriate.
-Please include one or more of the statements below, deleting those which do not apply.
-This section should appear directly before the Acknowledgments section.
-
-\begin{itemize}
-\item Some or all data, models, or code generated or used during the study are available in a repository online in accordance with funder data retention policies (provide full citations that include URLS or DOIs)
-\item Some or all data, models, or code used during the study were provided by a third party (list items).
-Direct requests for these materials may be made to the provider as indicated in the Acknowledgements.
-\item Some or all data, models, or code that support the findings of this study are available from the corresponding author upon reasonable request (list items).
-\item Some or all data, models, or code generated or used during the study are proprietary or confidential in nature and may only be provided with restrictions (e.g.~anonymized data) (List items and restrictions).
-\item All data, models, and code generated or used during the study appear in the submitted article.
-\item No data, models, or code were generated or used during the study (e.g., opinion or data-less paper).
-\end{itemize}
+All data, models, or code generated or used during the study are available in a repository online in accordance with funder data retention policies (provide full citations that include URLS or DOIs)
 

### 2021-02-09 bin/python/label-image-files.py
updated resolution test to read parameters from dict; and implemented raw image plotting in label-image-files.py

@@ -6,8 +6,12 @@ import numpy as np
 import os # used for navigating to image path
 import imageio # used for writing images
 import re # for matching image file name classes
+import matplotlib.pyplot as plt
+import random
+import ntpath
 
 raw_image_dir = '../../data/raw/Pictures for AI'
+tidy_image_dir = '../../data/tidy/labeled_images/'
 
 def getListOfFiles(dirName):
     """Returns single list of the filepath of each of the training image files"""
@@ -32,7 +36,6 @@ def saveImageFiles(image_file_list):
     improbable_counter = 1
     possible_counter = 1
     probable_counter = 1
-    tidy_image_dir = '../../data/tidy/labeled_images/'
     if not os.path.exists(tidy_image_dir):
         os.makedirs(tidy_image_dir)
     for filename in image_file_list:
@@ -51,10 +54,33 @@ def saveImageFiles(image_file_list):
     print('Number of possible images saved:', possible_counter - 1)
     print('Number of probable images saved:', probable_counter - 1)
 
+def plotRawImages(image_file_list, images_per_class = 2):
+    fig, axarr = plt.subplots(images_per_class, 3, sharex=True, sharey=True)
+    class_index = 0
+    class_list = ['probable', 'possible', 'improbable']
+    #print(image_file_list)
+    for image_class in class_list:
+        class_images = [i for i in image_file_list if ntpath.basename(i).startswith(image_class)]
+        print(len(class_images))
+        random.seed(111)
+        random_class_selection = random.choices(class_images, k = images_per_class)
+        for i in range(images_per_class):
+            axarr[i,class_index].imshow((Image.open(random_class_selection[i])).transpose(Image.TRANSPOSE), aspect='auto', 
+                interpolation='antialiased', extent = [0, 3024, 0, 4032])
+        class_index += 1
+    for ax, col in zip(axarr[0,:], [i.title() for i in class_list]):
+        ax.set_title(col, size=15)
+    image_filename = '../../figures/raw_input_images_' + str(images_per_class) + '.png'
+    plt.xticks([0,3024])
+    plt.yticks([0,4032])
+    fig.savefig(image_filename, dpi=120)
+
 def main():    
     tree_image_list = getListOfFiles(raw_image_dir)
     saveImageFiles(tree_image_list)
-    
+    tree_image_list_labeled = getListOfFiles(tidy_image_dir)
+    plotRawImages(tree_image_list_labeled, images_per_class = 2)
+
 if __name__ == "__main__":
     main()
 

### 2021-02-09 docs/manuscript/manuscript.tex
finished adding plot function to preprocess-images.py

@@
 Individual tree risk assessment can be costly because of the time it requires.
-In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{e.thomassmiley2017best}.
+In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{smiley2017best}.
 Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}.
@@
 ESRI ArcMaps was used to randomly distribute sampling sites across the state.
-Likelihood of failure assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.
+Likelihood of failure assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{smiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{goodfellow2020best}.
 This method is commonly used to assess trees in the United States.
 A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
-Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.
-The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{e.thomassmiley2017best}:
+Level 2 or Level 3 \cite{smiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.
+The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{smiley2017best}:
 \begin{itemize}
@@
 In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category.
+Two randomly chosen examples are shown for each category in \autoref{fig:raw_images}.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=.7\textwidth]{raw_input_images_2}
+  \caption{Random examples of the raw training images (2 are shown per class). Images are antialiased in this figure for greater clarity.}
+  \label{fig:raw_images}
+\end{figure}
 
@@
 Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{processed_input_images_Pr_Po_Im_64_px}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{pr_po_im_64}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_64_px}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{pr_im_64}
+  \end{subfigure}%
+
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_64_px}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{prpo_im_64}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_64_px}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{pr_poim_64}
+  \end{subfigure}%
+
+  \caption{Random examples of the processed training images under each classification scenario; 64 pixels.}
+  \label{fig:processed_images}
+\end{figure}
 

### 2021-02-09 bin/python/label-image-files.py

@@ -61,7 +61,6 @@ def plotRawImages(image_file_list, images_per_class = 2):
     #print(image_file_list)
     for image_class in class_list:
         class_images = [i for i in image_file_list if ntpath.basename(i).startswith(image_class)]
-        print(len(class_images))
         random.seed(111)
         random_class_selection = random.choices(class_images, k = images_per_class)
         for i in range(images_per_class):

### 2021-02-09 bin/python/preprocess-images.py

@@ -14,7 +14,6 @@ import os # used for navigating to image path
 import imageio # used for writing images
 import random
 import matplotlib.pyplot as plt
-#from timeit import default_timer as timer
 
 ## GLOBAL VARIABLES
 IMG_SIZE = 108
@@ -26,7 +25,7 @@ EXPANSION_FACTOR = 5 # of augmented images
 LABELED_IMAGES_DIR = '../../data/tidy/labeled_images'
 PROCESSED_IMAGES_DIR = '../../data/tidy/preprocessed_images'
 SEED = 10 # seed for repeatability
-
+NUM_PLOT_IMAGES_PER_CLASS = 4
 
 def getImageOneHotVector(image_file_name, classification_scenario = "Pr_Im"):
     """Returns one-hot vector encoding for each image based on specified classification scenario:
@@ -67,7 +66,7 @@ def getImageOneHotVector(image_file_name, classification_scenario = "Pr_Im"):
         else :
             return np.array([0, 0]) # if label is not present for current image        
 
-def processImageData(img_size, expansion_factor, class_scenario, seed_value, channels=1): # original size 4032 × 3024 px
+def processImageData(img_size, expansion_factor, class_scenario, seed_value, channels=1, images_per_class=4): # original size 4032 × 3024 px
     data = []
     image_list = os.listdir(LABELED_IMAGES_DIR)
     random.seed(seed_value) #seed for repeatability
@@ -100,36 +99,50 @@ def processImageData(img_size, expansion_factor, class_scenario, seed_value, cha
             #img = img.crop((left=400, top=0, r=3424, b=3024)) 
             #img = img.resize((img_size, img_size), Image.BICUBIC)
             #TODO: random cropping, flipping and resizing
+
+    random_image_selection_class_0 = random.choices([i[0] for i in data if i[1][0] == 1], k = images_per_class)
+    random_image_selection_class_1 = random.choices([i[0] for i in data if i[1][1] == 1], k = images_per_class)
+    image_selection_array = [random_image_selection_class_0, random_image_selection_class_1]
+    if class_scenario == "Pr_Im":
+        class_list = ["Probable", "Improbable"]
+    elif class_scenario == "PrPo_Im":
+        class_list = ["Probable/Possible", "Improbable"]
+    elif class_scenario == "Pr_PoIm":
[…]

### 2021-02-09 docs/manuscript/manuscript.tex
reorganized data section of manuscript

@@
 %=== GRAPHICS PATH ===========
-\graphicspath{{./images/}}
+\graphicspath{{./figures/}}
 % Marginpar width
@@
 Two randomly chosen examples are shown for each category in \autoref{fig:raw_images}.
+In the original set of training images, the class distribution is given in \autoref{tab:classdist}.
 
@@
 \end{figure}
-
-\subsection{Pre-processing and data augmentation}
-In the original set of training images, the class distribution is given in \ref{tab:classdist}.
 
@@
 
+\subsection{Classification scenarios}
+We define four classification scenarios in \autoref{tab:scenarios}.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l l}\toprule
+    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
+    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
+    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
+    \texttt{PrPo\_Im}        & \{Probable|Possible, Improbable\}  & 2 \\
+    \texttt{Pr\_PoIm}        & \{Probable, Possible|Impossible\}  & 2 \\\bottomrule
+    \end{tabular}
+    \caption{Classification scenarios}
+    \label{tab:scenarios}
+\end{table}
+
+\subsection{Pre-processing and data augmentation}
 To achieve robustness in training, and given the relatively small number of training images, we randomly cropped each image on either axis to 3024 $\times$ 3024 pixels, generating five instances for each one.
@@
 Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
+Random sets of images from each class across each of the four classification scenarios are shown in \autoref{fig:processed_images}.
 
@@
 
-\subsection{Classification experiments}
-
-We define four classification scenarios in \ref{tab:scenarios}.
-
-\begin{table}[h!]
-    \centering
-    \begin{tabular}{l l l}\toprule
-    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
-    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
-    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
-    \texttt{PrPo\_Im}        & \{Probable|Possible, Improbable\}  & 2 \\
-    \texttt{Pr\_PoIm}        & \{Probable, Possible|Impossible\}  & 2 \\\bottomrule
-    \end{tabular}
-    \caption{Classification scenarios}
-    \label{tab:scenarios}
-\end{table}
-
 \subsection{Sensitivity to training resolution}
@@
 \subsection{Comparison with state-of-the-art architectures}
-We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \ref{tab:comp}.
+We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \autoref{tab:comp}.
 

### 2021-02-11 bin/python/cnn-hyperparameter-optimization.py
created cnn-hyperparameter-optimization.py script

@@ -1,0 +1,173 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+from PIL import Image # used for loading images
+import numpy as np
+import os # used for navigating to image path
+import imageio # used for writing images
+import random
+import tensorflow as tf
+from tensorflow.keras import datasets, layers, models
+from timeit import default_timer as timer
+from tensorflow.keras import backend as K
+
+import sys
+sys.path.append("../python/")
+from helpers import *
+
+"""HP Tuning"""
+from kerastuner import HyperModel
+from kerastuner.tuners import RandomSearch
+
+import IPython
+
+import kerastuner as kt
+from sklearn import ensemble
+from sklearn import datasets
+from sklearn import linear_model
+from sklearn import metrics
+from sklearn import model_selection
+
+
+# In[ ]:
+
+
+NUM_CHANNELS = 1
+RESOLUTION_LIST = [64, 128, 224, 384]
+SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
+OPTIMAL_HYPERPARAMETERS_PATH = '../../results/optimal-hyperparameters/'
+HYPERBAND_MAX_EPOCHS = 10 #10
+EXECUTIONS_PER_TRIAL = 3 #5
+HYPERBAND_ITER = 5 #80
+
+
+# In[ ]:
+
+
+image_dict = dict.fromkeys(RESOLUTION_LIST)
+for p in RESOLUTION_LIST:
+    image_dict[p] = dict.fromkeys(SCENARIO_LIST)
+    for s in SCENARIO_LIST:
+        image_dict[p][s] = np.load('../../data/tidy/preprocessed_images/size' + str(p) + '_exp5_' + s + '.npy', allow_pickle = True)
+
+
+# In[ ]:
+
+
+class CNNHyperModel(HyperModel):
+    def __init__(self, input_image_shape, num_classes):
+        self.input_image_shape = input_image_shape
+        self.num_classes = num_classes
+
+    def build(self, hp):
+        model = models.Sequential()
+
+        ## Vary kernel size in first Conv Layer between 5 and 7
+        hp_k_size = hp.Choice('kernel_size', values = [5, 7])
+
+        # Tune the number of units in the first and second Dense layers
+        # Choose an optimal value between 32-512
+        dense_units_l = hp.Int('units_1', min_value = 32, max_value = 512, step = 32, default = 128)
[…]

### 2021-02-11 bin/python/cnn-hyperparameter-optimization.py
fixed minor issues in optimization python script

@@ -140,10 +140,12 @@ def optimizeCNNHyperparameters(scenario, image_size, seed_val = 1, save_results=
         scenario_path = os.path.join(OPTIMAL_HYPERPARAMETERS_PATH, str(image_size), s)
         if not os.path.exists(scenario_path):
             os.makedirs(scenario_path)
+        with open(os.path.join(scenario_path, 'console.txt'), 'w') as f:
+            f.write(json.dumps(console_printout))
+        with open(os.path.join(scenario_path, 'results-summary.txt'), 'w') as f:
+            f.write(json.dumps(results))               
         with open(os.path.join(scenario_path, 'hyperparameters.txt'), 'w') as f:
-            f.write(json.dumps(best_hps_dict))
-        with open(os.path.join(scenario_path, 'performance-summary.txt'), 'w') as f:
-            f.write(json.dumps(summary))            
+            f.write(json.dumps(best_hps_dict))         
     return #(console_printout, summary, best_hps_dict)
 
 

### 2021-02-12 bin/python/cnn-hyperparameter-optimization.py
further updates to cnn-hyperparameter-optimization.py and improvements to resolution-sensitivity.ipynb

@@ -28,30 +28,19 @@ from sklearn import linear_model
 from sklearn import metrics
 from sklearn import model_selection
 
-
-# In[ ]:
-
-
 NUM_CHANNELS = 1
 RESOLUTION_LIST = [64, 128, 224, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 OPTIMAL_HYPERPARAMETERS_PATH = '../../results/optimal-hyperparameters/'
 HYPERBAND_MAX_EPOCHS = 10 #10
-EXECUTIONS_PER_TRIAL = 3 #5
-HYPERBAND_ITER = 5 #80
-
-
-# In[ ]:
-
+EXECUTIONS_PER_TRIAL = 2 #5
+HYPERBAND_ITER = 3 #80
 
 image_dict = dict.fromkeys(RESOLUTION_LIST)
 for p in RESOLUTION_LIST:
     image_dict[p] = dict.fromkeys(SCENARIO_LIST)
     for s in SCENARIO_LIST:
         image_dict[p][s] = np.load('../../data/tidy/preprocessed_images/size' + str(p) + '_exp5_' + s + '.npy', allow_pickle = True)
-
-
-# In[ ]:
 
 
 class CNNHyperModel(HyperModel):
@@ -107,16 +96,9 @@ class CNNHyperModel(HyperModel):
         return model
 
 
-# In[ ]:
-
-
 class ClearTrainingOutput(tf.keras.callbacks.Callback):
     def on_train_end(*args, **kwargs):
         IPython.display.clear_output(wait = True)
-
-
-# In[ ]:
-
 
 def optimizeCNNHyperparameters(scenario, image_size, seed_val = 1, save_results=True):
     if scenario=="Pr_Po_Im":
@@ -132,30 +114,18 @@ def optimizeCNNHyperparameters(scenario, image_size, seed_val = 1, save_results=
     training_images, training_labels = getImageAndLabelArrays(training_images_and_labels)
     validation_images, validation_labels = getImageAndLabelArrays(test_images_and_labels)
     
-    console_printout = tuner.search(training_images, training_labels, validation_data = (validation_images, validation_labels), callbacks = [ClearTrainingOutput()])
-    results = tuner.results_summary()
+    tuner.search(training_images, training_labels, validation_data = (validation_images, validation_labels), callbacks = [ClearTrainingOutput()])
[…]

### 2021-02-12 docs/manuscript/manuscript.tex
created cnn-hyperparameter-optimization.py script

@@
 %\subsection{AI and risk assessment}
-AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances}.
+AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances, wang2019novel}.
 Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citeN{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.

### 2021-02-15 docs/manuscript/manuscript.tex

@@
-\documentclass[Journal,letterpaper, SingleSpace]{ascelike-new}
+\documentclass[Journal,letterpaper, SingleSpace, InsideFigs]{ascelike-new}
 %\usepackage{etex}
@@
 \begin{itemize}
-    \item kernel size in first convolutional layer
-    \item number of units in first densely connected layer
-    \item dropout rate applied to outputs of first dense layer
-    \item activation function for first dense layer
-    \item number of units in second densely connected layer
-    \item dropout rate applied to outputs of second dense layer
-    \item activation function for second dense layer    
-    \item learning rate (of Adam optimizer)
+    \item kernel size in first convolutional layer $ks$
+    \item number of units in first densely connected layer $u_1$
+    \item dropout rate applied to outputs of first dense layer $d_1$
+    \item activation function for first dense layer $a_1$
+    \item number of units in second densely connected layer $u_2$
+    \item dropout rate applied to outputs of second dense layer $d_2$
+    \item activation function for second dense layer $a_2$
+    \item learning rate (of Adam optimizer) $\lambda$
 \end{itemize}
 
+\begin{table}[h!]\small
+    \centering
+    \begin{tabular}{|l | l l l l| l l l l| l l l l | l l l l|}\hline
+    %\bf Scenario 
+        &  \multicolumn{4}{c|}{\texttt{Pr\_Po\_Im}} 
+            & \multicolumn{4}{c|}{\texttt{Pr\_Im}} 
+                &  \multicolumn{4}{c|}{\texttt{PrPo\_Im}} 
+                    &  \multicolumn{4}{c|}{\texttt{Pr\_PoIm}}  \\\hline
+    %\bf Resolution
+        & 64 & 128 & 224 & 384 
+            & 64 & 128 & 224 & 384 
+                & 64 & 128 & 224 & 384
+                    & 64 & 128 & 224 & 384  \\\hline
+    %\bf Hyperparameter 
+    $ks$    &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
+    $u_1$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
+    $d_1$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
+    $a_1$   & ReLu  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
+    $u_2$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\ 
+    $d_2$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
+    $a_2$   & ReLu &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
+    $\lambda$  
+            &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\\hline
+    \end{tabular}
+    \caption{Optimal hyperparameters found using the Hyperband search algorithm for all classification scenario and input image resolution combinations.}
+    \label{tab:opt-hyp}
+\end{table}
 \section{Results}

### 2021-02-15 bin/python/helpers.py
added resolution-scenario-sensitivity.py script

@@ -43,6 +43,21 @@ def getImageAndLabelArrays(image_label_tuple_array, num_channels = 1):
 	label_array = np.array([x[1] for x in image_label_tuple_array])
 	return(image_array, label_array)
 
+
+def createResolutionScenarioImageDict(resolution_list, scenario_list):
+    image_dict = dict.fromkeys(resolution_list)
+    for p in resolution_list:
+        image_dict[p] = dict.fromkeys(scenario_list)
+        for s in scenario_list:
+            image_dict[p][s] = np.load('../../data/tidy/preprocessed_images/size' + str(p) + '_exp5_' + s + '.npy', allow_pickle = True)
+    return(image_dict)
+
+def getOptCNNHyperparams(image_size, scenario):
+    with open('../../results/optimal-hyperparameters/' + str(image_size) + '/' + scenario + '/hyperparameters.txt') as f: 
+        data = f.read() 
+    opt_params_dict = json.loads(data)   
+    return(opt_params_dict)
+
 #https://github.com/keras-team/keras/issues/5400#issuecomment-408743570
 def check_units(y_true, y_pred):
 	if y_pred.shape[1] != 1:

### 2021-02-15 bin/python/resolution-scenario-sensitivity.py

@@ -1,0 +1,108 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+# This script explores CNN model performance based on training image resolution.
+# We test the following training image sizes:
+# 64 x 64, 128 x 128, 224 x 224, 384 x 384
+
+import numpy as np
+from timeit import default_timer as timer
+import random
+
+import tensorflow as tf
+from tensorflow.keras import datasets, layers, models
+from tensorflow.keras.optimizers import SGD
+from tensorflow.keras import backend as K
+from tensorflow.keras.callbacks import Callback
+
+import json
+import pickle
+import os
+import sys
+sys.path.append("../python/")
+from helpers import *
+
+
+# Globals
+NUM_CHANNELS = 1
+RESOLUTION_LIST = [64, 128] #, 224, 384]
+SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
+NUM_MODEL_RUNS = 10
+NUM_EPOCHS = 10
+RESOLUTION_PERFORMANCE_METRICS_DIR = '../../results/resolution-tests'
+
+
+def constructBaseCNN(image_size, scenario):
+    image_shape = (image_size, image_size, NUM_CHANNELS)
+    p_dict = getOptCNNHyperparams(image_size, scenario)
+    if scenario=="Pr_Po_Im":
+        num_classes = 3
+    else:
+        num_classes = 2
+    base_model = models.Sequential([
+        layers.Conv2D(filters = 64, kernel_size = p_dict['kernel_size'], strides = 2, activation="relu", padding="same", input_shape = image_shape),
+        layers.MaxPooling2D(2),
+        layers.Conv2D(128, 3, activation="relu", padding="same"),
+        layers.Conv2D(128, 3, activation="relu", padding="same"),
+        layers.MaxPooling2D(2),
+        layers.Conv2D(256, 3, activation="relu", padding="same"),
+        layers.Conv2D(256, 3, activation="relu", padding="same"),
+        layers.MaxPooling2D(2),
+        layers.Flatten(),
+        
+        layers.Dense(p_dict['units_1'], activation = p_dict['activation_1']),
+        layers.BatchNormalization(),
+        layers.Dropout(p_dict['dropout_1']), 
+        
+        layers.Dense(p_dict['units_2'], activation = p_dict['activation_2']), 
+        layers.Dropout(p_dict['dropout_2']),
+        
[…]

### 2021-02-15 bin/python/resolution-scenario-sensitivity.py
fixed error in sensitivity testing script

@@ -29,7 +29,7 @@ RESOLUTION_LIST = [64, 128] #, 224, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 NUM_MODEL_RUNS = 10
 NUM_EPOCHS = 10
-RESOLUTION_PERFORMANCE_METRICS_DIR = '../../results/resolution-tests'
+RESOLUTION_PERFORMANCE_METRICS_DIR = '../../results/sensitivity-tests'
 
 
 def constructBaseCNN(image_size, scenario):
@@ -94,7 +94,7 @@ def main(num_trials = NUM_MODEL_RUNS):
     for s in SCENARIO_LIST:
         for p in RESOLUTION_LIST:
             for i in range(num_trials):
-                print("Conducting performance test: Scenario- " + s + "; Resolution - " + p + "; Trial: " + str(i+1))
+                print("Conducting performance test: Scenario - " + s + "; Resolution - " + str(p) + "px; Trial - " + str(i+1))
                 scenario_performance_dict = testResolutionScenarioPerformance(p, s, num_epochs = NUM_EPOCHS, trial_seed = 1 + i) #ultimately should be averaged across trials       
                 scenario_filename = "scenario_resolution_performance_" + s + "_trial_" + str(i+1) + ".txt"
                 hist_filename = "scenario_resolution_hist_" + s + "_trial_" + str(i+1) + ".txt"

### 2021-02-15 bin/python/helpers.py
added scenario-resolution-performance plot

@@ -1,3 +1,4 @@
+import os
 import numpy as np
 from timeit import default_timer as timer
 import matplotlib.pyplot as plt
@@ -15,6 +16,24 @@ from sklearn.metrics import recall_score, classification_report
 from sklearn.datasets import make_classification
 
 import json
+
+# Handling files
+def getListOfFiles(dirName):
+    """Returns single list of the filepath of each of the training image files"""
+    # source: https://thispointer.com/python-how-to-get-list-of-files-in-directory-and-sub-directories/
+    listOfFile = os.listdir(dirName)
+    allFiles = list()
+    # Iterate over all the entries
+    for entry in listOfFile:
+        # Create full path
+        fullPath = os.path.join(dirName, entry)
+        # If entry is a directory then get the list of files in this directory 
+        if os.path.isdir(fullPath):
+            allFiles = allFiles + getListOfFiles(fullPath)
+        else:
+            allFiles.append(fullPath)
+                
+    return allFiles
 
 def splitData(image_array, prop = 0.80, seed_num = 111):
 	"""Returns training and test arrays of images with specified proportion - prop:1-prop"""

### 2021-02-15 bin/python/resolution-scenario-sensitivity.py

@@ -96,8 +96,7 @@ def main(num_trials = NUM_MODEL_RUNS):
             for i in range(num_trials):
                 print("Conducting performance test: Scenario - " + s + "; Resolution - " + str(p) + "px; Trial - " + str(i+1))
                 scenario_performance_dict = testResolutionScenarioPerformance(p, s, num_epochs = NUM_EPOCHS, trial_seed = 1 + i) #ultimately should be averaged across trials       
-                scenario_filename = "scenario_resolution_performance_" + s + "_trial_" + str(i+1) + ".txt"
-                hist_filename = "scenario_resolution_hist_" + s + "_trial_" + str(i+1) + ".txt"
+                scenario_filename = "scenario_resolution_performance_" + s + str(p) + "px_trial_" + str(i+1) + ".txt"
                 with open(os.path.join(RESOLUTION_PERFORMANCE_METRICS_DIR, scenario_filename), 'w') as f:
                    f.write(json.dumps(scenario_performance_dict )) # use `json.loads` to do the reverse)
     return

### 2021-02-16 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
     \centering
-    \begin{tabular}{l l l}\toprule
+    \begin{tabular}{l l c}\toprule
     \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
@@
 For efficiency, we converted the images to grayscale and scaled the pixel values from 0 to 1.
-Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
+Finally, we downsampled the images to the following resolutions (pixels): $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
 Random sets of images from each class across each of the four classification scenarios are shown in \autoref{fig:processed_images}.
@@
     \centering
-    \begin{tabular}{|l | l l l l| l l l l| l l l l | l l l l|}\hline
-    %\bf Scenario 
-        &  \multicolumn{4}{c|}{\texttt{Pr\_Po\_Im}} 
-            & \multicolumn{4}{c|}{\texttt{Pr\_Im}} 
-                &  \multicolumn{4}{c|}{\texttt{PrPo\_Im}} 
-                    &  \multicolumn{4}{c|}{\texttt{Pr\_PoIm}}  \\\hline
-    %\bf Resolution
-        & 64 & 128 & 224 & 384 
-            & 64 & 128 & 224 & 384 
-                & 64 & 128 & 224 & 384
-                    & 64 & 128 & 224 & 384  \\\hline
-    %\bf Hyperparameter 
-    $ks$    &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
-    $u_1$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
-    $d_1$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
-    $a_1$   & ReLu  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
-    $u_2$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\ 
-    $d_2$   &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
-    $a_2$   & ReLu &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\
-    $\lambda$  
-            &  &  &  &  &   &  &  &  &   &  &  &  &   &  &  &   \\\hline
+    \begin{tabular}{l  c  p{1.75cm} p{1.75cm}  p{1.75cm} p{1.75cm}}\toprule
+    \bf Scenario & \bf Hyperparameter & \multicolumn{4}{c}{\bf Resolution ($n\times n$ pixels)} \\
+    &         & 64 & 128 & 224 & 384 \\\midrule
+    \multirow{8}{*}{\texttt{Pr\_Po\_Im}} 
+        & $ks$    & 7 &  &  &   \\
+        & $u_1$   & 128 &  &  &   \\
+        & $d_1$   & 0.25 &  &  &   \\
+        & $a_1$   & ReLu  &  &  &   \\
+        & $u_2$   & 320 &  &  &   \\
+        & $d_2$   & 0.30 &  &  &   \\
+        & $a_2$   & ReLu &  &  &   \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \midrule
+    \multirow{8}{*}{\texttt{Pr\_Im}} 
+        & $ks$    & 7 &  &  &   \\
+        & $u_1$   & 128 &  &  &   \\
+        & $d_1$   & 0.25 &  &  &   \\
+        & $a_1$   & ReLu  &  &  &   \\
+        & $u_2$   & 320 &  &  &   \\
+        & $d_2$   & 0.30 &  &  &   \\
+        & $a_2$   & ReLu &  &  &   \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \midrule
+    \multirow{8}{*}{\texttt{PrPo\_Im}} 
+        & $ks$    & 7 &  &  &   \\
+        & $u_1$   & 128 &  &  &   \\
+        & $d_1$   & 0.25 &  &  &   \\
+        & $a_1$   & ReLu  &  &  &   \\
+        & $u_2$   & 320 &  &  &   \\
+        & $d_2$   & 0.30 &  &  &   \\
+        & $a_2$   & ReLu &  &  &   \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \midrule
+    \multirow{8}{*}{\texttt{Pr\_PoIm}}  
+        & $ks$    & 7 &  &  &   \\
+        & $u_1$   & 128 &  &  &   \\
+        & $d_1$   & 0.25 &  &  &   \\
+        & $a_1$   & ReLu  &  &  &   \\
+        & $u_2$   & 320 &  &  &   \\
+        & $d_2$   & 0.30 &  &  &   \\
+        & $a_2$   & ReLu &  &  &   \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \bottomrule
     \end{tabular}

### 2021-02-16 docs/manuscript/manuscript.tex

@@
     \begin{tabular}{l  c  p{1.75cm} p{1.75cm}  p{1.75cm} p{1.75cm}}\toprule
-    \bf Scenario & \bf Hyperparameter & \multicolumn{4}{c}{\bf Resolution ($n\times n$ pixels)} \\
-    &         & 64 & 128 & 224 & 384 \\\midrule
-    \multirow{8}{*}{\texttt{Pr\_Po\_Im}} 
-        & $ks$    & 7 &  &  &   \\
-        & $u_1$   & 128 &  &  &   \\
-        & $d_1$   & 0.25 &  &  &   \\
-        & $a_1$   & ReLu  &  &  &   \\
-        & $u_2$   & 320 &  &  &   \\
-        & $d_2$   & 0.30 &  &  &   \\
-        & $a_2$   & ReLu &  &  &   \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \midrule
-    \multirow{8}{*}{\texttt{Pr\_Im}} 
-        & $ks$    & 7 &  &  &   \\
-        & $u_1$   & 128 &  &  &   \\
-        & $d_1$   & 0.25 &  &  &   \\
-        & $a_1$   & ReLu  &  &  &   \\
-        & $u_2$   & 320 &  &  &   \\
-        & $d_2$   & 0.30 &  &  &   \\
-        & $a_2$   & ReLu &  &  &   \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \midrule
-    \multirow{8}{*}{\texttt{PrPo\_Im}} 
-        & $ks$    & 7 &  &  &   \\
-        & $u_1$   & 128 &  &  &   \\
-        & $d_1$   & 0.25 &  &  &   \\
-        & $a_1$   & ReLu  &  &  &   \\
-        & $u_2$   & 320 &  &  &   \\
-        & $d_2$   & 0.30 &  &  &   \\
-        & $a_2$   & ReLu &  &  &   \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \midrule
-    \multirow{8}{*}{\texttt{Pr\_PoIm}}  
-        & $ks$    & 7 &  &  &   \\
-        & $u_1$   & 128 &  &  &   \\
-        & $d_1$   & 0.25 &  &  &   \\
-        & $a_1$   & ReLu  &  &  &   \\
-        & $u_2$   & 320 &  &  &   \\
-        & $d_2$   & 0.30 &  &  &   \\
-        & $a_2$   & ReLu &  &  &   \\
+    \bf Scenario & \bf Hyperparameter & \multicolumn{3}{c}{\bf Resolution ($n\times n$ pixels)} \\
+    &         & 64 & 128 & 224   \\\midrule
+    \multirow{7}{*}{\texttt{Pr\_Po\_Im}} 
+        & $ks$    & 7 &  &     \\
+        & $u_1$   & 128 &  &     \\
+        & $d_1$   & 0.25 &  &     \\
+        & $a_1$   & ReLu  &  &      \\
+        & $u_2$   & 320 &  &    \\
+        & $d_2$   & 0.30 &  &     \\
+        & $a_2$   & ReLu &  &     \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
+    \multirow{7}{*}{\texttt{Pr\_Im}} 
+        & $ks$    & 7 &  &      \\
+        & $u_1$   & 128 &  &     \\
+        & $d_1$   & 0.25 &  &      \\
+        & $a_1$   & ReLu  &  &      \\
+        & $u_2$   & 320 &  &      \\
+        & $d_2$   & 0.30 &  &     \\
+        & $a_2$   & ReLu &  &      \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
+    \multirow{7}{*}{\texttt{PrPo\_Im}} 
+        & $ks$    & 7 &  &    \\
+        & $u_1$   & 128 &  &     \\
+        & $d_1$   & 0.25 &  &      \\
+        & $a_1$   & ReLu  &  &     \\
+        & $u_2$   & 320 &  &     \\
+        & $d_2$   & 0.30 &  &     \\
+        & $a_2$   & ReLu &  &      \\
+        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &     \\ \midrule
+    \multirow{7}{*}{\texttt{Pr\_PoIm}}  
+        & $ks$    & 7 &  &      \\
+        & $u_1$   & 128 &  &      \\
+        & $d_1$   & 0.25 &  &      \\
+        & $a_1$   & ReLu  &  &      \\
+        & $u_2$   & 320 &  &      \\
+        & $d_2$   & 0.30 &  &     \\
+        & $a_2$   & ReLu &  &      \\
         & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \bottomrule
@@
 
+\begin{figure}
+    \centering
+    \includegraphics{}
+    \caption{Caption}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{table}[h!]\small
+    \centering
+    \begin{tabular}{ l c r r}\toprule
+\bf Scenario & \bf Resolution & \multicolumn{2}{c}{\bf Validation Accuracy}\\
+& & Mean & SD \\\midrule
+    \texttt{PrPo\_Im} & 64  & 0.96 & 0.04 \\
+                      & 128 & 0.96 & 0.05 \\
+    \texttt{Pr\_Im}   & 64  & 0.98 & 0.03 \\
+                      & 128 & 0.97 & 0.03 \\
+    \texttt{Pr\_PoIm} & 64  & 0.96 & 0.04 \\
+                      & 128 & 0.97 & 0.04 \\
+    \texttt{Pr\_Po\_Im}& 64  & 0.94 & 0.07 \\
+                      & 128 & 0.95 & 0.07 \\ \bottomrule
+    \end{tabular}
+    \caption{Summary validation accuracy (mean and standard deviation) for each scenario and input image resolution training instance.
+There is no statistically significant difference between the performance for each resolution case under each scenario.}
+    \label{tab:meanvalacc}
+\end{table}
 \subsection{Model visualization and inference}

### 2021-02-16 docs/manuscript/manuscript.tex

@@
 %\subsection{AI and risk assessment}
-AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances, wang2019novel}.
+AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances,wang2019novel}.
 Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citeN{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.

### 2021-02-16 bin/python/tabulate-optimal-hyperparameters.py
converted hyperparameter summary notebook to script

@@ -1,0 +1,41 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+import os
+import sys
+sys.path.append("../python/")
+from helpers import *
+
+import numpy as np
+import pandas as pd
+
+RESOLUTION_LIST = [64, 128]# 224] #, 384]
+SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
+
+def getOptHyperparamsSummary(scenario_list, resolution_list, orientation="landscape"):
+    df_all = pd.DataFrame()
+    for s in scenario_list:
+        for p in resolution_list:
+            p_dict = getOptCNNHyperparams(p, s)
+            #print(p_dict)
+            df = pd.DataFrame.from_dict(p_dict, orient='index', columns=["Values"])#, columns= ["Hyperparameters", "Values"])
+            df['Scenario'] = s
+            df['Resolution'] = p
+            df = df.drop(["tuner/bracket", "tuner/epochs", "tuner/initial_epoch", "tuner/round"])
+            df_all = df_all.append(df)
+    df_all['Hyperparameters'] = df_all.index
+    df_all = df_all.reset_index() #['Scenario','Resolution'])
+    pd.set_option('display.float_format', '{:.2E}'.format)
+    if orientation=="landscape":
+        df_all = df_all.pivot(index=['Hyperparameters'], columns=['Scenario','Resolution'], values='Values')
+    elif orientation=="portrait":
+        df_all = df_all.pivot(index=['Scenario','Hyperparameters'], columns=['Resolution'], values='Values')
+    file = '../../results/opt-hyperparameter-summary-table-' + orientation + '.csv'
+    df_all.to_csv(file, index=True, float_format = ':.2E')
+    return(df_all)
+
+
+if __name__ == "__main__":
+    getOptHyperparamsSummary(SCENARIO_LIST, RESOLUTION_LIST)
+    getOptHyperparamsSummary(SCENARIO_LIST, RESOLUTION_LIST, orientation='portrait')
+

### 2021-02-16 docs/manuscript/manuscript.tex

@@
 \usepackage{enumerate}
+\usepackage{siunitx}
 
@@
 \author[1]{Author 1}
-\author[2]{Author 2}
+\author[1]{Author 2}
 \author[2]{Author 3}
@@
 
-\affil[1]{First affiliation address, with corresponding author email. Email: author.one@umass.edu}
-\affil[2]{Second affiliation address}
+\affil[1]{Department of Civil and Environmental Engineering, University of Massachusetts Amherst, MA 01003, USA} %. Email: author.one@umass.edu}
+\affil[2]{Department of Ecological Conservation, University of Massachusetts Amherst, MA 01003, USA}
 
@@
 For efficiency, we converted the images to grayscale and scaled the pixel values from 0 to 1.
-Finally, we downsampled the images to the following resolutions (pixels): $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
+Finally, we downsampled the images to the following resolutions (pixels): $64 \times 64$, $128 \times 128$ and $224 \times 224$, creating a training set for each case.
 Random sets of images from each class across each of the four classification scenarios are shown in \autoref{fig:processed_images}.
@@
 \subsection{Hyperparameter optimization}
-We used the Hyperband approach \cite{li2018hyperband} to perform a grid search to find the optimal values of the following hyperparameters:
+We used the Hyperband approach \cite{li2018hyperband} to perform a guided grid search to find the optimal hyperparameter values across each scenario and input resolution instance.
+Thus, we performed 12 searches on the following hyperparameters in our CNN:
 \begin{itemize}
@@
 \end{itemize}
+Each search was conducted using 90 trials of unique hyperparameter combinations.
 
@@
     \centering
-    \begin{tabular}{l  c  p{1.75cm} p{1.75cm}  p{1.75cm} p{1.75cm}}\toprule
-    \bf Scenario & \bf Hyperparameter & \multicolumn{3}{c}{\bf Resolution ($n\times n$ pixels)} \\
-    &         & 64 & 128 & 224   \\\midrule
-    \multirow{7}{*}{\texttt{Pr\_Po\_Im}} 
-        & $ks$    & 7 &  &     \\
-        & $u_1$   & 128 &  &     \\
-        & $d_1$   & 0.25 &  &     \\
-        & $a_1$   & ReLu  &  &      \\
-        & $u_2$   & 320 &  &    \\
-        & $d_2$   & 0.30 &  &     \\
-        & $a_2$   & ReLu &  &     \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
-    \multirow{7}{*}{\texttt{Pr\_Im}} 
-        & $ks$    & 7 &  &      \\
-        & $u_1$   & 128 &  &     \\
-        & $d_1$   & 0.25 &  &      \\
-        & $a_1$   & ReLu  &  &      \\
-        & $u_2$   & 320 &  &      \\
-        & $d_2$   & 0.30 &  &     \\
-        & $a_2$   & ReLu &  &      \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
-    \multirow{7}{*}{\texttt{PrPo\_Im}} 
-        & $ks$    & 7 &  &    \\
-        & $u_1$   & 128 &  &     \\
-        & $d_1$   & 0.25 &  &      \\
-        & $a_1$   & ReLu  &  &     \\
-        & $u_2$   & 320 &  &     \\
-        & $d_2$   & 0.30 &  &     \\
-        & $a_2$   & ReLu &  &      \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &     \\ \midrule
-    \multirow{7}{*}{\texttt{Pr\_PoIm}}  
-        & $ks$    & 7 &  &      \\
-        & $u_1$   & 128 &  &      \\
-        & $d_1$   & 0.25 &  &      \\
-        & $a_1$   & ReLu  &  &      \\
-        & $u_2$   & 320 &  &      \\
-        & $d_2$   & 0.30 &  &     \\
-        & $a_2$   & ReLu &  &      \\
-        & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \bottomrule
+    \sisetup{round-mode=places,
+        round-precision=2,
+        scientific-notation=fixed,
+        exponent-product = \cdot,
+        table-omit-exponent,
+        fixed-exponent=-4
+        }
+ % Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+ \begin{tabular}{@{}llr r@{}}
+\toprule
+& & \multicolumn{2}{c}{\bf Resolution}\\
+\textbf{Scenario}   & \textbf{Hyperparameters} & 64                & 128               \\ \midrule
+PrPo\_Im   & activation\_1   & relu              & tanh              \\
+   & activation\_2   & relu              & relu              \\
+   & dropout\_1      & 0.1               & 0.25              \\
+   & dropout\_2      & 0.3               & 0.35              \\
+   & kernel\_size    & 7                 & 5                 \\
+   & learning\_rate  & \num{0.000102956216062} & \num{0.000109137476524} \\
+   & units\_1        & 480               & 384               \\
+   & units\_2        & 448               & 256               \\
+Pr\_Im     & activation\_1   & relu              & relu              \\
+      & activation\_2   & relu              & relu              \\
+      & dropout\_1      & 0.1               & 0.1               \\
+     & dropout\_2      & 0.3               & 0.3               \\
+      & kernel\_size    & 7                 & 7                 \\
+      & learning\_rate  & \num{0.000102956216062} & \num{0.000102956216062} \\
+      & units\_1        & 480               & 480               \\
+      & units\_2        & 448               & 448               \\
+Pr\_PoIm   & activation\_1   & relu              & relu              \\
+   & activation\_2   & tanh              & relu              \\
+   & dropout\_1      & 0.25              & 0.1               \\
+   & dropout\_2      & 0.3               & 0.3               \\
+   & kernel\_size    & 7                 & 7                 \\
+   & learning\_rate  & \num{0.000175822328104} & \num{0.000102956216062} \\
+   & units\_1        & 128               & 480               \\
+   & units\_2        & 320               & 448               \\
+Pr\_Po\_Im & activation\_1   & relu              & tanh              \\
+  & activation\_2   & tanh              & relu              \\
+  & dropout\_1      & 0.25              & 0.25              \\
+  & dropout\_2      & 0.3               & 0.35              \\
+  & kernel\_size    & 7                 & 5                 \\
+  & learning\_rate  & \num{0.000175822328104} & \num{0.000109137476524} \\
+  & units\_1        & 128               & 384               \\
+  & units\_2        & 320               & 256              \\\bottomrule
 \end{tabular}
+     % \begin{tabular}{l  c  p{1.75cm} p{1.75cm}  p{1.75cm} p{1.75cm}}\toprule
+    % \bf Scenario & \bf Hyperparameter & \multicolumn{3}{c}{\bf Resolution ($n\times n$ pixels)} \\
+    % &         & 64 & 128 & 224   \\\midrule
+    % \multirow{7}{*}{\texttt{Pr\_Po\_Im}} 
+    %     & $ks$    & 7 &  &     \\
+    %     & $u_1$   & 128 &  &     \\
+    %     & $d_1$   & 0.25 &  &     \\
+    %     & $a_1$   & ReLu  &  &      \\
+    %     & $u_2$   & 320 &  &    \\
+    %     & $d_2$   & 0.30 &  &     \\
+    %     & $a_2$   & ReLu &  &     \\
+    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
+    % \multirow{7}{*}{\texttt{Pr\_Im}} 
+    %     & $ks$    & 7 &  &      \\
+    %     & $u_1$   & 128 &  &     \\
+    %     & $d_1$   & 0.25 &  &      \\
+    %     & $a_1$   & ReLu  &  &      \\
+    %     & $u_2$   & 320 &  &      \\
+    %     & $d_2$   & 0.30 &  &     \\
+    %     & $a_2$   & ReLu &  &      \\
+    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
+    % \multirow{7}{*}{\texttt{PrPo\_Im}} 
+    %     & $ks$    & 7 &  &    \\
+    %     & $u_1$   & 128 &  &     \\
+    %     & $d_1$   & 0.25 &  &      \\
+    %     & $a_1$   & ReLu  &  &     \\
+    %     & $u_2$   & 320 &  &     \\
+    %     & $d_2$   & 0.30 &  &     \\
+    %     & $a_2$   & ReLu &  &      \\
+    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &     \\ \midrule
+    % \multirow{7}{*}{\texttt{Pr\_PoIm}}  
+    %     & $ks$    & 7 &  &      \\
+    %     & $u_1$   & 128 &  &      \\
+    %     & $d_1$   & 0.25 &  &      \\
+    %     & $a_1$   & ReLu  &  &      \\
+    %     & $u_2$   & 320 &  &      \\
+    %     & $d_2$   & 0.30 &  &     \\
+    %     & $a_2$   & ReLu &  &      \\
+    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \bottomrule
+    % \end{tabular}
     \caption{Optimal hyperparameters found using the Hyperband search algorithm for all classification scenario and input image resolution combinations.}
@@
 
-\begin{figure}
+\begin{figure}[ht]
     \centering
-    \includegraphics{}
-    \caption{Caption}
-    \label{fig:my_label}
+    \includegraphics[width=\textwidth]{figures/scenario_resolution_performance.png}
+    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and confidence intervals (trials, $n=10$) are shown for 10 epochs in each case.}
+    \label{fig:sens-overall}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/scenario_resolution_validation_accuracy.png}
+    \caption{Mean validation accuracy values with error bars for each scenario and resolution combination.}
+    \label{fig:sens-val-acc}
 \end{figure}

### 2021-02-16 bin/python/helpers.py
hyperparams 224

@@ -77,6 +77,36 @@ def getOptCNNHyperparams(image_size, scenario):
     opt_params_dict = json.loads(data)   
     return(opt_params_dict)
 
+def constructBaseCNN(image_size, scenario, num_channels = 1):
+    image_shape = (image_size, image_size, num_channels)
+    p_dict = getOptCNNHyperparams(image_size, scenario)
+    if scenario=="Pr_Po_Im":
+        num_classes = 3
+    else:
+        num_classes = 2
+    base_model = models.Sequential([
+        layers.Conv2D(filters = 64, kernel_size = p_dict['kernel_size'], strides = 2, activation="relu", padding="same", input_shape = image_shape),
+        layers.MaxPooling2D(2),
+        layers.Conv2D(128, 3, activation="relu", padding="same"),
+        layers.Conv2D(128, 3, activation="relu", padding="same"),
+        layers.MaxPooling2D(2),
+        layers.Conv2D(256, 3, activation="relu", padding="same"),
+        layers.Conv2D(256, 3, activation="relu", padding="same"),
+        layers.MaxPooling2D(2),
+        layers.Flatten(),
+        
+        layers.Dense(p_dict['units_1'], activation = p_dict['activation_1']),
+        layers.BatchNormalization(),
+        layers.Dropout(p_dict['dropout_1']), 
+        
+        layers.Dense(p_dict['units_2'], activation = p_dict['activation_2']), 
+        layers.Dropout(p_dict['dropout_2']),
+
+        layers.Dense(num_classes, activation="softmax")   
+        ])
+    return(base_model)
+
+
 #https://github.com/keras-team/keras/issues/5400#issuecomment-408743570
 def check_units(y_true, y_pred):
 	if y_pred.shape[1] != 1:

### 2021-02-16 bin/python/resolution-scenario-sensitivity.py

@@ -32,34 +32,34 @@ NUM_EPOCHS = 10
 RESOLUTION_PERFORMANCE_METRICS_DIR = '../../results/sensitivity-tests'
 
 
-def constructBaseCNN(image_size, scenario):
-    image_shape = (image_size, image_size, NUM_CHANNELS)
-    p_dict = getOptCNNHyperparams(image_size, scenario)
-    if scenario=="Pr_Po_Im":
-        num_classes = 3
-    else:
-        num_classes = 2
-    base_model = models.Sequential([
-        layers.Conv2D(filters = 64, kernel_size = p_dict['kernel_size'], strides = 2, activation="relu", padding="same", input_shape = image_shape),
-        layers.MaxPooling2D(2),
-        layers.Conv2D(128, 3, activation="relu", padding="same"),
-        layers.Conv2D(128, 3, activation="relu", padding="same"),
-        layers.MaxPooling2D(2),
-        layers.Conv2D(256, 3, activation="relu", padding="same"),
-        layers.Conv2D(256, 3, activation="relu", padding="same"),
-        layers.MaxPooling2D(2),
-        layers.Flatten(),
+# def constructBaseCNN(image_size, scenario):
+#     image_shape = (image_size, image_size, NUM_CHANNELS)
+#     p_dict = getOptCNNHyperparams(image_size, scenario)
+#     if scenario=="Pr_Po_Im":
+#         num_classes = 3
+#     else:
+#         num_classes = 2
+#     base_model = models.Sequential([
+#         layers.Conv2D(filters = 64, kernel_size = p_dict['kernel_size'], strides = 2, activation="relu", padding="same", input_shape = image_shape),
+#         layers.MaxPooling2D(2),
+#         layers.Conv2D(128, 3, activation="relu", padding="same"),
+#         layers.Conv2D(128, 3, activation="relu", padding="same"),
+#         layers.MaxPooling2D(2),
+#         layers.Conv2D(256, 3, activation="relu", padding="same"),
+#         layers.Conv2D(256, 3, activation="relu", padding="same"),
+#         layers.MaxPooling2D(2),
+#         layers.Flatten(),
         
-        layers.Dense(p_dict['units_1'], activation = p_dict['activation_1']),
-        layers.BatchNormalization(),
-        layers.Dropout(p_dict['dropout_1']), 
+#         layers.Dense(p_dict['units_1'], activation = p_dict['activation_1']),
+#         layers.BatchNormalization(),
[…]

### 2021-02-17 bin/python/cnn-hyperparameter-optimization.py
final hyperparam results added

@@ -3,7 +3,7 @@
 
 from PIL import Image # used for loading images
 import numpy as np
-import os # used for navigating to image path
+import os 
 import imageio # used for writing images
 import random
 import tensorflow as tf
@@ -29,7 +29,7 @@ from sklearn import metrics
 from sklearn import model_selection
 
 NUM_CHANNELS = 1
-RESOLUTION_LIST = [64, 128, 224, 384]
+RESOLUTION_LIST = [64, 128, 224] #, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 OPTIMAL_HYPERPARAMETERS_PATH = '../../results/optimal-hyperparameters/'
 HYPERBAND_MAX_EPOCHS = 10 #10
@@ -131,9 +131,9 @@ def optimizeCNNHyperparameters(scenario, image_size, seed_val = 1, save_results=
 def main():
     for i in RESOLUTION_LIST:
         for s in SCENARIO_LIST:     
-            print("Beginning search for scenario: " + s)
+            print("Beginning search for scenario: " + s + ", resolution: " + i)
             optimizeCNNHyperparameters(s, i, seed_val = 1, save_results = True)
-            print("Search for scenario: " + s + " is complete.")
+            print("Search for scenario: " + s + ", resolution: " + i + " is complete.")
     return
 
 if __name__ == "__main__":

### 2021-02-17 bin/python/tabulate-optimal-hyperparameters.py

@@ -9,7 +9,7 @@ from helpers import *
 import numpy as np
 import pandas as pd
 
-RESOLUTION_LIST = [64, 128]# 224] #, 384]
+RESOLUTION_LIST = [64, 128, 224] #, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 
 def getOptHyperparamsSummary(scenario_list, resolution_list, orientation="landscape"):

### 2021-02-17 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
 
-We can apply cut-out (occluding portions of the image) for improved performance \cite{devries2017improved}.
+% We can apply cut-out (occluding portions of the image) for improved performance \cite{devries2017improved}.
 Also, it has been shown that training with lower resolution improves performance on higher resolution test images \cite{touvron2019fixing}.
@@
 % \usepackage{booktabs}
- \begin{tabular}{@{}llr r@{}}
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% \usepackage{multirow}
+\begin{tabular}{@{}lllll@{}}
 \toprule
-& & \multicolumn{2}{c}{\bf Resolution}\\
-\textbf{Scenario}   & \textbf{Hyperparameters} & 64                & 128               \\ \midrule
-PrPo\_Im   & activation\_1   & relu              & tanh              \\
-   & activation\_2   & relu              & relu              \\
-   & dropout\_1      & 0.1               & 0.25              \\
-   & dropout\_2      & 0.3               & 0.35              \\
-   & kernel\_size    & 7                 & 5                 \\
-   & learning\_rate  & \num{0.000102956216062} & \num{0.000109137476524} \\
-   & units\_1        & 480               & 384               \\
-   & units\_2        & 448               & 256               \\
-Pr\_Im     & activation\_1   & relu              & relu              \\
-      & activation\_2   & relu              & relu              \\
-      & dropout\_1      & 0.1               & 0.1               \\
-     & dropout\_2      & 0.3               & 0.3               \\
-      & kernel\_size    & 7                 & 7                 \\
-      & learning\_rate  & \num{0.000102956216062} & \num{0.000102956216062} \\
-      & units\_1        & 480               & 480               \\
-      & units\_2        & 448               & 448               \\
-Pr\_PoIm   & activation\_1   & relu              & relu              \\
-   & activation\_2   & tanh              & relu              \\
-   & dropout\_1      & 0.25              & 0.1               \\
-   & dropout\_2      & 0.3               & 0.3               \\
-   & kernel\_size    & 7                 & 7                 \\
-   & learning\_rate  & \num{0.000175822328104} & \num{0.000102956216062} \\
-   & units\_1        & 128               & 480               \\
-   & units\_2        & 320               & 448               \\
-Pr\_Po\_Im & activation\_1   & relu              & tanh              \\
-  & activation\_2   & tanh              & relu              \\
-  & dropout\_1      & 0.25              & 0.25              \\
-  & dropout\_2      & 0.3               & 0.35              \\
-  & kernel\_size    & 7                 & 5                 \\
-  & learning\_rate  & \num{0.000175822328104} & \num{0.000109137476524} \\
-  & units\_1        & 128               & 384               \\
-  & units\_2        & 320               & 256              \\\bottomrule
+\multirow{2}{*}{\textbf{Scenario}} & \multirow{2}{*}{\textbf{Hyperparameters}} & \multicolumn{3}{c}{\textbf{Resolution}} \\ \cmidrule(l){3-5} 
+                            &                & 64                & 128               & 224               \\ \midrule
+\multirow{8}{*}{PrPo\_Im}   & activation\_1  & relu              & tanh              & relu              \\
+                            & activation\_2  & relu              & relu              & relu              \\
+                            & dropout\_1     & 0.1               & 0.25              & 0.1               \\
+                            & dropout\_2     & 0.3               & 0.35              & 0.3               \\
+                            & kernel\_size   & 7                 & 5                 & 7                 \\
+                            & learning\_rate & \num{0.000102956216062} & \num{0.000109137476524} & \num{0.000102956216062} \\
+                            & units\_1       & 480               & 384               & 480               \\
+                            & units\_2       & 448               & 256               & 448               \\\midrule
+\multirow{8}{*}{Pr\_Im}     & activation\_1  & relu              & relu              & relu              \\
+                            & activation\_2  & relu              & relu              & relu              \\
+                            & dropout\_1     & 0.1               & 0.1               & 0.1               \\
+                            & dropout\_2     & 0.3               & 0.3               & 0.3               \\
+                            & kernel\_size   & 7                 & 7                 & 7                 \\
+                            & learning\_rate & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\
+                            & units\_1       & 480               & 480               & 480               \\
+                            & units\_2       & 448               & 448               & 448               \\\midrule
+\multirow{8}{*}{Pr\_PoIm}   & activation\_1  & relu              & relu              & relu              \\
+                            & activation\_2  & tanh              & relu              & tanh              \\
+                            & dropout\_1     & 0.25              & 0.1               & 0.2               \\
+                            & dropout\_2     & 0.3               & 0.3               & 0.15              \\
+                            & kernel\_size   & 7                 & 7                 & 7                 \\
+                            & learning\_rate & \num{0.000175822328104} & \num{0.000102956216062} & \num{0.000114656409879} \\
+                            & units\_1       & 128               & 480               & 416               \\
+                            & units\_2       & 320               & 448               & 416               \\
+\multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & relu              & tanh              & relu              \\\midrule
+                            & activation\_2  & tanh              & relu              & relu              \\
+                            & dropout\_1     & 0.25              & 0.25              & 0.1               \\
+                            & dropout\_2     & 0.3               & 0.35              & 0.3               \\
+                            & kernel\_size   & 7                 & 5                 & 7                 \\
+                            & learning\_rate & \num{0.000175822328104} & \num{0.000109137476524} & \num{0.000102956216062} \\
+                            & units\_1       & 128               & 384               & 480               \\
+                            & units\_2       & 320               & 256               & 448               \\ \bottomrule
 \end{tabular}

### 2021-02-17 bin/python/resolution-scenario-sensitivity.py
added CNN diagram in figures/net2vis/

@@ -25,7 +25,7 @@ from helpers import *
 
 # Globals
 NUM_CHANNELS = 1
-RESOLUTION_LIST = [64, 128] #, 224, 384]
+RESOLUTION_LIST = [224] # 64, 128] #, 224, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 NUM_MODEL_RUNS = 10
 NUM_EPOCHS = 10
@@ -105,3 +105,8 @@ def main(num_trials = NUM_MODEL_RUNS):
 if __name__ == "__main__":
     main()
 
+
+# predictions = model.predict_classes(x_val)
+# predictions = predictions.reshape(1,-1)[0]
+
+# print(classification_report(y_val, predictions, target_names = ['Rugby (Class 0)','Soccer (Class 1)']))

### 2021-02-18 docs/manuscript/manuscript.tex
formatted hyperparameter table
Updates from Overleaf

@@
 
+The structure of the CNN is shown in \autoref{fig:cnn}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{net2vis/graph.pdf}
+
+    \medskip
+
+    \includegraphics[width=.7\textwidth]{figures/net2vis-name-labels/legend}
+    \caption{Diagram of convolutional neural network structure.
+Hyperparameters that are optimized include the number of units in the penultimate dense layers.
+(Generated using automated framework \cite{bauerle2021net2vis})}
+    \label{fig:cnn}
+\end{figure}
+
 \subsection{Hyperparameter optimization}
@@
 \end{tabular}
-     % \begin{tabular}{l  c  p{1.75cm} p{1.75cm}  p{1.75cm} p{1.75cm}}\toprule
-    % \bf Scenario & \bf Hyperparameter & \multicolumn{3}{c}{\bf Resolution ($n\times n$ pixels)} \\
-    % &         & 64 & 128 & 224   \\\midrule
-    % \multirow{7}{*}{\texttt{Pr\_Po\_Im}} 
-    %     & $ks$    & 7 &  &     \\
-    %     & $u_1$   & 128 &  &     \\
-    %     & $d_1$   & 0.25 &  &     \\
-    %     & $a_1$   & ReLu  &  &      \\
-    %     & $u_2$   & 320 &  &    \\
-    %     & $d_2$   & 0.30 &  &     \\
-    %     & $a_2$   & ReLu &  &     \\
-    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
-    % \multirow{7}{*}{\texttt{Pr\_Im}} 
-    %     & $ks$    & 7 &  &      \\
-    %     & $u_1$   & 128 &  &     \\
-    %     & $d_1$   & 0.25 &  &      \\
-    %     & $a_1$   & ReLu  &  &      \\
-    %     & $u_2$   & 320 &  &      \\
-    %     & $d_2$   & 0.30 &  &     \\
-    %     & $a_2$   & ReLu &  &      \\
-    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &      \\ \midrule
-    % \multirow{7}{*}{\texttt{PrPo\_Im}} 
-    %     & $ks$    & 7 &  &    \\
-    %     & $u_1$   & 128 &  &     \\
-    %     & $d_1$   & 0.25 &  &      \\
-    %     & $a_1$   & ReLu  &  &     \\
-    %     & $u_2$   & 320 &  &     \\
-    %     & $d_2$   & 0.30 &  &     \\
-    %     & $a_2$   & ReLu &  &      \\
-    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &     \\ \midrule
-    % \multirow{7}{*}{\texttt{Pr\_PoIm}}  
-    %     & $ks$    & 7 &  &      \\
-    %     & $u_1$   & 128 &  &      \\
-    %     & $d_1$   & 0.25 &  &      \\
-    %     & $a_1$   & ReLu  &  &      \\
-    %     & $u_2$   & 320 &  &      \\
-    %     & $d_2$   & 0.30 &  &     \\
-    %     & $a_2$   & ReLu &  &      \\
-    %     & $\lambda$  & $1.76\cdot 10^{-4}$ &  &  &   \\ \bottomrule
-    % \end{tabular}
+% Table edited at: https://www.tablesgenerator.com/latex_tables#
 \caption{Optimal hyperparameters found using the Hyperband search algorithm for all classification scenario and input image resolution combinations.}

### 2021-02-20 bin/python/helpers.py
implemented multiclass performance metrics (precision, recall, f1 score), along with confusion matrix and heatmap

@@ -12,7 +12,7 @@ from tensorflow.keras import backend as K
 from tensorflow.keras.callbacks import Callback
 
         
-from sklearn.metrics import recall_score, classification_report
+from sklearn.metrics import recall_score, classification_report, accuracy_score, precision_score, recall_score, confusion_matrix
 from sklearn.datasets import make_classification
 
 import json
@@ -62,6 +62,16 @@ def getImageAndLabelArrays(image_label_tuple_array, num_channels = 1):
 	label_array = np.array([x[1] for x in image_label_tuple_array])
 	return(image_array, label_array)
 
+def getClassLabels(scenario):
+    if scenario=="Pr_Po_Im":
+        labels = ["Probable", "Possible", "Improbable"]
+    elif scenario=="Pr_Im":
+        labels = ["Probable", "Improbable"]
+    elif scenario=="PrPo_Im":
+        labels = ["Probable/Possible", "Improbable"]
+    elif scenario=="Pr_PoIm":
+        labels = ["Probable", "Possible/Improbable"]
+    return(labels)	
 
 def createResolutionScenarioImageDict(resolution_list, scenario_list):
     image_dict = dict.fromkeys(resolution_list)

### 2021-02-20 docs/manuscript/manuscript.tex

@@
 
-\begin{table}[h!]\small
+\section{Results}
+
+\subsection{Optimal hyperparameters}
+The results of the 8 hyperparameters optimized in each of the 12 training instances is shown in \autoref{tab:opt-hyp}.
+\begin{table}[ht!]\small
     \centering
@@
         }
- % Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-% \usepackage{multirow}
 \begin{tabular}{@{}lllll@{}}
@@
 \end{table}
-\section{Results}
-
 \subsection{Sensitivity to training resolution}
+We trained the CNN for each of the 4 classification scenarios.
+In each scenarios, we also trained on three image resolution sets (64, 108 and 224).
+The goal was to determine the best performing resolution, given the tradeoff between performance and computational expenditure.
+Thus, twelve learning instances were generated, using a training-validation ratio of 80:20.
+In each instance, we performed 10 trials, each consisting of 10 epochs.
+Validation sets were randomly selected in each trial, but consistent across each of the 12 training instances.
+The trajectories of the training/validation loss and accuracy are shown in \autoref{fig:sens-overall}.
+Although the 64-pixel case is the slowest to converge in all scenarios, it appears to have a similar performance with respect to the other two resolutions (128px and 224px).
 
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/scenario_resolution_performance.png}
+    \includegraphics[width=1.05\textwidth]{scenario_resolution_performance.png}
     \caption{Performance metrics for each scenario and input resolution instance. Average metrics and confidence intervals (trials, $n=10$) are shown for 10 epochs in each case.}
@@
 
+\autoref{fig:sens-val-acc} shows the boxplots of the validation accuracy in all twelve instances ($n=10$).
+We further summarize the statistics (mean and standard deviation) of the validation accuracies in \autoref{tab:meanvalacc}.
+Using the Welch test for pairwise comparisons of resolution performance within each scenario, we find no statistically significant difference between the validation accuracy for each resolution ($p\ge 0.57$).
 \begin{figure}[ht]
     \centering
-    \includegraphics[width=\textwidth]{figures/scenario_resolution_validation_accuracy.png}
-    \caption{Mean validation accuracy values with error bars for each scenario and resolution combination.}
+    \includegraphics[width=\textwidth]{figures/scenario_resolution_validation_accuracy_boxplot.png}
+    \caption{Boxplots of validation accuracy values with error bars for each scenario and resolution combination. Mean values are depicted as white squares in each box.}
     \label{fig:sens-val-acc}
 \end{figure}
-
 \begin{table}[h!]\small
@@
                       & 128 & 0.96 & 0.05 \\
+                      &224  &      0.96&  0.05  \\ \midrule        
     \texttt{Pr\_Im}   & 64  & 0.98 & 0.03 \\
                       & 128 & 0.97 & 0.03 \\
+                      & 224 & 0.97 & 0.03\\\midrule       
     \texttt{Pr\_PoIm} & 64  & 0.96 & 0.04 \\
                       & 128 & 0.97 & 0.04 \\
+                      & 224 & 0.97 & 0.04 \\\midrule       
     \texttt{Pr\_Po\_Im}& 64  & 0.94 & 0.07 \\
-                      & 128 & 0.95 & 0.07 \\ \bottomrule
+                      & 128 & 0.95 & 0.07 \\
+                      & 224 & 0.95 & 0.06 \\\bottomrule
     \end{tabular}
-    \caption{Summary validation accuracy (mean and standard deviation) for each scenario and input image resolution training instance.
+    \caption{Summary validation accuracy (mean and standard deviation to 2 decimal places) for each scenario and input image resolution training instance.
 There is no statistically significant difference between the performance for each resolution case under each scenario.}
@@
 \end{table}
+
+Given that there was no significant difference in performance among the three resolutions tested, we focus our attention on the 64-px case, in the interest of efficiency.
+We re-trained the model on all the four scenarios and then evaluated performance on the following validation metrics: average precision, average recall and accuracy.
+
 \subsection{Model visualization and inference}
@@
 \section{Acknowledgments}
+The authors would like to thank...
 

### 2021-02-22 bin/python/cnn-performance.py
executable script for cnn-performance

@@ -1,0 +1,344 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+# In[2]:
+
+
+import numpy as np
+np.seterr(divide='ignore', invalid='ignore')
+import pandas as pd
+
+from timeit import default_timer as timer
+import random
+
+import tensorflow as tf
+from tensorflow.keras import datasets, layers, models
+from tensorflow.keras.optimizers import SGD
+from tensorflow.keras import backend as K
+from tensorflow.keras.callbacks import Callback
+from tensorflow.keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import seaborn as sns
+import matplotlib.ticker as tkr
+
+from matplotlib import cm
+from tf_keras_vis.gradcam import Gradcam
+from tf_keras_vis.saliency import Saliency
+from tf_keras_vis.utils import normalize
+
+import json
+import pickle
+import os
+import sys
+sys.path.append("../python/")
+from helpers import *
+from sklearn.metrics import classification_report, accuracy_score, f1_score, precision_score, recall_score, confusion_matrix
+from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
+#from sklearn.preprocessing import OrdinalEncoder
+#enc = OrdinalEncoder()
+
+#from contextlib import redirect_stdout
+
+# Globals
+NUM_CHANNELS = 1
+RESOLUTION_LIST = [64] # 64, 128] #, 224, 384]
+SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
+NUM_EPOCHS = 20
+SAVED_MODEL_DIR = '../../results/models/'
+MODEL_PERFORMANCE_METRICS_DIR = '../../results/model-performance/'
+
+
+# In[3]:
+
+
+image_sets = createResolutionScenarioImageDict(RESOLUTION_LIST, SCENARIO_LIST)
+
+
+# In[4]:
+
+
+
+# class ReturnBestEarlyStopping(EarlyStopping):
+#     def __init__(self, **kwargs):
+#         super(ReturnBestEarlyStopping, self).__init__(**kwargs)
+
+#     def on_train_end(self, logs=None):
+#         if self.stopped_epoch > 0:
+#             if self.verbose > 0:
+#                 print(f'\nEpoch {self.stopped_epoch + 1}: early stopping')
+#         elif self.restore_best_weights:
+#             if self.verbose > 0:
[…]

### 2021-02-22 bin/python/helpers.py

@@ -167,24 +167,6 @@ class Histories(Callback):
 		self.losses.append(logs.get('loss'))
 		self.accuracies.append(logs.get('accuracy'))
 
-# Callback to find metrics at epoch end (not perfectly implemented; TODO) # Source: https://stackoverflow.com/a/56485026/3023033
-class Metrics(Callback):
-	def __init__(self, x, y):
-		self.x = x
-		self.y = y if (y.ndim == 1 or y.shape[1] == 1) else np.argmax(y, axis=1)
-		self.reports = []
-
-	def on_epoch_end(self, epoch, logs={}):
-		y_hat = np.asarray(self.model.predict(self.x))
-		y_hat = np.where(y_hat > 0.5, 1, 0) if (y_hat.ndim == 1 or y_hat.shape[1] == 1)  else np.argmax(y_hat, axis=1)
-		report = classification_report(self.y,y_hat,output_dict=True)
-		self.reports.append(report)
-		return
-   
-	# Utility method
-	def get(self, metrics, of_class):
-		return [report[str(of_class)][metrics] for report in self.reports]
-
 
 def plot_model_accuracy(hist):
 	plt.plot(hist.history["accuracy"])

### 2021-02-22 bin/python/preprocess-images.py

@@ -17,11 +17,11 @@ import matplotlib.pyplot as plt
 
 ## GLOBAL VARIABLES
 IMG_SIZE = 108
-IMG_SIZE_LIST = [64, 128, 224, 384]
+IMG_SIZE_LIST = [64]#, 128, 224]#, 384]
 NUM_CHANNELS = 1
 CLASSIFICATION_SCENARIO = "Pr_Im"
-CLASSIFICATION_SCENARIO_LIST = ["Pr_Po_Im", "Pr_Im", "PrPo_Im", "Pr_PoIm"]  
-EXPANSION_FACTOR = 5 # of augmented images
+CLASSIFICATION_SCENARIO_LIST = ["Pr_Po_Im"]#, "Pr_Im", "PrPo_Im", "Pr_PoIm"]  
+EXPANSION_FACTOR = 5 #5 of augmented images
 LABELED_IMAGES_DIR = '../../data/tidy/labeled_images'
 PROCESSED_IMAGES_DIR = '../../data/tidy/preprocessed_images'
 SEED = 10 # seed for repeatability
@@ -70,6 +70,7 @@ def processImageData(img_size, expansion_factor, class_scenario, seed_value, cha
     data = []
     image_list = os.listdir(LABELED_IMAGES_DIR)
     random.seed(seed_value) #seed for repeatability
+    print("Preprocessing images for scenario " + class_scenario + "and resolution " + str(img_size))
     for img in image_list:
         label = getImageOneHotVector(img, class_scenario)
         if label.sum() == 0: # if image unlabeled, move to next one
@@ -99,7 +100,7 @@ def processImageData(img_size, expansion_factor, class_scenario, seed_value, cha
             #img = img.crop((left=400, top=0, r=3424, b=3024)) 
             #img = img.resize((img_size, img_size), Image.BICUBIC)
             #TODO: random cropping, flipping and resizing
-
+    print("Images:", class_scenario, (np.array([x[1] for x in data])).sum(axis=0) )
     random_image_selection_class_0 = random.choices([i[0] for i in data if i[1][0] == 1], k = images_per_class)
     random_image_selection_class_1 = random.choices([i[0] for i in data if i[1][1] == 1], k = images_per_class)
     image_selection_array = [random_image_selection_class_0, random_image_selection_class_1]
@@ -113,10 +114,11 @@ def processImageData(img_size, expansion_factor, class_scenario, seed_value, cha
[…]

### 2021-02-22 bin/python/cnn-performance.py
cnn performance metrics

@@ -1,8 +1,6 @@
 #!/usr/bin/env python
 # coding: utf-8
 
-# In[2]:
-
 
 import numpy as np
 np.seterr(divide='ignore', invalid='ignore')
@@ -41,35 +39,16 @@ from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
 
 # Globals
 NUM_CHANNELS = 1
-RESOLUTION_LIST = [64] # 64, 128] #, 224, 384]
+RESOLUTION_LIST = [64, 128] # 64, 128] #, 224, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 NUM_EPOCHS = 20
 SAVED_MODEL_DIR = '../../results/models/'
 MODEL_PERFORMANCE_METRICS_DIR = '../../results/model-performance/'
 
 
-# In[3]:
-
-
 image_sets = createResolutionScenarioImageDict(RESOLUTION_LIST, SCENARIO_LIST)
 
 
-# In[4]:
-
-
-
-# class ReturnBestEarlyStopping(EarlyStopping):
-#     def __init__(self, **kwargs):
-#         super(ReturnBestEarlyStopping, self).__init__(**kwargs)
-
-#     def on_train_end(self, logs=None):
-#         if self.stopped_epoch > 0:
-#             if self.verbose > 0:
-#                 print(f'\nEpoch {self.stopped_epoch + 1}: early stopping')
-#         elif self.restore_best_weights:
-#             if self.verbose > 0:
-#                 print('Restoring model weights from the end of the best epoch.')
-#             self.model.set_weights(self.best_weights
                                    
 # https://github.com/keisen/tf-keras-vis/blob/master/examples/attentions.ipynb
 # Metrics2 modified from https://stackoverflow.com/a/61856587/3023033
@@ -101,39 +80,25 @@ class Metrics(Callback):
         return
 
 
-# In[55]:
-
-
-## TEST
-# imd = np.load('../../data/tidy/preprocessed_images/size64_exp5_Pr_Po_Im.npy',allow_pickle = True)
-# trimg,vaimg,trlab,valab =  train_test_split(np.array([np.expand_dims(x[0],axis=2) for x in imd]), 
-#                  np.array([x[1] for x in imd]), stratify= np.array([x[1] for x in imd]), test_size=.2, random_state = 1  )
-# print(trlab.sum(axis=0))
[…]

### 2021-02-22 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
 
+\subsection{Analysis of classification strategies}
+We investigated in the detail the performance of the four classification scenarios earlier outlined, using the 64px resolution images.
+The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
+In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the weighted averages of the precision, recall and $F_1$ score metrics.
+
+\begin{align}
+\text{Weighted average precision: }    Pr^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c TP_{c} }{TP_{c} + FP_c}\right) \\
+\text{Weighted average recall: }    Re^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c TP_{c} }{TP_{c} + FN_c}\right) \\
+\text{Weighted average $F_1$ score: }    F_1^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{2 n_c Pr_{c} Re_{c}}{Pr_c + Re_c}\right)
+\end{align}
+where the class-based metrics are given by:
+\begin{align}
+    TP_c &= \sum_{i \in c} I(\hat{y}_i = y_i ) \\
+    FP_c &= \sum_{i \in c} I(\hat{y}_i \in c | y_i \notin c) \\
+    FN_c &= \sum_{i \in c} I(\hat{y}_i \notin c | y_i \in c)
+\end{align}
 \subsection{Model visualization and inference}
@@
 \section{Data Availability Statement}
-All data, models, or code generated or used during the study are available in a repository online in accordance with funder data retention policies (provide full citations that include URLS or DOIs)
+All data, models, or code generated or used during the study are available in a GitHub repository online at \url{https://github.com/narslab/tree-risk-ai}.
 

### 2021-02-22 docs/manuscript/manuscript.tex
results (new figures and text)

@@
     \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_64_px}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \caption{Scenario \texttt{Pr\_Im}}
     \label{pr_im_64}
@@
     \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_64_px}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \caption{Scenario \texttt{PrPo\_Im}}
     \label{prpo_im_64}
@@
     \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_64_px}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \caption{Scenario \texttt{Pr\_PoIm}}
     \label{pr_poim_64}
@@
 
-Given that there was no significant difference in performance among the three resolutions tested, we focus our attention on the 64-px case, in the interest of efficiency.
+Given that there was no significant difference in performance among the three resolutions tested, we focus our attention on the 128-px case, as a tradeoff between efficiency and performance.
 We re-trained the model on all the four scenarios and then evaluated performance on the following validation metrics: average precision, average recall and accuracy.
@@
 In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the weighted averages of the precision, recall and $F_1$ score metrics.
+The precision score $\mathrm{Pr}$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is a important measure of how good a classifier is.
+Meanwhile, the recall $\mathrm{Re}$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
+The $F_1$ metric is given as the harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
+These three metrics can also be weighted by the number of observations in each class in order to better indicate class-specific performance.
+We formally define these metrics in the following equations:
 
 \begin{align}
-\text{Weighted average precision: }    Pr^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c TP_{c} }{TP_{c} + FP_c}\right) \\
-\text{Weighted average recall: }    Re^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c TP_{c} }{TP_{c} + FN_c}\right) \\
-\text{Weighted average $F_1$ score: }    F_1^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{2 n_c Pr_{c} Re_{c}}{Pr_c + Re_c}\right)
+\text{Weighted average precision: }    \mathrm{Pr}^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FP}_c}\right) \\
+\text{Weighted average recall: }    \mathrm{Re}^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FN}_c}\right) \\
+\text{Weighted average $F_1$ score: }    F_1^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{2 n_c \mathrm{Pr}_{c} \mathrm{Re}_{c}}{\mathrm{Pr}_{c} + \mathrm{Re}_c}\right)
 \end{align}
@@
 \begin{align}
-    TP_c &= \sum_{i \in c} I(\hat{y}_i = y_i ) \\
-    FP_c &= \sum_{i \in c} I(\hat{y}_i \in c | y_i \notin c) \\
-    FN_c &= \sum_{i \in c} I(\hat{y}_i \notin c | y_i \in c)
+    \mathrm{TP}_c &= \sum_{i \in c} I(\hat{y}_i = y_i ) \\
+    \mathrm{FP}_c &= \sum_{i \in c} I(\hat{y}_i \in c | y_i \notin c) \\
+    \mathrm{FN}_c &= \sum_{i \in c} I(\hat{y}_i \notin c | y_i \in c)
 \end{align}
+
+Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
+The metrics are shown in \autoref{fig:cnn-perf}.
+In each scenario, the model was trained over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
+That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
+From this figure, we see that Pr\_Po\_Im performed significant worse than the other three.
+This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
+Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=1.05\textwidth]{cnn-performance-metrics-128px-plot}
+    \caption{Model performance across four classification scenarios for a single training instance in each case.
+Training resolution was 128px.
+All weighted average metrics were computed on the validation set.}
+    \label{fig:cnn-perf}
+\end{figure}
+
+We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
+
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{pr_po_im_cm}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Im}}
+    \label{pr_im_cm}
+  \end{subfigure}%
+
+   \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
+    \caption{Scenario \texttt{PrPo\_Im}}
+    \label{prpo_im_cm}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
+    \caption{Scenario \texttt{Pr\_PoIm}}
+    \label{pr_poim_cm}
+  \end{subfigure}%
+    \caption{Confusion matrix for final model performance using the 128px images across all scenarios.}
+    \label{fig:confusion-matrix}
+\end{figure}
+
+\begin{table}[ht]
+    \centering
+    \begin{tabular}{c|c}
+         &  \\
+         & 
+    \end{tabular}
+    \caption{Performance metric summaries across the four scenarios}
+    \label{tab:my_label}
+\end{table}
+
 \subsection{Model visualization and inference}
+Model explainability is increasingly critical in the application of CNNs to an ever-expanding set of problems.
+
+\begin{figure}[ht]
+    \centering
+
+    \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-gradcam-Pr_Im-128-px-4-images.png}
+    \caption{Scenario \texttt{Pr\_Im}}
+    \label{prpo_im_cm}
+    \end{subfigure}%   
+    \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-gradcam-PrPo_Im-128-px-4-images.png}
+    \caption{Scenario \texttt{PrPo\_Im}}
+    \label{prpo_im_cm}
+    \end{subfigure}%
+
+    \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-gradcam-Pr_PoIm-128-px-4-images.png}
+    \caption{Scenario \texttt{Pr\_PoIm}}
+    \label{prpo_im_cm}
+    \end{subfigure}%
+    \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-gradcam-Pr_Po_Im-128-px-4-images.png}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{prpo_im_cm}
+    \end{subfigure}%
+
+    \caption{Gradient-weighted class activation maps (Grad-CAMs) for randomly selected validation images in each class. (Resolution: 128px)}
+    \label{fig:my_label}
+\end{figure}
 

### 2021-02-23 bin/python/cnn-performance.py
fixes to cnn-performance.py

@@ -98,7 +98,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='loss', patience=3, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_loss', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -148,20 +148,20 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     ## Confusion matrix
     con_mat = tf.math.confusion_matrix(labels=np.argmax(validation_labels, axis=-1), predictions=y_pred).numpy()
     con_mat_norm = np.around(con_mat.astype('float') / con_mat.sum(axis=1)[:, np.newaxis], decimals=2)
-    con_mat_df = pd.DataFrame(con_mat, index = class_labels, columns = class_labels)
+    con_mat_df = pd.DataFrame(con_mat_norm, index = class_labels, columns = class_labels)
     #print("Confusion matrix for scenario " + scenario + ", resolution: " + str(image_size) + ":")
     #print(con_mat_df)
     ## Confusion matrix heatmap
-    figure = plt.figure()#figsize=(4, 4))
-    ax = sns.heatmap(con_mat_df, annot=True, cmap=plt.cm.Blues, fmt='g', cbar = False)
-    plt.tight_layout()
-    plt.ylabel('True')
-    ax.set_yticklabels(class_labels, ha='center')
-    ax.set_xticklabels(class_labels, ha='center')
-    plt.xlabel('Predicted')
+    figure = plt.figure(figsize=(4, 4))
+    ax = sns.heatmap(con_mat_df, annot=True, cmap=plt.cm.Blues, fmt='g', cbar = False, annot_kws={"size": 16})
+    figure.tight_layout()
+    plt.ylabel('True',fontsize=16)
+    ax.set_yticklabels(class_labels,va='center',fontsize=14)
+    ax.set_xticklabels(class_labels, ha='center',fontsize=14)
+    plt.xlabel('Predicted',fontsize=16)
     #plt.show()
     con_mat_heatmap_file = "../../figures/opt-confusion-matrix-" + scenario + "-" + str(image_size) + "-px.png"
[…]

### 2021-02-23 docs/manuscript/manuscript.tex

@@
 
-\begin{table}[ht]
+\begin{table}[ht]\small
     \centering
-    \begin{tabular}{c|c}
-         &  \\
-         & 
+\begin{tabular}{@{}llllll@{}}
+\toprule
+\textbf{Scenario} & \textbf{Averaging method} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-score} & \textbf{Support} \\ \midrule
+Pr\_Im    & Probable            & 0.9  & 0.75 & 0.81 & 103 \\
+         & Improbable          & 0.92 & 0.97 & 0.95 & 322 \\
+         & accuracy            &      &      & 0.92 & 425 \\
+         & weighted avg        & 0.92 & 0.92 & 0.92 & 425 \\ \midrule
+Pr\_Po\_Im & Probable            & 0.82 & 0.56 & 0.67 & 103 \\
+         & Possible            & 0.43 & 0.74 & 0.54 & 80  \\
+         & Improbable          & 0.93 & 0.86 & 0.89 & 322 \\
+         & accuracy            &      &      & 0.78 & 505 \\
+         & weighted avg        & 0.83 & 0.78 & 0.79 & 505 \\ \midrule
+Pr\_PoIm  & Probable            & 0.77 & 0.75 & 0.76 & 103 \\
+         & Possible/Improbable & 0.94 & 0.94 & 0.94 & 402 \\
+         & accuracy            &      &      & 0.9  & 505 \\
+         & weighted avg        & 0.9  & 0.9  & 0.9  & 505 \\ \midrule
+PrPo\_Im  & Probable/Possible   & 0.85 & 0.96 & 0.9  & 183 \\
+         & Improbable          & 0.97 & 0.9  & 0.94 & 322 \\
+         & accuracy            &      &      & 0.92 & 505 \\
+         & weighted avg        & 0.93 & 0.92 & 0.92 & 505 \\ \bottomrule
 \end{tabular}
-    \caption{Performance metric summaries across the four scenarios}
-    \label{tab:my_label}
+    \caption{Performance metrics across the four scenarios}
+    \label{tab:performance}
 \end{table}
@@
 
+\eject
 \bibliography{ai-trees-references}

### 2021-02-23 bin/python/cnn-performance.py
new images

@@ -66,9 +66,9 @@ class Metrics(Callback):
         xVal, yVal = self.validation_data
         val_pred = np.argmax(np.asarray(self.model.predict(xVal)), axis=1)
         val_true = np.argmax(yVal, axis=1)        
-        _val_f1 = f1_score(val_true, val_pred, average='weighted', zero_division = 0)
-        _val_precision = precision_score(val_true, val_pred, average='weighted', zero_division = 0)
-        _val_recall = recall_score(val_true, val_pred, average='weighted', zero_division = 0)
+        _val_f1 = f1_score(val_true, val_pred, average='macro', zero_division = 0)
+        _val_precision = precision_score(val_true, val_pred, average='macro', zero_division = 0)
+        _val_recall = recall_score(val_true, val_pred, average='macro', zero_division = 0)
 
         self.val_f1s.append(_val_f1)
         self.val_recalls.append(_val_recall)
@@ -98,7 +98,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_loss', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='loss', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -114,7 +114,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     model.compile(loss='categorical_crossentropy',  optimizer = opt, metrics =  ['accuracy']) 
     hist = model.fit(training_images, training_labels, batch_size = 32, epochs = num_epochs, verbose=1, 
                      validation_data=(validation_images, validation_labels),
-                     callbacks = [model_metrics]) #, early_stopping])     
+                     callbacks = [model_metrics, early_stopping])     
     
     # SAVE MODEL, SUMMARY AND PERFORMANCE
     model_name = "opt-cnn-" + scenario + "-" +str(image_size) + "-px"
@@ -165,12 +165,12 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     return(model, hist) #performance_dict)
 
 
-def visualizeCNN(model, scenario, image_size, images_per_class = 4, seed_num = 1, saliency=False):
[…]

### 2021-02-23 bin/python/plot-processed-images.py

@@ -1,0 +1,27 @@
+#TODO (in order to not have to regenerate images each time)
+def plotProcessedImages(class_scenario, image_array, class_list, images_per_class, resolution):
+    num_cols = images_per_class
+    num_rows = len(class_list)
+    fig, axarr = plt.subplots(num_rows, num_cols, sharex=True, sharey=True)
+    #print(image_file_list)
+    for i in range(num_rows):
+        for j in range(num_cols):
+            axarr[i, j].imshow(image_array[i][j], cmap = 'gist_gray', extent = [0, resolution, 0, resolution])
+    for ax, row in zip(axarr[:, 0], [i for i in class_list]):
+        ax.set_ylabel(row, size=15)
+    image_filename = '../../figures/processed_input_images_' + str(class_scenario) + '_' + str(resolution) + '_px.png'
+    #plt.xticks([0,3024])
+    #plt.yticks([0,4032])
+    fig.savefig(image_filename, dpi=180)
+    return
+
+def main():
+    for scenario in CLASSIFICATION_SCENARIO_LIST:
+        for image_size in IMG_SIZE_LIST:
+            imd = np.load('../../data/tidy/preprocessed_images/size128_exp5_Pr_Po_Im.npy',allow_pickle = True)
+            array_random_images = 
+            plotProcessedImages(scenario, array_random_images, classes, images_per_class=NUM_PLOT_IMAGES_PER_CLASS, resolution=image_size)
+    return
+
+if __name__ == "__main__":
+    main()

### 2021-02-23 bin/python/preprocess-images.py

@@ -17,15 +17,15 @@ import matplotlib.pyplot as plt
 
 ## GLOBAL VARIABLES
 IMG_SIZE = 108
-IMG_SIZE_LIST = [64]#, 128, 224]#, 384]
+IMG_SIZE_LIST = [64, 128]#, 224]#, 384]
 NUM_CHANNELS = 1
 CLASSIFICATION_SCENARIO = "Pr_Im"
-CLASSIFICATION_SCENARIO_LIST = ["Pr_Po_Im"]#, "Pr_Im", "PrPo_Im", "Pr_PoIm"]  
+CLASSIFICATION_SCENARIO_LIST = ["Pr_Po_Im", "Pr_Im", "PrPo_Im", "Pr_PoIm"]  
 EXPANSION_FACTOR = 5 #5 of augmented images
 LABELED_IMAGES_DIR = '../../data/tidy/labeled_images'
 PROCESSED_IMAGES_DIR = '../../data/tidy/preprocessed_images'
-SEED = 10 # seed for repeatability
-NUM_PLOT_IMAGES_PER_CLASS = 4
+SEED = 100  # 10 seed for repeatability
+NUM_PLOT_IMAGES_PER_CLASS = 1 #4
 
 def getImageOneHotVector(image_file_name, classification_scenario = "Pr_Im"):
     """Returns one-hot vector encoding for each image based on specified classification scenario:
@@ -66,7 +66,7 @@ def getImageOneHotVector(image_file_name, classification_scenario = "Pr_Im"):
         else :
             return np.array([0, 0]) # if label is not present for current image        
 
-def processImageData(img_size, expansion_factor, class_scenario, seed_value, channels=1, images_per_class=4): # original size 4032 × 3024 px
+def processImageData(img_size, expansion_factor, class_scenario, seed_value, channels=1, images_per_class=4, save_image_binary_files=True): # original size 4032 × 3024 px
     data = []
     image_list = os.listdir(LABELED_IMAGES_DIR)
     random.seed(seed_value) #seed for repeatability
@@ -101,8 +101,8 @@ def processImageData(img_size, expansion_factor, class_scenario, seed_value, cha
             #img = img.resize((img_size, img_size), Image.BICUBIC)
             #TODO: random cropping, flipping and resizing
     print("Images:", class_scenario, (np.array([x[1] for x in data])).sum(axis=0) )
[…]

### 2021-02-24 bin/python/cnn-performance.py
new images

@@ -91,14 +91,14 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     # training_images, validation_images, training_labels, validation_labels =  train_test_split(np.array([np.expand_dims(x[0],axis=2) for x in image_sets[image_size][scenario]]), 
     #                                                                                            np.array([x[1] for x in image_sets[image_size][scenario]]), 
     #                                                                                            stratify= np.array([x[1] for x in image_sets[image_size][scenario]]), 
-    #                                                                                            test_size = .2, random_state = 1)
+    #                                                                                            test_size = .2, random_state = trial_seed)
 
     print("Number of class training images:", training_labels.sum(axis=0), "total: ", training_labels.sum())
     print("Number of class validation images:", validation_labels.sum(axis=0), "total: ", validation_labels.sum())
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='loss', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_loss', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -114,7 +114,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     model.compile(loss='categorical_crossentropy',  optimizer = opt, metrics =  ['accuracy']) 
     hist = model.fit(training_images, training_labels, batch_size = 32, epochs = num_epochs, verbose=1, 
                      validation_data=(validation_images, validation_labels),
-                     callbacks = [model_metrics, early_stopping])     
+                     callbacks = [model_metrics,early_stopping])     
     
     # SAVE MODEL, SUMMARY AND PERFORMANCE
     model_name = "opt-cnn-" + scenario + "-" +str(image_size) + "-px"
@@ -238,7 +238,7 @@ def visualizeCNN(model, scenario, image_size, images_per_class = 4, trial_seed =
         for j in np.arange(images_per_class):
             print(i, j, image_counter)
             heatmap = np.uint8(cm.jet(cam[image_counter])[..., :3] * 255)            
-            ax[i, j].imshow(gradcam_images[image_counter])
[…]

### 2021-02-24 bin/python/cnn-performance.py
new trial of cnn-performance results

@@ -45,9 +45,14 @@ NUM_EPOCHS = 20
 SAVED_MODEL_DIR = '../../results/models/'
 MODEL_PERFORMANCE_METRICS_DIR = '../../results/model-performance/'
 
+#
+#mage_sets = createResolutionScenarioImageDict(RESOLUTION_LIST, SCENARIO_LIST)
 
-image_sets = createResolutionScenarioImageDict(RESOLUTION_LIST, SCENARIO_LIST)
-
+image_sets = dict.fromkeys(RESOLUTION_LIST)
+for p in RESOLUTION_LIST:
+    image_sets[p] = dict.fromkeys(SCENARIO_LIST)
+    for s in SCENARIO_LIST:
+        image_sets[p][s] = np.load('../../data/tidy/preprocessed_images/size' + str(p) + '_exp5_' + s + '.npy', allow_pickle = True)
 
                                    
 # https://github.com/keisen/tf-keras-vis/blob/master/examples/attentions.ipynb

### 2021-02-24 bin/python/cnn-performance.py
added testing functionality to cnn-performance.py to avoid overwriting results during tests

@@ -46,13 +46,13 @@ SAVED_MODEL_DIR = '../../results/models/'
 MODEL_PERFORMANCE_METRICS_DIR = '../../results/model-performance/'
 
 #
-#mage_sets = createResolutionScenarioImageDict(RESOLUTION_LIST, SCENARIO_LIST)
+image_sets = createResolutionScenarioImageDict(RESOLUTION_LIST, SCENARIO_LIST)
 
-image_sets = dict.fromkeys(RESOLUTION_LIST)
-for p in RESOLUTION_LIST:
-    image_sets[p] = dict.fromkeys(SCENARIO_LIST)
-    for s in SCENARIO_LIST:
-        image_sets[p][s] = np.load('../../data/tidy/preprocessed_images/size' + str(p) + '_exp5_' + s + '.npy', allow_pickle = True)
+# image_sets = dict.fromkeys(RESOLUTION_LIST)
+# for p in RESOLUTION_LIST:
+#     image_sets[p] = dict.fromkeys(SCENARIO_LIST)
+#     for s in SCENARIO_LIST:
+#         image_sets[p][s] = np.load('../../data/tidy/preprocessed_images/size' + str(p) + '_exp5_' + s + '.npy', allow_pickle = True)
 
                                    
 # https://github.com/keisen/tf-keras-vis/blob/master/examples/attentions.ipynb
@@ -85,7 +85,7 @@ class Metrics(Callback):
         return
 
 
-def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_seed = 1): 
+def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_seed = 1, testing = True): 
     
     # IMAGES (former approach)
     training_images_and_labels, test_images_and_labels = splitData(image_sets[image_size][scenario], prop = 0.8, seed_num = trial_seed)
@@ -107,38 +107,33 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
-    ## shape of images
-    input_shape = (image_size, image_size, NUM_CHANNELS)
-    ## learning rate
-    opt_learning_rate = getOptCNNHyperparams(image_size, scenario)['learning_rate']
+    input_shape = (image_size, image_size, NUM_CHANNELS) ## shape of images
[…]

### 2021-02-24 bin/python/cnn-performance.py
updated bib file and new results

@@ -103,7 +103,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_loss', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_accuracy', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -278,7 +278,7 @@ def visualizeCNN(model, scenario, image_size, images_per_class = 4, trial_seed =
     return
 
 
-def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1):
+def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_boolean = True):
     df = pd.DataFrame()
     for s in SCENARIO_LIST:
         m, h = trainModelWithDetailedMetrics(res, s, num_epochs, trial_seed = seed_val, testing = test_boolean)
@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1):
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=15, seed_val = 1, testing=True)
+    getScenarioModelPerformance(res=128, num_epochs=20, seed_val = 111, test_boolean=False)

### 2021-02-24 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
 \affil[1]{Department of Civil and Environmental Engineering, University of Massachusetts Amherst, MA 01003, USA} %. Email: author.one@umass.edu}
-\affil[2]{Department of Ecological Conservation, University of Massachusetts Amherst, MA 01003, USA}
+\affil[2]{Department of Environmental Conservation, University of Massachusetts Amherst, MA 01003, USA}
 
@@
 \section{Introduction}
-Power outages due to contact between tree parts and power lines annually cause tens of billions of dollars in economic and other disruptions throughout the United States, despite extensive efforts by utilities to mitigate and prevent those contacts.
+Despite extensive efforts by utilities to prevent them, contacts between tree parts and power lines cause outages that annually result in tens of billions of dollars in economic costs throughout the United States.
 Presently, the identification of potential contact between trees and power lines is labor intensive and time-consuming.
-This paper describes an artificial intelligence and machine learning approach that automatically classifies trees, using only a single photograph and with a high degree of accuracy, into categories corresponding to probable failure, possible failure and improbable failure-categories corresponding to those used by utility arborists to assess risk of contact between trees and power lines.
-This preliminary study demonstrates the possible efficacy of AI approaches to tree risk assessment and, following further development of the approach has the potential to reduce power outages and utility costs by allowing utilities to more effectively target their pruning and mitigation efforts.
+This paper describes an artificial intelligence and machine learning approach that automatically classifies trees, using only a single photograph and with a high degree of accuracy, into categories used by utility arborists to describe the likelihood of tree failure: probable, possible, and improbable.
+This preliminary study demonstrates the possible efficacy of AI approaches to tree risk assessment and, following further development of the approach, has the potential to reduce power outages and utility costs by allowing utilities to more effectively target their pruning and mitigation efforts.
 
-Contact between tree parts and power lines can take several forms: tree branches can grow into lines; branches can fail and fall onto lines; whole-tree failure can occur due to uprooting or trunk failure.
-A study in the state of Connecticut, USA provides some context for the amount of economic disruption, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.
+Contact between tree parts and power lines can take three forms: tree branches can grow into lines; branches can fail and fall onto lines; whole-tree failure can occur due to uprooting or trunk failure.
+A study in Connecticut, USA provides some context for the amount of economic disruption, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.
 That extremely high cost occurred despite extensive efforts on the part of utilities to mitigate conflicts between trees and power lines through active and aggressive pruning programs that, on their own cost billions of dollars annually \cite{guggenmoos2003effects}.
 
-Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called ``preventable'' contact incidents between trees and power lines.
-For example, in Massachusetts, USA, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}.
+Despite its high cost, pruning trees to maintain clearance from power lines is an effective way to reduce outages due to so-called ``preventable'' contacts between trees and power lines.
+For example, in Massachusetts, USA, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}; similar results were found in a study conducted in Connecticut \cite{parent2019analysis}.
 The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \cite{nateghi2014power}.
@@
 Even effective pruning cannot, however, completely eliminate tree-caused outages.
-Failure of trees away from the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}.
-The proportion of tree failures away from the wires that causes outages varies and has not been rigorously quantified.
+Failure of trees outside the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}.
+The proportion of outages caused by failure of trees outside the right-of-way has not been rigorously quantified.
 \citeN{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citeN{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
@@
 Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{smiley2017best,goodfellow2020best}.
-Risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences.
+Estimating tree risk includes assessing the likelihood of tree failure, the likelihood of impact of the failed tree (or tree part) on a target, and the severity of consequences of the impact.
 The likelihood of failure depends on the anticipated loads on the tree and its load-bearing capacity.
 The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees.
-Severity of consequences depends on the damage done to the infrastructure---which, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure---and, more importantly in some cases, the economic costs and disruption associated with outages.
+Severity of consequences depends on the damage done to the infrastructure---which, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure---and, more importantly in some cases, the economic costs and disruption associated with electrical outages.
 
@@
 In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{smiley2017best}.
-Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}.
-The utility of Level 1 assessments in these states suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.
+Studies in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency} have shown that, compared to more time-consuming risk assessments, Level 1 risk assessments successfully identified trees with a higher degree of risk--precisely the trees that arborists prioritize for risk mitigation.
+The utility of Level 1 assessments demonstrated in these studies suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment while still identifying high risk trees.
 
 The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible, and improbable.
-The data used for training, testing and illustration of the method consists of 505 tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.
+The data used for training, testing and illustration of the method consists of 505 tree images that have been classified by the authors according to best management practices used by utility arborists \cite{goodfellow2020best}.
 
@@
 AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances,wang2019novel}.
-Neural networks have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citeN{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
+Neural networks, which comprise a major category of AI frameworks, have been widely applied in the field of earthquake risk assessment (an excellent review is provided by \citeN{xie2020promise}), but the authors are not aware of attempts to operate directly on, for example, building images in the absence of technical structural data to predict seismic risk.
 Neural networks have also been used to interrogate remote sensing data of the landscape to assess landslide risk \cite{su2020deep}.
-
-A relevant application for tree species identification using a convolutional neural network (CNN) was recently demonstrated \cite{fricker2019convolutional}.
-Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
+A few recent efforts have demonstrated the potential for AI-based tree recognition from drone imagery \cite{santos2019assessment, egli2020cnnbased}.
+Furthermore, an application of a convolutional neural network (CNN) to tree species identification using was recently demonstrated by \citeN{fricker2019convolutional}.
+Yet, AI has yet to be applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.
 
@@
 However, the breakthrough came with the ``neocognitron'' \cite{fukushima1980neocognitron}, which
-was a self-learning neural network for pattern recognition that was robust to changes in position and shape distortion, a problem that plagued earlier efforts, including ``cognitron'' \cite{fukushima1975cognitron} proposed a few years earlier.
+was a self-learning neural network for pattern recognition that was robust to changes in position and shape distortion, a problem that plagued earlier efforts, including the ``cognitron'' also proposed by \citeN{fukushima1975cognitron}.
+%proposed a few years earlier.
 
@@
 extraction.
-\cite{lecun1989handwritten} soon afterward introduced a multilayer neural network that mapped a feature in each neuron (representing a ``local receptive field'') via convolution.
+\citeN{lecun1989handwritten} soon afterward introduced a multilayer neural network that mapped a feature in each neuron (representing a ``local receptive field'') via convolution.
 This network could also be trained by backpropagation like other existing neural networks and featured pooling operations for better distortion and
@@
 
-Along with these developments in their architectures, CNNs have demonstrated viability for applications to image
-classification, object and text detection, object and document tracking, labeling, speech, among several other related
-fields \cite{gu2018recent}. %\hl{Paragraph on applications to be completed---with particular attention paid to tree-related efforts.}
+Along with these developments in their architectures, CNNs have demonstrated viability for applications ranging from image classification, object and text detection to document tracking, labeling, speech, among several other related
+fields \cite{gu2018recent}. 
+In this study, we show that a relatively simple CNN architecture coupled with state-of-the-art approaches for model training and regularization is capable of efficiently and effectively predicting tree failure classes.
+%\hl{Paragraph on applications to be completed---with particular attention paid to tree-related efforts.}
 
@@
 A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
-Level 2 or Level 3 \cite{smiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.
+Level 2 or Level 3 \cite{smiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility right-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{goodfellow2020best}; and (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.
 The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{smiley2017best}:
@@
 In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category.
-Two randomly chosen examples are shown for each category in \autoref{fig:raw_images}.
+Randomly chosen examples are shown for each category in \autoref{fig:raw_images}.
 In the original set of training images, the class distribution is given in \autoref{tab:classdist}.
@@
     \centering
-    \begin{tabular}{l l}\toprule
-    \bf Class Label     & \bf Number of images  \\ \midrule
+    \begin{tabular}{l r}\toprule
+    \bf Category     & \bf Number of images  \\ \midrule
     Improbable & 322\\
@@
     \end{tabular}
-    \caption{Class distribution of images in training set}
+    \caption{Category distribution of images in the original set of input images}
     \label{tab:classdist}
@@
 
-\subsection{Classification scenarios}
-We define four classification scenarios in \autoref{tab:scenarios}.
+\subsection{Pre-processing and data augmentation}
+Data augmentation refers to the variety of methods that are employed for synthetically generating more samples in a training dataset in order to improve model performance \cite{wong2016understanding}.
+Augmentation is desired, particularly in situations where the number of original observations is small, and the effectiveness of various relevant techniques in this domain has been amply demonstrated \cite{shorten2019survey}.
 
-\begin{table}[h!]
-    \centering
-    \begin{tabular}{l l c}\toprule
-    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
-    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
-    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
-    \texttt{PrPo\_Im}        & \{Probable|Possible, Improbable\}  & 2 \\
-    \texttt{Pr\_PoIm}        & \{Probable, Possible|Impossible\}  & 2 \\\bottomrule
-    \end{tabular}
-    \caption{Classification scenarios}
-    \label{tab:scenarios}
-\end{table}
-
-\subsection{Pre-processing and data augmentation}
-To achieve robustness in training, and given the relatively small number of training images, we randomly cropped each image on either axis to 3024 $\times$ 3024 pixels, generating five instances for each one.
+In order to achieve robustness in our model, and given the relatively small number of training images, we randomly cropped each image on either axis to 3024 $\times$ 3024 pixels, generating five instances for each one.
 Thus, we increased the size of our training set from 505 to 2525 images.
@@
 \subsection{Convolutional neural network}
-
 % We can apply cut-out (occluding portions of the image) for improved performance \cite{devries2017improved}.
 Also, it has been shown that training with lower resolution improves performance on higher resolution test images \cite{touvron2019fixing}.
+We employ a convolutional neural network (CNN) as the AI framework for tree risk failure likelihood prediction.
+Like other neural networks, the CNN is an arrangement of neurons within layers, each neuron performing an operation that maps from a pixel in an input image to the final output. 
+The input into each neuron is a weighted sum from the previous layer, while the output from each neuron is modulated by an activation function.
+The activation function in the final layer of an CNN is typically the softmax function, which gives the class probabilities of a given input image. 
+A class is assigned to the image based on the one with the corresponding maximum probability.
 
+Unlike other neural networks, however, the CNN performs fundamental pixel-mapping operations in its "convolutional" layers.
+Each convolutional layer is defined by a stack of feature maps, which result from a dot product of a filter and correspondinlgy-sized local receptive fields from the input image or preceding layer.
+The numeric values of the filters correspond to weights whose optimal values are learned during the training of the CNN.
+The size of the filter in each convolutional layer is given by the "kernel size."
+
+In this paper, we employ a relatively simple CNN architecture, historically inspired by AlexNet \cite{krizhevsky2012imageneta}.
 The structure of the CNN is shown in \autoref{fig:cnn}.
-
+After the input layer (a matrix of pixels from the input image), we use a 64-filter convolutional layer.
+The output is downsampled using a pooling layer that returns the maximum ouput from a $2\times 2$ subsample from the previous layer.
+Next, we stack two successive convolutional layers each with a depth of 128 filters.
+We follow these with a maximum-pooling layer and then two further 256-filter convolutional layers.
+A final maximum-pooling layer is used before we "flatten" all outputs into a one-dimensional (fully-connected) array of neurons. 
+After flattening the outputs, we use a dense layer to reduce the number of outputs a specified number of units.
+Batch normalization \cite{ioffe2015batch} is employed after the first dense layer to improve training efficiency.
+A second dense layer is used prior to the output layer, with the number of outputs corresponding to the number of classes in the dataset.
+A regularization technique known as "dropout" is used after each dense layer.
+In each dropout layer, a proportion of the neurons are randomly zeroed during training in order to improve the robustness of the model.
 \begin{figure}
@@
 
-\subsection{Hyperparameter optimization}
-We used the Hyperband approach \cite{li2018hyperband} to perform a guided grid search to find the optimal hyperparameter values across each scenario and input resolution instance.
-Thus, we performed 12 searches on the following hyperparameters in our CNN:
-\begin{itemize}
-    \item kernel size in first convolutional layer $ks$
-    \item number of units in first densely connected layer $u_1$
-    \item dropout rate applied to outputs of first dense layer $d_1$
-    \item activation function for first dense layer $a_1$
-    \item number of units in second densely connected layer $u_2$
-    \item dropout rate applied to outputs of second dense layer $d_2$
-    \item activation function for second dense layer $a_2$
-    \item learning rate (of Adam optimizer) $\lambda$
-\end{itemize}
-Each search was conducted using 90 trials of unique hyperparameter combinations.
+The various hyperparameters in the model are summarized in \autoref{tab:hyp}.
+We optimized eight of these parameters using the Hyperband approach \cite{li2018hyperband}.
+The others, however, were fixed.
+
+\begin{table}[]
+    \centering
+    \begin{tabular}{}
+         &  \\
+         & 
+    \end{tabular}
+    \caption{Caption}
+    \label{tab:my_label}
+\end{table}
+
+\subsection{Classification scenarios}
+In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, 
+we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
+Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
+Thus, 
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l c}\toprule
+    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
+    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
+    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
+    \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
+    \texttt{Pr\_PoIm}        & \{Probable, Possible + Impossible\}  & 2 \\\bottomrule
+    \end{tabular}
+    \caption{Classification scenarios}
+    \label{tab:scenarios}
+\end{table}
+
+\subsection{Model training and assessment}
+In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the macro-averages of the precision, recall and $F_1$ score metrics.
+The precision score $Pr$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is a important measure of how good a classifier is.
+Meanwhile, the recall $Re$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
+The $F_1$ metric is given as the harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
+These three metrics are macro-averaged for each class for assessing model performing.
+Thus, each class is given equal weight, ensuring that misclassifications within the smaller classes ("Probable" and "Possible") are adequately represented in the aggregation.
+These metrics are formally defined as follows:
+\begin{align}
+\text{Macro-average precision: }    Pr^m &= \frac{1}{|C|}\sum_{c\in C} \left(  \frac{ \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FP}_c}\right) \\
+\text{Macro-average recall: }    Re^m &= \frac{1}{|C|} \sum_{c\in C} \left(  \frac{ \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FN}_c}\right) \\
+\text{Macro-average $F_1$ score: }    F_1^m &= \frac{1}{|C|} \sum_{c\in C} \left(  \frac{2  \mathrm{Pr}_{c} \mathrm{Re}_{c}}{\mathrm{Pr}_{c} + \mathrm{Re}_c}\right)
+\end{align}
+% \begin{align}
+% \text{Weighted average precision: }    \mathrm{Pr}^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FP}_c}\right) \\
+% \text{Weighted average recall: }    \mathrm{Re}^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FN}_c}\right) \\
+% \text{Weighted average $F_1$ score: }    F_1^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{2 n_c \mathrm{Pr}_{c} \mathrm{Re}_{c}}{\mathrm{Pr}_{c} + \mathrm{Re}_c}\right)
+% \end{align}
+where $c$ is the index of a class in the set $C$ and $|C|$ the number of classes in the dataset. 
+The class-specific prediction metrics are given by:
+\begin{align}
+    \text{True positives for class $c$: } \mathrm{TP}_c &= \sum_{i \in c} I(\hat{y}_i = y_i ) \\
+    \text{False positives for class $c$: } \mathrm{FP}_c &= \sum_{i \in c} I(\hat{y}_i \in c | y_i \notin c) \\
+    \text{False negatives for class $c$: } \mathrm{FN}_c &= \sum_{i \in c} I(\hat{y}_i \notin c | y_i \in c)
+\end{align}
 
@@
 
-\subsection{Optimal hyperparameters}
-The results of the 8 hyperparameters optimized in each of the 12 training instances is shown in \autoref{tab:opt-hyp}.
+\subsection{Hyperparameter optimization}
+We performed a guided grid search to find the optimal hyperparameter values across each of the 12 unique classification scenario and image resolution instances.
+Each search was conducted using 90 trials of unique hyperparameter combinations. The results of the 8 hyperparameters optimized in each of the 12 training instances is shown in \autoref{tab:opt-hyp}.
 \begin{table}[ht!]\small
@@
 \end{table}
+
+\subsection{Analysis of classification strategies}
+We investigated in the detail the performance of the four classification scenarios earlier outlined, using the 64px resolution images.
+The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
+
+Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
+The metrics are shown in \autoref{fig:cnn-perf}.
+In each scenario, the model was trained over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
+That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
+From this figure, we see that Pr\_Po\_Im performed significant worse than the other three.
+This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
+Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=1.05\textwidth]{cnn-performance-metrics-128px-plot}
+    \caption{Model performance across four classification scenarios for a single training instance in each case.
+Training resolution was 128px.
+All weighted average metrics were computed on the validation set.}
+    \label{fig:cnn-perf}
+\end{figure}
+
+We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
+
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{pr_po_im_cm}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Im}}
+    \label{pr_im_cm}
+  \end{subfigure}%
+
+   \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
+    \caption{Scenario \texttt{PrPo\_Im}}
+    \label{prpo_im_cm}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
+    \caption{Scenario \texttt{Pr\_PoIm}}
+    \label{pr_poim_cm}
+  \end{subfigure}%
+    \caption{Confusion matrix for final model performance using the 128px images across all scenarios.}
+    \label{fig:confusion-matrix}
+\end{figure}
+
 \subsection{Sensitivity to training resolution}
@@
 We re-trained the model on all the four scenarios and then evaluated performance on the following validation metrics: average precision, average recall and accuracy.
-
-\subsection{Analysis of classification strategies}
-We investigated in the detail the performance of the four classification scenarios earlier outlined, using the 64px resolution images.
-The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
-In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the weighted averages of the precision, recall and $F_1$ score metrics.
-The precision score $\mathrm{Pr}$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is a important measure of how good a classifier is.
-Meanwhile, the recall $\mathrm{Re}$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
-The $F_1$ metric is given as the harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
-These three metrics can also be weighted by the number of observations in each class in order to better indicate class-specific performance.
-We formally define these metrics in the following equations:
-
-\begin{align}
-\text{Weighted average precision: }    \mathrm{Pr}^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FP}_c}\right) \\
-\text{Weighted average recall: }    \mathrm{Re}^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{n_c \mathrm{TP}_{c} }{\mathrm{TP}_{c} + \mathrm{FN}_c}\right) \\
-\text{Weighted average $F_1$ score: }    F_1^w &= \frac{1}{\sum_{c\in C} n_c} \sum_{c\in C} \left(  \frac{2 n_c \mathrm{Pr}_{c} \mathrm{Re}_{c}}{\mathrm{Pr}_{c} + \mathrm{Re}_c}\right)
-\end{align}
-where the class-based metrics are given by:
-\begin{align}
-    \mathrm{TP}_c &= \sum_{i \in c} I(\hat{y}_i = y_i ) \\
-    \mathrm{FP}_c &= \sum_{i \in c} I(\hat{y}_i \in c | y_i \notin c) \\
-    \mathrm{FN}_c &= \sum_{i \in c} I(\hat{y}_i \notin c | y_i \in c)
-\end{align}
-
-Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
-The metrics are shown in \autoref{fig:cnn-perf}.
-In each scenario, the model was trained over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
-That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
-From this figure, we see that Pr\_Po\_Im performed significant worse than the other three.
-This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
-Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=1.05\textwidth]{cnn-performance-metrics-128px-plot}
-    \caption{Model performance across four classification scenarios for a single training instance in each case.
-Training resolution was 128px.
-All weighted average metrics were computed on the validation set.}
-    \label{fig:cnn-perf}
-\end{figure}
-
-We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
-
-\begin{figure}[ht]
-  \centering
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
-    \label{pr_po_im_cm}
-  \end{subfigure}%
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
-    \caption{Scenario \texttt{Pr\_Im}}
-    \label{pr_im_cm}
-  \end{subfigure}%
-
-   \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
-    \caption{Scenario \texttt{PrPo\_Im}}
-    \label{prpo_im_cm}
-  \end{subfigure}%
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
-    \caption{Scenario \texttt{Pr\_PoIm}}
-    \label{pr_poim_cm}
-  \end{subfigure}%
-    \caption{Confusion matrix for final model performance using the 128px images across all scenarios.}
-    \label{fig:confusion-matrix}
-\end{figure}
 
@@
 
-\subsection{Comparison with state-of-the-art architectures}
-We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \autoref{tab:comp}.
+% \subsection{Comparison with state-of-the-art architectures}
+% We compare the performance of our selected model with existing high-performance architectures. The results are summarized in \autoref{tab:comp}.
 
-\begin{table}[h!]\small
-  \centering
-  \begin{tabular}{l l l l l l l }\toprule
-    \bf Model & \multicolumn{3}{c}{\bf Training metrics} &\multicolumn{3}{c}{\bf Validation metrics}  \\\midrule
-    & Error & Precision & Recall     & Error & Precision & Recall \\
-    SafeTree & & & & & & \\
-    GoogleNet (InceptionV3) & & & & & & \\
-    ResNet50 & & & & & & \\
-        VGGNet & & & & & & \\
-    AlexNet & & & & & & \\\bottomrule
-  \end{tabular}
-  \caption{Comparing our model SafeTree with state-of-the-art CNN architectures trained on our data}
-  \label{tab:comp}
-\end{table}
+% \begin{table}[h!]\small
+%   \centering
+%   \begin{tabular}{l l l l l l l }\toprule
+%     \bf Model & \multicolumn{3}{c}{\bf Training metrics} &\multicolumn{3}{c}{\bf Validation metrics}  \\\midrule
+%     & Error & Precision & Recall     & Error & Precision & Recall \\
+%     SafeTree & & & & & & \\
+%     GoogleNet (InceptionV3) & & & & & & \\
+%     ResNet50 & & & & & & \\
+%         VGGNet & & & & & & \\
+%     AlexNet & & & & & & \\\bottomrule
+%   \end{tabular}
+%   \caption{Comparing our model SafeTree with state-of-the-art CNN architectures trained on our data}
+%   \label{tab:comp}
+% \end{table}
+
 \section{Conclusion}

### 2021-02-24 docs/manuscript/manuscript.tex
Extensive updates to Data and Methods section

@@
 \usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
+\usepackage[table]{xcolor}
 
@@
 
-\begin{table}[]
+\begin{table}[]\small
+\rowcolors{1}{white}{gray!35}
     \centering
-    \begin{tabular}{}
-         &  \\
-         & 
+    \begin{tabular}{l c c c c c c c}\toprule
+    \bf Layer & \bf Layer \# & \bf No.\ Filters & \bf Kernel Size & \bf Strides &
+     &  & \bf Activation \\\midrule
+    Convolutional  & 1         & 64                & $k$   & 2           &&& ReLu \\
+    Max. Pooling   & 1         &                   & 2               &&&             &      \\
+    Convolutional  & 2         & 128               & 3   & 1           &&& ReLu \\
+    Convolutional  & 3         & 128               & 3   & 1           &&& ReLu \\
+    Max. Pooling   & 2         &                   & 2               &&&             &      \\    
+    Convolutional  & 4         & 256               & 3   & 1           &&& ReLu \\
+    Convolutional  & 5         & 256               & 3   & 1           &&& ReLu \\
+    Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
+    Flatten & & & & &&& \\\midrule
+     &  &   &  & &
+     \bf Rate & \bf No.\ Units & \\\midrule
+    Dense           & 1  & & &  & & $u_1^*$ & $a_1^*$ \\
+    Dropout         & 1  & & &  & $r_1^*$ & &  \\
+    Dense           & 2  & & &  & & $u_2^*$ & $a_2^*$ \\
+    Dropout         & 2  & & &  & $r_2^*$ & & \\ \bottomrule
     \end{tabular}
-    \caption{Caption}
-    \label{tab:my_label}
+    \caption{Summary of the convolutional neural network hyperparameters. 
+    Those indicated by an asterisked symbol are optimized using a guided search.}
+    \label{tab:hyp}
 \end{table}
@@
 \subsection{Model training and assessment}
-In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the macro-averages of the precision, recall and $F_1$ score metrics.
+The CNN is trained using a variant of the stochastic gradient algorithm, Adam \cite{kingma2017adam}.
+The goal of the training procedure is to learn the optimal weights and bias terms for the CNN by minimizing a loss function.
+In this case, we use the categorical cross-entropy loss function, which is given by:
+\begin{equation}
+    L_{CE} = - \sum_{c\in C} y \log( f(c)_i)
+\end{equation}
+The softmax function returns the class prediction probabilities for a given observation $i$. It is defined for each class $c$ as:
+\begin{equation}
+    f(c)_i = \frac{e^{s^c_i}}{\sum_{c \in C} e^{s^c_i}}
+\end{equation}
+where $s^c_i$ is class score for the $i$th observation.
+Training is iteratively performed, with gradient updates averaged over a batch of input images.
+Here, we use a batch size of 32.
+The learning rate $\la$ of the optimization algorithm is an important hyperparameter that affects training performance.
+We optimized for this in the hyperparameter search as discussed.
+Furthermore, the CNN is trained over multiple passes through the entire training set.
+Each such pass is referred to as an epoch.
+
+In real terms, we measure the performance of the trained CNN by how accurately it predicts the classes in a validation set excluded from the training set.
+For this paper, we use a randomly sampled validation that was 20\% of the size of the input dataset of 2525 images in each training instance.
+Thus, we define the accuracy as the overall proportion of correct predictions across all classes.
+This metric was computed both for the training and validation sets in each epoch
+In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the macro-averages of the precision, recall and $F_1$ score metrics computed over the validation set in each epoch.
 The precision score $Pr$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is a important measure of how good a classifier is.
@@
 \end{align}
+where $\hat y_i$ is the predicted class and $y_i$ the observed (true) class for a given image $i$ in class $c$.
+The indicator function $I(\cdot)$ returns 1 if the corresponding condition is true, and 0 otherwise.
 
@@
 \subsection{Analysis of classification strategies}
-We investigated in the detail the performance of the four classification scenarios earlier outlined, using the 64px resolution images.
+We investigated in the detail the performance of the four classification scenarios earlier outlined, using the 128-pixel resolution images.
 The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
-
 Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
@@
 That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
-From this figure, we see that Pr\_Po\_Im performed significant worse than the other three.
+From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three.
 This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
@@
 We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
-
+Each row of the confusion mat
 \begin{figure}[ht]
   \centering
-  \begin{subfigure}[t]{.5\linewidth}
+  \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Im}}
+    \label{pr_im_cm}
+  \end{subfigure}%
+   \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
+    \caption{Scenario \texttt{PrPo\_Im}}
+    \label{prpo_im_cm}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
+    \caption{Scenario \texttt{Pr\_PoIm}}
+    \label{pr_poim_cm}
+\end{subfigure}%
+  \begin{subfigure}[t]{.25\linewidth}
     \centering
@@
   \end{subfigure}%
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
-    \caption{Scenario \texttt{Pr\_Im}}
-    \label{pr_im_cm}
-  \end{subfigure}%
-
-   \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
-    \caption{Scenario \texttt{PrPo\_Im}}
-    \label{prpo_im_cm}
-  \end{subfigure}%
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
-    \caption{Scenario \texttt{Pr\_PoIm}}
-    \label{pr_poim_cm}
-  \end{subfigure}%
-    \caption{Confusion matrix for final model performance using the 128px images across all scenarios.}
+    \caption{Confusion matrices for CNN performance using the 128px images across all scenarios.
+    Each row of a confusion matrix indicates the proportional distribution of class predictions for the true members of each class.
+Thus, the diagonals indicate the recall for each class $Re_c$.
+The average of the diagonal values gives the macro-average $Re^m$ for each scenario.}
     \label{fig:confusion-matrix}

### 2021-02-24 bin/python/cnn-performance.py
new results

@@ -103,7 +103,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_accuracy', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_loss', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=20, seed_val = 111, test_boolean=False)
+    getScenarioModelPerformance(res=128, num_epochs=16, seed_val = 5, test_boolean=False)

### 2021-02-25 bin/python/cnn-performance.py
new results and bib file

@@ -103,14 +103,14 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_loss', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_accuracy', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
-    input_shape = (image_size, image_size, NUM_CHANNELS) ## shape of images
-    opt_learning_rate = getOptCNNHyperparams(image_size, scenario)['learning_rate']    ## learning rate
-    opt = tf.keras.optimizers.Adam(learning_rate = opt_learning_rate)
+    #input_shape = (image_size, image_size, NUM_CHANNELS) ## shape of images
     model = constructBaseCNN(image_size, scenario, num_channels = NUM_CHANNELS)    ## get model
+    opt_learning_rate = getOptCNNHyperparams(image_size, scenario)['learning_rate']    ## learning rate
+    opt = tf.keras.optimizers.Adam(learning_rate = opt_learning_rate)    
     reset_weights(model) # re-initialize model weights
     model.compile(loss='categorical_crossentropy',  optimizer = opt, metrics =  ['accuracy'])     ## compile and fit
     hist = model.fit(training_images, training_labels, batch_size = 32, epochs = num_epochs, verbose=1, 
@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=16, seed_val = 5, test_boolean=False)
+    getScenarioModelPerformance(res=128, num_epochs=15, seed_val = 1, test_boolean=False)

### 2021-02-25 bin/python/preprocess-images.py

@@ -17,7 +17,7 @@ import matplotlib.pyplot as plt
 
 ## GLOBAL VARIABLES
 IMG_SIZE = 108
-IMG_SIZE_LIST = [64, 128]#, 224]#, 384]
+IMG_SIZE_LIST = [128]#, 224]#, 384]
 NUM_CHANNELS = 1
 CLASSIFICATION_SCENARIO = "Pr_Im"
 CLASSIFICATION_SCENARIO_LIST = ["Pr_Po_Im", "Pr_Im", "PrPo_Im", "Pr_PoIm"]  
@@ -70,7 +70,7 @@ def processImageData(img_size, expansion_factor, class_scenario, seed_value, cha
     data = []
     image_list = os.listdir(LABELED_IMAGES_DIR)
     random.seed(seed_value) #seed for repeatability
-    print("Preprocessing images for scenario " + class_scenario + "and resolution " + str(img_size))
+    print("Preprocessing images for scenario " + class_scenario + "; resolution " + str(img_size))
     for img in image_list:
         label = getImageOneHotVector(img, class_scenario)
         if label.sum() == 0: # if image unlabeled, move to next one
@@ -143,6 +143,7 @@ def plotProcessedImages(class_scenario, image_array, class_list, images_per_clas
     image_filename = '../../figures/processed_input_images_' + str(class_scenario) + '_' + str(resolution) + '_px.png'
     #plt.xticks([0,3024])
     #plt.yticks([0,4032])
+    plt.tight_layout()
     fig.savefig(image_filename, dpi=180)
     return
 

### 2021-02-25 docs/manuscript/manuscript.tex

@@
 
-\subsection{Pre-processing and data augmentation}
+\subsection{Classification scenarios}
+In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, 
+we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
+Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
+Thus, 
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l c}\toprule
+    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
+    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
+    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
+    \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
+    \texttt{Pr\_PoIm}        & \{Probable, Possible + Impossible\}  & 2 \\\bottomrule
+    \end{tabular}
+    \caption{Classification scenarios}
+    \label{tab:scenarios}
+\end{table}
+
+\subsection{Image pre-processing and augmentation}
 Data augmentation refers to the variety of methods that are employed for synthetically generating more samples in a training dataset in order to improve model performance \cite{wong2016understanding}.
@@
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_Pr_Po_Im_64_px}
+    \includegraphics[width=\linewidth]{processed_input_images_Pr_Po_Im_128_px}
     \caption{Scenario \texttt{Pr\_Po\_Im}}
@@
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_64_px}
+    \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_128_px}
     \caption{Scenario \texttt{Pr\_Im}}
@@
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_64_px}
+    \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_128_px}
     \caption{Scenario \texttt{PrPo\_Im}}
@@
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_64_px}
+    \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_128_px}
     \caption{Scenario \texttt{Pr\_PoIm}}
@@
 The various hyperparameters in the model are summarized in \autoref{tab:hyp}.
-We optimized eight of these parameters using the Hyperband approach \cite{li2018hyperband}.
-The others, however, were fixed.
-
 \begin{table}[]\small
@@
      &  & \bf Activation \\\midrule
-    Convolutional  & 1         & 64                & $k$   & 2           &&& ReLu \\
+    Convolutional  & 1         & 64                & $k^*$   & 2           &&& ReLu \\
     Max. Pooling   & 1         &                   & 2               &&&             &      \\
@@
 
-\subsection{Classification scenarios}
-In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, 
-we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
-Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
-Thus, 
-\begin{table}[h!]
+\subsection{Hyperparameter optimization}
+We optimized eight of the CNN hyperparameters using Hyperband \cite{li2018hyperband}, an efficient guided grid-search algorithm.
+Twelve searches were performed for each classification scenario and image resolution combination.
+Each search was conducted using 90 trials of unique hyperparameter combinations. The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}.
+\begin{table}[ht!]\small
+%\rowcolors{7}{white}{gray!35}
     \centering
-    \begin{tabular}{l l c}\toprule
-    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
-    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
-    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
-    \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
-    \texttt{Pr\_PoIm}        & \{Probable, Possible + Impossible\}  & 2 \\\bottomrule
+    \sisetup{round-mode=places,
+        round-precision=2,
+        scientific-notation=fixed,
+        exponent-product = \cdot,
+        table-omit-exponent,
+        fixed-exponent=-4
+        }
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% \usepackage{multirow}
+\begin{tabular}{@{}lllrrr@{}}
+\toprule
+\multirow{2}{*}{\textbf{Scenario}} & \multirow{2}{*}{\textbf{Hyperparameter}} & \multirow{2}{*}{\textbf{Range}} & \multicolumn{3}{c}{\textbf{Resolution}} \\ \cmidrule(l){4-6} 
+                          &                              &                           & \textbf{64}             & \textbf{128}            & \textbf{224}            \\ \midrule
+\multirow{8}{*}{PrPo\_Im}  & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 5                       & 7                       \\
+                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & tanh                    & relu                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
+                          & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.1                     & 0.25                    & 0.1                     \\
+                          & 2nd dropout rate, $r_2$      & $\{0, .05, \ldots, 5\}$   & 0.3                     & 0.35                    & 0.3                     \\
+                          & 1st dense layer units, $u_1$ & $\{32, 64, \ldots, 512\}$ & 480                     & 384                     & 480                     \\
+                          & 2nd dense layer units, $u_2$ & $\{32, 64, \ldots, 512\}$ & 448                     & 256                     & 448                     \\
+                          & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000102956216062} & \num{0.000109137476524} & \num{0.000102956216062} \\\midrule
+\multirow{8}{*}{Pr\_Im}    & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
+                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
+                          & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.1                     & 0.1                     & 0.1                     \\
+                          & 2nd dropout rate, $r_2$      & $\{0, .05, \ldots, 5\}$   & 0.3                     & 0.3                     & 0.3                     \\
+                          & 1st dense layer units, $u_1$ & $\{32, 64, \ldots, 512\}$ & 480                     & 480                     & 480                     \\
+                          & 2nd dense layer units, $u_2$ & $\{32, 64, \ldots, 512\}$ & 448                     & 448                     & 448                     \\
+                          & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\\midrule
+\multirow{8}{*}{Pr\_PoIm}  & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
+                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & tanh                    & relu                    & tanh                    \\
+                          & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.25                    & 0.1                     & 0.2                     \\
+                          & 2nd dropout rate, $r_2$      & $\{0, .05, \ldots, 5\}$   & 0.3                     & 0.3                     & 0.15                    \\
+                          & 1st dense layer units, $u_1$ & $\{32, 64, \ldots, 512\}$ & 128                     & 480                     & 416                     \\
+                          & 2nd dense layer units, $u_2$ & $\{32, 64, \ldots, 512\}$ & 320                     & 448                     & 416                     \\
+                          & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000175822328104} & \num{0.000102956216062} & \num{0.000114656409879} \\\midrule
+\multirow{8}{*}{Pr\_Po\_Im} & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 5                       & 7                       \\
+                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & tanh                    & relu                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & tanh                    & relu                    & relu                    \\
+                          & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.25                    & 0.25                    & 0.1                     \\
+                          & 2nd dropout rate, $r_2$      & $\{0, .05, \ldots, 5\}$   & 0.3                     & 0.35                    & 0.3                     \\
+                          & 1st dense layer units, $u_1$ & $\{32, 64, \ldots, 512\}$ & 128                     & 384                     & 480                     \\
+                          & 2nd dense layer units, $u_2$ & $\{32, 64, \ldots, 512\}$ & 320                     & 256                     & 448                     \\
+                          & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000175822328104} & \num{0.000109137476524} & \num{0.000102956216062} \\ \bottomrule
 \end{tabular}
-    \caption{Classification scenarios}
-    \label{tab:scenarios}
+% \begin{tabular}{@{}lllll@{}}
+% \toprule
+% \multirow{2}{*}{\textbf{Scenario}} & \multirow{2}{*}{\textbf{Hyperparameters}} & \multicolumn{3}{c}{\textbf{Resolution}} \\ \cmidrule(l){3-5} 
+%                             &                & 64                & 128               & 224               \\ \midrule
+% \multirow{8}{*}{PrPo\_Im}   & activation\_1  & relu              & tanh              & relu              \\
+%                             & activation\_2  & relu              & relu              & relu              \\
+%                             & dropout\_1     & 0.1               & 0.25              & 0.1               \\
+%                             & dropout\_2     & 0.3               & 0.35              & 0.3               \\
+%                             & kernel\_size   & 7                 & 5                 & 7                 \\
+%                             & learning\_rate & \num{0.000102956216062} & \num{0.000109137476524} & \num{0.000102956216062} \\
+%                             & units\_1       & 480               & 384               & 480               \\
+%                             & units\_2       & 448               & 256               & 448               \\\midrule
+% \multirow{8}{*}{Pr\_Im}     & activation\_1  & relu              & relu              & relu              \\
+%                             & activation\_2  & relu              & relu              & relu              \\
+%                             & dropout\_1     & 0.1               & 0.1               & 0.1               \\
+%                             & dropout\_2     & 0.3               & 0.3               & 0.3               \\
+%                             & kernel\_size   & 7                 & 7                 & 7                 \\
+%                             & learning\_rate & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\
+%                             & units\_1       & 480               & 480               & 480               \\
+%                             & units\_2       & 448               & 448               & 448               \\\midrule
+% \multirow{8}{*}{Pr\_PoIm}   & activation\_1  & relu              & relu              & relu              \\
+%                             & activation\_2  & tanh              & relu              & tanh              \\
+%                             & dropout\_1     & 0.25              & 0.1               & 0.2               \\
+%                             & dropout\_2     & 0.3               & 0.3               & 0.15              \\
+%                             & kernel\_size   & 7                 & 7                 & 7                 \\
+%                             & learning\_rate & \num{0.000175822328104} & \num{0.000102956216062} & \num{0.000114656409879} \\
+%                             & units\_1       & 128               & 480               & 416               \\
+%                             & units\_2       & 320               & 448               & 416               \\
+% \multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & relu              & tanh              & relu              \\\midrule
+%                             & activation\_2  & tanh              & relu              & relu              \\
+%                             & dropout\_1     & 0.25              & 0.25              & 0.1               \\
+%                             & dropout\_2     & 0.3               & 0.35              & 0.3               \\
+%                             & kernel\_size   & 7                 & 5                 & 7                 \\
+%                             & learning\_rate & \num{0.000175822328104} & \num{0.000109137476524} & \num{0.000102956216062} \\
+%                             & units\_1       & 128               & 384               & 480               \\
+%                             & units\_2       & 320               & 256               & 448               \\ \bottomrule
+% \end{tabular}
+% Table edited at: https://www.tablesgenerator.com/latex_tables#
+\caption{Optimal hyperparameters found using the Hyperband search algorithm for 12 classification scenario and input image resolution combinations.}
+    \label{tab:opt-hyp}
 \end{table}
@@
 \subsection{Model training and assessment}
+The softmax activation function in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$:
+\begin{equation}
+    f(s_c) = \frac{e^{s^c}}{\sum\limits_{c' \in C} e^{s_{c'}}}
+\end{equation}
+Thus for the $i$th observation, the softmax activation return the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
 The CNN is trained using a variant of the stochastic gradient algorithm, Adam \cite{kingma2017adam}.
 The goal of the training procedure is to learn the optimal weights and bias terms for the CNN by minimizing a loss function. 
-In this case, we use the categorical cross-entropy loss function, which is given by:
+In this case, we use the categorical cross-entropy loss function, which for a single observation can be simply defined as:
 \begin{equation}
-    L_{CE} = - \sum_{c\in C} y \log( f(c)_i)
+    L_{i}^{CE} = -  \log( p_{i,c}) = - \log(f(s_c^i))
 \end{equation}
-The softmax function returns the class prediction probabilities for a given observation $i$. It is defined for each class $c$ as:
-\begin{equation}
-    f(c)_i = \frac{e^{s^c_i}}{\sum_{c \in C} e^{s^c_i}}
-\end{equation}
-where $s^c_i$ is class score for the $i$th observation.
-Training is iteratively performed, with gradient updates averaged over a batch of input images.
+Training is iteratively performed, with gradient of the loss function computed and averaged over a batch of input images.
 Here, we use a batch size of 32.
-The learning rate $\la$ of the optimization algorithm is an important hyperparameter that affects training performance.
+The learning rate $\lambda$ of the optimization algorithm is an important hyperparameter that affects training performance.
 We optimized for this in the hyperparameter search as discussed.
@@
 The $F_1$ metric is given as the harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
-These three metrics are macro-averaged for each class for assessing model performing.
+These three metrics are macro-averaged.
 Thus, each class is given equal weight, ensuring that misclassifications within the smaller classes ("Probable" and "Possible") are adequately represented in the aggregation.
@@
 
-\subsection{Hyperparameter optimization}
-We performed a guided grid search to find the optimal hyperparameter values across each of the 12 unique classification scenario and image resolution instances.
-Each search was conducted using 90 trials of unique hyperparameter combinations. The results of the 8 hyperparameters optimized in each of the 12 training instances is shown in \autoref{tab:opt-hyp}.
-\begin{table}[ht!]\small
-    \centering
-    \sisetup{round-mode=places,
-        round-precision=2,
-        scientific-notation=fixed,
-        exponent-product = \cdot,
-        table-omit-exponent,
-        fixed-exponent=-4
-        }
-\begin{tabular}{@{}lllll@{}}
-\toprule
-\multirow{2}{*}{\textbf{Scenario}} & \multirow{2}{*}{\textbf{Hyperparameters}} & \multicolumn{3}{c}{\textbf{Resolution}} \\ \cmidrule(l){3-5} 
-                            &                & 64                & 128               & 224               \\ \midrule
-\multirow{8}{*}{PrPo\_Im}   & activation\_1  & relu              & tanh              & relu              \\
-                            & activation\_2  & relu              & relu              & relu              \\
-                            & dropout\_1     & 0.1               & 0.25              & 0.1               \\
-                            & dropout\_2     & 0.3               & 0.35              & 0.3               \\
-                            & kernel\_size   & 7                 & 5                 & 7                 \\
-                            & learning\_rate & \num{0.000102956216062} & \num{0.000109137476524} & \num{0.000102956216062} \\
-                            & units\_1       & 480               & 384               & 480               \\
-                            & units\_2       & 448               & 256               & 448               \\\midrule
-\multirow{8}{*}{Pr\_Im}     & activation\_1  & relu              & relu              & relu              \\
-                            & activation\_2  & relu              & relu              & relu              \\
-                            & dropout\_1     & 0.1               & 0.1               & 0.1               \\
-                            & dropout\_2     & 0.3               & 0.3               & 0.3               \\
-                            & kernel\_size   & 7                 & 7                 & 7                 \\
-                            & learning\_rate & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\
-                            & units\_1       & 480               & 480               & 480               \\
-                            & units\_2       & 448               & 448               & 448               \\\midrule
-\multirow{8}{*}{Pr\_PoIm}   & activation\_1  & relu              & relu              & relu              \\
-                            & activation\_2  & tanh              & relu              & tanh              \\
-                            & dropout\_1     & 0.25              & 0.1               & 0.2               \\
-                            & dropout\_2     & 0.3               & 0.3               & 0.15              \\
-                            & kernel\_size   & 7                 & 7                 & 7                 \\
-                            & learning\_rate & \num{0.000175822328104} & \num{0.000102956216062} & \num{0.000114656409879} \\
-                            & units\_1       & 128               & 480               & 416               \\
-                            & units\_2       & 320               & 448               & 416               \\
-\multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & relu              & tanh              & relu              \\\midrule
-                            & activation\_2  & tanh              & relu              & relu              \\
-                            & dropout\_1     & 0.25              & 0.25              & 0.1               \\
-                            & dropout\_2     & 0.3               & 0.35              & 0.3               \\
-                            & kernel\_size   & 7                 & 5                 & 7                 \\
-                            & learning\_rate & \num{0.000175822328104} & \num{0.000109137476524} & \num{0.000102956216062} \\
-                            & units\_1       & 128               & 384               & 480               \\
-                            & units\_2       & 320               & 256               & 448               \\ \bottomrule
-\end{tabular}
-% Table edited at: https://www.tablesgenerator.com/latex_tables#
-\caption{Optimal hyperparameters found using the Hyperband search algorithm for all classification scenario and input image resolution combinations.}
-    \label{tab:opt-hyp}
-\end{table}
+%\subsection{Hyperparameter optimization}
 

### 2021-02-25 bin/python/cnn-performance.py
added figures to man

@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=15, seed_val = 1, test_boolean=False)
+    getScenarioModelPerformance(res=64, num_epochs=15, seed_val = 1, test_boolean=False)

### 2021-02-26 bin/python/cnn-performance.py
new figures and results

@@ -31,7 +31,7 @@ import sys
 sys.path.append("../python/")
 from helpers import *
 from sklearn.metrics import classification_report, accuracy_score, f1_score, precision_score, recall_score, confusion_matrix
-from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
+#from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
 #from sklearn.preprocessing import OrdinalEncoder
 #enc = OrdinalEncoder()
 
@@ -39,7 +39,7 @@ from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
 
 # Globals
 NUM_CHANNELS = 1
-RESOLUTION_LIST = [64, 128] # 64, 128] #, 224, 384]
+RESOLUTION_LIST = [64, 128, 224] # 64, 128] #, 224, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 NUM_EPOCHS = 20
 SAVED_MODEL_DIR = '../../results/models/'
@@ -103,7 +103,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_accuracy', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_accuracy', patience=6, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -115,7 +115,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     model.compile(loss='categorical_crossentropy',  optimizer = opt, metrics =  ['accuracy'])     ## compile and fit
     hist = model.fit(training_images, training_labels, batch_size = 32, epochs = num_epochs, verbose=1, 
                      validation_data=(validation_images, validation_labels),
-                     callbacks = [model_metrics,early_stopping])     
+                     callbacks = [model_metrics, early_stopping])     
     
     # SAVE MODEL, SUMMARY AND PERFORMANCE
     if testing == True:
@@ -183,17 +183,17 @@ def visualizeCNN(model, scenario, image_size, images_per_class = 4, trial_seed =
     print(validation_labels.sum(axis=0))    
     
     # GRAD CAM
[…]

### 2021-02-26 bin/python/preprocess-images.py

@@ -143,7 +143,7 @@ def plotProcessedImages(class_scenario, image_array, class_list, images_per_clas
     image_filename = '../../figures/processed_input_images_' + str(class_scenario) + '_' + str(resolution) + '_px.png'
     #plt.xticks([0,3024])
     #plt.yticks([0,4032])
-    plt.tight_layout()
+    #plt.tight_layout()
     fig.savefig(image_filename, dpi=180)
     return
 

### 2021-02-26 docs/manuscript/manuscript.tex

@@
   \centering
-  \begin{subfigure}[t]{.5\linewidth}
+  \begin{subfigure}[t]{.75\linewidth}
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_Pr_Po_Im_128_px}
+    \includegraphics[width=.8\linewidth, trim={0 3cm 0 3cm}, clip]{processed_input_images_Pr_Po_Im_128_px}
     \caption{Scenario \texttt{Pr\_Po\_Im}}
@@
   \end{subfigure}%
+
   \begin{subfigure}[t]{.5\linewidth}
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_128_px}
+    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_Pr_Im_128_px}
     \caption{Scenario \texttt{Pr\_Im}}
@@
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_128_px}
+    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_PrPo_Im_128_px}
     \caption{Scenario \texttt{PrPo\_Im}}
@@
     \centering
-    \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_128_px}
+    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_Pr_PoIm_128_px}
     \caption{Scenario \texttt{Pr\_PoIm}}
@@
 
-  \caption{Random examples of the processed training images under each classification scenario; 64 pixels.}
+  \caption{Random examples of the processed training images under each classification scenario; 128 pixels.}
   \label{fig:processed_images}
@@
      &  & \bf Activation \\\midrule
-    Convolutional  & 1         & 64                & $k^*$   & 2           &&& ReLu \\
+    Convolutional  & 1         & 64                & $k^*$   & 2           &&& ReLU \\
     Max. Pooling   & 1         &                   & 2               &&&             &      \\
-    Convolutional  & 2         & 128               & 3   & 1           &&& ReLu \\
-    Convolutional  & 3         & 128               & 3   & 1           &&& ReLu \\
+    Convolutional  & 2         & 128               & 3   & 1           &&& ReLU \\
+    Convolutional  & 3         & 128               & 3   & 1           &&& ReLU \\
     Max. Pooling   & 2         &                   & 2               &&&             &      \\    
-    Convolutional  & 4         & 256               & 3   & 1           &&& ReLu \\
-    Convolutional  & 5         & 256               & 3   & 1           &&& ReLu \\
+    Convolutional  & 4         & 256               & 3   & 1           &&& ReLU \\
+    Convolutional  & 5         & 256               & 3   & 1           &&& ReLU \\
     Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
@@
 Twelve searches were performed for each classification scenario and image resolution combination.
-Each search was conducted using 90 trials of unique hyperparameter combinations. The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}.
+Each search was conducted using 90 trials of unique hyperparameter combinations.
+The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}.
+For the kernel size in the first convolutional layer, we allowed for a choice between a $5\times 5$ and a $7\times 7$ kernel.
+The activation function in both dense layers was specified as a choice between the rectified linear unit (ReLU) function and the hyperbolic tangent ($\tanh$).
+The ReLU was introduced to address the so-called "vanishing gradient" problem and has been shown to improve performance in CNNs \cite{glorot2011deep}.
+Nevertheless, the $\tanh$ function remains a viable option, as well.
+The dropout rates were allowed to range from 0 to 5 in steps of 0.05, while the number of neurons or units in each dense layer varied from 32 to 512 in steps of 32.
+Finally, we uniformly sampled learning rates $\la$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
+
 \begin{table}[ht!]\small
@@
 % \usepackage{multirow}
-\begin{tabular}{@{}lllrrr@{}}
+\begin{tabular}{@{}lllccc@{}}
 \toprule
@@
 \multirow{8}{*}{PrPo\_Im}  & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 5                       & 7                       \\
-                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & tanh                    & relu                    \\
-                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
+                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & tanh                    & ReLU                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
                           & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.1                     & 0.25                    & 0.1                     \\
@@
 \multirow{8}{*}{Pr\_Im}    & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
-                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
-                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
+                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
                           & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.1                     & 0.1                     & 0.1                     \\
@@
 \multirow{8}{*}{Pr\_PoIm}  & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
-                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
-                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & tanh                    & relu                    & tanh                    \\
+                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & tanh                    & ReLU                    & tanh                    \\
                           & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.25                    & 0.1                     & 0.2                     \\
@@
 \multirow{8}{*}{Pr\_Po\_Im} & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 5                       & 7                       \\
-                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & tanh                    & relu                    \\
-                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & tanh                    & relu                    & relu                    \\
+                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & tanh                    & ReLU                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & tanh                    & ReLU                    & ReLU                    \\
                           & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.25                    & 0.25                    & 0.1                     \\
@@
 %                             &                & 64                & 128               & 224               \\ \midrule
-% \multirow{8}{*}{PrPo\_Im}   & activation\_1  & relu              & tanh              & relu              \\
-%                             & activation\_2  & relu              & relu              & relu              \\
+% \multirow{8}{*}{PrPo\_Im}   & activation\_1  & ReLU              & tanh              & ReLU              \\
+%                             & activation\_2  & ReLU              & ReLU              & ReLU              \\
 %                             & dropout\_1     & 0.1               & 0.25              & 0.1               \\
@@
 %                             & units\_2       & 448               & 256               & 448               \\\midrule
-% \multirow{8}{*}{Pr\_Im}     & activation\_1  & relu              & relu              & relu              \\
-%                             & activation\_2  & relu              & relu              & relu              \\
+% \multirow{8}{*}{Pr\_Im}     & activation\_1  & ReLU              & ReLU              & ReLU              \\
+%                             & activation\_2  & ReLU              & ReLU              & ReLU              \\
 %                             & dropout\_1     & 0.1               & 0.1               & 0.1               \\
@@
 %                             & units\_2       & 448               & 448               & 448               \\\midrule
-% \multirow{8}{*}{Pr\_PoIm}   & activation\_1  & relu              & relu              & relu              \\
-%                             & activation\_2  & tanh              & relu              & tanh              \\
+% \multirow{8}{*}{Pr\_PoIm}   & activation\_1  & ReLU              & ReLU              & ReLU              \\
+%                             & activation\_2  & tanh              & ReLU              & tanh              \\
 %                             & dropout\_1     & 0.25              & 0.1               & 0.2               \\
@@
 %                             & units\_2       & 320               & 448               & 416               \\
-% \multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & relu              & tanh              & relu              \\\midrule
-%                             & activation\_2  & tanh              & relu              & relu              \\
+% \multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & ReLU              & tanh              & ReLU              \\\midrule
+%                             & activation\_2  & tanh              & ReLU              & ReLU              \\
 %                             & dropout\_1     & 0.25              & 0.25              & 0.1               \\
@@
 
-\subsection{Model visualization and inference}
-Model explainability is increasingly critical in the application of CNNs to an ever-expanding set of problems.
+% \subsection{Model visualization and inference}
+% Model explainability is increasingly critical in the application of CNNs to an ever-expanding set of problems.
 
-\begin{figure}[ht]
-    \centering
+% \begin{figure}[ht]
+%     \centering
 
-    \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-gradcam-Pr_Im-128-px-4-images.png}
-    \caption{Scenario \texttt{Pr\_Im}}
-    \label{prpo_im_cm}
-    \end{subfigure}%   
-    \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-gradcam-PrPo_Im-128-px-4-images.png}
-    \caption{Scenario \texttt{PrPo\_Im}}
-    \label{prpo_im_cm}
-    \end{subfigure}%
+%     \begin{subfigure}[t]{.45\linewidth}
+%     \centering
+%     \includegraphics[width=\linewidth]{opt-gradcam-Pr_Im-128-px-4-images.png}
+%     \caption{Scenario \texttt{Pr\_Im}}
+%     \label{prpo_im_cm}
+%     \end{subfigure}%   
+%     \begin{subfigure}[t]{.45\linewidth}
+%     \centering
+%     \includegraphics[width=\linewidth]{opt-gradcam-PrPo_Im-128-px-4-images.png}
+%     \caption{Scenario \texttt{PrPo\_Im}}
+%     \label{prpo_im_cm}
+%     \end{subfigure}%
 
-    \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-gradcam-Pr_PoIm-128-px-4-images.png}
-    \caption{Scenario \texttt{Pr\_PoIm}}
-    \label{prpo_im_cm}
-    \end{subfigure}%
-    \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-gradcam-Pr_Po_Im-128-px-4-images.png}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
-    \label{prpo_im_cm}
-    \end{subfigure}%
+%     \begin{subfigure}[t]{.45\linewidth}
+%     \centering
+%     \includegraphics[width=\linewidth]{opt-gradcam-Pr_PoIm-128-px-4-images.png}
+%     \caption{Scenario \texttt{Pr\_PoIm}}
+%     \label{prpo_im_cm}
+%     \end{subfigure}%
+%     \begin{subfigure}[t]{.45\linewidth}
+%     \centering
+%     \includegraphics[width=\linewidth]{opt-gradcam-Pr_Po_Im-128-px-4-images.png}
+%     \caption{Scenario \texttt{Pr\_Po\_Im}}
+%     \label{prpo_im_cm}
+%     \end{subfigure}%
 
-    \caption{Gradient-weighted class activation maps (Grad-CAMs) for randomly selected validation images in each class. (Resolution: 128px)}
-    \label{fig:my_label}
-\end{figure}
+%     \caption{Gradient-weighted class activation maps (Grad-CAMs) for randomly selected validation images in each class. (Resolution: 128px)}
+%     \label{fig:my_label}
+% \end{figure}
 
@@
 \section{Conclusion}
+We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood (Probable, Possible and Improbable) with respect to utility infrastructure.
+Specifically, we developed a convolutional neural network with state-of-the-art configurations.
+We applied data augmentation and pre-processing strategies to increase the size of our dataset by a factor of 5, thus generating 2525 images.
+From an initial resolution of $4032 \times 3024$ pixels, we created three sets of image resolutions: $64 \times 64$, $128 \times 128$ and $224 \times 224$.
+We then optimized eight of our CNN hyperparameters for 12  classification scenario and image resolution combinations. 
+We conducted ten model training trials in each of the 12 cases, using a 20\% validation set to measure model performance.
+In these trials, the average accuracy of classification was $\geq 0.94$, with a maximum standard deviation of 0.7.
+There was no statistically significant difference between the performance across image resolutions.
+
+We further conducted more detailed classification performance analyses over single training instances for the 128-pixel case.
+Our results indicate that the CNN performed best at recalling Probable vs. Improbable cases.
+The Possible category appeared to be a confounding case.
+This was not unexpected, given the uncertainty and subjectivity in assessing these trees to begin with.
+The CNN, however, performed better at distinguishing between Possible/Improbable and Probable, than between Probable/Possible and Improbable.
+This might be an indicator that the Possible cases are more likely identified as Improbable.
+When all three failure-likelihood categories were predicted individually, the Possible case had the lowest recall score.
+
+Nevertheless, given the relatively small input dataset of original images, these preliminary results are extremely promising for future improvements.
+First, we can train better models with more data.
+We also plan to rigorously quantify the uncertainty in ground truth category assignments by incorporating predictions from multiple experts on the same images.
+In order to better understand how the CNN is classifying each image, we will conduct extensive visual inference in further work, for instance, using the gradient-weighted class activation mapping approach \cite{zeiler2014visualizing}.
 

### 2021-02-26 bin/python/cnn-performance.py
new figures and results

@@ -103,7 +103,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_accuracy', patience=6, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_accuracy', patience=5, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=13, seed_val = 2, test_boolean=False)
+    getScenarioModelPerformance(res=128, num_epochs=20, seed_val = 111, test_boolean=False)

### 2021-02-26 docs/manuscript/manuscript.tex
manuscript updates

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\NameTag{Author1LastName, \today}
+\NameTag{Oke, \today}
 
@@
 % ALL CAPS TITLE REQUIRED
-\title{Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network}
-\author[1]{Author 1}
-\author[1]{Author 2}
-\author[2]{Author 3}
-\author[2]{Author 4}
-\author[2]{Author 5}
+\title{Predicting tree failure likelihood for utility risk mitigation via artificial intelligence}
+\author[1]{Jimi Oke}
+\author[1]{Nasko Apostolov}
+\author[2]{Ryan Suttle}
+\author[1]{Sanjay Arwade}
+\author[2]{Brian Kane}
 
@@
 \begin{abstract}
-    Abstract here.
+  Critical to the resilience of utility power lines, tree failure assessments have historically been performed via
+  manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
+  likelihood categories under four classification scenarios.  Starting with an original set of expert-labeled 525 images, we perform
+  preprocessing and augmentation tasks to increase the number of samples. We optimize several hyperparameters in our CNN
+  and test its performance for three different image resolutions. The trained classifier has an average validation
+  accuracy of at least 0.94 across all scenario-resolution combinations.  Thus, via this novel framework, we
+  demonstrate the potential of artificial intelligence to automate and consequently reduce the costs of tree failure
+  likelihood assessments, thereby promoting sustainable infrastructure.
 \end{abstract}
@@
 Neural networks have also been used to interrogate remote sensing data of the landscape to assess landslide risk \cite{su2020deep}.
-A few recent efforts have demonstrated the potential for AI-based tree recognition from drone imagery \cite{santos2019assessment, egli2020cnnbased}.
+A few recent efforts have demonstrated the potential for AI-based tree recognition from drone imagery \cite{santos2019assessment,egli2020cnnbased}.
 Furthermore, an application of a convolutional neural network (CNN) to tree species identification using was recently demonstrated by \citeN{fricker2019convolutional}.
@@
 
-\subsection{Analysis of classification strategies}
-We investigated in the detail the performance of the four classification scenarios earlier outlined, using the 128-pixel resolution images.
-The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
-Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
-The metrics are shown in \autoref{fig:cnn-perf}.
-In each scenario, the model was trained over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
-That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
-From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three.
-This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
-Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=1.05\textwidth]{cnn-performance-metrics-128px-plot}
-    \caption{Model performance across four classification scenarios for a single training instance in each case.
-Training resolution was 128px.
-All weighted average metrics were computed on the validation set.}
-    \label{fig:cnn-perf}
-\end{figure}
-
-We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
-Each row of the confusion mat
-\begin{figure}[ht]
-  \centering
-  \begin{subfigure}[t]{.25\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
-    \caption{Scenario \texttt{Pr\_Im}}
-    \label{pr_im_cm}
-  \end{subfigure}%
-   \begin{subfigure}[t]{.25\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
-    \caption{Scenario \texttt{PrPo\_Im}}
-    \label{prpo_im_cm}
-  \end{subfigure}%
-  \begin{subfigure}[t]{.25\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
-    \caption{Scenario \texttt{Pr\_PoIm}}
-    \label{pr_poim_cm}
-\end{subfigure}%
-  \begin{subfigure}[t]{.25\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
-    \label{pr_po_im_cm}
-  \end{subfigure}%
-    \caption{Confusion matrices for CNN performance using the 128px images across all scenarios.
-    Each row of a confusion matrix indicates the proportional distribution of class predictions for the true members of each class.
-Thus, the diagonals indicate the recall for each class $Re_c$.
-The average of the diagonal values gives the macro-average $Re^m$ for each scenario.}
-    \label{fig:confusion-matrix}
-\end{figure}
-
 \subsection{Sensitivity to training resolution}
@@
 
+\subsection{Analysis of classification strategies}
+We investigated in detail the performance of the four classification scenarios earlier outlined, using the 128-pixel resolution images.
+The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
+Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
+The metrics are shown in \autoref{fig:cnn-perf}.
+In each scenario, the model was trained over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
+That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
+From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three.
+This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
+Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=1.05\textwidth]{cnn-performance-metrics-128px-plot}
+    \caption{Model performance across four classification scenarios for a single training instance in each case.
+Training resolution was 128px.
+All weighted average metrics were computed on the validation set.}
+    \label{fig:cnn-perf}
+\end{figure}
+
+We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
+Each row of the confusion matrix indicates the proportion of observation in a given class that are predicted to be in the classes across the columns.
+Thus, the diagonal entry in each matrix represents the class-specific recall score, $Re_{c}$.
+The matrices show that generally, the ``Possible'' category is difficult to predict accurately as an individual class.
+The classifier performs best when it only has to distinguish between ``Probable'' and ``Improbable'' (\autoref{pr_im_cm}).
+Performance only suffers slightly when ``Possible'' is grouped with ``Probable'' in the PrPo\_Im scenario (\autoref{prpo_im_cm}).
+
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Im}}
+    \label{pr_im_cm}
+  \end{subfigure}%
+   \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
+    \caption{Scenario \texttt{PrPo\_Im}}
+    \label{prpo_im_cm}
+  \end{subfigure}%
+  \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
+    \caption{Scenario \texttt{Pr\_PoIm}}
+    \label{pr_poim_cm}
+\end{subfigure}%
+  \begin{subfigure}[t]{.25\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
+    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \label{pr_po_im_cm}
+  \end{subfigure}%
+    \caption{Confusion matrices for CNN performance using the 128px images across all scenarios.
+    Each row of a confusion matrix indicates the proportional distribution of class predictions for the true members of each class.
+Thus, the diagonals indicate the recall for each class $Re_c$.
+The average of the diagonal values gives the macro-average $Re^m$ for each scenario.}
+    \label{fig:confusion-matrix}
+  \end{figure}
+
+  The detailed performance metrics for the 128-pixel case are summarized in \autoref{tab:performance}.
+
 \begin{table}[ht]\small
@@
 \end{tabular}
-    \caption{Performance metrics across the four scenarios}
+    \caption{Detailed performance metrics across the four scenarios in the 128-pixel case (single trial)}
     \label{tab:performance}
@@
 
+\eject
 \section{Conclusion}
@@
 
-Please also see the guidelines at: \url{https://ascelibrary.org/page/dataavailability}.
-
 \section{Acknowledgments}
-The authors would like to thank...
+The authors acknowledge the support of Eversource.
 

### 2021-02-26 bin/python/cnn-performance.py

@@ -40,7 +40,7 @@ from sklearn.metrics import classification_report, accuracy_score, f1_score, pre
 # Globals
 NUM_CHANNELS = 1
 RESOLUTION_LIST = [64, 128, 224] # 64, 128] #, 224, 384]
-SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
+SCENARIO_LIST = ["Pr_Po_Im"] #"Pr_Im", "PrPo_Im", "Pr_PoIm", 
 NUM_EPOCHS = 20
 SAVED_MODEL_DIR = '../../results/models/'
 MODEL_PERFORMANCE_METRICS_DIR = '../../results/model-performance/'
@@ -86,7 +86,7 @@ class Metrics(Callback):
 
 
 def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_seed = 1, testing = True): 
-    
+    random.reset_seeds()
     # IMAGES (former approach)
     training_images_and_labels, test_images_and_labels = splitData(image_sets[image_size][scenario], prop = 0.8, seed_num = trial_seed)
     training_images, training_labels = getImageAndLabelArrays(training_images_and_labels)
@@ -103,9 +103,10 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_accuracy', patience=5, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_accuracy', patience=15, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
+    del model
     K.clear_session()
     #input_shape = (image_size, image_size, NUM_CHANNELS) ## shape of images
     model = constructBaseCNN(image_size, scenario, num_channels = NUM_CHANNELS)    ## get model
@@ -295,4 +296,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=20, seed_val = 111, test_boolean=False)
+    getScenarioModelPerformance(res=128, num_epochs=100, seed_val = 1, test_boolean=True)

### 2021-02-26 bin/python/resolution-scenario-sensitivity.py

@@ -104,9 +104,3 @@ def main(num_trials = NUM_MODEL_RUNS):
 
 if __name__ == "__main__":
     main()
-
-
-# predictions = model.predict_classes(x_val)
-# predictions = predictions.reshape(1,-1)[0]
-
-# print(classification_report(y_val, predictions, target_names = ['Rugby (Class 0)','Soccer (Class 1)']))

### 2021-03-01 bin/python/cnn-performance.py
mres updates

@@ -31,7 +31,7 @@ import sys
 sys.path.append("../python/")
 from helpers import *
 from sklearn.metrics import classification_report, accuracy_score, f1_score, precision_score, recall_score, confusion_matrix
-#from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
+from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
 #from sklearn.preprocessing import OrdinalEncoder
 #enc = OrdinalEncoder()
 
@@ -86,17 +86,16 @@ class Metrics(Callback):
 
 
 def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_seed = 1, testing = True): 
-    random.reset_seeds()
     # IMAGES (former approach)
-    training_images_and_labels, test_images_and_labels = splitData(image_sets[image_size][scenario], prop = 0.8, seed_num = trial_seed)
-    training_images, training_labels = getImageAndLabelArrays(training_images_and_labels)
-    validation_images, validation_labels = getImageAndLabelArrays(test_images_and_labels)
+    # training_images_and_labels, test_images_and_labels = splitData(image_sets[image_size][scenario], prop = 0.8, seed_num = trial_seed)
+    # training_images, training_labels = getImageAndLabelArrays(training_images_and_labels)
+    # validation_images, validation_labels = getImageAndLabelArrays(test_images_and_labels)
     class_labels = getClassLabels(scenario)
     print("Class labels:", class_labels)
-    # training_images, validation_images, training_labels, validation_labels =  train_test_split(np.array([np.expand_dims(x[0],axis=2) for x in image_sets[image_size][scenario]]), 
-    #                                                                                            np.array([x[1] for x in image_sets[image_size][scenario]]), 
-    #                                                                                            stratify= np.array([x[1] for x in image_sets[image_size][scenario]]), 
-    #                                                                                            test_size = .2, random_state = trial_seed)
+    training_images, validation_images, training_labels, validation_labels =  train_test_split(np.array([np.expand_dims(x[0],axis=2) for x in image_sets[image_size][scenario]]), 
+                                                                                               np.array([x[1] for x in image_sets[image_size][scenario]]), 
+                                                                                               stratify= np.array([x[1] for x in image_sets[image_size][scenario]]), 
[…]

### 2021-03-01 bin/python/resolution-scenario-sensitivity.py

@@ -22,80 +22,118 @@ import sys
 sys.path.append("../python/")
 from helpers import *
 
+from sklearn.model_selection import StratifiedKFold
+from sklearn.metrics import accuracy_score, f1_score, precision_score, recall_score
+
 
 # Globals
 NUM_CHANNELS = 1
-RESOLUTION_LIST = [224] # 64, 128] #, 224, 384]
+RESOLUTION_LIST = [64, 128, 224] #, 224, 384]
 SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
-NUM_MODEL_RUNS = 10
-NUM_EPOCHS = 10
-RESOLUTION_PERFORMANCE_METRICS_DIR = '../../results/sensitivity-tests'
+NUM_MODEL_RUNS = 5
+NUM_EPOCHS = 15
+RESOLUTION_PERFORMANCE_METRICS_DIR = '../../results/sensitivity-tests-02272021' #'../../results/sensitivity-tests'
 
 
-# def constructBaseCNN(image_size, scenario):
-#     image_shape = (image_size, image_size, NUM_CHANNELS)
-#     p_dict = getOptCNNHyperparams(image_size, scenario)
-#     if scenario=="Pr_Po_Im":
-#         num_classes = 3
-#     else:
-#         num_classes = 2
-#     base_model = models.Sequential([
-#         layers.Conv2D(filters = 64, kernel_size = p_dict['kernel_size'], strides = 2, activation="relu", padding="same", input_shape = image_shape),
-#         layers.MaxPooling2D(2),
-#         layers.Conv2D(128, 3, activation="relu", padding="same"),
-#         layers.Conv2D(128, 3, activation="relu", padding="same"),
-#         layers.MaxPooling2D(2),
-#         layers.Conv2D(256, 3, activation="relu", padding="same"),
-#         layers.Conv2D(256, 3, activation="relu", padding="same"),
-#         layers.MaxPooling2D(2),
-#         layers.Flatten(),
-        
-#         layers.Dense(p_dict['units_1'], activation = p_dict['activation_1']),
-#         layers.BatchNormalization(),
-#         layers.Dropout(p_dict['dropout_1']), 
-        
-#         layers.Dense(p_dict['units_2'], activation = p_dict['activation_2']), 
-#         layers.Dropout(p_dict['dropout_2']),
-        
-#         layers.Dense(num_classes, activation="softmax")
-#     ])
[…]

### 2021-03-02 bin/python/cnn-performance.py
new results

@@ -40,7 +40,7 @@ from sklearn.model_selection import train_test_split, StratifiedShuffleSplit
 # Globals
 NUM_CHANNELS = 1
 RESOLUTION_LIST = [64, 128, 224] # 64, 128] #, 224, 384]
-SCENARIO_LIST = ["Pr_Po_Im"] #"Pr_Im", "PrPo_Im", "Pr_PoIm", 
+SCENARIO_LIST = ["Pr_Im", "PrPo_Im", "Pr_PoIm", "Pr_Po_Im"]
 NUM_EPOCHS = 20
 SAVED_MODEL_DIR = '../../results/models/'
 MODEL_PERFORMANCE_METRICS_DIR = '../../results/model-performance/'

### 2021-03-02 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
   manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
-  likelihood categories under four classification scenarios.  Starting with an original set of expert-labeled 525 images, we perform
+  likelihood categories under four classification scenarios.  Starting with an original set of 525 expert-labeled images, we perform
   preprocessing and augmentation tasks to increase the number of samples. We optimize several hyperparameters in our CNN
@@
 
-The remainder of the paper provides a brief history and background of AI and its use in earthquake risk assessment and tree identification (section 2); describes the methods used to train and validate a novel CNN to categorize likelihood of tree failure (section 3); and presents and discusses the output of the novel CNN (sections 4 and 5).
+The remainder of the paper provides a brief history and background of AI and its use in infrastructure risk assessment and tree identification (section 2); describes the methods used to train and validate a novel CNN to categorize likelihood of tree failure (section 3); and presents and discusses the output of the novel CNN (sections 4 and 5).
 The goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.
@@
 \begin{itemize}
-\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
-\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
-\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
+\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions;
+\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions;
+\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame;
 \item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load.
-This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect people from harm
+This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect targets from impact.
 \end{itemize}
@@
 The dropout rates were allowed to range from 0 to 5 in steps of 0.05, while the number of neurons or units in each dense layer varied from 32 to 512 in steps of 32.
-Finally, we uniformly sampled learning rates $\la$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
+Finally, we uniformly sampled learning rates $\lambda$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
 
@@
 \begin{equation}
-    f(s_c) = \frac{e^{s^c}}{\sum\limits_{c' \in C} e^{s_{c'}}}
+    f(s_c) = \frac{e^{s_c}}{\sum\limits_{c' \in C} e^{s_{c'}}}
 \end{equation}
@@
 Thus, we define the accuracy as the overall proportion of correct predictions across all classes.
-This metric was computed both for the training and validation sets in each epoch
+This metric was computed both for the training and validation sets in each epoch.
 In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the macro-averages of the precision, recall and $F_1$ score metrics computed over the validation set in each epoch.
@@
 \section{Results}
-
-%\subsection{Hyperparameter optimization}
-
 \subsection{Sensitivity to training resolution}
 We trained the CNN for each of the 4 classification scenarios.
-In each scenarios, we also trained on three image resolution sets (64, 108 and 224).
+In each scenario, we also trained on three image resolution sets (64, 108 and 224).
 The goal was to determine the best performing resolution, given the tradeoff between performance and computational expenditure.
@@
 \end{figure}
-\begin{table}[h!]\small
-    \centering
-    \begin{tabular}{ l c r r}\toprule
-\bf Scenario & \bf Resolution & \multicolumn{2}{c}{\bf Validation Accuracy}\\
-& & Mean & SD \\\midrule
-    \texttt{PrPo\_Im} & 64  & 0.96 & 0.04 \\
-                      & 128 & 0.96 & 0.05 \\
-                      &224  &      0.96&  0.05  \\ \midrule        
-    \texttt{Pr\_Im}   & 64  & 0.98 & 0.03 \\
-                      & 128 & 0.97 & 0.03 \\
-                      & 224 & 0.97 & 0.03\\\midrule       
-    \texttt{Pr\_PoIm} & 64  & 0.96 & 0.04 \\
-                      & 128 & 0.97 & 0.04 \\
-                      & 224 & 0.97 & 0.04 \\\midrule       
-    \texttt{Pr\_Po\_Im}& 64  & 0.94 & 0.07 \\
-                      & 128 & 0.95 & 0.07 \\
-                      & 224 & 0.95 & 0.06 \\\bottomrule
-    \end{tabular}
-    \caption{Summary validation accuracy (mean and standard deviation to 2 decimal places) for each scenario and input image resolution training instance.
+% \begin{table}[h!]\small
+%     \centering
+%     \begin{tabular}{ l c r r}\toprule
+% \bf Scenario & \bf Resolution & \multicolumn{2}{c}{\bf Validation Accuracy}\\
+% & & Mean & SD \\\midrule
+%     \texttt{PrPo\_Im} & 64  & 0.96 & 0.04 \\
+%                       & 128 & 0.96 & 0.05 \\
+%                       &224  &      0.96&  0.05  \\ \midrule        
+%     \texttt{Pr\_Im}   & 64  & 0.98 & 0.03 \\
+%                       & 128 & 0.97 & 0.03 \\
+%                       & 224 & 0.97 & 0.03\\\midrule       
+%     \texttt{Pr\_PoIm} & 64  & 0.96 & 0.04 \\
+%                       & 128 & 0.97 & 0.04 \\
+%                       & 224 & 0.97 & 0.04 \\\midrule       
+%     \texttt{Pr\_Po\_Im}& 64  & 0.94 & 0.07 \\
+%                       & 128 & 0.95 & 0.07 \\
+%                       & 224 & 0.95 & 0.06 \\\bottomrule
+%     \end{tabular}
+%     \caption{Summary validation accuracy (mean and standard deviation to 2 decimal places) for each scenario and input image resolution training instance.
 There is no statistically significant difference between the performance for each resolution case under each scenario.}
-    \label{tab:meanvalacc}
-\end{table}
+%     \label{tab:meanvalacc}
+% \end{table}
 
@@
 \subsection{Analysis of classification strategies}
-We investigated in detail the performance of the four classification scenarios earlier outlined, using the 128-pixel resolution images.
+We investigated in detail the performance of the four classification scenarios outlined earlier, using the 128-pixel resolution images.
 The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
@@
 From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three.
-This indicates the uncertainty surrounding the subjective risk classification of the trees to begin.
+This indicates the uncertainty surrounding the subjective risk classification of the trees to begin with.
 Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
@@
 
-We also plot the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
-Each row of the confusion matrix indicates the proportion of observation in a given class that are predicted to be in the classes across the columns.
+We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
+Each row of the confusion matrix indicates the proportion of observations in a given class that are predicted to be in the classes across the columns.
 Thus, the diagonal entry in each matrix represents the class-specific recall score, $Re_{c}$.
 The matrices show that generally, the ``Possible'' category is difficult to predict accurately as an individual class.
-The classifier performs best when it only has to distinguish between ``Probable'' and ``Improbable'' (\autoref{pr_im_cm}).
-Performance only suffers slightly when ``Possible'' is grouped with ``Probable'' in the PrPo\_Im scenario (\autoref{prpo_im_cm}).
+The classifier performed best when it only had to distinguish between ``Probable'' and ``Improbable'' (\autoref{pr_im_cm}).
+Performance only suffered slightly when ``Possible'' was grouped with ``Probable'' in the PrPo\_Im scenario (\autoref{prpo_im_cm}).
 

### 2021-03-02 bin/python/cnn-performance.py
new results

@@ -152,9 +152,9 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     con_mat_df = pd.DataFrame(con_mat_norm, index = class_labels, columns = class_labels)
     #print("Confusion matrix for scenario " + scenario + ", resolution: " + str(image_size) + ":")
     #print(con_mat_df)
-    figure = plt.figure(figsize=(4, 4))    ## Confusion matrix heatmap
+    figure = plt.figure()#figsize=(4, 4))    ## Confusion matrix heatmap
     ax = sns.heatmap(con_mat_df, annot=True, cmap=plt.cm.Blues, fmt='g', cbar = False, annot_kws={"size": 16})
-    figure.tight_layout()
+    #figure.tight_layout()
     plt.ylabel('True',fontsize=16)
     ax.set_yticklabels(class_labels,va='center',fontsize=14)
     ax.set_xticklabels(class_labels, ha='center',fontsize=14)
@@ -164,7 +164,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
         con_mat_heatmap_file = "../../figures/test-opt-confusion-matrix-" + scenario + "-" + str(image_size) + "-px.png"
     else:
         con_mat_heatmap_file = "../../figures/opt-confusion-matrix-" + scenario + "-" + str(image_size) + "-px.png"
-    figure.savefig(con_mat_heatmap_file, dpi=180, bbox_inches='tight')
+    figure.savefig(con_mat_heatmap_file, dpi=180)#, bbox_inches='tight')
     return(model, hist) 
 
 
@@ -281,11 +281,12 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     df = pd.DataFrame()
     for s in SCENARIO_LIST:
         m, h = trainModelWithDetailedMetrics(res, s, num_epochs, trial_seed = seed_val, testing = test_boolean)
-        visualizeCNN(m, s, res, images_per_class = 4, trial_seed = seed_val, testing = test_boolean)       
+        #visualizeCNN(m, s, res, images_per_class = 4, trial_seed = seed_val, testing = test_boolean)       
         perf = pd.DataFrame.from_dict(h.history)
         perf[['Scenario']] = s
         perf['epoch'] = perf.index + 1
         df = df.append(perf, ignore_index=True)
+        #del m
     if test_boolean == True:
[…]

### 2021-03-02 docs/manuscript/manuscript.tex
manuscript updates

@@
   \centering
-  \begin{subfigure}[t]{.25\linewidth}
+  \begin{subfigure}[t]{.45\linewidth}
     \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Im-128-px.png}
+    \includegraphics[width=\linewidth, trim={0 0 1cm 1cm}, clip]{opt-confusion-matrix-Pr_Im-128-px.png}
     \caption{Scenario \texttt{Pr\_Im}}
@@
   \end{subfigure}%
-   \begin{subfigure}[t]{.25\linewidth}
+   \begin{subfigure}[t]{.45\linewidth}
     \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-PrPo_Im-128-px.png}
+    \includegraphics[width=\linewidth, trim={0 0 1cm 1cm}, clip]{opt-confusion-matrix-PrPo_Im-128-px.png}
     \caption{Scenario \texttt{PrPo\_Im}}
@@
   \end{subfigure}%
-  \begin{subfigure}[t]{.25\linewidth}
+
+  \begin{subfigure}[t]{.45\linewidth}
     \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_PoIm-128-px.png}
+    \includegraphics[width=\linewidth, trim={0 0 1cm 1cm}, clip]{opt-confusion-matrix-Pr_PoIm-128-px.png}
     \caption{Scenario \texttt{Pr\_PoIm}}
@@
 \end{subfigure}%
-  \begin{subfigure}[t]{.25\linewidth}
+  \begin{subfigure}[t]{.45\linewidth}
     \centering
-    \includegraphics[width=\linewidth]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
+    \includegraphics[width=\linewidth, trim={0 0 1cm 1cm}, clip]{opt-confusion-matrix-Pr_Po_Im-128-px.png}
     \caption{Scenario \texttt{Pr\_Po\_Im}}

### 2021-03-02 bin/python/cnn-performance.py

@@ -102,7 +102,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     
     # CALLBACKS
     model_metrics = Metrics(val_data=(validation_images, validation_labels))
-    early_stopping = EarlyStopping(monitor='val_accuracy', patience=15, restore_best_weights=True)
+    early_stopping = EarlyStopping(monitor='val_accuracy', patience=3, restore_best_weights=True)
     
     # INIT MODEL AND PARAMS, FIT
     K.clear_session()
@@ -114,7 +114,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     model.compile(loss='categorical_crossentropy',  optimizer = opt, metrics =  ['accuracy'])     ## compile and fit
     hist = model.fit(training_images, training_labels, batch_size = 32, epochs = num_epochs, verbose=1, 
                      validation_data=(validation_images, validation_labels),
-                     callbacks = [model_metrics])#, early_stopping])     
+                     callbacks = [model_metrics, early_stopping])     
     
     # SAVE MODEL, SUMMARY AND PERFORMANCE
     if testing == True:
@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=12, seed_val = 1, test_boolean=False)
+    getScenarioModelPerformance(res=128, num_epochs=13, seed_val = 1, test_boolean=False)

### 2021-03-03 bin/python/cnn-performance.py
latest results

@@ -114,7 +114,7 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     model.compile(loss='categorical_crossentropy',  optimizer = opt, metrics =  ['accuracy'])     ## compile and fit
     hist = model.fit(training_images, training_labels, batch_size = 32, epochs = num_epochs, verbose=1, 
                      validation_data=(validation_images, validation_labels),
-                     callbacks = [model_metrics, early_stopping])     
+                     callbacks = [model_metrics]) #, early_stopping])     
     
     # SAVE MODEL, SUMMARY AND PERFORMANCE
     if testing == True:
@@ -295,4 +295,4 @@ def getScenarioModelPerformance(res = 64, num_epochs = 15, seed_val = 1, test_bo
     return df
 
 if __name__ == "__main__":
-    getScenarioModelPerformance(res=128, num_epochs=13, seed_val = 1, test_boolean=False)
+    getScenarioModelPerformance(res=128, num_epochs=13, seed_val = 2, test_boolean=False)

### 2021-03-03 docs/manuscript/manuscript.tex
Updates from Overleaf

--- 2021-03-02 docs/manuscript/manuscript.tex
@@
 
-%\subsection{AI and risk assessment}
 AI-based image analysis is relatively widely used, even in engineering applications, such as earthquake risk assessment \cite{jiao2020artificial,salehi2018emerging} and structural health monitoring \cite{spencer2019advances,wang2019novel}.
@@
 
-%Neural networks developed over the past several decades from the introduction of the single-layer perceptron.
 The groundbreaking study of \citeN{hubel1959receptive} showed that visual perception in cats was a result of the
@@
 ESRI ArcMaps was used to randomly distribute sampling sites across the state.
-Likelihood of failure assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{smiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{goodfellow2020best}.
+Field assessments of trees to classify likelihood of failure followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{smiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{goodfellow2020best}.
 This method is commonly used to assess trees in the United States.
-A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
+A Level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e.
 Level 2 or Level 3 \cite{smiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility right-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{goodfellow2020best}; and (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.
@@
 \end{itemize}
-In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category.
+In this study, only images of trees assigned to the 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were included in modeling, while images of \textit{imminent} trees were excluded due to their rarity.
 Randomly chosen examples are shown for each category in \autoref{fig:raw_images}.
@@
 \subsection{Model training and assessment}
-The softmax activation function in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$:
+The softmax activation function  $f(\cdot)$ in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$:
 \begin{equation}
@@
 \end{equation}
-Thus for the $i$th observation, the softmax activation return the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
+Thus for the $i$th observation, the softmax activation returns the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
 The CNN is trained using a variant of the stochastic gradient algorithm, Adam \cite{kingma2017adam}.
@@
 
-In real terms, we measure the performance of the trained CNN by how accurately it predicts the classes in a validation set excluded from the training set.
-For this paper, we use a randomly sampled validation that was 20\% of the size of the input dataset of 2525 images in each training instance.
+In real terms, we measured the performance of the trained CNN by how accurately it predicts the classes in a validation set excluded from the training set.
+For this paper, we used a randomly sampled validation that was 20\% of the size of the input dataset of 2525 images in each training instance.
 Thus, we define the accuracy as the overall proportion of correct predictions across all classes.
@@
 In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the macro-averages of the precision, recall and $F_1$ score metrics computed over the validation set in each epoch.
-The precision score $Pr$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is a important measure of how good a classifier is.
-Meanwhile, the recall $Re$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
+The precision score $Pr$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is an important measure of how good a classifier is.
+The recall $Re$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
 The $F_1$ metric is given as the harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
 These three metrics are macro-averaged.
-Thus, each class is given equal weight, ensuring that misclassifications within the smaller classes ("Probable" and "Possible") are adequately represented in the aggregation.
+Thus, each category is given equal weight, ensuring that misclassifications within the smaller classes ("Probable" and "Possible") are adequately represented in the aggregation.
 These metrics are formally defined as follows:
@@
 The goal was to determine the best performing resolution, given the tradeoff between performance and computational expenditure.
-Thus, twelve learning instances were generated, using a training-validation ratio of 80:20.
-In each instance, we performed 10 trials, each consisting of 10 epochs.
-Validation sets were randomly selected in each trial, but consistent across each of the 12 training instances.
+Thus, we generated twelve learning cases.
+In each case, we performed a five-fold cross-validation (resulting in a training-validation ratio of 80:20), each consisting of 10 epochs.
 The trajectories of the training/validation loss and accuracy are shown in \autoref{fig:sens-overall}.
@@
     \includegraphics[width=1.05\textwidth]{scenario_resolution_performance.png}
-    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and confidence intervals (trials, $n=10$) are shown for 10 epochs in each case.}
+    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and 95\% confidence intervals (trials, $n=5) are shown for 10 epochs in each case.}
     \label{fig:sens-overall}
@@
 \autoref{fig:sens-val-acc} shows the boxplots of the validation accuracy in all twelve instances ($n=10$).
-We further summarize the statistics (mean and standard deviation) of the validation accuracies in \autoref{tab:meanvalacc}.
-Using the Welch test for pairwise comparisons of resolution performance within each scenario, we find no statistically significant difference between the validation accuracy for each resolution ($p\ge 0.57$).
+% We further summarize the statistics (mean and standard deviation) of the validation accuracies in \autoref{tab:meanvalacc}.
+Using the Welch test for pairwise comparisons of resolution performance within each scenario, we found no statistically significant difference between the validation accuracy for each resolution ($p\ge 0.57$).
 \begin{figure}[ht]
@@
     \includegraphics[width=\textwidth]{figures/scenario_resolution_validation_accuracy_boxplot.png}
-    \caption{Boxplots of validation accuracy values with error bars for each scenario and resolution combination. Mean values are depicted as white squares in each box.}
+    \caption{Boxplots of validation accuracy values with error bars for each scenario and resolution combination.
+Mean values are depicted as white squares in each box.
+The upper and lower bounds of the box are the first and third quartiles; the whiskers are a standard deviation (SD); and the diamonds are outliers.} % >X SDs from the mean}
     \label{fig:sens-val-acc}
@@
 The metrics are shown in \autoref{fig:cnn-perf}.
-In each scenario, the model was trained over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
+In each scenario, we trained the model over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
 That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.

### 2021-03-03 docs/manuscript/manuscript.tex
man updates
Updates from Overleaf

@@
   \centering
-  \includegraphics[width=.7\textwidth]{raw_input_images_2}
-  \caption{Random examples of the raw training images (2 are shown per class). Images are antialiased in this figure for greater clarity.}
+  \includegraphics[width=.7\textwidth]{Labeled example trees.png}
+  \caption{Examples of the raw training images (2 are shown per class).
+Trees in the left column were categorized as improbable due to their lack of structural defects as well as good physiological health.
+Trees in the center column were categorized as possible due to weak branch unions and dieback of crown extremities.
+Trees in the right column were categorized as probable for extensive sections of dead wood in the upper photograph, and a standing dead tree in the lower photograph.
+Images are antialiased in this figure for greater clarity.}
   \label{fig:raw_images}
@@
     \includegraphics[width=1.05\textwidth]{scenario_resolution_performance.png}
-    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and 95\% confidence intervals (trials, $n=5) are shown for 10 epochs in each case.}
+    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and 95\% confidence intervals (trials, $n=5$) are shown for 10 epochs in each case.}
     \label{fig:sens-overall}

### 2021-03-03 docs/manuscript/manuscript.tex

--- 2021-03-03 docs/manuscript/manuscript.tex
@@
 \end{itemize}
-In this study, only images of trees assigned to the 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were included in modeling, while images of \textit{imminent} trees were excluded due to their rarity.
-Randomly chosen examples are shown for each category in \autoref{fig:raw_images}.
+In this study, only images of trees assigned to the 3 likelihood of failure categories of \textit{Improbable}, \textit{Possible},and \textit{Probable} were included in modeling, while images of \textit{Imminent} trees were excluded due to their rarity.
+Typical examples are shown for each category in \autoref{fig:raw_images}.
 In the original set of training images, the class distribution is given in \autoref{tab:classdist}.
 
-\begin{figure}[h!]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=.7\textwidth]{raw_input_images_2}
-  \caption{Random examples of the raw training images (2 are shown per class). Images are antialiased in this figure for greater clarity.}
+  % \includegraphics[width=\textwidth]{raw_input_images_2}
+  \begin{subfigure}[t]{.32\linewidth}\centering
+    \includegraphics[width=\textwidth,angle=-90]{probable-example-1} %probable_standing dead.JPG
+
+    \includegraphics[width=\textwidth,angle=-90]{probable-example-2} %Probable (deadwood above line hieght, vine)
+    \caption{Probable}
+  \end{subfigure}
+    \begin{subfigure}[t]{.32\linewidth}\centering
+    \includegraphics[width=\textwidth,angle=-90]{possible-example-1} % Possible (spruce, codominant and dieback)
+
+    \includegraphics[width=\textwidth,angle=-90]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
+    \caption{Possible}   
+  \end{subfigure}
+  \begin{subfigure}[t]{.32\linewidth}\centering
+    \includegraphics[width=\textwidth,angle=-90]{improbable-example-1} %Improbable_3_london plane
+
+    \includegraphics[width=\textwidth,angle=-90]{improbable-example-2}
+    \caption{Improbable}
+  \end{subfigure}
+  \caption{Examples of training images in each of the three tree risk categories considered in this study: Probable, Possible and Improbable. Two are shown in each case.}
   \label{fig:raw_images}
@@
     \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
-    \texttt{Pr\_PoIm}        & \{Probable, Possible + Impossible\}  & 2 \\\bottomrule
+    \texttt{Pr\_PoIm}        & \{Probable, Possible + Improbable\}  & 2 \\\bottomrule
     \end{tabular}
@@
 
-Unlike other neural networks, however, the CNN performs fundamental pixel-mapping operations in its "convolutional" layers.
-Each convolutional layer is defined by a stack of feature maps, which result from a dot product of a filter and correspondinlgy-sized local receptive fields from the input image or preceding layer.
-The numeric values of the filters correspond to weights whose optimal values are learned during the training of the CNN.
-The size of the filter in each convolutional layer is given by the "kernel size."
+Unlike other neural networks, however, the CNN performs fundamental pixel-mapping operations in its convolutional
+layers. Each convolutional layer is defined by a stack of feature maps, which result from a dot product of a filter and
+correspondinlgy-sized local receptive fields from the input image or preceding layer.  The numeric values of the filters
+correspond to weights whose optimal values are learned during the training of the CNN.  The size of the filter in each
+convolutional layer is given by the \textit{kernel size}.
 
 In this paper, we employ a relatively simple CNN architecture, historically inspired by AlexNet \cite{krizhevsky2012imageneta}.
-The structure of the CNN is shown in \autoref{fig:cnn}.
+The structure of the CNN is shown in \autoref{fig:cnn} (generated using an automated  framework \cite{bauerle2021net2vis}).
 After the input layer (a matrix of pixels from the input image), we use a 64-filter convolutional layer.
@@
     \includegraphics[width=.7\textwidth]{figures/net2vis-name-labels/legend}
-    \caption{Diagram of convolutional neural network structure.
-Hyperparameters that are optimized include the number of units in the penultimate dense layers.
-(Generated using automated framework \cite{bauerle2021net2vis})}
+    \caption{Diagram of convolutional neural network structure. Hyperparameters that are optimized include the number of units in the penultimate dense layers.}
     \label{fig:cnn}
@@
     \bf Layer & \bf Layer \# & \bf No.\ Filters & \bf Kernel Size & \bf Strides &
-     &  & \bf Activation \\\midrule
-    Convolutional  & 1         & 64                & $k^*$   & 2           &&& ReLU \\
+    \bf Activation &   \bf Rate & \bf No.\ Units  \\\midrule
+    Convolutional  & 1         & 64                & $k^*$   & 2           & ReLU && \\
     Max. Pooling   & 1         &                   & 2               &&&             &      \\
-    Convolutional  & 2         & 128               & 3   & 1           &&& ReLU \\
-    Convolutional  & 3         & 128               & 3   & 1           &&& ReLU \\
+    Convolutional  & 2         & 128               & 3   & 1           & ReLU && \\
+    Convolutional  & 3         & 128               & 3   & 1           & ReLU && \\
     Max. Pooling   & 2         &                   & 2               &&&             &      \\    
-    Convolutional  & 4         & 256               & 3   & 1           &&& ReLU \\
-    Convolutional  & 5         & 256               & 3   & 1           &&& ReLU \\
+    Convolutional  & 4         & 256               & 3   & 1           & ReLU && \\
+    Convolutional  & 5         & 256               & 3   & 1           & ReLU && \\
     Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
     Flatten & & & & &&& \\\midrule
-     &  &   &  & &
-     \bf Rate & \bf No.\ Units & \\\midrule
-    Dense           & 1  & & &  & & $u_1^*$ & $a_1^*$ \\
-    Dropout         & 1  & & &  & $r_1^*$ & &  \\
-    Dense           & 2  & & &  & & $u_2^*$ & $a_2^*$ \\
-    Dropout         & 2  & & &  & $r_2^*$ & & \\ \bottomrule
+%     &  &   &  & & &
+%     \bf Rate & \bf No.\ Units  \\\midrule
+    Dense           & 1  & & &  &  $a_1^*$ & & $u_1^*$ \\
+    Dropout         & 1  & & &  && $r_1^*$ &   \\
+    Dense           & 2  & & &  &  $a_2^*$ & & $u_2^*$  \\
+    Dropout         & 2  & & &  && $r_2^*$ &  \\ \bottomrule
     \end{tabular}
@@
                           &                              &                           & \textbf{64}             & \textbf{128}            & \textbf{224}            \\ \midrule
+\multirow{8}{*}{Pr\_Im}    & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
+                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
+                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
+                          & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.1                     & 0.1                     & 0.1                     \\
+                          & 2nd dropout rate, $r_2$      & $\{0, .05, \ldots, 5\}$   & 0.3                     & 0.3                     & 0.3                     \\
+                          & 1st dense layer units, $u_1$ & $\{32, 64, \ldots, 512\}$ & 480                     & 480                     & 480                     \\
+                          & 2nd dense layer units, $u_2$ & $\{32, 64, \ldots, 512\}$ & 448                     & 448                     & 448                     \\
+                          & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\\midrule
 \multirow{8}{*}{PrPo\_Im}  & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 5                       & 7                       \\
@@
                           & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000102956216062} & \num{0.000109137476524} & \num{0.000102956216062} \\\midrule
-\multirow{8}{*}{Pr\_Im}    & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
-                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
-                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
-                          & 1st dropout rate, $r_1$      & $\{0, .05, \ldots, 5\}$   & 0.1                     & 0.1                     & 0.1                     \\
-                          & 2nd dropout rate, $r_2$      & $\{0, .05, \ldots, 5\}$   & 0.3                     & 0.3                     & 0.3                     \\
-                          & 1st dense layer units, $u_1$ & $\{32, 64, \ldots, 512\}$ & 480                     & 480                     & 480                     \\
-                          & 2nd dense layer units, $u_2$ & $\{32, 64, \ldots, 512\}$ & 448                     & 448                     & 448                     \\
-                          & learning rate, $\lambda$     & $[10^{-4}, 10^{-2}]$      & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\\midrule
 \multirow{8}{*}{Pr\_PoIm}  & 1st conv. kernel size, $k$   & $\{5, 7\}$                & 7                       & 7                       & 7                       \\
@@
 \section{Results}
-\subsection{Sensitivity to training resolution}
 We trained the CNN for each of the 4 classification scenarios.
@@
 Thus, we generated twelve learning cases.
-In each case, we performed a five-fold cross-validation (resulting in a training-validation ratio of 80:20), each consisting of 10 epochs.
-The trajectories of the training/validation loss and accuracy are shown in \autoref{fig:sens-overall}.
-Although the 64-pixel case is the slowest to converge in all scenarios, it appears to have a similar performance with respect to the other two resolutions (128px and 224px).
+In each case, we performed a 5-fold cross-validation (resulting in a training-validation ratio of 80:20 in each fold).
+We trained the CNN over 20 epochs in all the cases.
+The trajectories of the  loss and accuracy are shown in \autoref{fig:sens-overall} for both the training and validation sets.
+% Although the 64-pixel case is the slowest to converge in all scenarios, it appears to have a similar performance with respect to the other two resolutions (128px and 224px).
 
@@
     \includegraphics[width=1.05\textwidth]{scenario_resolution_performance.png}
-    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and 95\% confidence intervals (trials, $n=5) are shown for 10 epochs in each case.}
+    \caption{Performance metrics for each scenario and input resolution instance. Average metrics and 95\% confidence intervals (trials, $n=5$) are shown for 20 epochs in each case.}
     \label{fig:sens-overall}
@@
 
-\autoref{fig:sens-val-acc} shows the boxplots of the validation accuracy in all twelve instances ($n=10$).
+\subsection{Sensitivity to training resolution}
+\autoref{fig:sens-val-acc} shows the boxplots of the validation performance metrics for all twelve cases considered.
+The metrics are: accuracy and the macro-averages of precision, recall and the $F_{1}$ score.
+%With respect to the accuracy, there are no significant differences between the 
+Taking all metrics into consideration, Welch pairwise tests showed that there are no significant differences in performance among the 3 training resolutions, with one exception.
+In the scenario Pr\_PoIm, the accuracy, $F_{1}^{m}$, $Re^{m}$ for the 128-pixel case are greater than those for the 224-pixel case ($p<.001$).
+This difference in performance is not as stark between the 128-pixel and 64-pixel cases in the same scenario.
+This outcome implies that we can achieve efficiency by  training at lower resolutions without significant losses in performance.
+
 % We further summarize the statistics (mean and standard deviation) of the validation accuracies in \autoref{tab:meanvalacc}.
-Using the Welch test for pairwise comparisons of resolution performance within each scenario, we found no statistically significant difference between the validation accuracy for each resolution ($p\ge 0.57$).
-\begin{figure}[ht]
+% Using the Welch test for pairwise comparisons of resolution performance within each scenario, we found no statistically significant difference between the validation accuracy for each resolution ($p\ge 0.57$).
+
+\begin{figure}[ht!]
     \centering
-    \includegraphics[width=\textwidth]{figures/scenario_resolution_validation_accuracy_boxplot.png}
-    \caption{Boxplots of validation accuracy values with error bars for each scenario and resolution combination.
-Mean values are depicted as white squares in each box.
-The upper and lower bounds of the box are the first and third quartiles; the whiskers are a standard deviation (SD); and the diamonds are outliers.} % >X SDs from the mean}
+    \includegraphics[width=\textwidth]{figures/scenario_resolution_performance_metrics_facet_boxplot.png}
+    \caption{Boxplots of validation performance metrics for each scenario and resolution combination.
+      The upper and lower bounds of each box are the first and third quartiles; the whiskers are three standard deviations apart; and the diamonds are outliers.
+Mean values are depicted as white squares in each box.}
  \label{fig:sens-val-acc}
@@
 
+\subsection{Scenario sensitivity}
+We conducted pairwise tests between each scenario combination for all resolutions.
+The results indicated that considering all metrics, the scenarios are statistically significantly different from each other ($p < .005$) except Pr\_Im compared to PrPo\_Im and Pr\_PoIm compared to Pr\_Po\_Im, with respect to certain metrics.
+If the accuracy is not taken into account, then there is no strong evidence that Pr\_Im differs in performance when compared with PrPo\_Im.
+In the case of Pr\_PoIm versus Pr\_Po\_Im, there is greater evidence to support their similarity in performance 
+with respect to  $F_{1}^{m}$ and $Re^{m}$. %and $Pr^{m}$
+Otherwise, the performance between these two scenarios is significantly different.
+From \autoref{fig:sens-val-acc}, we observe the best performances in Pr\_Im and PrPo\_Im, with all metrics greater than or equal to 0.90
+The performance in Pr\_PoIm is lower, and Pr\_Po\_Im has the lowest performance of all four.
+
+\subsection{Analysis of classification strategies}
 Given that there was no significant difference in performance among the three resolutions tested, we focus our attention on the 128-px case, as a tradeoff between efficiency and performance.
 We re-trained the model on all the four scenarios and then evaluated performance on the following validation metrics: average precision, average recall and accuracy.
-
-\subsection{Analysis of classification strategies}
-We investigated in detail the performance of the four classification scenarios outlined earlier, using the 128-pixel resolution images.
-The goal of this analysis was to explore the efficiencies of various grouping strategies for tree failure risk.
+%We investigated in detail the performance of the four classification scenarios outlined earlier, using the 128-pixel resolution images.
+The goal of this analysis was to explore the efficiencies of various classification strategies for tree failure risk.
 Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.

### 2021-03-03 docs/manuscript/manuscript.tex
merges and updates

@@
   \end{subfigure}
-  \caption{Examples of training images in each of the three tree risk categories considered in this study: Probable, Possible and Improbable. Two are shown in each case.}
+  \caption{Examples of training images in each of the three tree risk categories considered in this study.
+  Trees in the left columnwere categorized as improbable due to their lack of structural defects as well as good physiologicalhealth.
+Trees in the center column were categorized as possible due to weak branch unions anddieback of crown extremities.
+Trees in the right column were categorized as probable for extensivesections of dead wood in the upper photograph, and a standing dead tree in the lower photograph.Images are antialiased in this figure for greater clarity}
   \label{fig:raw_images}

### 2021-03-04 docs/manuscript/manuscript.tex
Symbols description table

--- 2021-03-03 docs/manuscript/manuscript.tex
@@
   manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
-  likelihood categories under four classification scenarios.  Starting with an original set of 525 expert-labeled images, we perform
+  likelihood categories under four classification scenarios.  Starting with an original set of 505 expert-labeled images, we perform
   preprocessing and augmentation tasks to increase the number of samples. We optimize several hyperparameters in our CNN
@@
   \end{subfigure}
-  \caption{Examples of training images in each of the three tree risk categories considered in this study: Probable, Possible and Improbable. Two are shown in each case.}
+  \caption{Examples of training images in each of the three tree risk categories considered in this study.
+  Trees in the right-hand column were categorized as improbable due to their lack of structural defects as well as good physiological health.
+Trees in the center column were categorized as possible due to weak branch unions and crown dieback.
+Trees in the left-hand column were categorized as probable because they were dead.
+Leaves in the bottom image in the left-hand column are from vines attached to the dead tree.
+Images are antialiased in this figure for greater clarity}
   \label{fig:raw_images}
@@
 
+In \autoref{tab:scenarios}, the first scenario considered each likelihood of failure category separately, as an arborist would do in practice.
+The second scenario considered only the highest and lowest likelihood of failure categories used in the study to clearly distinguish between categories.
+Since Previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure (KOESER), in the third scenario, we pooled trees in the probable and possible categories and compared them to trees in the improbable category.
+In the fourth scenario, we pooled trees in the possible and improbable categories and compared them to trees in the probable scenario.
+In practice, this scenario is less likely because arborists typically distinguish trees with an improbable likelihood of failure as those with minimal or no structural defects.
+It requires additional judgment to distinguish trees with probable or possible likelihood of failure because an arborist must assess the severity of structural defects, the presence of response growth, and the expected loads \cite{smiley2017best}.
 \subsection{Image pre-processing and augmentation}
@@
 Finally, we uniformly sampled learning rates $\lambda$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l c}\toprule
+    \bf Symbol            & \bf Definition  \\\midrule
+    \texttt i     &  \ Given observation \\
+    \texttt c          & \ Class\\
+    \texttt s_c        & \ Class-specified score \\
+    \texttt $\hat y_i$       & \ Predicted class of a given observation \\
+    \texttt y_i        & \ Observed (true) class of a given observation \\
+    \texttt f(s_c)        & \ Softmax activation function \\
+    \texttt p_{i,c}        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class c \\
+    \texttt L^{CE}_i        & \ Categorical cross entropy loss function of a single observation \\
+    \textttt Pr & \ Precision \\
+    \textttt Re & \ Recall \\
+    \textttt F_1 & \ F_1 score \\\bottomrule
+    \end{tabular}
+    \caption{Symbols description}
+    \label{tab:symbols}
+\end{table}
 

### 2021-03-04 docs/manuscript/manuscript.tex

@@
   \centering
-  % \includegraphics[width=\textwidth]{raw_input_images_2}
   \begin{subfigure}[t]{.32\linewidth}\centering
-    \includegraphics[width=\textwidth,angle=-90]{probable-example-1} %probable_standing dead.JPG
+    \includegraphics[width=\textwidth,keepaspectratio=true]{probable-example-1} %probable_standing dead.JPG
 
-    \includegraphics[width=\textwidth,angle=-90]{probable-example-2} %Probable (deadwood above line hieght, vine)
+    \includegraphics[width=\textwidth,keepaspectratio=true]{probable-example-2} %Probable (deadwood above line hieght, vine)
     \caption{Probable}
@@
     \begin{subfigure}[t]{.32\linewidth}\centering
-    \includegraphics[width=\textwidth,angle=-90]{possible-example-1} % Possible (spruce, codominant and dieback)
+    \includegraphics[width=\textwidth,keepaspectratio=true]{possible-example-1} % Possible (spruce, codominant and dieback)
 
-    \includegraphics[width=\textwidth,angle=-90]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
+    \includegraphics[width=\textwidth,keepaspectratio=true]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
     \caption{Possible}   
@@
   \begin{subfigure}[t]{.32\linewidth}\centering
-    \includegraphics[width=\textwidth,angle=-90]{improbable-example-1} %Improbable_3_london plane
+    \includegraphics[width=\textwidth,keepaspectratio=true]{improbable-example-1} %Improbable_3_london plane
 
-    \includegraphics[width=\textwidth,angle=-90]{improbable-example-2}
+    \includegraphics[width=\textwidth,keepaspectratio=true]{improbable-example-2}
     \caption{Improbable}
@@
 Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
-Thus, 
 \begin{table}[h!]
@@
 It requires additional judgment to distinguish trees with probable or possible likelihood of failure because an arborist must assess the severity of structural defects, the presence of response growth, and the expected loads \cite{smiley2017best}.
+
 \subsection{Image pre-processing and augmentation}
@@
 The dropout rates were allowed to range from 0 to 5 in steps of 0.05, while the number of neurons or units in each dense layer varied from 32 to 512 in steps of 32.
-Finally, we uniformly sampled learning rates $\lambda$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
+Finally, we uniformly sampled learning rates for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
 
@@
     \bf Symbol            & \bf Definition  \\\midrule
-    \texttt i     &  \ Given observation \\
-    \texttt c          & \ Class\\
-    \texttt s_c        & \ Class-specified score \\
-    \texttt $\hat y_i$       & \ Predicted class of a given observation \\
-    \texttt y_i        & \ Observed (true) class of a given observation \\
-    \texttt f(s_c)        & \ Softmax activation function \\
-    \texttt p_{i,c}        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class c \\
-    \texttt L^{CE}_i        & \ Categorical cross entropy loss function of a single observation \\
-    \textttt Pr & \ Precision \\
-    \textttt Re & \ Recall \\
-    \textttt F_1 & \ F_1 score \\\bottomrule
+    $i$     &  \ Given observation \\
+    $c$          & \ Class\\
+    $s_c$        & \ Class-specified score \\
+    $\hat y_i$       & \ Predicted class of a given observation \\
+    $y_i$        & \ Observed (true) class of a given observation \\
+    $f(s_c)$        & \ Softmax activation function \\
+    $p_{i,c}$        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class c \\
+    $L^{CE}_i$        & \ Categorical cross entropy loss function of a single observation \\
+    $Pr$ & \ Precision \\
+    $Re$ & \ Recall \\
+    $F_1$ & \ F_1 score \\\bottomrule
     \end{tabular}
@@
 Here, we use a batch size of 32.
-The learning rate $\lambda$ of the optimization algorithm is an important hyperparameter that affects training performance.
+The learning rate  of the optimization algorithm is an important hyperparameter that affects training performance.
 We optimized for this in the hyperparameter search as discussed.

### 2021-03-04 docs/manuscript/manuscript.tex
Replace monospace with proper notation

@@
 The second scenario considered only the highest and lowest likelihood of failure categories used in the study to clearly distinguish between categories.
-Since Previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure (KOESER), in the third scenario, we pooled trees in the probable and possible categories and compared them to trees in the improbable category.
+Since Previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure \cite{koeser2020can}, in the third scenario, we pooled trees in the probable and possible categories and compared them to trees in the improbable category.
 In the fourth scenario, we pooled trees in the possible and improbable categories and compared them to trees in the probable scenario.
@@
     $Re$ & \ Recall \\
-    $F_1$ & \ F_1 score \\\bottomrule
+    $F_1$ & \ $F_1$ score \\\bottomrule
     \end{tabular}

### 2021-03-04 docs/manuscript/manuscript.tex
minor edits to italicize categories and improve formatting.

--- 2021-03-04 docs/manuscript/manuscript.tex
@@
 \begin{itemize}
-\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions;
-\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions;
-\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame;
-\item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load.
+\item \textit{Improbable}: failure unlikely either during normal or extreme weather conditions;
+\item \textit{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions;
+\item \textit{Probable}: failure expected under normal weather conditions within a given time frame;
+\item \textit{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load.
 This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect targets from impact.
 \end{itemize}
-In this study, only images of trees assigned to the 3 likelihood of failure categories of \textit{Improbable}, \textit{Possible},and \textit{Probable} were included in modeling, while images of \textit{Imminent} trees were excluded due to their rarity.
+In this study, only images of trees assigned to the likelihood of failure categories of \textit{Improbable}, \textit{Possible} and \textit{Probable} were included in modeling.
+Images of \textit{Imminent} trees were excluded due to their rarity.
 Typical examples are shown for each category in \autoref{fig:raw_images}.
@@
 \subsection{Classification scenarios}
-In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, 
-we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
+In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
 Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
@@
     \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
-    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
-    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
-    \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
-    \texttt{Pr\_PoIm}        & \{Probable, Possible + Improbable\}  & 2 \\\bottomrule
+    Pr\_Im          & \{Probable, Improbable\}           & 2 \\
+    PrPo\_Im        & \{Probable + Possible, Improbable\}  & 2 \\
+    Pr\_PoIm        & \{Probable, Possible + Improbable\}  & 2 \\
+    Pr\_Po\_Im     &  \{Probable, Possible, Improbable\} & 3 \\
+    \bottomrule
     \end{tabular}
@@
 
-In \autoref{tab:scenarios}, the first scenario considered each likelihood of failure category separately, as an arborist would do in practice.
-The second scenario considered only the highest and lowest likelihood of failure categories used in the study to clearly distinguish between categories.
-Since Previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure (KOESER), in the third scenario, we pooled trees in the probable and possible categories and compared them to trees in the improbable category.
-In the fourth scenario, we pooled trees in the possible and improbable categories and compared them to trees in the probable scenario.
-In practice, this scenario is less likely because arborists typically distinguish trees with an improbable likelihood of failure as those with minimal or no structural defects.
+Scenario Pr\_Im considered only the highest and lowest likelihood of failure categories used in the study to clearly distinguish between categories.
+Since previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure \cite{koeser2020can}, in scenario PrPo\_Im, we pooled trees in the \textit{Probable} and \textit{Possible} categories and compared them to trees in the \textit{Improbable} category.
+In scenario Pr\_PoIm, we pooled trees in the \textit{Possible} and \textit{Improbable} categories and compared them to trees in the \textit{Probable} category.
+In practice, this scenario is less likely because arborists typically distinguish trees with an \textit{Improbable} likelihood of failure as those with minimal or no structural defects.
 It requires additional judgment to distinguish trees with probable or possible likelihood of failure because an arborist must assess the severity of structural defects, the presence of response growth, and the expected loads \cite{smiley2017best}.
+Scenario \texttt{Pr\_Po\_Im} considered each likelihood of failure category separately, as an arborist would do in practice.
 
@@
     $f(s_c)$        & \ Softmax activation function \\
-    $p_{i,c}$        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class c \\
+    $p_{i,c}$        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class $c$ \\
     $L^{CE}_i$        & \ Categorical cross entropy loss function of a single observation \\
@@
 \end{tabular}
-% \begin{tabular}{@{}lllll@{}}
-% \toprule
-% \multirow{2}{*}{\textbf{Scenario}} & \multirow{2}{*}{\textbf{Hyperparameters}} & \multicolumn{3}{c}{\textbf{Resolution}} \\ \cmidrule(l){3-5} 
-%                             &                & 64                & 128               & 224               \\ \midrule
-% \multirow{8}{*}{PrPo\_Im}   & activation\_1  & ReLU              & tanh              & ReLU              \\
-%                             & activation\_2  & ReLU              & ReLU              & ReLU              \\
-%                             & dropout\_1     & 0.1               & 0.25              & 0.1               \\
-%                             & dropout\_2     & 0.3               & 0.35              & 0.3               \\
-%                             & kernel\_size   & 7                 & 5                 & 7                 \\
-%                             & learning\_rate & \num{0.000102956216062} & \num{0.000109137476524} & \num{0.000102956216062} \\
-%                             & units\_1       & 480               & 384               & 480               \\
-%                             & units\_2       & 448               & 256               & 448               \\\midrule
-% \multirow{8}{*}{Pr\_Im}     & activation\_1  & ReLU              & ReLU              & ReLU              \\
-%                             & activation\_2  & ReLU              & ReLU              & ReLU              \\
-%                             & dropout\_1     & 0.1               & 0.1               & 0.1               \\
-%                             & dropout\_2     & 0.3               & 0.3               & 0.3               \\
-%                             & kernel\_size   & 7                 & 7                 & 7                 \\
-%                             & learning\_rate & \num{0.000102956216062} & \num{0.000102956216062} & \num{0.000102956216062} \\
-%                             & units\_1       & 480               & 480               & 480               \\
-%                             & units\_2       & 448               & 448               & 448               \\\midrule
-% \multirow{8}{*}{Pr\_PoIm}   & activation\_1  & ReLU              & ReLU              & ReLU              \\
-%                             & activation\_2  & tanh              & ReLU              & tanh              \\
-%                             & dropout\_1     & 0.25              & 0.1               & 0.2               \\
-%                             & dropout\_2     & 0.3               & 0.3               & 0.15              \\
-%                             & kernel\_size   & 7                 & 7                 & 7                 \\
-%                             & learning\_rate & \num{0.000175822328104} & \num{0.000102956216062} & \num{0.000114656409879} \\
-%                             & units\_1       & 128               & 480               & 416               \\
-%                             & units\_2       & 320               & 448               & 416               \\
-% \multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & ReLU              & tanh              & ReLU              \\\midrule
-%                             & activation\_2  & tanh              & ReLU              & ReLU              \\
-%                             & dropout\_1     & 0.25              & 0.25              & 0.1               \\
-%                             & dropout\_2     & 0.3               & 0.35              & 0.3               \\
-%                             & kernel\_size   & 7                 & 5                 & 7                 \\
-%                             & learning\_rate & \num{0.000175822328104} & \num{0.000109137476524} & \num{0.000102956216062} \\
-%                             & units\_1       & 128               & 384               & 480               \\
-%                             & units\_2       & 320               & 256               & 448               \\ \bottomrule
-% \end{tabular}
-% Table edited at: https://www.tablesgenerator.com/latex_tables#
 \caption{Optimal hyperparameters found using the Hyperband search algorithm for 12 classification scenario and input image resolution combinations.}
@@
 \subsection{Model training and assessment}
-The softmax activation function  $f(\cdot)$ in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$:
+The softmax activation function  $f(\cdot)$ in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$ as:
 \begin{equation}
@@
 \end{equation}
+where $C$ is the set of classes and $c, c'$ are indices for a given class.
 Thus for the $i$th observation, the softmax activation returns the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
@@
   \centering
-  \begin{subfigure}[t]{.45\linewidth}
+  \begin{subfigure}[t]{.4\linewidth}
     \centering
@@
   \end{subfigure}%
-   \begin{subfigure}[t]{.45\linewidth}
+   \begin{subfigure}[t]{.4\linewidth}
     \centering
@@
 
-  \begin{subfigure}[t]{.45\linewidth}
+  \begin{subfigure}[t]{.4\linewidth}
     \centering
@@
 \end{subfigure}%
-  \begin{subfigure}[t]{.45\linewidth}
+  \begin{subfigure}[t]{.4\linewidth}
     \centering

### 2021-03-04 docs/manuscript/manuscript.tex
Summarized macro-average metrics

@@
     $L^{CE}_i$        & \ Categorical cross entropy loss function of a single observation \\
-    $Pr$ & \ Precision \\
-    $Re$ & \ Recall \\
-    $F_1$ & \ $F_1$ score \\\bottomrule
+    $Pr_c$ & \ Class-specific precision \\
+    $Re_c$ & \ Class-specific recall \\
+    $Pr^m$ & \ Macro-average precision \\
+    $Re^m$ & \ Macro-average recall \\
+    $F_1^m$ & \ Macro-average $F_1$ score \\\bottomrule
     \end{tabular}

### 2021-03-04 docs/manuscript/manuscript.tex
generated images for figure 2

--- 2021-03-04 docs/manuscript/manuscript.tex
@@
   \begin{subfigure}[t]{.32\linewidth}\centering
-    \includegraphics[width=\textwidth,keepaspectratio=true]{probable-example-1} %probable_standing dead.JPG
+    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{probable-example-1} %probable_standing dead.JPG
 
-    \includegraphics[width=\textwidth,keepaspectratio=true]{probable-example-2} %Probable (deadwood above line hieght, vine)
+    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{probable-example-2} %Probable (deadwood above line hieght, vine)
     \caption{Probable}
@@
   \begin{subfigure}[t]{.32\linewidth}\centering
-    \includegraphics[width=\textwidth,keepaspectratio=true]{possible-example-1} % Possible (spruce, codominant and dieback)
+    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{possible-example-1} % Possible (spruce, codominant and dieback)
 
-    \includegraphics[width=\textwidth,keepaspectratio=true]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
+    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
     \caption{Possible}   
@@
   \begin{subfigure}[t]{.32\linewidth}\centering
-    \includegraphics[width=\textwidth,keepaspectratio=true]{improbable-example-1} %Improbable_3_london plane
+    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{improbable-example-1} %Improbable_3_london plane
 
-    \includegraphics[width=\textwidth,keepaspectratio=true]{improbable-example-2}
+    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{improbable-example-2}
     \caption{Improbable}

### 2021-03-04 docs/manuscript/manuscript.tex
extensive edits to manuscript

--- 2021-03-04 docs/manuscript/manuscript.tex
@@
   \centering
-  \begin{subfigure}[t]{.32\linewidth}\centering
+  \begin{subfigure}[t]{.325\linewidth}\centering
     \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{probable-example-1} %probable_standing dead.JPG
@@
   \end{subfigure}
-  \begin{subfigure}[t]{.32\linewidth}\centering
+  \begin{subfigure}[t]{.325\linewidth}\centering
     \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{possible-example-1} % Possible (spruce, codominant and dieback)
@@
   \end{subfigure}
-  \begin{subfigure}[t]{.32\linewidth}\centering
+  \begin{subfigure}[t]{.325\linewidth}\centering
     \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{improbable-example-1} %Improbable_3_london plane
@@
     \bf Category     & \bf Number of images  \\ \midrule
-    Improbable & 322\\
-    Possible & 80 \\
-    Probable & 56 \\\midrule
+    \textit{Improbable} & 322\\
+    \textit{Possible} & 80 \\
+    \textit{Probable} & 56 \\\midrule
     Total & 505 \\\bottomrule
     \end{tabular}
-    \caption{Category distribution of images in the original set of input images}
+    \caption{Category distribution  in the set of raw input images}
     \label{tab:classdist}
@@
     \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
-    Pr\_Im          & \{Probable, Improbable\}           & 2 \\
-    PrPo\_Im        & \{Probable + Possible, Improbable\}  & 2 \\
-    Pr\_PoIm        & \{Probable, Possible + Improbable\}  & 2 \\
-    Pr\_Po\_Im     &  \{Probable, Possible, Improbable\} & 3 \\
+    Pr\_Im          & \{\textit{Probable}, \textit{Improbable}\}           & 2 \\
+    PrPo\_Im        & \{\textit{Probable} + \textit{Possible}, \textit{Improbable}\}  & 2 \\
+    Pr\_PoIm        & \{\textit{Probable}, \textit{Possible} + \textit{Improbable}\}  & 2 \\
+    Pr\_Po\_Im     &  \{\textit{Probable}, \textit{Possible}, \textit{Improbable}\} & 3 \\
     \bottomrule
@@
 It requires additional judgment to distinguish trees with probable or possible likelihood of failure because an arborist must assess the severity of structural defects, the presence of response growth, and the expected loads \cite{smiley2017best}.
-Scenario \texttt{Pr\_Po\_Im} considered each likelihood of failure category separately, as an arborist would do in practice.
+Scenario Pr\_Po\_Im considered each likelihood of failure category separately, as an arborist would do in practice.
 
@@
 \begin{figure}[h!]
+  \begin{subfigure}[b]{.24\linewidth}
     \centering
-  \begin{subfigure}[t]{.75\linewidth}
+    \textit{\footnotesize Probable}
+
+    \includegraphics[width=\textwidth]{processed-probable-example-1}
+
+    \textit{\footnotesize Improbable}
+
+    \includegraphics[width=\textwidth]{processed-improbable-example-1}
+    \caption{Scenario Pr\_Im}
+    \label{pr_im_64}
+  \end{subfigure}%
+  \begin{subfigure}[b]{.24\linewidth}
     \centering
-    \includegraphics[width=.8\linewidth, trim={0 3cm 0 3cm}, clip]{processed_input_images_Pr_Po_Im_128_px}
-    \caption{Scenario \texttt{Pr\_Po\_Im}}
+    \textit{\footnotesize Probable/Possible}
+
+    \includegraphics[width=\textwidth]{processed-probable-example-2-flipped}
+
+    \textit{\footnotesize Improbable}
+
+    \includegraphics[width=\textwidth]{processed-improbable-example-2}
+    \caption{Scenario  PrPo\_Im}
+    \label{prpo_im_64}
+  \end{subfigure}%  
+  \begin{subfigure}[b]{.24\linewidth}
+    \centering
+    \textit{\footnotesize Probable}
+
+    \includegraphics[width=\textwidth]{processed-probable-example-1-flipped}
+
+    \textit{\footnotesize Possible/Improbable}
+
+    \includegraphics[width=\textwidth]{processed-possible-example-2-flipped}
+    \caption{Scenario Pr\_PoIm}
+    \label{pr_poim_64}
+  \end{subfigure}%  
+  \begin{subfigure}[b]{.24\linewidth}
+    \centering
+    % \includegraphics[width=.8\linewidth, trim={0 3cm 0 3cm}, clip]{processed_input_images_Pr_Po_Im_128_px}
+    \textit{\footnotesize Probable}
+
+    \includegraphics[width=\textwidth]{processed-probable-example-2}
+
+    \textit{\footnotesize Possible}
+
+    \includegraphics[width=\textwidth]{processed-possible-example-1}
+
+    \textit{\footnotesize Improbable}
+
+    \includegraphics[width=\textwidth]{processed-improbable-example-2-flipped}
+    \caption{Scenario Pr\_Po\_Im}
     \label{pr_po_im_64}
@@
 
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_Pr_Im_128_px}
-    \caption{Scenario \texttt{Pr\_Im}}
-    \label{pr_im_64}
-  \end{subfigure}%
-
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_PrPo_Im_128_px}
-    \caption{Scenario \texttt{PrPo\_Im}}
-    \label{prpo_im_64}
-  \end{subfigure}%
-  \begin{subfigure}[t]{.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_Pr_PoIm_128_px}
-    \caption{Scenario \texttt{Pr\_PoIm}}
-    \label{pr_poim_64}
-  \end{subfigure}%
-
-  \caption{Random examples of the processed training images under each classification scenario; 128 pixels.}
+  \caption{Selected processed training images under each classification scenario.
+    The images shown are processed versions of those shown in \autoref{fig:raw_images}.
+    All were randomly cropped along the vertical axis and some horizontally flipped as well.
+  }
   \label{fig:processed_images}
@@
 We follow these with a maximum-pooling layer and then two further 256-filter convolutional layers.
-A final maximum-pooling layer is used before we "flatten" all outputs into a one-dimensional (fully-connected) array of neurons. 
+A final maximum-pooling layer is used before we ``flatten'' all outputs into a one-dimensional (fully-connected) array of neurons. 
 After flattening the outputs, we use a dense layer to reduce the number of outputs a specified number of units.
@@
 A second dense layer is used prior to the output layer, with the number of outputs corresponding to the number of classes in the dataset.
-A regularization technique known as "dropout" is used after each dense layer.
+A regularization technique known as ''dropout'' is used after each dense layer.
 In each dropout layer, a proportion of the neurons are randomly zeroed during training in order to improve the robustness of the model.
@@
 The activation function in both dense layers was specified as a choice between the rectified linear unit (ReLU) function and the hyperbolic tangent ($\tanh$).
-The ReLU was introduced to address the so-called "vanishing gradient" problem and has been shown to improve performance in CNNs \cite{glorot2011deep}.
+The ReLU was introduced to address the so-called ``vanishing gradient'' problem and has been shown to improve performance in CNNs \cite{glorot2011deep}.
 Nevertheless, the $\tanh$ function remains a viable option, as well.
@@
 Finally, we uniformly sampled learning rates for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.
-
-\begin{table}[h!]
-    \centering
-    \begin{tabular}{l l c}\toprule
-    \bf Symbol            & \bf Definition  \\\midrule
-    $i$     &  \ Given observation \\
-    $c$          & \ Class\\
-    $s_c$        & \ Class-specified score \\
-    $\hat y_i$       & \ Predicted class of a given observation \\
-    $y_i$        & \ Observed (true) class of a given observation \\
-    $f(s_c)$        & \ Softmax activation function \\
-    $p_{i,c}$        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class $c$ \\
-    $L^{CE}_i$        & \ Categorical cross entropy loss function of a single observation \\
-    $Pr$ & \ Precision \\
-    $Re$ & \ Recall \\
-    $F_1$ & \ $F_1$ score \\\bottomrule
-    \end{tabular}
-    \caption{Symbols description}
-    \label{tab:symbols}
-\end{table}
 
@@
 \subsection{Model training and assessment}
+In this subsection, we provide an overview of the learning procedure for the convolutional neural network.
+As a reference, all the symbols used here are summarized in \autoref{tab:symbols}
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{l l c}\toprule
+    \bf Symbol            & \bf Definition  \\\midrule
+      $c$          & \ Index of given class\\
+      $f(s_c)$        & \ Softmax activation function \\
+      $F_{1c}$ & \ Class-specific  $F_1$ score \\      
+      $F_1^m$ & \ Macro-average $F_1$ score \\
+      $i$     &  \ Index of given image observation \\           
+      $L^{CE}_i$        & \ Categorical cross entropy loss function of a single observation \\
+      $\hat{p}_{i,c}$        & \ Predicted probability that the $i$\textsuperscript{th} observation belongs to class $c$ \\      
+    $Pr_c$ & \ Class-specific precision \\
+      $Pr^m$ & \ Macro-average precision \\
+      $Re_c$ & \ Class-specific recall \\
+      $Re^m$ & \ Macro-average recall \\
+          $s_c$        & \ Class-specific score \\
+    $y_i$        & \ Observed (true) class of a given observation \\
+      $\hat y_i$       & \ Predicted class of a given observation \\
+      \bottomrule
+    \end{tabular}
+    \caption{Summary of symbols related to the training and assessment of the convolutional neural network}
+    \label{tab:symbols}
+\end{table}
+
 The softmax activation function  $f(\cdot)$ in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$ as:
@@
 where $C$ is the set of classes and $c, c'$ are indices for a given class.
-Thus for the $i$th observation, the softmax activation returns the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
+Thus for the $i$th observation, the softmax activation returns the predicted probability $\hat{p}_{i,c}$ that the $i$th observation belongs to class $c$.
 The CNN is trained using a variant of the stochastic gradient algorithm, Adam \cite{kingma2017adam}.
@@
 \begin{equation}
-    L_{i}^{CE} = -  \log( p_{i,c}) = - \log(f(s_c^i))
+    L_{i}^{CE} = -  \log( \hat{p}_{i,c}) = - \log(f(s_c^i))
 \end{equation}
@@
 In addition to overall accuracy of making correct classifications, we assessed the models trained under these scenarios based on the macro-averages of the precision, recall and $F_1$ score metrics computed over the validation set in each epoch.
-The precision score $Pr$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is an important measure of how good a classifier is.
-The recall $Re$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
-The $F_1$ metric is given as the harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
+The precision score $Pr_{c}$ captures the proportion of correct predictions for a certain class relative to all the predictions for that class, and is an important measure of how good a classifier is.
+The recall $Re_{c}$ captures the ability of a classifier to correctly predict observations for a certain class relative to all the true observations in that class.
+The $F_{1c}$ metric is given as the class-specific harmonic mean of the precision and recall, and is thus more sensitive than the overall accuracy score.
 These three metrics are macro-averaged.
-Thus, each category is given equal weight, ensuring that misclassifications within the smaller classes ("Probable" and "Possible") are adequately represented in the aggregation.
+Thus, each category is given equal weight, ensuring that misclassifications within the smaller classes (\textit{Probable} and \textit{Possible}) are adequately represented in the aggregation.
 These metrics are formally defined as follows:
@@
 Otherwise, the performance between these two scenarios is significantly different.
-From \autoref{fig:sens-val-acc}, we observe the best performances in Pr\_Im and PrPo\_Im, with all metrics greater than or equal to 0.90
+From \autoref{fig:sens-val-acc}, we observe the best performances in Pr\_Im and PrPo\_Im, with all metrics greater than or equal to 0.90.
 The performance in Pr\_PoIm is lower, and Pr\_Po\_Im has the lowest performance of all four.
@@
 The matrices show that generally, the ``Possible'' category is difficult to predict accurately as an individual class.
-The classifier performed best when it only had to distinguish between ``Probable'' and ``Improbable'' (\autoref{pr_im_cm}).
-Performance only suffered slightly when ``Possible'' was grouped with ``Probable'' in the PrPo\_Im scenario (\autoref{prpo_im_cm}).
+The classifier performed best when it only had to distinguish between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}).
+Performance only suffered slightly when \textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}).
 
@@
   \centering
-  \begin{subfigure}[t]{.4\linewidth}
+  \begin{subfigure}[t]{.45\linewidth}
     \centering
@@
   \end{subfigure}%
-   \begin{subfigure}[t]{.4\linewidth}
+   \begin{subfigure}[t]{.45\linewidth}
     \centering
@@
 
-  \begin{subfigure}[t]{.4\linewidth}
+  \begin{subfigure}[t]{.45\linewidth}
     \centering
@@
 \end{subfigure}%
-  \begin{subfigure}[t]{.4\linewidth}
+  \begin{subfigure}[t]{.45\linewidth}
     \centering
@@
   \end{subfigure}%
-    \caption{Confusion matrices for CNN performance using the 128px images across all scenarios.
-    Each row of a confusion matrix indicates the proportional distribution of class predictions for the true members of each class.
-Thus, the diagonals indicate the recall for each class $Re_c$.
-The average of the diagonal values gives the macro-average $Re^m$ for each scenario.}
+  \caption{Scenario confusion matrices for a single training instance using the 128-pixel images.  Each row of a
+    confusion matrix indicates the proportional distribution of class predictions for the true members of each
+    class. Thus, the diagonals indicate the recall for each class $Re_c$. The average of the diagonal values gives the
+    macro-average $Re^m$ for each scenario.}
     \label{fig:confusion-matrix}
@@
 
-  The detailed performance metrics for the 128-pixel case are summarized in \autoref{tab:performance}.
+%   The detailed performance metrics for the 128-pixel case are summarized in \autoref{tab:performance}.
 
-\begin{table}[ht]\small
-    \centering
-\begin{tabular}{@{}llllll@{}}
-\toprule
-\textbf{Scenario} & \textbf{Averaging method} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-score} & \textbf{Support} \\ \midrule
-Pr\_Im    & Probable            & 0.9  & 0.75 & 0.81 & 103 \\
-         & Improbable          & 0.92 & 0.97 & 0.95 & 322 \\
-         & accuracy            &      &      & 0.92 & 425 \\
-         & weighted avg        & 0.92 & 0.92 & 0.92 & 425 \\ \midrule
-Pr\_Po\_Im & Probable            & 0.82 & 0.56 & 0.67 & 103 \\
-         & Possible            & 0.43 & 0.74 & 0.54 & 80  \\
-         & Improbable          & 0.93 & 0.86 & 0.89 & 322 \\
-         & accuracy            &      &      & 0.78 & 505 \\
-         & weighted avg        & 0.83 & 0.78 & 0.79 & 505 \\ \midrule
-Pr\_PoIm  & Probable            & 0.77 & 0.75 & 0.76 & 103 \\
-         & Possible/Improbable & 0.94 & 0.94 & 0.94 & 402 \\
-         & accuracy            &      &      & 0.9  & 505 \\
-         & weighted avg        & 0.9  & 0.9  & 0.9  & 505 \\ \midrule
-PrPo\_Im  & Probable/Possible   & 0.85 & 0.96 & 0.9  & 183 \\
-         & Improbable          & 0.97 & 0.9  & 0.94 & 322 \\
-         & accuracy            &      &      & 0.92 & 505 \\
-         & weighted avg        & 0.93 & 0.92 & 0.92 & 505 \\ \bottomrule
-\end{tabular}
-    \caption{Detailed performance metrics across the four scenarios in the 128-pixel case (single trial)}
-    \label{tab:performance}
-\end{table}
+% \begin{table}[ht]\small
+%     \centering
+% \begin{tabular}{@{}llllll@{}}
+% \toprule
+% \textbf{Scenario} & \textbf{Averaging method} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-score} & \textbf{Support} \\ \midrule
+% Pr\_Im    & Probable            & 0.9  & 0.75 & 0.81 & 103 \\
+%          & Improbable          & 0.92 & 0.97 & 0.95 & 322 \\
+%          & accuracy            &      &      & 0.92 & 425 \\
+%          & weighted avg        & 0.92 & 0.92 & 0.92 & 425 \\ \midrule
+% Pr\_Po\_Im & Probable            & 0.82 & 0.56 & 0.67 & 103 \\
+%          & Possible            & 0.43 & 0.74 & 0.54 & 80  \\
+%          & Improbable          & 0.93 & 0.86 & 0.89 & 322 \\
+%          & accuracy            &      &      & 0.78 & 505 \\
+%          & weighted avg        & 0.83 & 0.78 & 0.79 & 505 \\ \midrule
+% Pr\_PoIm  & Probable            & 0.77 & 0.75 & 0.76 & 103 \\
+%          & Possible/Improbable & 0.94 & 0.94 & 0.94 & 402 \\
+%          & accuracy            &      &      & 0.9  & 505 \\
+%          & weighted avg        & 0.9  & 0.9  & 0.9  & 505 \\ \midrule
+% PrPo\_Im  & Probable/Possible   & 0.85 & 0.96 & 0.9  & 183 \\
+%          & Improbable          & 0.97 & 0.9  & 0.94 & 322 \\
+%          & accuracy            &      &      & 0.92 & 505 \\
+%          & weighted avg        & 0.93 & 0.92 & 0.92 & 505 \\ \bottomrule
+% \end{tabular}
+%     \caption{Detailed performance metrics across the four scenarios in the 128-pixel case (single trial)}
+%     \label{tab:performance}
+% \end{table}
 
@@
 \section{Conclusion}
-We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood (Probable, Possible and Improbable) with respect to utility infrastructure.
+We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood (\textit{Probable}, \textit{Possible} and \textit{Improbable}) with respect to utility infrastructure.
 Specifically, we developed a convolutional neural network with state-of-the-art configurations.
@@
 We further conducted more detailed classification performance analyses over single training instances for the 128-pixel case.
-Our results indicate that the CNN performed best at recalling Probable vs. Improbable cases.
-The Possible category appeared to be a confounding case.
+Our results indicate that the CNN performed best at recalling \textit{Probable} vs. \textit{Improbable} cases.
+The \textit{Possible} category appeared to be a confounding case.
 This was not unexpected, given the uncertainty and subjectivity in assessing these trees to begin with.
-The CNN, however, performed better at distinguishing between Possible/Improbable and Probable, than between Probable/Possible and Improbable.
-This might be an indicator that the Possible cases are more likely identified as Improbable.
-When all three failure-likelihood categories were predicted individually, the Possible case had the lowest recall score.
+The CNN, however, performed better at distinguishing between \textit{Possible}/\textit{Improbable} and \textit{Probable}, than between \textit{Probable}/\textit{Possible} and \textit{Improbable}.
+This might be an indicator that trees in the \textit{Possible} category are more likely to be identified as \textit{Improbable}.
+When all three failure-likelihood categories were predicted individually, the \textit{Possible} category had the lowest recall score.
 

### 2021-03-04 docs/manuscript/manuscript.tex
penultimate manuscript revisions

@@
 \usepackage{amsmath}
-%\usepackage{rotating}
 \usepackage{newtxtext,newtxmath}
@@
 \usepackage[table]{xcolor}
-
-%% Cite Title
-%\usepackage[style=numeric,backend=biber,natbib,sorting=none,maxcitenames=2,maxbibnames=99,doi=false,isbn=false,url=false,eprint=false]{biblatex}
-%\addbibresource{../ai-trees-references.bib}
-
-%%% Counters
-% \usepackage{chngcntr,mathtools}
-% \counterwithout{figure}{section}
-% \counterwithout{table}{section}
-
-% \numberwithin{equation}{section}
 
@@
 %\usepackage{colortbl}
-%\usepackage{color}
-
-%\DeclareRobustCommand{\hlcyan}[1]{{\sethlcolor{cyan}\hl{#1}}}
 
@@
 \usepackage{multirow,array,multicol}
-
-%
-%\definecolor{slblue}{rgb}{0,.3,.62}
-% \hypersetup{
-%     colorlinks,%
-%     citecolor=blue,%
-%     filecolor=blue,%
-%     linkcolor=blue,
-%     urlcolor=slblue
-% }
-
-%%%TIKZ
-% \usepackage{tikz}
-% \usepackage{pgfplots}
-% \usepackage{pgfplotstable}
-% \usepackage{pgfgantt}
-% \pgfplotsset{compat=newest}
-
-% \usetikzlibrary{arrows,shapes,positioning}
-% \usetikzlibrary{decorations.markings}
-% \usetikzlibrary{shadows,automata}
-% \usetikzlibrary{patterns,fit}
-% %\usetikzlibrary{circuits.ee.IEC}
-% \usetikzlibrary{decorations.text}
-% % For Sagnac Picture
-% \usetikzlibrary{%
-%     decorations.pathreplacing,%
-%     decorations.pathmorphing%
-% }
-
-% \newcommand{\osn}{\oldstylenums}
-% \newcommand{\dg}{^{\circ}}
-% \newcommand{\lt}{\left}
-% \newcommand{\rt}{\right}
-% \newcommand{\pt}{\phantom}
-% \newcommand{\tf}{\therefore}
-% \newcommand{\?}{\stackrel{?}{=}}
-% \newcommand{\fr}{\frac}
-% \newcommand{\dfr}{\dfrac}
-% %\newcommand{\ul}{\underline}
-% \newcommand{\tn}{\tabularnewline}
-% \newcommand{\nl}{\newline}
-% \newcommand\relph[1]{\mathrel{\phantom{#1}}}
-% \newcommand{\cm}{\checkmark}
-% \newcommand{\ol}{\overline}
-% \newcommand{\rd}{\color{red}}
-% \newcommand{\bl}{\color{blue}}
-% \newcommand{\pl}{\color{purple}}
-% \newcommand{\og}{\color{orange!90!black}}
-% \newcommand{\gr}{\color{green!40!black}}
-% \newcommand{\nin}{\noindent}
-% \newcommand{\la}{\lambda}
-% \renewcommand{\th}{\theta}
-% \newcommand{\al}{\alpha}
-% \newcommand{\G}{\Gamma}
-% \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-%             \node[shape=circle,draw,thick,inner sep=1pt] (char) {\small #1};}}
-
-% \newcommand{\p}{\partial}
-% \newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
-% \newcommand{\dpd}[2]{\dfrac{\partial{#1}}{\partial{#2}}}
-% \newcommand{\pdd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
-  likelihood categories under four classification scenarios.  Starting with an original set of 505 expert-labeled images, we perform
-  preprocessing and augmentation tasks to increase the number of samples. We optimize several hyperparameters in our CNN
-  and test its performance for three different image resolutions. The trained classifier has an average validation
-  accuracy of at least 0.94 across all scenario-resolution combinations.  Thus, via this novel framework, we
-  demonstrate the potential of artificial intelligence to automate and consequently reduce the costs of tree failure
-  likelihood assessments, thereby promoting sustainable infrastructure.
+  likelihood categories (\textit{Probable}, \textit{Possible}, \textit{Improbable}) under four classification scenarios.
+  Starting with an original set of 505 expert-labeled images, we perform preprocessing and augmentation tasks to
+  increase the number of samples. We optimize several hyperparameters in our CNN and then train and test its performance
+  for three different image resolutions. The trained CNN produced a validation accuracy of at least 0.94
+  ($\hat\sigma = 0.1)$ in the best-performing, yet hypothetical scenario as it excludes one of the categories.  The
+  second-best performing scenario, which includes all categories and therefore more practical, resulted in a validation
+  accuracy of 0.92 ($\hat\sigma = 0.1$). Thus, via this novel framework, we demonstrate the potential of artificial
+  intelligence to automate and consequently reduce the costs of tree failure likelihood assessments, thereby promoting
+  sustainable infrastructure.
 \end{abstract}
@@
   \end{subfigure}
-  \caption{Examples of training images in each of the three tree risk categories considered in this study.
-  Trees in the right-hand column were categorized as improbable due to their lack of structural defects as well as good physiological health.
-Trees in the center column were categorized as possible due to weak branch unions and crown dieback.
-Trees in the left-hand column were categorized as probable because they were dead.
-Leaves in the bottom image in the left-hand column are from vines attached to the dead tree.
-Images are antialiased in this figure for greater clarity}
+  \caption{Examples of training images in each of the three tree risk categories considered in this study.  Trees in  
+     column (a) were categorized as \textit{Improbable} due to their lack of structural defects as well as good
+    physiological health. Trees in the center column were categorized as \textit{Possible} due to weak branch unions and crown
+    dieback. Trees in column (c)  were categorized as \textit{Probable} because they were dead. Leaves in the bottom
+    image in the left-hand column are from vines attached to the dead tree.}
   \label{fig:raw_images}
@@
     \bf Category     & \bf Number of images  \\ \midrule
-    \textit{Improbable} & 322\\
+    \textit{Probable} & 56 \\
       \textit{Possible} & 80 \\
-    \textit{Probable} & 56 \\\midrule
+      \textit{Improbable} & 322\\\midrule
     Total & 505 \\\bottomrule
@@
     The images shown are processed versions of those shown in \autoref{fig:raw_images}.
-    All were randomly cropped along the vertical axis and some horizontally flipped as well.
+    All were randomly cropped along the vertical axis and 50\% were horizontally flipped (including some in this figure).
   }
@@
 Also, it has been shown that training with lower resolution improves performance on higher resolution test images \cite{touvron2019fixing}.
-We employ a convolutional neural network (CNN) as the AI framework for tree risk failure likelihood prediction.
-Like other neural networks, the CNN is an arrangement of neurons within layers, each neuron performing an operation that maps from a pixel in an input image to the final output. 
-The input into each neuron is a weighted sum from the previous layer, while the output from each neuron is modulated by an activation function.
-The activation function in the final layer of an CNN is typically the softmax function, which gives the class probabilities of a given input image. 
-A class is assigned to the image based on the one with the corresponding maximum probability.
+We employ a convolutional neural network (CNN) as the AI framework for tree risk failure likelihood prediction.  Like
+other neural networks, the CNN is an arrangement of neurons within layers, each neuron performing an operation that maps
+from a pixel in an input image to the final output.  The input into each neuron is a weighted sum from the previous
+layer, while the output from each neuron is modulated by an activation function.  The activation function in the final
+layer of an CNN is typically the softmax function, which gives the class probabilities of a given input image.  A class
+is assigned to the image based on the one with the corresponding maximum probability.
 
@@
 After the input layer (a matrix of pixels from the input image), we use a 64-filter convolutional layer.
-The output is downsampled using a pooling layer that returns the maximum ouput from a $2\times 2$ subsample from the previous layer.
+The output is downsampled using a pooling layer that returns the maximum output from a $2\times 2$ subsample from the previous layer.
 Next, we stack two successive convolutional layers each with a depth of 128 filters.
@@
 
-    \medskip
+    \bigskip
 
-    \includegraphics[width=.7\textwidth]{figures/net2vis-name-labels/legend}
-    \caption{Diagram of convolutional neural network structure. Hyperparameters that are optimized include the number of units in the penultimate dense layers.}
+    \includegraphics[width=.5\textwidth, trim={4cm 0 0 0}, clip]{figures/net2vis-name-labels/legend}
+    \caption{Diagram of convolutional neural network structure (excluding the input layer). Hyperparameters that are
+      optimized include the number of units in the penultimate dense layers. Here, 480 and 448 units are used, respectively.}
     \label{fig:cnn}
@@
 
+Scenario Pr\_Im unsurprisingly has the best performance (greatest accuracy and $Pr^{m}$ of 0.95), since the
+classifier only has to predict two extreme categories of \textit{Probable} and \textit{Improbable}.  However, the
+\textit{Possible} category cannot be evaded in reality.  Thus, from the perspective of practical application, scenario
+PrPo\_Im, which is the next best performing scenario (best accuracy of 0.92, $\hat\sigma=0.01$), demonstrates that it is
+most viable to group \textit{Probable} with \textit{Possible} for the best CNN performance.
+
 \subsection{Analysis of classification strategies}
-Given that there was no significant difference in performance among the three resolutions tested, we focus our attention on the 128-px case, as a tradeoff between efficiency and performance.
-We re-trained the model on all the four scenarios and then evaluated performance on the following validation metrics: average precision, average recall and accuracy.
-%We investigated in detail the performance of the four classification scenarios outlined earlier, using the 128-pixel resolution images.
+Given that there was no significant difference in performance among the three resolutions tested, we focused our
+attention on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all the four
+scenarios using this resolution and then evaluated performance on the afore-mentioned validation metrics: accuracy and the macro-averages of
+class-specific precision, recall and $F_{1}$ score.
 The goal of this analysis was to explore the efficiencies of various classification strategies for tree failure risk.
-Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios.
-The metrics are shown in \autoref{fig:cnn-perf}.
-In each scenario, we trained the model over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3.
-That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations.
-From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three.
-This indicates the uncertainty surrounding the subjective risk classification of the trees to begin with.
-Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.
+
+Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the
+respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios. The
+metrics are shown in \autoref{fig:cnn-perf}.  In each scenario, we trained the model over 13 epochs. % but we employed
+% early stopping using validation loss as the criterion with a patience of 3.
+% That is, the training routine would set the
+% weights of the final model at the epoch from which there was no further improvement in validation loss after three
+% subsequent iterations.
+From this figure, we see that Pr\_Po\_Im performs significantly worse than the other three. This indicates the
+uncertainty surrounding the expert, yet subjective, failure likelihood assessment of the trees to begin
+with, particularly when the category is deemed to be \textit{Possible}. % Given the results
+% from the sensitivity tests, however, it is possible that a different training instance may have provided better
+% results, or perhaps, further iterations were required to achieve convergence.
 
@@
 
-We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
-Each row of the confusion matrix indicates the proportion of observations in a given class that are predicted to be in the classes across the columns.
-Thus, the diagonal entry in each matrix represents the class-specific recall score, $Re_{c}$.
-The matrices show that generally, the ``Possible'' category is difficult to predict accurately as an individual class.
-The classifier performed best when it only had to distinguish between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}).
-Performance only suffered slightly when \textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}).
+We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix} to further investigate the
+performance of the classifier.  Each row of the confusion matrix indicates the proportion of observations in a given
+class that are predicted to be in the classes across the columns.  Thus, the diagonal entry in each matrix represents
+the class-specific recall score, $Re_{c}$.  The matrices show that generally, the \textit{Possible} category is
+difficult to predict accurately as an individual class.  The classifier performs best when it only has to distinguish
+between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}). Performance only suffers slightly when
+\textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}), which we would
+select as the best strategy from a practical standpoint.  These two scenarios (Pr\_Im and PrPo\_Im) have the highest
+class-specific recall scores $Re_{c}$ ($\ge 0.90$) compared to the other two scenarios Pr\_PoIm ($Re_{c} \ge 0.58$) and
+Pr\_Po\_Im ($Re_{c} \ge 0.54)$. In scenario Pr\_PoIm, we see that even grouping \textit{Possible} with
+\textit{Improbable} worsens the ability of the classifier to distinguish the \textit{Probable} category.
 
@@
 \section{Conclusion}
-We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood (\textit{Probable}, \textit{Possible} and \textit{Improbable}) with respect to utility infrastructure.
-Specifically, we developed a convolutional neural network with state-of-the-art configurations.
-We applied data augmentation and pre-processing strategies to increase the size of our dataset by a factor of 5, thus generating 2525 images.
-From an initial resolution of $4032 \times 3024$ pixels, we created three sets of image resolutions: $64 \times 64$, $128 \times 128$ and $224 \times 224$.
-We then optimized eight of our CNN hyperparameters for 12  classification scenario and image resolution combinations. 
-We conducted ten model training trials in each of the 12 cases, using a 20\% validation set to measure model performance.
-In these trials, the average accuracy of classification was $\geq 0.94$, with a maximum standard deviation of 0.7.
-There was no statistically significant difference between the performance across image resolutions.
+We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood
+(\textit{Probable}, \textit{Possible} and \textit{Improbable}) with respect to utility infrastructure. Specifically, we
+developed a convolutional neural network with state-of-the-art configurations.  We applied data augmentation and
+pre-processing strategies to increase the size of our dataset by a factor of 5, thus generating 2525 images.  From an
+initial resolution of $4032 \times 3024$ pixels, we created three sets of image resolutions: $64 \times 64$,
+$128 \times 128$ and $224 \times 224$. We then defined four classification scenarios to investigate the performance of
+various groupings of the three categories: Pr\_Im (\textit{Probable} vs.\ \textit{Improbable}); PrPo\_Im
+(\textit{Probable} and \textit{Possible} vs.\ \textit{Improbable}); Pr\_PoIm (\textit{Probable} vs.\ \textit{Possible}
+and \textit{Improbable}); and Pr\_Po\_Im (\textit{Probable} vs.\ \textit{Possible} vs. \textit{Improbable}).
+We then optimized eight of our CNN hyperparameters for 12 classification
+scenario and image resolution combinations.  We conducted five-fold cross-validation for each of the 12 cases, and
+assessed model performance based on accuracy and the macro-averages of precision, recall and $F_{1}$ score.
 
-We further conducted more detailed classification performance analyses over single training instances for the 128-pixel case.
-Our results indicate that the CNN performed best at recalling \textit{Probable} vs. \textit{Improbable} cases.
-The \textit{Possible} category appeared to be a confounding case.
-This was not unexpected, given the uncertainty and subjectivity in assessing these trees to begin with.
-The CNN, however, performed better at distinguishing between \textit{Possible}/\textit{Improbable} and \textit{Probable}, than between \textit{Probable}/\textit{Possible} and \textit{Improbable}.
-This might be an indicator that trees in the \textit{Possible} category are more likely to be identified as \textit{Improbable}.
-When all three failure-likelihood categories were predicted individually, the \textit{Possible} category had the lowest recall score.
+Our results indicated that there generally was no significant difference in performance between the resolutions.  Among
+the scenarios, however, Pr\_Im performed the best with a top accuracy and $Pr^{m}$ of 0.94 ($\hat\sigma = 0.01)$.
+The next best-performing scenario was PrPo\_Im with a top accuracy and $Pr^{m}$ of 0.92 ($\hat\sigma = 0.01$).  For
+practical applications, scenario PrPo\_Im is more realistic, as it includes all three categories.  Thus, we deemed this
+as the most viable.  Scenarios Pr\_PoIm and Pr\_Po\_Im had best accuracy scores of 0.91 ($\hat\sigma = 0.01$) and 0.84
+($\hat\sigma = 0.02$), respectively.  But their performance across the other metrics was considerably worse.
 
-Nevertheless, given the relatively small input dataset of original images, these preliminary results are extremely promising for future improvements.
-First, we can train better models with more data.
-We also plan to rigorously quantify the uncertainty in ground truth category assignments by incorporating predictions from multiple experts on the same images.
-In order to better understand how the CNN is classifying each image, we will conduct extensive visual inference in further work, for instance, using the gradient-weighted class activation mapping approach \cite{zeiler2014visualizing}.
+We further conducted more detailed classification performance analyses over single training instances for the 128-pixel
+case.  Our results indicate that the CNN performed best at recalling \textit{Probable} vs. \textit{Improbable} or
+\textit{Probable}/\textit{Possible} vs.\ \textit{Improbable}.  The \textit{Possible} category appeared to be a
+confounding for the classifier.  This was not unexpected, given the uncertainty and subjectivity in assessing these trees
+to begin with.  The CNN, however, performed better at distinguishing between \textit{Possible}/\textit{Improbable} and
+\textit{Probable}, than between \textit{Probable}/\textit{Possible} and \textit{Improbable}.  This might be an indicator
+that trees in the \textit{Possible} category are more likely to be identified as \textit{Improbable}.  When all three
+failure-likelihood categories were predicted individually in scenario Pr\_Po\_Im, the \textit{Possible} category had the
+lowest class-specific recall score.
+
+Nevertheless, given the relatively small input dataset of original images, these preliminary results are extremely
+promising for future improvements.  First, we can train better models with more data.  We also plan to rigorously
+quantify the uncertainty in ground truth category assignments by incorporating predictions from multiple experts on the
+same images.  In order to better understand how the CNN is classifying each image, we will conduct extensive visual
+inference in further work, for instance, using the gradient-weighted class activation mapping approach
+\cite{zeiler2014visualizing}.  There is also a potential for mapping the visual cues learned by the CNN to physical
+relationships governing tree structure, in order to gain greater insights into tree failure processes.  The automation
+for tree failure likelihood assessments can potentially supplement human decision-making for increased resilience and,
+in the future, reduce costs and improve reliability in tree assessments, thus leading to more sustainable communities.
 
@@
 \section{Acknowledgments}
-The authors acknowledge the support of Eversource.
+The authors acknowledge the partial support of Eversource in funding this work.
 

### 2021-03-05 docs/manuscript/manuscript.tex
Updates from Overleaf

@@
   likelihood categories (\textit{Probable}, \textit{Possible}, \textit{Improbable}) under four classification scenarios.
-  Starting with an original set of 505 expert-labeled images, we perform preprocessing and augmentation tasks to
-  increase the number of samples. We optimize several hyperparameters in our CNN and then train and test its performance
-  for three different image resolutions. The trained CNN produced a validation accuracy of at least 0.94
+  Starting with an original set of 505 expert-labeled images, we performed preprocessing and augmentation tasks to
+  increase the number of samples by a factor of 5. We optimized several hyperparameters in the CNN and then trained and assessed its performance via cross-validation
+  for three different image resolutions under all four scenarios. The CNN produced a validation accuracy of at least 0.94
   ($\hat\sigma = 0.1)$ in the best-performing, yet hypothetical scenario as it excludes one of the categories.  The
@@
     \bf Activation &   \bf Rate & \bf No.\ Units  \\\midrule
-    Convolutional  & 1         & 64                & $k^*$   & 2           & ReLU && \\
+    Convolutional  & 1         & 64                & $k$   & 2           & ReLU && \\
     Max. Pooling   & 1         &                   & 2               &&&             &      \\
@@
 %     \bf Rate & \bf No.\ Units  \\\midrule
-    Dense           & 1  & & &  &  $a_1^*$ & & $u_1^*$ \\
-    Dropout         & 1  & & &  && $r_1^*$ &   \\
-    Dense           & 2  & & &  &  $a_2^*$ & & $u_2^*$  \\
-    Dropout         & 2  & & &  && $r_2^*$ &  \\ \bottomrule
+    Dense           & 1  & & &  &  $a_1$ & & $u_1$ \\
+    Dropout         & 1  & & &  && $r_1$ &   \\
+    Dense           & 2  & & &  &  $a_2$ & & $u_2$  \\
+    Dropout         & 2  & & &  && $r_2$ &  \\ \bottomrule
     \end{tabular}
     \caption{Summary of the convolutional neural network hyperparameters. 
-    Those indicated by an asterisked symbol are optimized using a guided search.}
+    Those represented by a symbol are optimized using a guided search.}
     \label{tab:hyp}
@@
 Taking all metrics into consideration, Welch pairwise tests showed that there are no significant differences in performance among the 3 training resolutions, with one exception.
-In the scenario Pr\_PoIm, the accuracy, $F_{1}^{m}$, $Re^{m}$ for the 128-pixel case are greater than those for the 224-pixel case ($p<.001$).
+In the scenario Pr\_PoIm, the accuracy, $F_{1}^{m}$, and $Re^{m}$ for the 128-pixel case are greater than those for the 224-pixel case ($p<.001$).
 This difference in performance is not as stark between the 128-pixel and 64-pixel cases in the same scenario.
@@
 \subsection{Analysis of classification strategies}
-Given that there was no significant difference in performance among the three resolutions tested, we focused our
-attention on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all the four
-scenarios using this resolution and then evaluated performance on the afore-mentioned validation metrics: accuracy and the macro-averages of
+Given that there was no significant difference in performance among the three resolutions tested, we focused 
+on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all four
+scenarios using this resolution and then evaluated performance on the aforementioned validation metrics: accuracy and the macro-averages of
 class-specific precision, recall and $F_{1}$ score.
@@
 % subsequent iterations.
-From this figure, we see that Pr\_Po\_Im performs significantly worse than the other three. This indicates the
+From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three. This indicates the
 uncertainty surrounding the expert, yet subjective, failure likelihood assessment of the trees to begin
@@
 difficult to predict accurately as an individual class.  The classifier performs best when it only has to distinguish
-between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}). Performance only suffers slightly when
+between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}).
+This observation aligns with previous work \cite{koeser2020can} and is arboriculturally intuitive.
+The \textit{Improbable} category includes trees with no apparent structural defects, which is comparatively easy for a qualified arborist to assess.
+When a defect is present, however, an arborist must estimate the severity of the defect that is present, whether the tree has grown in such a way that structurally compensates for the presence of the defect, and the anticipated loads on the tree.
+Although there are guidelines to inform an arborist's assessment \cite{smiley2017best,goodfellow2020best}, evaluating each of these factors still depends on an arborist's judgment and experience, making it more difficult to distinguish between trees in the \textit{Possible} and \textit{Probable} categories.
+Thus, it was not surprising that performance only suffered slightly when
 \textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}), which we would
@@
 \section{Acknowledgments}
-The authors acknowledge the partial support of Eversource in funding this work.
+The authors acknowledge the partial support of Eversource Energy in funding this work.
 

### 2021-03-05 docs/manuscript/manuscript.tex
added appendix tables

@@
 %\usepackage{amssymb,amsmath,amsthm}
-%\usepackage{bm}
+\usepackage{bm}
 %\usepackage[mathscr]{eucal}
@@
 The metrics are: accuracy and the macro-averages of precision, recall and the $F_{1}$ score.
+For reference, we summarize the statistics (mean and standard deviation) of the validation performance metrics in \autoref{tab:meanstd}.
 %With respect to the accuracy, there are no significant differences between the 
-Taking all metrics into consideration, Welch pairwise tests showed that there are no significant differences in performance among the 3 training resolutions, with one exception.
+Taking all metrics into consideration, Welch's pairwise tests show that there are no significant differences in performance among the 3 training resolutions, with one exception.
 In the scenario Pr\_PoIm, the accuracy, $F_{1}^{m}$, and $Re^{m}$ for the 128-pixel case are greater than those for the 224-pixel case ($p<.001$).
@@
 This outcome implies that we can achieve efficiency by  training at lower resolutions without significant losses in performance.
-
-% We further summarize the statistics (mean and standard deviation) of the validation accuracies in \autoref{tab:meanvalacc}.
+(A complete summary of the p-values is provided in \autoref{tab:welch-resolution}.)
 % Using the Welch test for pairwise comparisons of resolution performance within each scenario, we found no statistically significant difference between the validation accuracy for each resolution ($p\ge 0.57$).
@@
 \subsection{Scenario sensitivity}
-We conducted pairwise tests between each scenario combination for all resolutions.
+We conducted pairwise tests between each scenario combination for all resolutions. (A complete summary of the results is given in \autoref{tab:welch-scenario}.
 The results indicated that considering all metrics, the scenarios are statistically significantly different from each other ($p < .005$) except Pr\_Im compared to PrPo\_Im and Pr\_PoIm compared to Pr\_Po\_Im, with respect to certain metrics.
@@
 \appendix
+\section{Summary of validation performance metrics}
+For further reference we tabulate the mean and standard deviation of the performance metrics obtained on the scenario-resolution sensitivity tests. These values are shown in \autoref{tab:meanstd}.
+The metrics are accuracy, macro-average $F_1$ ($F_1^m$), macro-average recall ($Re^m$) and macro-average precision ($Pr^m$).
+
+\begin{table}[ht]\small\centering
+\caption{Mean and standard deviation of the 5-fold cross-validation estimates of classifier performance metrics on the 12 scenario-resolution cases investigated.
+``SD'' represents ``standard deviation.''}
+\label{tab:meanstd}
+\begin{tabular}{llllllllll}\toprule
+\multirow{3}{*}{\textbf{Scenario}} &  
+\multirow{3}{*}{\textbf{Resolution}}  & \multicolumn{2}{c}{\textbf{Accuracy}} & \multicolumn{2}{c}{$\bm{F_1^m}$} 
+& \multicolumn{2}{c}{$\bm{Re^m}$} & \multicolumn{2}{c}{$\bm{Pr^m}$} \\%\cmidrule{3-10}
+                          &            & Mean & SD  & Mean & SD  & Mean & SD  & Mean & SD  \\\midrule
+\multirow{3}{*}{Pr\_Im}    & 64        & 0.94 & 0.01 & 0.92 & 0.02 & 0.9  & 0.02 & 0.94 & 0.01 \\
+                          & 128        & 0.95 & 0.01 & 0.93 & 0.02 & 0.91 & 0.03 & 0.95 & 0.01 \\
+                          & 224        & 0.94 & 0.01 & 0.92 & 0.01 & 0.91 & 0.01 & 0.93 & 0.02 \\\midrule
+\multirow{3}{*}{PrPo\_Im}  & 64         & 0.92 & 0.01 & 0.91 & 0.01 & 0.91 & 0.02 & 0.92 & 0.01 \\
+                          & 128        & 0.92 & 0.01 & 0.91 & 0.01 & 0.9  & 0.01 & 0.92 & 0.01 \\
+                          & 224        & 0.91 & 0.01 & 0.91 & 0.01 & 0.9  & 0.01 & 0.91 & 0.01 \\\midrule
+\multirow{3}{*}{Pr\_PoIm}  & 64         & 0.88 & 0.01 & 0.8  & 0.04 & 0.79 & 0.07 & 0.85 & 0.05 \\
+                          & 128        & 0.91 & 0.01 & 0.85 & 0.02 & 0.84 & 0.02 & 0.87 & 0.03 \\
+                          & 224        & 0.88 & 0.01 & 0.79 & 0.02 & 0.76 & 0.02 & 0.84 & 0.03 \\\midrule
+\multirow{3}{*}{Pr\_Po\_Im} & 64         & 0.82 & 0.03 & 0.74 & 0.03 & 0.73 & 0.02 & 0.75 & 0.04 \\
+                          & 128        & 0.84 & 0.02 & 0.76 & 0.03 & 0.74 & 0.03 & 0.77 & 0.03 \\
+                          & 224        & 0.83 & 0.01 & 0.74 & 0.02 & 0.72 & 0.03 & 0.76 & 0.01\\ \bottomrule
+\end{tabular}
+\end{table}
+
+\section{Welch $t$-test results}
+Here, we summarize the p-values resulting from the Welch independent $t$-tests used to determine if the mean values of the various performance metrics are statistically significant or not.
+Strong evidence of significant difference is given by a low p-value (a typical threshold is $< 0.01$).
+
+\begin{table}[ht]\small\centering
+\caption{Welch's $t$-test p-value results for pairwise comparisons between image resolutions across all four scenarios.
+Low p-values indicates statistically different mean values for the validation metrics considered.}
+\label{tab:welch-resolution}
+\sisetup{scientific-notation=true, round-precision = 3,round-mode=figures,table-format = 1.2e-1,
+exponent-product=\ensuremath{\cdot}}%output-exponent-marker = \text{e}, 
+\begin{tabular}{@{}lll SSSS@{}}
+\toprule
+\multirow{2}{*}{\textbf{Scenario}} &
+\multirow{2}{*}{\textbf{Resolution 1}} &
+\multirow{2}{*}{\textbf{Resolution 2}} &
+\multicolumn{4}{c}{\textbf{p-values}} \\ %\cmidrule(l){4-7} 
+    & & & 
+  \multicolumn{1}{c}{\textbf{Accuracy}} &
+  \multicolumn{1}{c}{$\bm{F_1^m}$} &
+  \multicolumn{1}{c}{$\bm{Re^m}$} &
+  \multicolumn{1}{r}{$\bm{Pr^m}$} \\ \midrule
+\multirow{3}{*}{Pr\_Im}    & 64  & 128 & 0.427766 & 0.472000 & 0.638283 & 0.243409 \\
+                          & 64  & 224 & 0.810526 & 0.957770 & 0.728743 & 0.400994 \\
+                          & 128 & 224 & 0.333005 & 0.421449 & 0.781772 & 0.119129 \\\midrule
+\multirow{3}{*}{PrPo\_Im}  & 64  & 128 & 0.343214 & 0.334622 & 0.280111 & 0.607451 \\
+                          & 64  & 224 & 0.112543 & 0.153209 & 0.299562 & 0.131835 \\
+                          & 128 & 224 & 0.465132 & 0.566733 & 0.978346 & 0.253573 \\\midrule
+\multirow{3}{*}{Pr\_PoIm}  & 64  & 128 & 0.006932 & 0.037836 & 0.139458 & 0.345129 \\
+                          & 64  & 224 & 0.836653 & 0.715576 & 0.494841 & 0.697297 \\
+                          & 128 & 224 & 0.001099 & 0.000207 & 0.000467 & 0.076234 \\\midrule
+\multirow{3}{*}{Pr\_Po\_Im} & 64  & 128 & 0.241587 & 0.249616 & 0.391389 & 0.311935 \\
+                          & 64  & 224 & 0.677163 & 0.937617 & 0.598361 & 0.650331 \\
+                          & 128 & 224 & 0.227783 & 0.258068 & 0.252791 & 0.366722 \\ \bottomrule
+\end{tabular}
+\end{table}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% \usepackage{multirow}
+\begin{table}[ht]\small\centering
+\caption{Welch's $t$-test p-value results for pairwise comparisons between scenarios resolutions for all three image resolutions.
+Low p-values indicates statistically different mean values for the validation metrics considered.}
+\label{tab:welch-scenario}
+\sisetup{scientific-notation=true, round-precision = 3,round-mode=figures,table-format = 1.2e-1,
+exponent-product=\ensuremath{\cdot}}%output-exponent-marker = \text{e},
+
+\begin{tabular}{@{}lllSSSS@{}}
+\toprule
+\multirow{2}{*}{\textbf{Scenario}} &
+\multirow{2}{*}{\textbf{Resolution 1}} &
+\multirow{2}{*}{\textbf{Resolution 2}} &
+\multicolumn{4}{c}{\textbf{p-values}} \\ %\cmidrule(l){4-7} 
+    & & & 
+  \multicolumn{1}{c}{\textbf{Accuracy}} &
+  \multicolumn{1}{c}{$\bm{F_1^m}$} &
+  \multicolumn{1}{c}{$\bm{Re^m}$} &
+  \multicolumn{1}{r}{$\bm{Pr^m}$} \\ \midrule
+\multirow{6}{*}{64}  & Pr\_Im   & PrPo\_Im  & 0.005913 & 0.714479 & 0.422914 & 0.010639 \\
+                     & Pr\_Im   & Pr\_PoIm  & 0.000068 & 0.001772 & 0.014794 & 0.008354 \\
+                     & Pr\_Im   & Pr\_Po\_Im & 0.000184 & 0.000004 & 0.000004 & 0.000193 \\
+                     & PrPo\_Im & Pr\_PoIm  & 0.000814 & 0.002534 & 0.010824 & 0.020167 \\
+                     & PrPo\_Im & Pr\_Po\_Im & 0.000553 & 0.000014 & 0.000000 & 0.000351 \\
+                     & Pr\_PoIm & Pr\_Po\_Im & 0.005380 & 0.029135 & 0.132204 & 0.007418 \\\midrule
+\multirow{6}{*}{128} & Pr\_Im   & PrPo\_Im  & 0.004705 & 0.137997 & 0.578580 & 0.002389 \\
+                     & Pr\_Im   & Pr\_PoIm  & 0.001249 & 0.000455 & 0.003295 & 0.000790 \\
+                     & Pr\_Im   & Pr\_Po\_Im & 0.000005 & 0.000004 & 0.000012 & 0.000060 \\
+                     & PrPo\_Im & Pr\_PoIm  & 0.169503 & 0.000446 & 0.001725 & 0.011313 \\
+                     & PrPo\_Im & Pr\_Po\_Im & 0.000099 & 0.000064 & 0.000050 & 0.000320 \\
+                     & Pr\_PoIm & Pr\_Po\_Im & 0.000132 & 0.000244 & 0.000244 & 0.000931 \\\midrule
+\multirow{6}{*}{224} & Pr\_Im   & PrPo\_Im  & 0.000716 & 0.088647 & 0.576529 & 0.055010 \\
+                     & Pr\_Im   & Pr\_PoIm  & 0.000002 & 0.000001 & 0.000025 & 0.000256 \\
+                     & Pr\_Im   & Pr\_Po\_Im & 0.000000 & 0.000005 & 0.000029 & 0.000000 \\
+                     & PrPo\_Im & Pr\_PoIm  & 0.000130 & 0.000003 & 0.000014 & 0.001828 \\
+                     & PrPo\_Im & Pr\_Po\_Im & 0.000003 & 0.000014 & 0.000017 & 0.000000 \\
+                     & Pr\_PoIm & Pr\_Po\_Im & 0.000144 & 0.003617 & 0.029345 & 0.001147 \\ \bottomrule
+\end{tabular}
+\end{table}
 

### 2021-03-05 docs/manuscript/manuscript.tex
final edits to first journal submission

@@
 
-\eject
 \section{Conclusion}
@@
 \section{Data Availability Statement}
-All data, models, or code generated or used during the study are available in a GitHub repository online at \url{https://github.com/narslab/tree-risk-ai}.
+All data, models, or code generated or used during the study are publicly available in a GitHub repository online at \url{https://github.com/narslab/tree-risk-ai}.
 
@@
 For further reference we tabulate the mean and standard deviation of the performance metrics obtained on the scenario-resolution sensitivity tests. These values are shown in \autoref{tab:meanstd}.
-The metrics are accuracy, macro-average $F_1$ ($F_1^m$), macro-average recall ($Re^m$) and macro-average precision ($Pr^m$).
+The metrics are: accuracy, macro-average $F_1$ ($F_1^m$), macro-average recall ($Re^m$) and macro-average precision ($Pr^m$).
 
@@
 Strong evidence of significant difference is given by a low p-value (a typical threshold is $< 0.01$).
+In \autoref{tab:welch-resolution}, we consider pairwise comparisons between different image resolutions in all four scenarios.
+In \autoref{tab:welch-scenario}, we compare different scenarios for the same image resolution.
 

### 2021-03-08 bin/python/cnn-hyperparameter-optimization.py
Improved README documentation of all scripts and notebooks

@@ -36,6 +36,7 @@ HYPERBAND_MAX_EPOCHS = 10 #10
 EXECUTIONS_PER_TRIAL = 2 #5
 HYPERBAND_ITER = 3 #80
 
+# TODO: make image_dict a function
 image_dict = dict.fromkeys(RESOLUTION_LIST)
 for p in RESOLUTION_LIST:
     image_dict[p] = dict.fromkeys(SCENARIO_LIST)

### 2021-03-08 bin/python/cnn-performance.py

@@ -168,115 +168,6 @@ def trainModelWithDetailedMetrics(image_size, scenario, num_epochs = 10, trial_s
     return(model, hist) 
 
 
-def visualizeCNN(model, scenario, image_size, images_per_class = 4, trial_seed = 1, saliency=False, testing=True):
-    #ALT: use train_test_split
-#    trimg,vaimg,trlab,valab =  train_test_split(np.array([np.expand_dims(x[0],axis=2) for x in image_sets[64]["Pr_Im"]]), 
-#                 np.array([x[1] for x in image_sets[64]["Pr_Im"]]), stratify= np.array([x[1] for x in image_sets[64]["Pr_Im"]]), test_size=.2, random_state = 1  )
-
-    training_images_and_labels, test_images_and_labels = splitData(image_sets[image_size][scenario], prop = 0.8, seed_num = trial_seed)
-    training_images, training_labels = getImageAndLabelArrays(training_images_and_labels)
-    validation_images, validation_labels = getImageAndLabelArrays(test_images_and_labels)
-    class_labels = getClassLabels(scenario)
-    print("Class labels:", class_labels)
-    print(training_labels.sum(axis=0))
-    print(validation_labels.sum(axis=0))    
-    
-    # GRAD CAM
-    #random.seed(trial_seed)
-    # Randomly sample images from each class
-    random_image_selection_class_0 = random.choices([i for i, j in enumerate(validation_labels) if np.argmax(j) == 0], k = images_per_class)    
-    #random.seed(trial_seed+1)
-    random_image_selection_class_1 = random.choices([i for i, j in enumerate(validation_labels) if np.argmax(j) == 1], k = images_per_class)
-    assert validation_labels[random_image_selection_class_0].mean(axis=0)[0] == 1 #assert that indices of class 0 labels are correct
-    assert validation_labels[random_image_selection_class_1].mean(axis=0)[1] == 1 #assert that indices of class 1 labels are correct
-    cam_list = random_image_selection_class_0 + random_image_selection_class_1 # join lists of indices in both classes
-    if scenario=="Pr_Po_Im": # in 3-class case
-        #random.seed(trial_seed+2)
-        random_image_selection_class_2 = random.choices([i for i, j in enumerate(validation_labels) if np.argmax(j) == 2], k = images_per_class)    
-        cam_list = cam_list + random_image_selection_class_2 # join to prior list of class 0 and class 1
[…]

### 2021-03-08 bin/python/label-image-files.py

@@ -10,8 +10,8 @@ import matplotlib.pyplot as plt
 import random
 import ntpath
 
-raw_image_dir = '../../data/raw/Pictures for AI'
-tidy_image_dir = '../../data/tidy/labeled_images/'
+RAW_IMAGE_DIR = '../../data/raw/Pictures for AI'
+TIDY_IMAGE_DIR = '../../data/tidy/labeled_images/'
 
 def getListOfFiles(dirName):
     """Returns single list of the filepath of each of the training image files"""
@@ -36,18 +36,18 @@ def saveImageFiles(image_file_list):
     improbable_counter = 1
     possible_counter = 1
     probable_counter = 1
-    if not os.path.exists(tidy_image_dir):
-        os.makedirs(tidy_image_dir)
+    if not os.path.exists(TIDY_IMAGE_DIR):
+        os.makedirs(TIDY_IMAGE_DIR)
     for filename in image_file_list:
         if '.JPG' in filename or '.jpg' in filename:        
             if any(re.findall(r'improbable', filename, re.IGNORECASE)):
-                save_name = tidy_image_dir + 'improbable' + '-' + str(improbable_counter) + '.jpg'
+                save_name = TIDY_IMAGE_DIR + 'improbable' + '-' + str(improbable_counter) + '.jpg'
                 improbable_counter += 1
             elif any(re.findall(r'probable', filename, re.IGNORECASE)):
-                save_name = tidy_image_dir + 'probable' + '-' + str(probable_counter) + '.jpg'
+                save_name = TIDY_IMAGE_DIR + 'probable' + '-' + str(probable_counter) + '.jpg'
                 probable_counter += 1 
             elif any(re.findall(r'possible', filename, re.IGNORECASE)):
-                save_name = tidy_image_dir + 'possible' + '-' + str(possible_counter) + '.jpg'
+                save_name = TIDY_IMAGE_DIR + 'possible' + '-' + str(possible_counter) + '.jpg'
                 possible_counter += 1 
             imageio.imwrite(save_name, np.array(Image.open(filename)))
     print('Number of improbable images saved:', improbable_counter - 1)    
@@ -75,9 +75,9 @@ def plotRawImages(image_file_list, images_per_class = 2):
     fig.savefig(image_filename, dpi=120)
 
 def main():    
-    tree_image_list = getListOfFiles(raw_image_dir)
[…]

### 2021-03-08 bin/python/plot-processed-images.py

@@ -1,26 +1,48 @@
-#TODO (in order to not have to regenerate images each time)
-def plotProcessedImages(class_scenario, image_array, class_list, images_per_class, resolution):
-    num_cols = images_per_class
-    num_rows = len(class_list)
-    fig, axarr = plt.subplots(num_rows, num_cols, sharex=True, sharey=True)
-    #print(image_file_list)
-    for i in range(num_rows):
-        for j in range(num_cols):
-            axarr[i, j].imshow(image_array[i][j], cmap = 'gist_gray', extent = [0, resolution, 0, resolution])
-    for ax, row in zip(axarr[:, 0], [i for i in class_list]):
-        ax.set_ylabel(row, size=15)
-    image_filename = '../../figures/processed_input_images_' + str(class_scenario) + '_' + str(resolution) + '_px.png'
-    #plt.xticks([0,3024])
-    #plt.yticks([0,4032])
-    fig.savefig(image_filename, dpi=180)
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+IMAGE_ARRAY = np.load('../../data/tidy/preprocessed_images/size128_exp5_Pr_Po_Im.npy',allow_pickle = True)
+
+def plotSelectedProcessedImages(image_array):
+    ## Generates the subfigures in Figure 2 in the paper.
+    example_image_list = [1790, 1792, 
+                          2235, 2237, 
+                          950, 951, 
+                          1180, 1181, 
+                          2110, 2111,
+                          1301, 1302
+                         ]
+    example_image_filenames = ['probable-example-1.png', 'probable-example-1-flipped.png',
+                              'probable-example-2.png', 'probable-example-2-flipped.png',
+                              'possible-example-1.png', 'possible-example-1-flipped.png',
+                              'possible-example-2.png', 'possible-example-2-flipped.png',   
+                              'improbable-example-1.png', 'improbable-example-1-flipped.png',
+                              'improbable-example-2.png', 'improbable-example-2-flipped.png',                            
+                              ]
+    for i,n in enumerate(example_image_list):
+        plt.imshow(image_array[n][0], cmap = 'gist_gray')
+        plt.tight_layout()
+        plt.yticks([0,40,80,120],fontsize=16)
+        plt.xticks([0,40,80,120],fontsize=16)
[…]
